### 2021-03-22 04_reports/manuscript.Rmd
my advances of monday 22

---
title: "Dissecting the causal question underlying the association between cancer and ADRD "
author: "L. Paloma Rojas-Saunero"
date: ""
output: [html_document]
bibliography: cancer_dementia.bib
---
```{r, echo =  FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align="center")
```

### Motivation

Several observational studies have found that individuals with history or incident cancer that survive over follow-up have a lower risk of Alzheimer´s disease or related dementias (ADRD) when compared to individuals with no history of cancer. Multiple mechanisms related to inflammatory response, immune response, etc etc are postulated to understand this phenomenon. However, the community has raised their concerned about the potential unmeasured confounding, selective survival and misclassification issues that could drive this results.

In the lasts few years there has been an important attention to unveiling the impact of death as a competing event, and the role of survival bias to explain this association. Previous studies alluded to choosing a particular summary measure or modeling approach to account for death (incidence rate ratios (ref), hazard ratios (ref), multistate models (ref)), though the underlying estimand remains unclear.

Overall, although observational studies have measured history of cancer and incident cancer as the key exposure, researchers refer to biological mechanisms which onset may be prior to cancer diagnosis and the effect of cancer treatment.

Pin1 is an enzime deeply involved in the pathogenesis of cancer and ADRD. One of the roles of Pin1 is regulate mitosis process and apoptosis. Over-expression of Pin1 has been found in several tumors and it's inhibition causes tumor regression. In contrast, since adult neuronal cells do not divide, Pin1 over-expression in brain tissue is used to restore tau and amyloid precursor protein to a functional shape, thus preventing the buildup of the toxic tau and beta-amyloid species associated with AD.

For this reason Pin1 regulation is proposed as one of the mechanisms that can explain the inverse association between cancer and AD found in observational studies.

Our aim is to outline one way to tie the etiological question of Pin1 with the questions that can be answered with observational data where Pin1 measurements are not available. We aim to draw the complexity of the research question by building progressively a directed acyclic graph (DAG) which help us connect the causal question to the observable data and the assumptions we rely on to identify the effect. We describe the challenges and illustrate the assumptions we have to fulfill to identify the effect.

## Overview of the causal structure and assumptions related to cancer diagnosis as a proxy for Pin1

Given the previous literature on the topic, we might be interested in answering the following question: _What is the direct effect of Pin1 overexpression at age of 65 in the risk of ADRD by age 85 compared to regulated expression of Pin1 by age of 65._ To identify this effect with real-world data, we will illustrate the causal question through a directed acyclic graphs (DAGs), spelling out the data generating mechanisms and assumptions tied to the data. _Maybe say something about dags here_

Figure 1. illustrates a situation where Pin1, represented as $P$, was measured in all participants on their 65 birthday. Based on the measurement of Pin1, participant's exposure was classified as 1) Over-expressed Pin1 or 2) Regulated Pin1. Through out the follow-up each participant followed a strict monitoring for ADRD onset, represented as $Y_{85}$, and no participant was lost to follow-up nor died during the study period.

Pin1 expression (over-expressed vs. regulated) was not an intervention randomized at age 65, therefore we have to consider a set of measured covariates $L$ for which conditional exchangeability holds. Previous studies have described age, sex, educational level and race/ethnicity as the minimal adjusting set of covariates (Ospina). However, environmental and behavioral factors such as smoking, alcohol intake, or chronic diseases which are known to cause microenvironmental changes such as inflammation and changes in tissue remodelling, may translate into Pin1 over-expression and are also related to the development of ADRD. Therefore, had we measured all $L$ we could identify the direct of $P$ in $Y_{85}$ by conditioning on $L$.

```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = "Direct effect of Pin1 in ADRD"}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
\node (p) {$P$};
\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$L$};
\node [right of = p] (y) {$Y_{85}$};
\draw[->] (l) -- (p);
\draw[->, violet] (p) -- (y);
\draw[->] (l) -- (y);
\end{tikzpicture}
```

Unfortunately, Pin1 expression measurement is not available in real-world settings, thus we can only rely on a proxy of this exposure. Since Pin1 over-expression is present in tumors and tumors are only measured through screening and diagnosis, we considered cancer diagnosis as the proxy for Pin1 over-expression, as most observational studies have defined (all refs). We depict this feature in Figure 2, where $P^*$ represents _incident cancer diagnosis_, the measured proxy of $P$. In this DAG we colored the path $P^* \leftarrow P \rightarrow Y_{85}$ because although we are measuring the association between $P^*$ and $Y_85$ in the observed data, we are assuming that the effect is only through $P$.

```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = "Direct effect of Pin1 in ADRD, with cancer diagnosis as proxy of Pin1"}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
\node (p) {$P$};
\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$L$};
\node [above of = p, yshift = -1cm](pstar) {$P^*$};
\node [right of = p] (y) {$Y_{85}$};
\draw[->, violet] (p) -- (pstar);
\draw[->, violet] (p) -- (y);
\draw[->] (l) -- (p);
\draw[->] (l) -- (y);

\end{tikzpicture}
```

A major challenge related to assigning cancer diagnosis as the proxy of Pin1 is defining time zero. Not everyone who had Pin1 overexpression will be diagnosed with cancer by age 65. For this reason prospective cohort studies have considered cancer as a time-varying or time-dependent exposure. This means that a participant who had cancer diagnosis at age 70 contributes to the "regulated Pin1" from age 65 to 70, and from age 70 and onwards, this participant contributes to "Over-expressed Pin1" arm. Alternatively, studies that have included participants with cancer diagnosis at the time of the diagnosis (for example from cancer registries such as SEER) and matched participants by age. These subtleties are relevant for the next point.

Since the population is elderly, the risk of death at age 65 is not trivial and it increases over time. This means that some individuals may die prior to cancer diagnosis due to other causes, such as cardiovascular disease which is one of the leading causes of death in this age group. As such, we can only measure $P^*$ in the subset of individuals who have survived long enough to have a cancer diagnosis. We zoom in the relationship of $P \rightarrow P^*$ in Figure 3. In this figure we include 
$D = 0$ between $P$ and $P^*$, this illustrates that to observe $P^*$, we condition (box around $D = )$ on surviving long enough to have a cancer diagnosis.

Several risk factors that increase the risk of cancer might also be shared causes of death prior to cancer diagnosis, for example smoking may cause lung cancer and chronic obstructive pulmonary disease (a leading cause of death in this age group). Therefore, to isolate the effect of $P$ in $P^*$ (the violet arrow), we need to block all shared causes of Pin1 and death from other causes, and the shared causes of death from other causes and cancer diagnosis, both represented in Figure 3 as $C_1$ and $C_2$. In this way we assume a hypothetical scenario in which we could prevent death prior to cancer diagnosis, by conditioning on a rich set of covariates. This assumption must hold regardless of whether we use incident cancer as a time-varying exposure, or match cancer patients to participants free of cancer by age.

```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = "Zoom in the association between P and P star"}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
\node (p) {$P$};
\node [draw, right of = p] (d) {$D = 0$};
\node [right of = d] (pstar) {$P^*$};
\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$C_1$};
\node [below of = d, yshift = 1cm, xshift = 1cm](c) {$C_2$};
\draw[->] (p) -- (d);
\draw[->, violet] (p) to [out=45,in=135, looseness=0.4] (pstar);
\draw[->] (d) -- (pstar);
\draw[->] (l) -- (p);
\draw[->] (l) -- (d);
\draw[->] (c) -- (pstar);
\draw[->] (c) -- (d);
\end{tikzpicture}
```

Up to this point we have outlined the underlying assumptions of using incident cancer diagnosis as a proxy for Pin1 expression. We continue outlining the additional assumptions related to death as a competing event of ADRD.

## Overview of the causal question and assumptions related to death as a competing event of ADRD

Death is a competing event of ADRD because if a participant dies prior to ADRD diagnosis, death prevents from observing ADRD at future time-points. The causal diagram in Figure 4 is an extension of Figure 2, including $D_{80}$ as a representation of Death by age 80. We include an arrow between $P$ and $D_{80}$ since $P$ may increase the risk, through tumor development or other shared common causes. Furthermore we include an arrow between $P^*$ and $D_{80}$ since cancer diagnosis, and subsequent treatment (or lack of treatment) may have an effect on death. The arrow from $D_{80}$ and $Y_{85}$ represents the key feature of a competing events data structure: an individual who dies by age 80 cannot subsequently develop ADRD. Last, since $D_{80}$ and $Y_{85}$ events that are related to the aging process, $C$ represent the shared causes related to both events (sch as cardiovascular conditions, fraility, other comorbidities, etc).

If we had measured all $L$, we could estimate the total effect of $P$ in the risk of ADRD under no additional assumptions. This effect includes all the pathways including those mediaded through death.

There are several causal estimands that can help us identify the direct of $P$ in $Y_{85}$ as presented by Young et al. 
However if we want to isolate the direct effect of P in Y (path in violet), we

```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = "Effect of Pin1 in the risk of ADRD by age 85, with cancer diagnosis as proxy of Pin1"}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
\node (p) {$P$};
\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$L$};
\node [above of = p, yshift = -1cm](pstar) {$P^*$};
\node [draw, right of = p] (d) {$D_{80}$};
\node [right of = d] (y) {$Y_{85}$};
\node [below of = d, yshift = 1cm, xshift = +1cm](c) {$C$};
\draw[->, violet] (p) -- (pstar);
\draw[->] (pstar) -- (d);
\draw[->] (p) -- (d);
\draw[->] (d) -- (y);
\draw[->] (l) -- (p);
\draw[->] (l) -- (y);
\draw[->] (c) -- (d);
\draw[->] (c) -- (y);
\draw[->, violet] (p) to [out=45,in=135, looseness=0.4] (y);
\end{tikzpicture}
```

## Example using the Rotterdam Study data

## Discussion

We need to figure out what the key aging-dependent changes are and how to modulate these factors safely

Causes such as aging, smoking, irradiation etc are causing micro environmental changes, like inflamation and changes in tissue remodelling. This promotes selection for adaptive mutations.

Efforts to prevent and treat cancer should converge with similar efforts to prevent other aging- associated diseases.

We need better means to

Youthful tissues

### 2021-03-23 04_reports/manuscript.Rmd
small updates

@@
 
-Given the previous literature on the topic, we might be interested in answering the following question: _What is the direct effect of Pin1 overexpression at age of 65 in the risk of ADRD by age 85 compared to regulated expression of Pin1 by age of 65._ To identify this effect with real-world data, we will illustrate the causal question through a directed acyclic graphs (DAGs), spelling out the data generating mechanisms and assumptions tied to the data.
-_Maybe say something about dags here_
+Given the previous literature on the topic, we focus the interest in answering the following question: _What is the direct effect of Pin1 over-expression at age of 65 in the risk of ADRD by age 85 compared to regulated expression of Pin1 by age of 65._ To identify this effect with real-world data, we will illustrate the causal question through a directed acyclic graphs (DAGs), spelling out the data generating mechanisms and assumptions tied to the data.
 
-Figure 1.
-illustrates a situation where Pin1, represented as $P$, was measured in all participants on their 65 birthday.
+The causal graph of Figure 1 illustrates a situation where Pin1, represented as $P$, was measured in all participants on their 65 birthday.
 Based on the measurement of Pin1, participant's exposure was classified as 1) Over-expressed Pin1 or 2) Regulated Pin1.
@@
 
-Pin1 expression (over-expressed vs.
-regulated) was not an intervention randomized at age 65, therefore we have to consider a set of measured covariates $L$ for which conditional exchangeability holds.
+was not an exposure or intervention randomized at age 65, therefore we have to consider a set of measured covariates $L$ for which conditional exchangeability holds.
 Previous studies have described age, sex, educational level and race/ethnicity as the minimal adjusting set of covariates (Ospina).
@@
 A major challenge related to assigning cancer diagnosis as the proxy of Pin1 is defining time zero.
-Not everyone who had Pin1 overexpression will be diagnosed with cancer by age 65.
+Not everyone who had Pin1 over-expression will be diagnosed with cancer by age 65.
 For this reason prospective cohort studies have considered cancer as a time-varying or time-dependent exposure.
-This means that a participant who had cancer diagnosis at age 70 contributes to the "regulated Pin1" from age 65 to 70, and from age 70 and onwards, this participant contributes to "Over-expressed Pin1" arm.
+This means that a participant who had cancer diagnosis at age 70 contributes to the "regulated Pin1" from age 65 to 70, and from age 70 and onwards to the "Over-expressed Pin1" arm.
 Alternatively, studies that have included participants with cancer diagnosis at the time of the diagnosis (for example from cancer registries such as SEER) and matched participants by age.
-These subtleties are relevant for the next point.
+If age at cancer diagnosis is considered as the time zero, we should be careful of adjusting for post-baseline covariates of $P$ or mediators between $P$ and $P^*$.
 
 Since the population is elderly, the risk of death at age 65 is not trivial and it increases over time.
-This means that some individuals may die prior to cancer diagnosis due to other causes, such as cardiovascular disease which is one of the leading causes of death in this age group.
+This means that some individuals may die prior to cancer diagnosis due to other leading causes of death at this age group, such as cardiovascular disease.
 As such, we can only measure $P^*$ in the subset of individuals who have survived long enough to have a cancer diagnosis.

### 2021-03-24 04_reports/manuscript.Rmd
update on competing events

@@
 
-Death is a competing event of ADRD because if a participant dies prior to ADRD diagnosis, death prevents from observing ADRD at future time-points.
-The causal diagram in Figure 4 is an extension of Figure 2, including $D_{80}$ as a representation of Death by age 80.
-We include an arrow between $P$ and $D_{80}$ since $P$ may increase the risk, through tumor development or other shared common causes.
-Furthermore we include an arrow between $P^*$ and $D_{80}$ since cancer diagnosis, and subsequent treatment (or lack of treatment) may have an effect on death.
-The arrow from $D_{80}$ and $Y_{85}$ represents the key feature of a competing events data structure: an individual who dies by age 80 cannot subsequently develop ADRD.
-Last, since $D_{80}$ and $Y_{85}$ events that are related to the aging process, $C$ represent the shared causes related to both events (sch as cardiovascular conditions, fraility, other comorbidities, etc).
+As illustrated in previous figures, we are interested in the direct effect of Pin1 in ADRD at 85 years, which means a length of follow-up of 20 years.
+However, mortality in elderly population between 65 and 85 is substancial, and a leading cause of death is cancer.
+This means that in order to measure ADRD by the age of 85, individuals have to had survive to cancer and to other causes.
+For this reason death is a competing event of ADRD because if a participant dies prior to ADRD diagnosis, death prevents from observing ADRD at future time-points.
 
-If we had measured all $L$, we could estimate the total effect of $P$ in the risk of ADRD under no additional assumptions. This effect includes all the pathways including those mediaded through death.
+We can visualize this data feature in the causal diagram in Figure 4.
+In this DAG we include $D_{80}$ as a representation of Death by age 80, an arrow between $P$ and $D_{80}$ since $P$ may increase the mortality risk irrespective of cancer diagnosis.
+We also include an arrow between $P^*$ and $D_{80}$ since cancer diagnosis, and subsequent treatment (or lack of treatment) may have an effect on death.
+Last, the arrow from $D_{80}$ and $Y_{85}$ represents the key feature of a competing events data structure: an individual who dies by age 80 cannot subsequently develop ADRD, and since $D_{80}$ and $Y_{85}$ are events related to aging, $C$ represent the shared causes of both events.
 
-There are several causal estimands that can help us identify the direct of $P$ in $Y_{85}$ as presented by Young et al. 
-However if we want to isolate the direct effect of P in Y (path in violet), we
+In this DAG we observe that, had we measured and adjusted for $L$, we could estimate the total effect of $P$ in $Y_{85}$ without further assumptions.
+However the total effect is a causal estimand that includes both the direct pathway between $P$ and $Y_{85}$.
+This means that if Pin1 has an effect on mortality through cancer or other mechanisms, we will observe a protective effect of Pin1 overexpression in ADRD, partially mediated through death.
+
+If the aim is to isolate the direct effect of $P$ in $Y_{85}$ as in Figure 4 (violet arrows), we first need to define the causal estimand of interest.
+There are at least three causal estimands, each one has different interpretation and relies on different assumptions: the controlled direct effect, the survivor average causal effect and the separable direct effect.
+Through the next section we will discuss the controlled direct effect since this estimand is tied to frequently used estimators.
+
+There are several ways to conceptualize direct effects and there are several causal parameters (estimands) to represent it, such as: the controlled direct effect (CDE), the survivor average causal effect and the separable direct effect.
+Through this section we will discuss the causal question, assumptions related to death and interpretation of the CDE since this estimand is connects to frequently used estimators.
+
+THe CDE represents the effect of Pin1 in ADRD in a setting where we could have prevented death over the study period.
 

### 2021-03-25 04_reports/manuscript.Rmd
rough on CE, move to examples

@@
 date: ""
-output: [html_document]
+output: [pdf_document]
 bibliography: cancer_dementia.bib
@@
 In this DAG we observe that, had we measured and adjusted for $L$, we could estimate the total effect of $P$ in $Y_{85}$ without further assumptions.
-However the total effect is a causal estimand that includes both the direct pathway between $P$ and $Y_{85}$.
-This means that if Pin1 has an effect on mortality through cancer or other mechanisms, we will observe a protective effect of Pin1 overexpression in ADRD, partially mediated through death.
+However the total effect includes all pathways between $P$ and $Y_{85}$.
+This means that if Pin1 has an effect on mortality through cancer or other mechanisms, we will observe a protective effect of Pin1 overexpression in ADRD, partially or fully mediated through death.
 
-If the aim is to isolate the direct effect of $P$ in $Y_{85}$ as in Figure 4 (violet arrows), we first need to define the causal estimand of interest.
-There are at least three causal estimands, each one has different interpretation and relies on different assumptions: the controlled direct effect, the survivor average causal effect and the separable direct effect.
-Through the next section we will discuss the controlled direct effect since this estimand is tied to frequently used estimators.
-
-There are several ways to conceptualize direct effects and there are several causal parameters (estimands) to represent it, such as: the controlled direct effect (CDE), the survivor average causal effect and the separable direct effect.
-Through this section we will discuss the causal question, assumptions related to death and interpretation of the CDE since this estimand is connects to frequently used estimators.
+Alternatively, there are several ways to conceptualize the direct effect of $P$ in $Y_{85}$ as in Figure 4 (violet arrows) and several causal parameters (estimands) to represent it, such as: the controlled direct effect (CDE), the survivor average causal effect and the natural separable direct effect.
+Through this section we will discuss the causal question, assumptions related to death and interpretation of the CDE since this estimand connects to frequently used estimators.
 
 THe CDE represents the effect of Pin1 in ADRD in a setting where we could have prevented death over the study period.
+It relies on the assumption that we have measured all $C$ to block the pathway $Y85 \leftarrow C  \rightarrow D_{80} \rightarrow P \rightarrow P*$.
+This assumption is defined as the independent censoring assumption conditional on covariates.
+In this setting death is treated as a censoring event and it can be interpreted as those who died would have the same risk of developing dementia if prevented from dying than those who remained alive and free of dementia at a given time point, conditional on shared causes of dementia and death.
 
@@
 
-## Example using the Rotterdam Study data
+## Motivating example with the RS data
+
+Descriptive statistics
+
+In a population of x, free of cancer and dementia at baseline...
+
+We first do counts:
+
+Time to cancer on average
+
+We show 3 kinds, ever vs. never, time-varying cancer, and weighting by cancer dx.
+
+We now show the risk if U, no U (Combined outcome, crude risk)
+
+We provide additional info on causes of death for both groups, and the extreme scenarios of bounds.
+
+## Discussion
+
+Through this example we describe how much do the direct effect of cancer changes if we measure and adjust for $C$ compared to not.
 

### 2021-03-29 02_R/01_wide.R
Updates, restrict to rs1, age, no trunc on 20y only by the end of long setup

@@ -10,24 +10,24 @@ library(magrittr)
 data <- import(here::here("01_data", "clean_data","complete_data.RData"))
 
 data <- data %>%  
+  rename(id = ergoid, cohort = rs_cohort) %>% 
   filter(!is.na(e1)) %>% 
   filter(dementia_inc != 2) %>% 
   filter(dementia_at_risk == 1) %>% 
-  #filter(cancer_prev != 1) %>% # KIM this one is not needed anymore
+  filter(cancer_prev != 1) %>% # Based on the ideal trial
   # filter(!is.na(education_three_levels)) %>% 
   # filter(!is.na(smoke1)) %>% 
-  filter(!is.na(bmi1)) %>% 
-  filter(!is.na(sbp1)) %>% 
-  # filter(!is.na(ht_drug1)) %>% 
+  filter(!is.na(bmi1)) %>%
+  filter(!is.na(sbp1)) %>%
+  filter(!is.na(ht_drug1)) %>%
   filter(!is.na(censor_date)) %>% 
-  rename(id = ergoid, cohort = rs_cohort) 
-
-#KIM: SELECT ONLY THOSE IN RS-1 AND RS-2
-data <- subset(data,data$cohort!=3)
+  filter(between(age_0, 60, 70)) %>% 
+  filter(cohort == 1)
 
 #KIM: SELECT ONLY THOSE WITH IC BASED ON STROKE DATASET
 data <- subset(data,data$ic_ok_2016==1)
 
+data %>% ggplot(aes(age_0)) + geom_histogram()
 
 ## truncate death in 2015
 data <- data %>%
@@ -71,6 +71,12 @@ data %<>%
     t2cancer_d = round(as.numeric(as.period((e1 %--% cancer_date), "days"
     ), "days"), 0))
 
+data %>% 
+  group_by(cohort) %>% 
+  summarize(
+    max(t2death_y),
+    max(t2dem_y, na.rm = TRUE),
+    max(t2cancer_y, na.rm = TRUE))
 
 ## t2dem and t2cancer are empty if date variables are empty,
 ## so use the end_fup_2015 to fill the fup time
@@ -96,8 +102,7 @@ data <- data %>%
 ## Cancer
 data <- data %>%
   mutate(
-    cancer = case_when(!is.na(cancer_date) & cancer_date <= end_fup_2015 ~ 1,
-      TRUE ~ 0),
[…]

### 2021-03-29 02_R/02_long.R

@@ -11,7 +11,7 @@ data <- import(here::here("01_data", "clean_data", "wide_noltfu.RData"))
 # Create long, as many rows as t2death in years + 1
 
 data_long <- data %>% 
-  select(id, cohort, dementia, cancer, cancer_20, cancer_prev, cancer_date,
+  select(id, cohort, dementia, cancer, cancer, cancer_prev, cancer_date,
          dementia_date, death_2015, starts_with("t2"),
          end_fup_2015, e1,
          sex, age_0, education, apoe4, 
@@ -21,16 +21,13 @@ data_long <- data %>%
   mutate(t2death_y = round(t2death_y,0)) %>% 
   group_by(id) %>% 
   slice(rep(1:n(), each = t2death_y + 2)) %>% #repeat rows by #years until death
-  ungroup() %>%
-  # select(everything(), - matches("[1,2,3,4,5]$"), apoe4) %>% ## delete all covariates
-  group_by(id) %>%
   mutate(year = year(e1),
          year = year + (row_number() - 1)) %>% #an easy roll for years + 1
   ungroup()
 
 
-## Completes with rows for all missing years (everyone until 2014)
-data_long_2014 <- data_long %>% 
+## Completes with rows for all missing years (everyone until 2016 because I extended an extra row)
+data_long_2015 <- data_long %>% 
   complete(id, year)
 
 ### Make dataset with visit data, and transform to long
@@ -58,18 +55,18 @@ rep_cov <- data %>%
 
 ## join and fill missign values (fills gaps between 2 visits by carrying forward)
 
-data_long_2014 <- data_long_2014 %>%
+data_long_2015 %<>% 
   left_join(rep_cov, by = c("id", "year"))
 
-var_names <- colnames(data_long_2014)
+var_names <- colnames(data_long_2015)
 
-data_long_2014 <- data_long_2014 %>%
+data_long_2015 %<>% 
   group_by(id) %>%
   fill(var_names[-1]) %>% 
   ungroup()
 
 ### remove rows prior to study entry
-data_long_2014 %<>% 
+data_long_2015 %<>% 
   group_by(id) %>% 
   filter(year >= year(e1)) %>% 
   ungroup()
@@ -78,7 +75,7 @@ data_long_2014 %<>%
[…]

### 2021-03-29 02_R/05_descriptives.R
descriptives for manuscript

@@ -1,0 +1,112 @@
+library(survival)
+library(tidyverse)
+
+data_long <-
+  rio::import(here::here("01_data", "clean_data", "dementia_long.RData"))
+
+data_wide <-
+  data_long %>%
+  group_by(id) %>% 
+  slice(max(row_number())) %>% 
+  ungroup()
+
+## Total n
+total_n <- dim(data_wide)[1]
+
+## Counts
+
+total_cancer <- data_wide %>% 
+  count(cancer_v) %>% 
+  mutate(prop = round(100*n/sum(n),0)) %>% 
+  mutate(cancer_count = paste0(prop, "%", " (n = ", n, ")")) %>% 
+  select(cancer_count, cancer_v)
+  
+no_cancer <- total_cancer %>% 
+  filter(cancer_v == 0) %>% 
+  pull(1)
+
+yes_cancer <- total_cancer %>% 
+  filter(cancer_v == 1) %>% 
+  pull(1)
+
+total_status <- data_wide %>%
+  group_by(cancer_v) %>% 
+  count(status) %>% 
+  mutate(prop = round(100*n/sum(n),0)) %>% 
+  mutate(total = paste0(prop, "%", " (n = ", n, ")")) %>% 
+  select(cancer_v, status, total)
+
+cancer_dem <- total_status %>% filter(cancer_v == 1, status == "Cancer and Dementia") %>% pull(3)
+cancer_death <- total_status %>% filter(cancer_v == 1, status == "Dead") %>% pull(3)
+cancer_alive <- total_status %>% filter(cancer_v == 1, status == "Only cancer") %>% pull(3)
+
+no_cancer_dem <- total_status %>% filter(cancer_v == 0, status == "Only dementia") %>% pull(3)
+no_cancer_death <- total_status %>% filter(cancer_v == 0, status == "Dead") %>% pull(3)
+no_cancer_alive <- total_status %>% filter(cancer_v == 0, status == "Alive, free of cancer and dementia") %>% pull(3)
+
+cancer_time <- data_long %>% 
+  filter(cancer_v == 1 & lag(cancer_v) == 0) %>% 
+  summarize(m = median(age_scale),
+            l = quantile(age_scale, 0.25),
+            u = quantile(age_scale, 0.75)) %>% 
+  mutate(temp = paste0(m, " (IQR: ", l, "-", u, ")")) %>% 
[…]

### 2021-03-29 02_R/03_plots.R
change to age scale, now focused on dementia_long

@@ -10,75 +10,80 @@ palette <- wes_palette("Moonrise3")
 
 data_wide <- import(here::here("01_data", "clean_data", "wide_noltfu.RData"))
 
-data_long_2014 <- import(here::here("01_data", "clean_data", "data_long_2014.RData"))
+data_long <- import(here::here("01_data", "clean_data", "data_long_2015.RData"))
+data_dem <- import(here::here("01_data", "clean_data", "dementia_long.RData"))
 
-data_long_2014 %<>%
-  mutate(age_scale = age_0 + (row_number() - 1))
+# data_long %>% select(age_0, age_scale) %>% View()
 
-data_long_2014 %>% 
-  filter(id == 17002) %>% 
-  select(id, cohort, year, e1, cancer, cancer_v, cancer_date, dementia, dem_v, dem_date, death_2015, death_v, death_year, status, t2death_y)%>% View() 
+# data_long %>%
+#   filter(id == 17002) %>%
+#   select(
+#     id,
+#     cohort,
+#     year,
+#     e1,
+#     cancer,
+#     cancer_v,
+#     cancer_date,
+#     dementia,
+#     dem_v,
+#     dem_date,
+#     death_2015,
+#     death_v,
+#     death_year,
+#     status,
+#     t2death_y
+#   ) %>% View()
 
 ## Plot distribution of events over time
 
 ## Keeps those who died
-data_long_2014 %>% 
-  # filter(between(age_0, 60, 75)) %>%
-  group_by(cohort, year) %>% 
+data_long %>% 
+  group_by(age_scale) %>% 
   count(status) %>%
-  mutate(prop = round(100*n/sum(n))) %>% 
+  mutate(prop = round(100*n/sum(n)),0) %>% 
   ungroup() %>% 
-  filter(cohort == 1) %>% 
-  ggplot(aes(year, prop, fill = status)) +
+  filter(age_scale <= 90) %>% 
+  ggplot(aes(age_scale, prop, fill = status)) +
   geom_area() + 
   scale_fill_manual(values = palette) +
-  # facet_wrap( . ~ cohort, scales = "free") +
   theme_minimal()
 
 data_wide %>% 
   ggplot(aes(age_0)) +
-  geom_histogram() +
-  facet_wrap(.~cohort)
+  geom_histogram()
 
 ## Removes those who died after year of death
-data_long_2014 %>% 
-  filter(between(age_0, 60, 75)) %>%
[…]

### 2021-03-29 02_R/03b_plot_causes_death.R

@@ -24,7 +24,7 @@ count_cancer_death <- data %>%
 
 count_cancer_death %>% 
   filter(category != "Alive") %>% 
-  mutate(cancer = ifelse(cancer == 0, "no history of cancer", "history or incident cancer"),
+  mutate(cancer = ifelse(cancer == 0, "No cancer", "Incident cancer"),
     category = as_factor(category),
          category = fct_reorder(category, prop),
          cancer = as_factor(cancer)) %>% 

### 2021-03-29 02_R/04_crude_net_risks.R
Updated descriptives

@@ -11,17 +11,16 @@ library(magrittr)
 # Import data -------------------------------------------------------------
 source(here::here("02_R", "04b_auxiliary_fx.R"))
 
-data_wide <-
-  import(here::here("01_data", "clean_data", "wide_noltfu.RData"))
-
 data_long <-
   import(here::here("01_data", "clean_data", "dementia_long.RData"))
 
-# 
-data_wide %>% count(cancer, cancer_20, cancer_prev, cancer_inc)
+data_wide <-
+  data_long %>%
+  group_by(id) %>% 
+  slice(n()) %>% 
+  ungroup()
 
 ## Use these numbers for chart
-data_wide %>% count(cancer_20, dementia_20)
 
 data_long %<>% 
   group_by(id) %>% 
@@ -39,28 +38,6 @@ data_long %<>%
                      competing_plr == 1 ~2,
                      TRUE ~ 0))
 
-long_in_wide <- data_long %>% 
-  group_by(id) %>% 
-  slice(n()) %>% 
-  ungroup()
-
-long_in_wide %>% 
-  count(cancer_v, both_outcomes)
-
-# id_a <- data_wide %>% filter(cancer_20 == 1) %>% pull(id)
-# 
-# id_b <- long_in_wide %>% filter(cancer_v == 1) %>% pull(id)
-# 
-# id_dif <- setdiff(id_b, id_a)
-
-# no_cancer <- data_wide %>% 
-#   filter(id %in% id_dif)
-# 
-# no_cancer %>% 
-#   select(id, year, time, end_dementia_death, cancer_date, dementia_date,
-#          end_fup_2015, cancer_v, outcome_plr, status) %>% 
-#   view()
-
 # Weights for death -------------------------------------------------------
 
 death_den <- glm(competing_plr ~ 
@@ -70,7 +47,7 @@ death_den <- glm(competing_plr ~
 
 summary(death_den)
 
-death_den %>% broom::tidy()
+death_den %>% broom::tidy() %>% View()
 
 death_num <- glm(competing_plr ~ 1, data = data_long, family = binomial)
 
@@ -98,7 +75,7 @@ data_long %>% ggplot(aes(x = as_factor(time), y = sw)) +
 # 1.1. Crude risks --------------------------------------------------------
 
 cancer_ever <-
-  survfit(Surv(t2dem_20, as.factor(dementia_20)) ~ cancer_20, data_wide)
[…]

### 2021-03-29 02_R/04_crude_net_risks.R
updated to new data structure (in wide) workflow needs to match analysis plan

@@ -20,8 +20,6 @@ data_wide <-
   slice(n()) %>% 
   ungroup()
 
-## Use these numbers for chart
-
 data_long %<>% 
   group_by(id) %>% 
   mutate(
@@ -38,7 +36,25 @@ data_long %<>%
                      competing_plr == 1 ~2,
                      TRUE ~ 0))
 
-# Weights for death -------------------------------------------------------
+data_wide %<>%
+  mutate(dementia_efu =
+           case_when(outcome_plr == 1 ~ 1,
+                     competing_plr == 1 ~ 2,
+                     TRUE ~ 0))
+
+
+data_wide %<>%
+  mutate(dementia_efu =
+           case_when(outcome_plr == 1 ~ 1,
+                     competing_plr == 1 ~ 2,
+                     TRUE ~ 0))
+
+data_wide %<>%
+  mutate(t2dem_efu =
+           ifelse(t2dem >= 240, 240, t2dem))
+
+
+# 0. Weights for death as competing event of cancer -------------------------------------------------------
 
 death_den <- glm(competing_plr ~ 
                    cancer_v + bs(time,3) + bs(age_0,3) + sex + education + apoe4 + 
@@ -72,10 +88,10 @@ data_long %>% ggplot(aes(x = as_factor(time), y = sw)) +
 
 # 1. EVER VS. NEVER -------------------------------------------------------
 
-# 1.1. Crude risks --------------------------------------------------------
+# 1.1. Crude risks unadjusted for cancer conf --------------------------------------------------------
 
 cancer_ever <-
-  survfit(Surv(t2dem, as.factor(outcome_plr)) ~ cancer_v, data_wide)
+  survfit(Surv(t2dem_efu, as.factor(dementia_efu)) ~ cancer_v, data_wide)
 
 survminer::ggcompetingrisks(cancer_ever) +
   ggthemes::scale_fill_tableau() +
@@ -89,10 +105,10 @@ plot_cif(cancer_ever, "Crude risk")
 
 data_wide %<>% 
   mutate(t2dem_no_death = 
-           ifelse(dementia_20 == 2, 240, t2dem_20))
+           ifelse(dementia_efu == 2, 240, t2dem_efu))
 
 cancer_ever_subcox <-
[…]

### 2021-03-29 02_R/04b_auxiliary_fx.R

@@ -1,4 +1,4 @@
-mytheme <- theme_minimal(base_family = "serif") +
+  mytheme <- theme_minimal(base_family = "serif") +
   theme(
     panel.grid.major.x = element_blank(),
     panel.grid.minor.x = element_blank(),
@@ -33,7 +33,7 @@ plot_cif <- function(model, title, ...){
     filter(state == "1") %>%
     select(time, strata, estimate, conf.high, conf.low) %>%
     rename(CIF = estimate) %>% 
-    mutate(strata = ifelse(strata == label_exp, "Free of cancer", "Ever cancer"))
+    mutate(strata = ifelse(strata == label_exp, "Free of cancer", "Incident cancer"))
   
   plot <- tidy %>% 
     ggplot(aes(time, CIF, group = strata)) +
@@ -65,7 +65,7 @@ plot_km <- function(model, title, ...){
     transmute(
       time = time, 
       strata = ifelse(strata == label_exp, 
-                               "Free of cancer", "Ever cancer"),
+                               "Free of cancer", "Incident cancer"),
       cif = 1 - estimate,
       conf.low2 = 1- conf.low,
       conf.high2 = 1 - conf.high) %>% 

### 2021-03-29 04_reports/manuscript.Rmd
monday sunshine progress

@@
 output: [pdf_document]
-bibliography: cancer_dementia.bib
+
 ---
 ```{r, echo =  FALSE}
-knitr::opts_chunk$set(echo = FALSE, fig.align="center")
+knitr::opts_chunk$set(echo = FALSE, message = FALSE, fig.align="center")
+source(here::here("02_R", "03_descriptives.R"))
 ```
@@
 In contrast, since adult neuronal cells do not divide, Pin1 over-expression in brain tissue is used to restore tau and amyloid precursor protein to a functional shape, thus preventing the buildup of the toxic tau and beta-amyloid species associated with AD.
-
 For this reason Pin1 regulation is proposed as one of the mechanisms that can explain the inverse association between cancer and AD found in observational studies.
@@
 
-Descriptive statistics
+In this section we illustrate how previous DAGS may reflect on the analysis performed to unveil the effect of Pin1 in all cause-dementia.
+We use data collected in the Rotterdam Study, a population-based prospective cohort study among persons living in the Ommoord district in Rotterdam, the Netherlands.
+Recruitment and initial assessments were held between 1990 and 1993, participants had follow-up visits between 1993-1995, 1997-1999, 2002-2005, and 2008-2010.
+All participants had data on incident cancer diagnosis and incident dementia diagnosis through follow-up, collected from medical records of general practitioners (including hospital discharge letters) and through linkage with national registries.
+Date and cause of death was collected on a weekly basis via municipal population registries.
+Data on clinical outcomes was available until 2015.
 
-In a population of x, free of cancer and dementia at baseline...
+To match the analysis to our initial question, we considered as inclusion criteria being between 60 and 70 years old at study entry, without history of cancer diagnosis, free of cognitive decline or with previous history of dementia.
+Out of 7983 persons who participated at baseline, `r total_n` were considered eligible.
+Time to cancer diagnosis, to dementia diagnosis and death status was measured for all participants.
+All participants were followed from study entry until dementia diagnosis, death or 20 years after their individual baseline date, whichever ocurred first.
 
-We first do counts:
+Participants had a mean age of `r mean_age`, and `r women_prop` were women.
+Further details on participants are presented in Table 1.
+Over follow up, `r yes_cancer` developed cancer and `r no_cancer` remain free of cancer diagnosis, the median age of cancer diagnosis was `r cancer_time`.
+From the total sample, `r dem_total` had dementia over follow-up and median time to dementia was `r dem_time`.
+Among participants with incident cancer, `r cancer_dem` had dementia diagnosis and `r cancer_death` died over follow-up, `r cancer_alive` remain alive at 20 years since study entry.
+In contrast, among participants free of cancer diagnosis over follow-up, `r no_cancer_dem` were diagnosed with dementia and `r no_cancer_death` died over follow-up, `r no_cancer_alive` were alive at the end of follow-up.
 
-Time to cancer on average
+We present the following analysis:
 
-We show 3 kinds, ever vs. never, time-varying cancer, and weighting by cancer dx.
+The difference in dementia risk treating cancer as ever vs. never (As if Pin1 and cancer dx. happened at the same time if looked retrospectively).
 
-We now show the risk if U, no U (Combined outcome, crude risk)
-
-We provide additional info on causes of death for both groups, and the extreme scenarios of bounds.
-
-## Discussion
-
-Through this example we describe how much do the direct effect of cancer changes if we measure and adjust for $C$ compared to not.
+The difference in dementia risk considering cancer as a time-varying exposure independent of death.
+For example.
+at five years from follow-up the risk of cancer was `r cancer_5y` and the risk of death was `r death_5y`, thus we assume that those who died would have the same risk of cancer had they not died, unconditionally.
 
@@
 Youthful tissues
+
+## Somewhere
+
+Knowing the cause of death provides information about the direction of missclassification.
+In patients with cancer we observed a slight % that died with dementia as a primary cause.
+On the other hand, among individuals free of cancer, we observed % of individuals who died with cancer as a primary cause.
+
+In cancer you are afraid of missing dementia
+in cancer free you are afraid of missing cancer dx

### 2021-03-29 02_R/03_descriptives.R

@@ -104,7 +104,7 @@ cancer_death_5y <- t2cancer_km %>%
   select(estimate, conf.low, conf.high, state) %>% 
   mutate_if(is.numeric, ~.*100) %>% 
   mutate_if(is.numeric, round, 1) %>%  
-  mutate(est_print = paste0(estimate, " (", conf.low, ", ", conf.high, ")")) %>% 
+  mutate(est_print = paste0(estimate, " (95%CI: ", conf.low, ", ", conf.high, ")")) %>% 
   select(est_print, state)
 
 death_5y <- cancer_death_5y %>% filter(state == 2) %>% pull(1)

### 2021-03-31 02_R/01_wide.R
changed inclusion criteria and missingness indicators

@@ -15,14 +15,13 @@ data <- data %>%
   filter(dementia_inc != 2) %>% 
   filter(dementia_at_risk == 1) %>% 
   filter(cancer_prev != 1) %>% # Based on the ideal trial
-  # filter(!is.na(education_three_levels)) %>% 
-  # filter(!is.na(smoke1)) %>% 
   filter(!is.na(bmi1)) %>%
   filter(!is.na(sbp1)) %>%
-  filter(!is.na(ht_drug1)) %>%
+  # filter(!is.na(oh1)) %>%
+  filter(!is.na(smoke1)) %>%
   filter(!is.na(censor_date)) %>% 
   filter(between(age_0, 60, 70)) %>% 
-  filter(cohort == 1)
+  filter(cohort != 3)
 
 #KIM: SELECT ONLY THOSE WITH IC BASED ON STROKE DATASET
 data <- subset(data,data$ic_ok_2016==1)
@@ -153,8 +152,11 @@ data %<>%
   mutate(
     education = ifelse(is.na(education), "Unknown", education),
     apoe4 = ifelse(is.na(apoe4),"Unknown" , apoe4),
+    smoke1 = ifelse(is.na(smoke1),3 , smoke1),
     ht_drug1 = ifelse(is.na(ht_drug1), 2, ht_drug1),
-    hd_prev = ifelse(is.na(hd_prev), 2, hd_prev),)
+    hd_prev = ifelse(is.na(hd_prev), 2, hd_prev),
+    diabetes_prev = ifelse(is.na(diabetes_prev), 2, diabetes_prev))
+
 
 # Wide truncated 20 years -------------------------------------------------
 
@@ -215,3 +217,4 @@ data %>%
   filter(t2dem <1) %>% 
   select(cancer_prev, cancer_inc,cancer, death_2015,
                               dementia, t2cancer, t2dem, t2death)
+

### 2021-03-31 02_R/03_descriptives.R

@@ -1,5 +1,6 @@
 library(survival)
 library(tidyverse)
+library(tableone)
 
 data_long <-
   rio::import(here::here("01_data", "clean_data", "dementia_long.RData"))
@@ -9,6 +10,17 @@ data_wide <-
   group_by(id) %>% 
   slice(max(row_number())) %>% 
   ungroup()
+
+# merge with baseline covariates ------------------------------------------
+
+baseline <-
+  rio::import(here::here("01_data", "clean_data", "wide_noltfu.RData")) %>% 
+  select(id, contains("1"))
+
+data_wide <- data_wide %>% 
+  left_join(baseline)
+
+rio::export(data_wide, here::here("01_data", "clean_data", "wide_after_truncation.RData"))
 
 ## Total n
 total_n <- dim(data_wide)[1]
@@ -82,7 +94,22 @@ women_prop <- data_wide %>%
   mutate(total = paste0(prop, "%", " (n = ", n, ")")) %>% 
   filter(sex == 1) %>% 
   pull(total)
-  
+
+## Tableone
+
+myvars <- c("sex", "age_0", "education", "apoe4", "smoke1", "bmi1", "oh1", "sbp1", 
+          "ht1","hd_prev", "hd_v", "diabetes_prev", "diab_v",
+          "stroke_prev", "stroke_v", "cancer_v") 
+num <- c("age_0","bmi1", "oh1", "sbp1" )
+
+cat <- myvars[!myvars %in% num]
+
+tableone <- CreateTableOne(vars = myvars, data = data_wide, factorVars = cat, strata = "cancer_v")
+
+tableone
+
+summary(tableone)
+
 # survminer::ggcompetingrisks(t2cancer_km) +
 #   ggthemes::scale_fill_tableau() +
 #   theme_minimal() +

### 2021-03-31 02_R/04_crude_net_risks.R
this is a draft I will have to erase

@@ -42,23 +42,16 @@ data_wide %<>%
                      competing_plr == 1 ~ 2,
                      TRUE ~ 0))
 
-
-data_wide %<>%
-  mutate(dementia_efu =
-           case_when(outcome_plr == 1 ~ 1,
-                     competing_plr == 1 ~ 2,
-                     TRUE ~ 0))
-
 data_wide %<>%
   mutate(t2dem_efu =
            ifelse(t2dem >= 240, 240, t2dem))
 
 
-# 0. Weights for death as competing event of cancer -------------------------------------------------------
+# 0. Weights for death as competing event of dementia -------------------------------------------------------
 
 death_den <- glm(competing_plr ~ 
                    cancer_v + bs(time,3) + bs(age_0,3) + sex + education + apoe4 + 
-                   bs(sbp,3) + bs(bmi,3 ) + hd_v + stroke_v + diab_v, 
+                   bs(sbp,3) + bs(bmi,3 ) = bs(oh,3 )+ hd_v + stroke_v + diab_v, 
                  data = data_long, family = quasibinomial)
 
 summary(death_den)
@@ -73,7 +66,7 @@ data_long$p_num = predict(death_num, data_long, type = "response")
 
 data_long %<>%
   group_by(id) %>%
-  mutate(sw = cumprod(1 - p_num) / cumprod(1 - p_denom),
+  mutate(w = cumprod(1 - p_num) / cumprod(1 - p_denom),
   ) %>% 
   ungroup()
 
@@ -108,12 +101,12 @@ data_wide %<>%
            ifelse(dementia_efu == 2, 240, t2dem_efu))
 
 cancer_ever_subcox <-
-  coxph(Surv(t2dem_no_death, dementia_efu == 1) ~ cancer_v, data_wide)
+  coxph(Surv(t2dem_no_death, dementia_efu == 1) ~ cancer_v + bs(age_0,3) + sex + education + apoe4, data_wide)
 
 ## risks and hazards
 tidy(cancer_ever_subcox, exponentiate = TRUE)
 
-risks_cif(cancer_ever)
+# risks_km(cancer_ever_subcox, strata =  "cancer_v")
 
 # 1.2. Net risks ----------------------------------------------------------
 
@@ -133,29 +126,47 @@ tidy(cancer_ever_cox, exponentiate = TRUE)
 risks_km(cancer_ever_km)
 
 
-# 1.2.b. Independent censoring conditional on tv covariates ---------------
[…]

### 2021-03-31 02_R/04_direct_effect.R
new version of results

@@ -1,0 +1,424 @@
+library(ggplot2)
+library(tidyverse)
+library(rio)
+library(wesanderson)
+library(survival)
+library(survminer)
+library(broom)
+library(splines)
+library(magrittr)
+library(WeightIt)
+library(cobalt)
+
+# Import data -------------------------------------------------------------
+source(here::here("02_R", "04b_auxiliary_fx.R"))
+
+data_long <-
+  import(here::here("01_data", "clean_data", "dementia_long.RData"))
+
+data_wide <-
+  import(here::here("01_data", "clean_data", "wide_after_truncation.RData"))
+
+data_long %<>% 
+  group_by(id) %>% 
+  mutate(
+    fuptime = time - 1,
+    tstart = lag(fuptime),
+    tstart = ifelse(is.na(tstart), -1, tstart)) %>% 
+  ungroup()
+
+## idea of how numbers look like over follow-up
+
+data_long %<>% 
+  mutate(both_outcomes = 
+           case_when(outcome_plr == 1 ~ 1,
+                     competing_plr == 1 ~2,
+                     TRUE ~ 0))
+
+data_wide %<>%
+  mutate(dementia_efu =
+           case_when(outcome_plr == 1 ~ 1,
+                     competing_plr == 1 ~ 2,
+                     TRUE ~ 0))
+
+data_wide %<>%
+  mutate(t2dem_efu =
+           ifelse(t2dem >= 240, 240, t2dem))
+
+
+# 1. Cancer Ever vs. Never ---------------------------------------------------
+
+
+# 1.1. IPTW (cancer ever vs. never) ------------------------------------------
+
+# cancer_den <-
+#   glm(
+#     cancer_v ~ bs(age_0) + sex + education + apoe4 + as.factor(smoke1),
+#     data = data_wide,
+#     family = binomial
+#   )
+# 
+# summary(cancer_den)
+# 
+# cancer_num <- glm(cancer_v ~ 1, data = data_wide)
+# 
+# summary(cancer_num)
+# 
+# data_wide <- data_wide %>%
+#   mutate(
+#     p_num = predict(cancer_num, type = "response"),
+#     p_denom = predict(cancer_den, type = "response"),
+#     w_cancer = ifelse(cancer_v == 1, p_num/p_denom, (1 - p_num)/(1- p_denom)))
+
+
+## Check standardized mean
+
[…]

### 2021-03-31 02_R/04b_auxiliary_fx.R

@@ -101,3 +101,8 @@ risks_km <- function(model){
            rr = .[[2]] / .[[1]]) %>% 
     mutate_at(c(1:3), ~.*100)} 
 
+tidy_hr <- function(model) {
+ model %>% 
+  broom::tidy(exponentiate = TRUE, conf.int = TRUE) %>%
+    select(term, estimate, contains("conf"))
+}

### 2021-03-31 04_reports/manuscript.Rmd
added ideas from melodem abstract, still rough but clearer aim (I hope)

@@
 For this reason Pin1 regulation is proposed as one of the mechanisms that can explain the inverse association between cancer and AD found in observational studies.
+
+Bias cannot be defined or assessed if the causal question is not explicitly specified.
+To define the underlying causal question there are two core challenges.
+First, specifying cancer diagnosis as the proxy for an underlying biological mechanism we would ideally intervene upon.
+
+Second, to rely on the independent censoring assumption in regards to the competing event of death when the inrestes is on the direct effect of cancer in dementia risk.
+
+With these considerations in mind, and illustrated through directed acyclic graphs, we describe a case study using data from the Rotterdam Study, a population-based prospective cohort with over 20 years of follow up.
+Through this case study we describe how can we connect the causal question to the analytic decisions with real-world data and demonstrate how point estimates may vary substantially depending on the assumptions related to previous points.
+
+How information on mortality and cause of death can provides insight about the direction of some sources of bias.
 
@@
 Time to cancer diagnosis, to dementia diagnosis and death status was measured for all participants.
-All participants were followed from study entry until dementia diagnosis, death or 20 years after their individual baseline date, whichever ocurred first.
+All participants were followed from study entry until dementia diagnosis, death or 20 years after their individual baseline date, whichever occurred first.
 
@@
 The difference in dementia risk treating cancer as ever vs. never (As if Pin1 and cancer dx. happened at the same time if looked retrospectively).
+
+We begin illustrating the most simple scenario (as Figure 1), were Pin1 overexpression is defined as "ever vs.
+never" and we consider death as an independent censoring mechanism.
+We consider baseline covariates for adjustment such as age at study entry, sex, educational attainment, smoking status, alcohol intake status.
 
@@
 in cancer free you are afraid of missing cancer dx
+
+Make a table of definition of weights based on cancer def. and

### 2021-03-31 04_reports/manuscript.Rmd
updated manuscript, prints kableextra

@@
 ```{r, echo =  FALSE}
-knitr::opts_chunk$set(echo = FALSE, message = FALSE, fig.align="center")
+knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.align="center")
+library(survival)
+library(tidyverse)
+library(tableone)
 source(here::here("02_R", "03_descriptives.R"))
@@
 Make a table of definition of weights based on cancer def. and
+
+## Table 1
+
+```{r}
+tableone_exp <-
+  print(
+    tableone,
+    explain = TRUE,
+    test = FALSE,
+    quote = FALSE,
+    noSpaces = TRUE,
+    printToggle = FALSE
+  )
+
+kableExtra::kable(tableone_exp, booktabs = T) %>% 
+  kableExtra::kable_styling(latex_options = c("striped", "hold_position"))
+
+```

### 2021-03-31 02_R/03_descriptives.R

@@ -1,7 +1,3 @@
-library(survival)
-library(tidyverse)
-library(tableone)
-
 data_long <-
   rio::import(here::here("01_data", "clean_data", "dementia_long.RData"))
 
@@ -18,7 +14,7 @@ baseline <-
   select(id, contains("1"))
 
 data_wide <- data_wide %>% 
-  left_join(baseline)
+  left_join(baseline, by = c("id", "death_2015", "end_fup_2015", "e1"))
 
 rio::export(data_wide, here::here("01_data", "clean_data", "wide_after_truncation.RData"))
 
@@ -106,9 +102,7 @@ cat <- myvars[!myvars %in% num]
 
 tableone <- CreateTableOne(vars = myvars, data = data_wide, factorVars = cat, strata = "cancer_v")
 
-tableone
-
-summary(tableone)
+# summary(tableone)
 
 # survminer::ggcompetingrisks(t2cancer_km) +
 #   ggthemes::scale_fill_tableau() +

### 2021-04-12 04_reports/manuscript.Rmd
Intro check, methodological considerations check, missing about bounds, lay out results and discussion

--- 2021-03-29 04_reports/manuscript.Rmd
@@
 ---
-```{r, echo =  FALSE}
-knitr::opts_chunk$set(echo = FALSE, message = FALSE, fig.align="center")
+```{r, echo =  FALSE, message = FALSE, warning = FALSE}
+knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.align="center")
+library(survival)
+library(tidyverse)
+library(tableone)
 source(here::here("02_R", "03_descriptives.R"))
@@
 
-### Motivation
+## Introduction
 
-Several observational studies have found that individuals with history or incident cancer that survive over follow-up have a lower risk of Alzheimer´s disease or related dementias (ADRD) when compared to individuals with no history of cancer.
-Multiple mechanisms related to inflammatory response, immune response, etc etc are postulated to understand this phenomenon.
-However, the community has raised their concerned about the potential unmeasured confounding, selective survival and misclassification issues that could drive this results.
+There is significant interest in understanding the association between cancer and dementia1.
+Many observational studies have consistently determined that individuals with a history or incident cancer have a lower risk of developing of Alzheimer´s disease or related dementias (ADRD) when compared to individuals with no history of cancer.
+These findings have motivated the identification of molecular and genetic mechanisms that explain this association.
+Specific molecular mechanisms, such as Pin1 enzyme over-expression promotes tumorigenesis, while it’s down-regulation is attributed to mechanisms that contribute to Alzheimer’s Disease.
+These findings are suggest new opportunities to repurpose cancer chemotherapeutics for Alzheimer´s Disease.
 
-In the lasts few years there has been an important attention to unveiling the impact of death as a competing event, and the role of survival bias to explain this association.
-Previous studies alluded to choosing a particular summary measure or modeling approach to account for death (incidence rate ratios (ref), hazard ratios (ref), multistate models (ref)), though the underlying estimand remains unclear.
+Nevertheless, observational studies are susceptible to several bias, and researchers have raised concerns related to the competing event of death as well as unmeasured confounding and ascertainment error  that could drive these results.
+Though, to understand sources of bias the causal question should be explicitly  specified.
+This particular question has at least two core challenges.
+First, defining the underlying biological mechanism we would ideally intervene upon and defining cancer diagnosis as the proxy for that mechanism.
+For example, researchers refer to biological mechanisms which onset may be prior to cancer diagnosis and to the effect of cancer treatment, both relate to different causal questions and would be translated into different study design and analytic choices.
+Second, defining the causal question (or estimand) of interest when the competing event of death is interest.
+Previous studies alluded to choosing a particular summary measure or modeling approach to account for death (incidence rate ratios (ref), hazard ratios (ref), multistate models (ref)).
+However, when competing events are present, there is more than one causal question that can be framed and each question has different interpretations and relies on different assumptions.
 
-Overall, although observational studies have measured history of cancer and incident cancer as the key exposure, researchers refer to biological mechanisms which onset may be prior to cancer diagnosis and the effect of cancer treatment.
-
-Pin1 is an enzime deeply involved in the pathogenesis of cancer and ADRD.
-One of the roles of Pin1 is regulate mitosis process and apoptosis.
-Over-expression of Pin1 has been found in several tumors and it's inhibition causes tumor regression.
-In contrast, since adult neuronal cells do not divide, Pin1 over-expression in brain tissue is used to restore tau and amyloid precursor protein to a functional shape, thus preventing the buildup of the toxic tau and beta-amyloid species associated with AD.
-For this reason Pin1 regulation is proposed as one of the mechanisms that can explain the inverse association between cancer and AD found in observational studies.
-
-Our aim is to outline one way to tie the etiological question of Pin1 with the questions that can be answered with observational data where Pin1 measurements are not available.
-We aim to draw the complexity of the research question by building progressively a directed acyclic graph (DAG) which help us connect the causal question to the observable data and the assumptions we rely on to identify the effect.
-We describe the challenges and illustrate the assumptions we have to fulfill to identify the effect.
+With these considerations in mind, we outline a case study focused on the question: _What is the direct effect of Pin1 over-expression at late-midlife in the risk of ADRD after 20 years of follow-up compared to having regulated expression of Pin1._ To identify this effect with real-world data, we illustrate the complexity of the research question by building progressively a directed acyclic graph (DAG) which help us connect the causal question to the observable data and the assumptions we rely on to identify the effect.
+We exemplify different scenarios with data collected from the Rotterdam Study, a population-based cohort study.
+We describe the challenges and how they translate into the analytic decisions.
+Last, we discuss how information on mortality and cause of death can provides insight about the direction of some sources of bias.
 
@@
 
-Given the previous literature on the topic, we focus the interest in answering the following question: _What is the direct effect of Pin1 over-expression at age of 65 in the risk of ADRD by age 85 compared to regulated expression of Pin1 by age of 65._ To identify this effect with real-world data, we will illustrate the causal question through a directed acyclic graphs (DAGs), spelling out the data generating mechanisms and assumptions tied to the data.
+We begin by illustrating a hypothetical scenario through the causal graph of Figure 1, that depicts a situation where Pin1, represented as $P$, was measured in all participants in late midlife (for example between age 50 and 60).
+Based on the measurement of Pin1, participant's exposure is classified as 1) Over-expressed Pin1 or 2) Regulated Pin1.
+Although Pin1 expression acts as a biomarker and is not a well-defined intervention, we can allude that sometime in the future people with over-expressed Pin1 could follow a specific treatment that lowers Pin1 under a specified threshold.
+Furthermore, through out the follow-up each participant followed a strict monitoring for ADRD onset, represented as $Y_{20}$, and no participant was lost to follow-up nor died during the study period (no competing events present).
 
-The causal graph of Figure 1 illustrates a situation where Pin1, represented as $P$, was measured in all participants on their 65 birthday.
-Based on the measurement of Pin1, participant's exposure was classified as 1) Over-expressed Pin1 or 2) Regulated Pin1.
-Through out the follow-up each participant followed a strict monitoring for ADRD onset, represented as $Y_{85}$, and no participant was lost to follow-up nor died during the study period.
-
-was not an exposure or intervention randomized at age 65, therefore we have to consider a set of measured covariates $L$ for which conditional exchangeability holds.
+Since Pin1 expression was not a randomized intervention, we consider a set of measured covariates $L$ for which conditional exchangeability holds.
 Previous studies have described age, sex, educational level and race/ethnicity as the minimal adjusting set of covariates (Ospina).
 However, environmental and behavioral factors such as smoking, alcohol intake, or chronic diseases which are known to cause microenvironmental changes such as inflammation and changes in tissue remodelling, may translate into Pin1 over-expression and are also related to the development of ADRD.
-Therefore, had we measured all $L$ we could identify the direct of $P$ in $Y_{85}$ by conditioning on $L$.
+Therefore, had we measured all $L$ we could identify the effect of $P$ in $Y_{20}$ by conditioning on $L$.
 
@@
 \node [below of = p, yshift = 1cm, xshift = -1cm](l) {$L$};
-\node [right of = p] (y) {$Y_{85}$};
+\node [right of = p] (y) {$Y_{20}$};
 \draw[->] (l) -- (p);
@@
 
-Unfortunately, Pin1 expression measurement is not available in real-world settings, thus we can only rely on a proxy of this exposure.
+Unfortunately, Pin1 expression measurement is not available at the moment in real-world settings, we can only rely on a proxy of this exposure.
 Since Pin1 over-expression is present in tumors and tumors are only measured through screening and diagnosis, we considered cancer diagnosis as the proxy for Pin1 over-expression, as most observational studies have defined (all refs).
 We depict this feature in Figure 2, where $P^*$ represents _incident cancer diagnosis_, the measured proxy of $P$.
-In this DAG we colored the path $P^* \leftarrow P \rightarrow Y_{85}$ because although we are measuring the association between $P^*$ and $Y_85$ in the observed data, we are assuming that the effect is only through $P$.
+In this DAG we colored the path $P^* \leftarrow P \rightarrow Y_{85}$ because although we are measuring the association between $P^*$ and $Y_{20}$ in the observed data, we are assuming that the captured effect is only through $P$.
 
@@
 A major challenge related to assigning cancer diagnosis as the proxy of Pin1 is defining time zero.
-Not everyone who had Pin1 over-expression will be diagnosed with cancer by age 65.
-For this reason prospective cohort studies have considered cancer as a time-varying or time-dependent exposure.
-This means that a participant who had cancer diagnosis at age 70 contributes to the "regulated Pin1" from age 65 to 70, and from age 70 and onwards to the "Over-expressed Pin1" arm.
-Alternatively, studies that have included participants with cancer diagnosis at the time of the diagnosis (for example from cancer registries such as SEER) and matched participants by age.
-If age at cancer diagnosis is considered as the time zero, we should be careful of adjusting for post-baseline covariates of $P$ or mediators between $P$ and $P^*$.
+Not everyone who had Pin1 over-expression will be diagnosed with cancer by late-midlife, but in late-life.
+For this reason prospective cohort studies have considered cancer diagnosis as a time-varying or time-dependent exposure.
+This means that a participant within a prospective cohort study contributes to the "regulated Pin1" since study entry up to the time of cancer diagnosis and later on to the "Over-expressed Pin1" arm.
+Other studies have included participants with cancer diagnosis at the time of the diagnosis (for example from cancer registries such as SEER) and matched participants by age.
+In both cases we must remember that the the main interest is on the unmeasured $P$ and we should only adjust for covariates prior to $P$, and be careful to adjust for post-baseline covariates of $P$ or mediators between $P$ and $P^*$.
 
-Since the population is elderly, the risk of death at age 65 is not trivial and it increases over time.
-This means that some individuals may die prior to cancer diagnosis due to other leading causes of death at this age group, such as cardiovascular disease.
-As such, we can only measure $P^*$ in the subset of individuals who have survived long enough to have a cancer diagnosis.
+One of the mediators between $P$ and $P^*$ is death prior to cancer diagnosis.
+Since individuals are at risk of dying from other causes (such as cardivascular death), we can only measure $P^*$ in the subset of individuals who have survived long enough to have a cancer diagnosis.
 We zoom in the relationship of $P \rightarrow P^*$ in Figure 3.
-In this figure we include
-$D = 0$ between $P$ and $P^*$, this illustrates that to observe $P^*$, we condition (box around $D = )$ on surviving long enough to have a cancer diagnosis.
-
-Several risk factors that increase the risk of cancer might also be shared causes of death prior to cancer diagnosis, for example smoking may cause lung cancer and chronic obstructive pulmonary disease (a leading cause of death in this age group).
-Therefore, to isolate the effect of $P$ in $P^*$ (the violet arrow), we need to block all shared causes of Pin1 and death from other causes, and the shared causes of death from other causes and cancer diagnosis, both represented in Figure 3 as $C_1$ and $C_2$.
+In this DAG we include
+$D = 0$ between $P$ and $P^*$, this illustrates that to observe $P^*$, we condition (box around $D = 0$) on surviving long enough to have a cancer diagnosis.
+Several risk factors that increase the risk of cancer might also cause death prior to cancer diagnosis, for example smoking may cause lung cancer and chronic obstructive pulmonary disease (a leading cause of death in this age group).
+Therefore, to isolate the effect of $P$ in $P^*$ (the violet arrow), we need to block all shared causes between $P$ and $D$ and between $D$ and $P*$ both represented in Figure 3 as $C_1$ and $C_2$.
 In this way we assume a hypothetical scenario in which we could prevent death prior to cancer diagnosis, by conditioning on a rich set of covariates.
@@
 
-As illustrated in previous figures, we are interested in the direct effect of Pin1 in ADRD at 85 years, which means a length of follow-up of 20 years.
-However, mortality in elderly population between 65 and 85 is substancial, and a leading cause of death is cancer.
-This means that in order to measure ADRD by the age of 85, individuals have to had survive to cancer and to other causes.
+As illustrated in previous figures, we are interested in the effect of Pin1 in late midlife in ADRD risk after 20 years of follow-up.
+Nevertheless, mortality is substantial and increases steeply in late life, and a leading cause of death is cancer.
+This means that in order to measure ADRD over follow-up individuals have to had survive to cancer and to other leading causes of death.
 For this reason death is a competing event of ADRD because if a participant dies prior to ADRD diagnosis, death prevents from observing ADRD at future time-points.
+We can visualize this data feature in the causal diagram in Figure 4.
+In this DAG we include $D_{15}$ as a representation of death at 15 years of follow-up, an arrow between $P$ and $D_{15}$ since $P$ may increase the mortality risk irrespective of cancer diagnosis.
+We also include an arrow between $P^*$ and $D_{15}$ since cancer diagnosis, and subsequent treatment (or lack of treatment) may have an effect on death.
+Last, the arrow from $D_{15}$ and $Y_{20}$ represents the key feature of a competing events data structure: an individual who dies at 15 years of follow-up cannot subsequently develop ADRD, and since $D_{15}$ and $Y_{20}$ are events related to aging, $C$ represent the shared causes of both events.
 
-We can visualize this data feature in the causal diagram in Figure 4.
-In this DAG we include $D_{80}$ as a representation of Death by age 80, an arrow between $P$ and $D_{80}$ since $P$ may increase the mortality risk irrespective of cancer diagnosis.
-We also include an arrow between $P^*$ and $D_{80}$ since cancer diagnosis, and subsequent treatment (or lack of treatment) may have an effect on death.
-Last, the arrow from $D_{80}$ and $Y_{85}$ represents the key feature of a competing events data structure: an individual who dies by age 80 cannot subsequently develop ADRD, and since $D_{80}$ and $Y_{85}$ are events related to aging, $C$ represent the shared causes of both events.
-
-In this DAG we observe that, had we measured and adjusted for $L$, we could estimate the total effect of $P$ in $Y_{85}$ without further assumptions.
-However the total effect includes all pathways between $P$ and $Y_{85}$.
-This means that if Pin1 has an effect on mortality through cancer or other mechanisms, we will observe a protective effect of Pin1 overexpression in ADRD, partially or fully mediated through death.
-
-Alternatively, there are several ways to conceptualize the direct effect of $P$ in $Y_{85}$ as in Figure 4 (violet arrows) and several causal parameters (estimands) to represent it, such as: the controlled direct effect (CDE), the survivor average causal effect and the natural separable direct effect.
-Through this section we will discuss the causal question, assumptions related to death and interpretation of the CDE since this estimand connects to frequently used estimators.
+In this DAG we observe that, had we measured and adjusted for $L$, we could estimate the total effect of $P$ in $Y_{20}$ without further assumptions.
+However the total effect includes all pathways between $P$ and $Y_{20}$.
+This means that if Pin1 has an effect on mortality through cancer or other mechanisms, we will observe a protective effect of Pin1 over-expression in ADRD, partially or fully mediated through death.
+For this reason estimators that are used to approximate a total effect (JY2020), such as the Aalen-Johansen estimator (underlying estimator for multistate models) or subdistribution hazard models, may not be useful in this setting.
+Since the question of interest is focused on the direct effect of $P$ in $Y_{20}$ as in Figure 4 (violet arrows), there are several causal questions (estimands) to represent this direct effect, such as: the controlled direct effect (CDE), the survivor average causal effect (SACE) and the natural separable direct effect.
+In this section we discuss the controlled direct effect as the causal question of interest since it translates to frequently used estimators in this literature (such as Kaplan-Meier estimator and Cox-proportional hazard model).
 
@@
 
+Since the conditional independent censoring assumption is untestable, when there are no other mechanisms of censoring, we can explore the extreme scenarios and estimate the bounds.
+
 ## Motivating example with the RS data
@@
 
-## Somewhere
-
 Knowing the cause of death provides information about the direction of missclassification.
@@
 in cancer free you are afraid of missing cancer dx
+
+Make a table of definition of weights based on cancer def. and
+
+Explicitly outlining the estimands and the assumptions that connect the causal question to the observed data provide an opportunity to improve the design of observational studies and the interpretation of their findings, plus better insight of potential sources of bias.
+This is a crucial since these studies are providing insights that are guiding other fields of research in the area, from bench science to biostatistics and epidemiological methods15.
+To guide future observational and methodologic studies, our aim is to outline the potential estimands of interest that represent the underlying causal questions behind the association between cancer and dementia
+
+## Table 1
+
+```{r}
+tableone_exp <-
+  print(
+    tableone,
+    explain = TRUE,
+    test = FALSE,
+    quote = FALSE,
+    noSpaces = TRUE,
+    printToggle = FALSE
+  )
+
+kableExtra::kable(tableone_exp, booktabs = T) %>% 
+  kableExtra::kable_styling(latex_options = c("striped", "hold_position"))
+
+```

### 2021-04-12 04_reports/manuscript.Rmd
New version looking goood....

@@
 source(here::here("02_R", "03_descriptives.R"))
+source(here::here("02_R", "05_plots.R"))
+source(here::here("02_R", "05b_plot_causes_death.R"))
 ```
 
-## Introduction
+## 1. Introduction
 
@@
 
-## Overview of the causal structure and assumptions related to cancer diagnosis as a proxy for Pin1
+## 2. Overview of the causal structure and assumptions related to cancer diagnosis as a proxy for Pin1
 
@@
 
-## Overview of the causal question and assumptions related to death as a competing event of ADRD
+## 3. Overview of the causal question and assumptions related to death as a competing event of ADRD
 
@@
 
-## Motivating example with the RS data
+## 4. Motivating example with the RS data
 
@@
 We use data collected in the Rotterdam Study, a population-based prospective cohort study among persons living in the Ommoord district in Rotterdam, the Netherlands.
-Recruitment and initial assessments were held between 1990 and 1993, participants had follow-up visits between 1993-1995, 1997-1999, 2002-2005, and 2008-2010.
+Recruitment and initial assessments were held between 1990 and 1993; it was later extended between 2000 and 2001 consisting of individuals who had reached the age of 55 years or who had moved into the study area.
+Participants from first subcohort had follow-up visits between 1993-1995, 1997-1999, 2002-2005, and 2008-2010, second subcohort had follow-up visits between 2004 and 2005, and between 2011 and 2012.
 All participants had data on incident cancer diagnosis and incident dementia diagnosis through follow-up, collected from medical records of general practitioners (including hospital discharge letters) and through linkage with national registries.
@@
 To match the analysis to our initial question, we considered as inclusion criteria being between 60 and 70 years old at study entry, without history of cancer diagnosis, free of cognitive decline or with previous history of dementia.
-Out of 7983 persons who participated at baseline, `r total_n` were considered eligible.
-Time to cancer diagnosis, to dementia diagnosis and death status was measured for all participants.
+Out of 10998 persons who participated at study entry, `r total_n` were considered eligible.
+Time to cancer diagnosis, time to dementia diagnosis and death status was measured for all participants.
 All participants were followed from study entry until dementia diagnosis, death or 20 years after their individual baseline date, whichever occurred first.
+
+### 4.1. Methods
+
+First we illustrate the most simple scenario (as Figure 1), were Pin1 over-expression is defined as "cancer ever vs.
+never" as if Pin1 and cancer diagnosis happened at the same time if looked retrospectively.
+Second, we consider cancer as a time-varying exposure independent of death.
+This can be represented as follows, at five years from study entry the observed unadjusted risk of cancer was `r cancer_5y` and the risk of death was `r death_5y` for the entire cohort, thus we assume that those who died would have the same risk of cancer had they not died, unconditionally, over follow-up.
+Third, we consider time to cancer as the
+
+To address confounding we fit inverse probability treatment weights including age at study entry, sex, educational attainment, smoking status and cohort as covariates.
+
+To identify the controlled direct effect we considered death as an independent censoring mechanism conditional on time-varying shared causes of death and dementia.
+
+The difference in dementia risk considering cancer as a time-varying exposure independent of death.
+For example.
+at five years from follow-up the risk of cancer was `r cancer_5y` and the risk of death was `r death_5y`, thus we assume that those who died would have the same risk of cancer had they not died, unconditionally.
+
+### 4.2. Results
 
@@
 
-We present the following analysis:
-
-The difference in dementia risk treating cancer as ever vs. never (As if Pin1 and cancer dx. happened at the same time if looked retrospectively).
-
-We begin illustrating the most simple scenario (as Figure 1), were Pin1 overexpression is defined as "ever vs.
-never" and we consider death as an independent censoring mechanism.
-We consider baseline covariates for adjustment such as age at study entry, sex, educational attainment, smoking status, alcohol intake status.
-
-The difference in dementia risk considering cancer as a time-varying exposure independent of death.
-For example.
-at five years from follow-up the risk of cancer was `r cancer_5y` and the risk of death was `r death_5y`, thus we assume that those who died would have the same risk of cancer had they not died, unconditionally.
-
-## Discussion
+## 5. Discussion
 
@@
 
-## Table 1
+\newpage
+
+## Tables
+
+### Table 1: Descriptive characteristics of individuals who had a cancer diagnosis and of those free of cancer diagnosis over follow up.
 
@@
 ```
+
+\newpage
+
+### Table 2. Risk difference and risk ratio for the risk of dementia
+
+\newpage
+
+## Figures
+
+### Figure 1: Distribution of participants under each health status, by age over follow-up
+
+```{r}
+plot_distribution
+```
+
+### Figure 2. Causes of death for participants with and without incident cancer
+
+```{r}
+causes_death
+```

### 2021-04-12 02_R/05_plots.R

@@ -1,10 +1,10 @@
 ## Plots 2021
-library(ggplot2)
 library(tidyverse)
 library(rio)
-library(wesanderson)
 library(magrittr)
-palette <- wes_palette("Moonrise3")
+# palette <- wes_palette("Moonrise3")
+safe_pal <- rcartocolor::carto_pal(12, "Safe")
+# ggthemes::scale_fill_colorblind()
 
 ## Import dataset that keeps deads until 2014
 
@@ -38,7 +38,7 @@ data_dem <- import(here::here("01_data", "clean_data", "dementia_long.RData"))
 ## Plot distribution of events over time
 
 ## Keeps those who died
-data_long %>% 
+plot_distribution <- data_long %>% 
   group_by(age_scale) %>% 
   count(status) %>%
   mutate(prop = round(100*n/sum(n)),0) %>% 
@@ -46,59 +46,72 @@ data_long %>%
   filter(age_scale <= 90) %>% 
   ggplot(aes(age_scale, prop, fill = status)) +
   geom_area() + 
-  scale_fill_manual(values = palette) +
-  theme_minimal()
-
-data_wide %>% 
-  ggplot(aes(age_0)) +
-  geom_histogram()
-
-## Removes those who died after year of death
-data_long %>% 
-  filter(year <= year(end_fup_2015)) %>%
-  group_by(age_scale) %>% 
-  count(status) %>%
-  mutate(prop = round(100*n/sum(n))) %>% 
-  ungroup() %>%
-  filter(age_scale <= 90) %>% 
-  ggplot(aes(age_scale, prop, fill = status)) +
-  geom_area() + 
-  scale_fill_manual(values = palette) +
-  theme_minimal()
-
-library(lubridate)
-
-## Removes those who had dementia after year of dementia dx
-data_dem %>% 
-  group_by(age_scale) %>% 
-  count(status) %>%
-  mutate(prop = round(100*n/sum(n))) %>% 
-  ungroup() %>% 
-  ggplot(aes(age_scale, prop, fill = status)) +
-  geom_area() + 
-  scale_fill_manual(values = palette) +
-  theme_minimal()
-
-##### Plot longitudinal distribution over time
-
-data_id <- data_long %>%ungroup() %>% count(id) %>% sample_n(100) %>% pull(id)
-
-data_long %>%
-  group_by(id) %>% 
-  mutate(age_death = max(age_scale)) %>% 
-  ungroup() %>% 
-  # filter(id %in% data_id) %>% 
[…]

### 2021-04-12 02_R/05b_plot_causes_death.R

@@ -1,6 +1,6 @@
 library(rio)
 library(tidyverse)
-library(wesanderson)
+
 ggthemes::scale_fill_colorblind()
 
 wide_data <- import(here::here("01_data", "clean_data", "wide_noltfu.RData"))
@@ -10,11 +10,11 @@ cases_death <- import(here::here("01_data", "raw_data", "causes_death.RData"))
 data <- wide_data %>% 
   left_join(cases_death, by = c("id" = "ergoid"))
 
-data %>%
-  group_by(cancer) %>%
-  count(dementia) %>% 
-mutate(prop = paste0(round(100*(n/sum(n)),0), "%")) %>% 
-  gt::gt()
+# data %>%
+#   group_by(cancer) %>%
+#   count(dementia) %>% 
+# mutate(prop = paste0(round(100*(n/sum(n)),0), "%")) %>% 
+#   gt::gt()
 
 count_cancer_death <- data %>% 
   filter(cohort == 1, dementia == 2) %>%
@@ -22,8 +22,9 @@ count_cancer_death <- data %>%
   count(category) %>% 
   mutate(prop = 100*(n/sum(n))) 
 
-count_cancer_death %>% 
+causes_death <- count_cancer_death %>% 
   filter(category != "Alive") %>% 
+  mutate(category = str_to_title(category)) %>% 
   mutate(cancer = ifelse(cancer == 0, "No cancer", "Incident cancer"),
     category = as_factor(category),
          category = fct_reorder(category, prop),
@@ -35,5 +36,13 @@ count_cancer_death %>%
   theme_minimal() +
   labs(y = "% of participants that died",
        x = NULL,
-       fill = NULL)
+       fill = NULL) +
+  theme(legend.position = "bottom",
+        legend.text = element_text(size=10)) +
+  theme(
+    strip.text.x = element_text(size = 10),
+    # strip.background = element_blank(),
+    strip.background = element_rect(fill=NA),
+    axis.text=element_text(size=10),
+    axis.title=element_text(size=10))
     

### 2021-04-15 02_R/02_long.R
cleaning table one

@@ -130,9 +130,23 @@ data_long_2015 %<>%
   ungroup()
 
 
-data_long_2015 %>% 
-  select(id, cohort, year, e1, cancer, cancer_v, cancer_date, dementia, dem_v, dem_date, death_2015, death_v, death_year, t2death_y)%>% View() 
-
+data_long_2015 %>%
+  select(
+    id,
+    cohort,
+    year,
+    e1,
+    cancer,
+    cancer_v,
+    cancer_date,
+    dementia,
+    dem_v,
+    dem_date,
+    death_2015,
+    death_v,
+    death_year,
+    t2death_y
+  ) %>% View()
 #### Create a "status" variable
 
 data_long_2015 %<>% 

### 2021-04-15 02_R/03_descriptives.R

@@ -18,6 +18,11 @@ data_wide <- data_wide %>%
 
 rio::export(data_wide, here::here("01_data", "clean_data", "wide_after_truncation.RData"))
 
+# ## merge with cancer data for type of cancer
+# 
+# cancer_data <-
+#   rio::import(here::here("01_data", "raw_data", "Oncodataset 19022020 R.RData"))
+
 ## Total n
 total_n <- dim(data_wide)[1]
 
@@ -96,11 +101,45 @@ women_prop <- data_wide %>%
 myvars <- c("sex", "age_0", "education", "apoe4", "smoke1", "bmi1", "oh1", "sbp1", 
           "ht1","hd_prev", "hd_v", "diabetes_prev", "diab_v",
           "stroke_prev", "stroke_v", "cancer_v") 
+
+data_tableone <- data_wide %>% 
+  select(myvars)
+
+data_tableone %<>% 
+  mutate(sex = ifelse(sex == 0, "Male", "Female"),
+         education = case_when(
+           education == 0 ~ "Lower",
+           education == 1 ~ "Intermediate",
+           education == 2 ~ "Higher",
+           TRUE ~ "Unknown"),
+         apoe4 = case_when(
+           apoe4 == 0 ~ "Not carrier",
+           apoe4 == 1 ~ "One allele carrier",
+           apoe4 == 2 ~ "Two allele carrier"),
+         smoke1 = case_when(
+           smoke1 == 0 ~ "Never",
+           smoke1 == 1 ~ "Former",
+           smoke1 == 2 ~ "Current"),
+         ht1 = ifelse(ht1 == 1, "History of hypertension", "No history of hypertension"),
+         hd_prev = case_when(hd_prev == 1 ~ "History of heart disease",
+                             hd_prev == 0 ~ "No history of heart disease"),
+         hd_v = case_when(hd_v == 1 ~ "Incident heart disease",
+                          hd_v == 0 ~ "No incident heart disease"),
+         diabetes_prev = case_when(diabetes_prev == 1 ~ "History of diabetes",
+                                   diabetes_prev == 0 ~ "No history of diabetes",
+                                   diabetes_prev == 2 ~ "Unknown"),
+         diab_v = ifelse(diab_v == 1, "Incident diabetes", "No incident diabetes"),
+         stroke_prev = ifelse(stroke_prev == 1, "History of stroke", "No history of stroke"),
[…]

### 2021-04-15 04_reports/manuscript.Rmd
new version of manuscript and results

@@
 output: [pdf_document]
-
+editor_options: 
+  chunk_output_type: console
 ---
@@
 library(tableone)
+library(magrittr)
+library(kableExtra)
 source(here::here("02_R", "03_descriptives.R"))
@@
 source(here::here("02_R", "05b_plot_causes_death.R"))
+table_results <- rio::import(here::here("02_R", "table_results.csv"))
+table_results_bounds <- rio::import(here::here("02_R", "table_results_bounds.csv"))
+models <- read_csv(here::here("04_reports", "models.csv")) %>% select(-6)
 ```
@@
 
-There is significant interest in understanding the association between cancer and dementia1.
+There is significant interest in understanding the association between cancer and dementia.
 Many observational studies have consistently determined that individuals with a history or incident cancer have a lower risk of developing of Alzheimer´s disease or related dementias (ADRD) when compared to individuals with no history of cancer.
 These findings have motivated the identification of molecular and genetic mechanisms that explain this association.
-Specific molecular mechanisms, such as Pin1 enzyme over-expression promotes tumorigenesis, while it’s down-regulation is attributed to mechanisms that contribute to Alzheimer’s Disease.
-These findings are suggest new opportunities to repurpose cancer chemotherapeutics for Alzheimer´s Disease.
+Specific molecular mechanisms, such as Pin1 enzyme over-expression promotes tumorigenesis, while its down-regulation is attributed to mechanisms that contribute to Alzheimer’s Disease.
+These findings suggest new opportunities to repurpose cancer chemotherapeutics for Alzheimer´s Disease.
 
@@
 Based on the measurement of Pin1, participant's exposure is classified as 1) Over-expressed Pin1 or 2) Regulated Pin1.
-Although Pin1 expression acts as a biomarker and is not a well-defined intervention, we can allude that sometime in the future people with over-expressed Pin1 could follow a specific treatment that lowers Pin1 under a specified threshold.
+Although Pin1 expression acts as a biomarker and could be considered an ill-defined intervention, we can allude that sometime in the future people with over-expressed Pin1 could follow a specific treatment that lowers Pin1 under a specified threshold.
 Furthermore, through out the follow-up each participant followed a strict monitoring for ADRD onset, represented as $Y_{20}$, and no participant was lost to follow-up nor died during the study period (no competing events present).
@@
 Previous studies have described age, sex, educational level and race/ethnicity as the minimal adjusting set of covariates (Ospina).
-However, environmental and behavioral factors such as smoking, alcohol intake, or chronic diseases which are known to cause microenvironmental changes such as inflammation and changes in tissue remodelling, may translate into Pin1 over-expression and are also related to the development of ADRD.
+However, environmental and behavioral factors such as smoking, which are known to cause microenvironmental changes such as inflammation and changes in tissue remodelling, may translate into Pin1 over-expression and are also related to the development of ADRD.
 Therefore, had we measured all $L$ we could identify the effect of $P$ in $Y_{20}$ by conditioning on $L$.
 
-```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = "Direct effect of Pin1 in ADRD"}
+```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = "Effect of Pin1 in ADRD"}
 \usetikzlibrary{arrows}
@@
 
-```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = "Direct effect of Pin1 in ADRD, with cancer diagnosis as proxy of Pin1"}
+```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = "Effect of Pin1 in ADRD, with cancer diagnosis as proxy of Pin1"}
 \usetikzlibrary{arrows}
@@
 Other studies have included participants with cancer diagnosis at the time of the diagnosis (for example from cancer registries such as SEER) and matched participants by age.
-In both cases we must remember that the the main interest is on the unmeasured $P$ and we should only adjust for covariates prior to $P$, and be careful to adjust for post-baseline covariates of $P$ or mediators between $P$ and $P^*$.
+In both cases we must remember that the the main interest is on the unmeasured $P$ thus we should only adjust for covariates prior to $P$ and be careful to adjust for post-baseline covariates of $P$ or mediators between $P$ and $P^*$.
 
 One of the mediators between $P$ and $P^*$ is death prior to cancer diagnosis.
-Since individuals are at risk of dying from other causes (such as cardivascular death), we can only measure $P^*$ in the subset of individuals who have survived long enough to have a cancer diagnosis.
+Since individuals are at risk of dying from other causes (such as cardiovascular death), we can only measure $P^*$ in the subset of individuals who have survived long enough to have a cancer diagnosis.
 We zoom in the relationship of $P \rightarrow P^*$ in Figure 3.
@@
 This means that if Pin1 has an effect on mortality through cancer or other mechanisms, we will observe a protective effect of Pin1 over-expression in ADRD, partially or fully mediated through death.
-For this reason estimators that are used to approximate a total effect (JY2020), such as the Aalen-Johansen estimator (underlying estimator for multistate models) or subdistribution hazard models, may not be useful in this setting.
+For this reason methods that are used to approximate a total effect, such as the Aalen-Johansen estimator (underlying estimator for multistate models) or subdistribution hazard models, may not be useful in this setting.
 Since the question of interest is focused on the direct effect of $P$ in $Y_{20}$ as in Figure 4 (violet arrows), there are several causal questions (estimands) to represent this direct effect, such as: the controlled direct effect (CDE), the survivor average causal effect (SACE) and the natural separable direct effect.
-In this section we discuss the controlled direct effect as the causal question of interest since it translates to frequently used estimators in this literature (such as Kaplan-Meier estimator and Cox-proportional hazard model).
+In this section we discuss the controlled direct effect as the causal question of interest since it translates to frequently used methods in this literature (such as Kaplan-Meier estimator and Cox-proportional hazard model).
 
 THe CDE represents the effect of Pin1 in ADRD in a setting where we could have prevented death over the study period.
-It relies on the assumption that we have measured all $C$ to block the pathway $Y85 \leftarrow C  \rightarrow D_{80} \rightarrow P \rightarrow P*$.
+It relies on the assumption that we have measured all $C$ to block the pathway $Y20 \leftarrow C  \rightarrow D_{15} \rightarrow P \rightarrow P*$.
 This assumption is defined as the independent censoring assumption conditional on covariates.
@@
 
-```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = "Effect of Pin1 in the risk of ADRD by age 85, with cancer diagnosis as proxy of Pin1"}
+```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = "Direct effect of Pin1 in the risk of ADRD at 20 years of follow-up, with cancer diagnosis as proxy of Pin1"}
 \usetikzlibrary{arrows}
@@
 \node [above of = p, yshift = -1cm](pstar) {$P^*$};
-\node [draw, right of = p] (d) {$D_{80}$};
-\node [right of = d] (y) {$Y_{85}$};
+\node [draw, right of = p] (d) {$D_{15}$};
+\node [right of = d] (y) {$Y_{20}$};
 \node [below of = d, yshift = 1cm, xshift = +1cm](c) {$C$};
@@
 
-Since the conditional independent censoring assumption is untestable, when there are no other mechanisms of censoring, we can explore the extreme scenarios and estimate the bounds.
-
-## 4. Motivating example with the RS data
+## 4. Application to the Rotterdam Study
 
@@
 Data on clinical outcomes was available until 2015.
+Further study details can be found in _Supplementary Material x_.
 
@@
 
-First we illustrate the most simple scenario (as Figure 1), were Pin1 over-expression is defined as "cancer ever vs.
-never" as if Pin1 and cancer diagnosis happened at the same time if looked retrospectively.
-Second, we consider cancer as a time-varying exposure independent of death.
-This can be represented as follows, at five years from study entry the observed unadjusted risk of cancer was `r cancer_5y` and the risk of death was `r death_5y` for the entire cohort, thus we assume that those who died would have the same risk of cancer had they not died, unconditionally, over follow-up.
-Third, we consider time to cancer as the
+We illustrate the association between cancer and dementia diagnosis under different scenarios that resemble the DAGs discussed above.
+First we considered the most simple scenario, were Pin1 over-expression is defined as _"cancer ever vs.
+never"_ as if Pin1 measurement and cancer diagnosis happened at the same time if looked retrospectively.
+Second, we considered cancer as a _"time-varying"_ exposure independent of death.
+This can be represented as follows, at five years from study entry the observed unadjusted risk of cancer was `r cancer_5y` and the risk of death was `r death_5y` for the entire cohort, this that those who died would have the same risk of cancer had they not died over follow-up.
+Third, we considered _"time to cancer"_ diagnosis as the proxy for Pin1 over-expression.
+_Not sure how to express this_
 
-To address confounding we fit inverse probability treatment weights including age at study entry, sex, educational attainment, smoking status and cohort as covariates.
+To address confounding we fit inverse probability treatment weights, stabilized and truncated at 99th percentile.
+Weight fitting was different for each scenario.
+For the scenario "ever vs.
+never" weights for cancer diagnosis were defined as the inverse of the probability of cancer diagnosis conditional on confounders, and for individuals free of cancer as the inverse of not having cancer conditional on covariates.
+The scenario "time-varying cancer diagnosis" is similar to the previous scenario, weights are the product of the time-fixed IPT weights above and year-specific.
+For the scenario "time to cancer", weights represent the product of the time-fix IPT weights for time to cancer diagnosis, weights turn to one at the moment of cancer diagnosis.
+In all cases we estimated these probabilities assuming a logistic regression model for cancer diagnosis as a function of the following covariates: age at study entry, sex, educational attainment, cohort, smoking status.
+Further details on modeling specifications and weights assessment are presented as **Supplementary material x**.
 
-To identify the controlled direct effect we considered death as an independent censoring mechanism conditional on time-varying shared causes of death and dementia.
+To estimate the controlled direct effect in time-varying settings we compared the complement of a weighted Kaplan-Meier survival estimator in participants with incident cancer vs.
+no incident of cancer with time indexed in years.
+The weights in this case are time-varying by follow-up year, defined as a product of the time-fixed IPT weights above and a year-specific inverse probability of censoring (IPC) by death weights.
+For an individual still alive in year t, the time t IPC weight is the product of the inverse probability of surviving in each year prior to t, conditional on measured common causes of death and dementia (that is, variables such as C in Figure 4).
+For an individual who has died by time t, the year t IPC weight is zero.
+We estimated survival probabilities using a logistic regression model for death as a function of baseline and time-varying covariates.
+Baseline covariates included age at study entry, sex, apoe4 status, educational attainment and the time-varying covariates smoking status, systolic blood pressure, BMI and prevalent and incident comorbidities such as: cancer, heart disease, stroke and diabetes.
+We additionally calculated the controlled direct effect considering death as an unconditional independent censoring event (as if there were no arrows from {C} to $D_{15}$ and $Y_{20}$) for an illustrative purpose.
 
-The difference in dementia risk considering cancer as a time-varying exposure independent of death.
-For example.
-at five years from follow-up the risk of cancer was `r cancer_5y` and the risk of death was `r death_5y`, thus we assume that those who died would have the same risk of cancer had they not died, unconditionally.
+Estimates of the controlled direct effect at 20 years of follow-up are presented as risk differences (RD), risk ratios (RR) and hazard ratios (HR).
+We note that hazards, unlike risks, inherently condition on surviving both dementia and death, as such they will not have a causal interpretation in this case.
+We present them for comparison against risk ratios.
+
+Since the conditional independent censoring assumption is untestable, we compute Peterson upper and lower bounds to represent: 1) the extreme scenario of independence, that refers to an scenario were those who died would never develop dementia (lower bound) and 2) complete dependency, that refers to an scenario where those who died would have a dementia prior to death (upper bound).
+The lower bound is calculated with the Aalen-Johansen estimator treating death as a competing event, and the upper bound is calculated with the Kaplan Meier estimator for the combined outcome of dementia or death.
+
+We additionally present the causes of death for participants who developed cancer and for participants free of cancer diagnosis over follow-up.
+This information provides insights about the association between cancer diagnosis and death and about the potential misclassification of individuals with no cancer diagnosis (cancer as a cause of death).
+
+All 95% confidence intervals were calculated using percentile-based bootstrapping based on 500 bootstrap samples.
+All analysis were performed using R, code is provided in supplementary material and available in https://github.com/palolili23/2021_cancer_dementia.
 
@@
 Further details on participants are presented in Table 1.
-Over follow up, `r yes_cancer` developed cancer and `r no_cancer` remain free of cancer diagnosis, the median age of cancer diagnosis was `r cancer_time`.
+Over follow-up, `r yes_cancer` developed cancer and `r no_cancer` remain free of cancer diagnosis, the median age of cancer diagnosis was `r cancer_time`.
 From the total sample, `r dem_total` had dementia over follow-up and median time to dementia was `r dem_time`.
@@
 
-## 5. Discussion
+Results for all scenarios are present in Table 2.
+Had we defined cancer diagnosis as _ever vs.
+never_, and relying on death as independent censoring event (unconditional), we observe a significant protective effect of ever having cancer in the risk of dementia [RR: `r table_results[2,6]` (95%CI), HR:  `r table_results[2,7]`].
+However this effect is diminished [RR: `r table_results[3,6]` (95%CI),
+HR: `r table_results[3,7]`)] if we relax the assumption of independent censoring conditional on baseline covariates related to death.
 
-We need to figure out what the key aging-dependent changes are and how to modulate these factors safely
+Had we defined cancer diagnosis as a _time-varying_ exposure we observe higher risk of dementia had participants had a cancer diagnosis over time, had we prevented death conditional on covariates [RR: `r table_results[6,6]` (95%CI), HR: `r table_results[6,7]`)].
 
-Causes such as aging, smoking, irradiation etc are causing micro environmental changes, like inflamation and changes in tissue remodelling. This promotes selection for adaptive mutations.
+## 5. Discussion (brainstorming points)
+
+- We observed how definition of cancer diagnosis as the proxy of Pin1 over-expression changes results.
+Most papers focus on a methodology but not on the question behind.
+Only with this in mind we can consider the confounders of the association.
+If the question was instead related to different cancer treatments it would require a different design and definition of confounders.
+As opposed to Ospina's paper that says this:
+
+_"Confounders that would explain the observed inverse cancer-AD association would be those that raise risk of cancer but reduce risk of AD, ruling out many common lifestyle and social factors associated with increased risk of both conditions, such as smoking or alcohol consumption.
+We considered age, sex, and educational level as sociodemographic factors that should be included in a minimal adjustment set in all studies on this association."_
+
+- Previous studies classified "competing risk bias" vs.
+"survival bias" which is unclear.
+We first need to pick an estimand.
+If we are interested in the CDE we rely a strong assumptions.
+But to consider that unconditional independecy makes no sense.
+Results change substantially if we relax this assumption with covariates.
+Also bounds show us the extent of extreme scenarios.
+Also a large proportion of cancer patients died prior to dementia diagnosis (`r cancer_death`), the leading cause of death was cancer in this group.
+
+- Besides, all estimands can be presented as risks, but depending on the estimand it treats death differently, and under different assumptions, and time-varying hazards (period specific hazards) are not useful.
 
 Efforts to prevent and treat cancer should converge with similar efforts to prevent other aging- associated diseases.
+We need to figure out what the key aging-dependent changes are and how to modulate these factors safely.
 
-We need better means to
+<!-- Causes such as aging, smoking, irradiation etc are causing micro environmental changes, like inflamation and changes in tissue remodelling.
+This promotes selection for adaptive mutations.Youthful tissues -->
 
-Youthful tissues
+- Knowing the cause of death provides information about the direction of missclassification. Among individuals free of cancer, we observed % of individuals who died with cancer as a primary cause.
 
-Knowing the cause of death provides information about the direction of missclassification.
-In patients with cancer we observed a slight % that died with dementia as a primary cause.
-On the other hand, among individuals free of cancer, we observed % of individuals who died with cancer as a primary cause.
+- Explicitly outlining the estimands and the assumptions that connect the causal question to the observed data provide an opportunity to improve the design of observational studies and the interpretation of their findings, plus better insight of potential sources of bias.
 
-In cancer you are afraid of missing dementia
-in cancer free you are afraid of missing cancer dx
+- This is a crucial since these studies are providing insights that are guiding other fields of research in the area, from bench science to biostatistics and epidemiological methods.
 
-Make a table of definition of weights based on cancer def. and
+- The CDE has an interpretation that relates to an scenario where death was eliminated. Future work on separable effects may help disentangle the different mechanisms that affect dementia and death.
 
-Explicitly outlining the estimands and the assumptions that connect the causal question to the observed data provide an opportunity to improve the design of observational studies and the interpretation of their findings, plus better insight of potential sources of bias.
-This is a crucial since these studies are providing insights that are guiding other fields of research in the area, from bench science to biostatistics and epidemiological methods15.
-To guide future observational and methodologic studies, our aim is to outline the potential estimands of interest that represent the underlying causal questions behind the association between cancer and dementia
+- In the future, we may be able to measure this biomarker and collect data retrospectively from stored blood samples, but we need to design the study very carefuly.
+
+- We could have change Pin1 to other molecular mechanism.
+This also extends to other questions that study the effect of one disease in the risk of another disease to understand the biological mechanisms behind the.
 
@@
 
-### Table 1: Descriptive characteristics of individuals who had a cancer diagnosis and of those free of cancer diagnosis over follow up.
+### Table 1: Descriptive characteristics of individuals who had a cancer diagnosis and of those free of cancer diagnosis over follow-up.
 
@@
 
+```{r}
+kableExtra::kable(table_results, booktabs = T) %>% 
+  kableExtra::kable_styling(latex_options = c("striped", "hold_position"))
+```
+
+### Table 3. Peterson bounds on the risk of dementia
+
+```{r}
+kableExtra::kable(table_results_bounds, booktabs = T) %>% 
+  kableExtra::kable_styling(latex_options = c("striped", "hold_position"))
+```
+
 \newpage
@@
 ```
+
+\newpage
+
+```{r}
+models <- models %>% pivot_longer(2:5, names_to = "Models", values_to = "Covariates")
+
+kableExtra::kable(models, booktabs = T) %>% 
+   kableExtra::kable_styling(latex_options = c("striped", "hold_position")) %>% 
+  kable_styling(full_width = F) %>%
+column_spec(3, width = "20em")
+```

### 2021-04-15 02_R/04_direct_effect_simplified.R

@@ -1,0 +1,663 @@
+library(tidyverse)
+library(rio)
+library(survival)
+library(survminer)
+library(broom)
+library(splines)
+library(magrittr)
+library(WeightIt)
+library(cobalt)
+
+# Import data -------------------------------------------------------------
+source(here::here("02_R", "04b_auxiliary_fx.R"))
+
+data_long <-
+  import(here::here("01_data", "clean_data", "dementia_long.RData"))
+
+data_wide <-
+  import(here::here("01_data", "clean_data", "wide_after_truncation.RData"))
+
+data_long %<>% 
+  group_by(id) %>% 
+  mutate(
+    fuptime = time - 1,
+    tstart = lag(fuptime),
+    tstart = ifelse(is.na(tstart), -1, tstart)) %>% 
+  ungroup()
+
+## idea of how numbers look like over follow-up
+
+data_long %<>% 
+  mutate(both_outcomes = 
+           case_when(outcome_plr == 1 ~ 1,
+                     competing_plr == 1 ~2,
+                     TRUE ~ 0))
+
+data_long %<>% 
+  mutate(combined_outcome = 
+           case_when(outcome_plr == 1 ~ 1,
+                     competing_plr == 1 ~ 1,
+                     TRUE ~ 0))
+
+data_wide %<>%
+  mutate(dementia_efu =
+           case_when(outcome_plr == 1 ~ 1,
+                     competing_plr == 1 ~ 2,
+                     TRUE ~ 0))
+
+data_wide %<>%
+  mutate(t2dem_efu =
+           ifelse(t2dem >= 240, 240, t2dem))
+
+## If competing_plr = 1, then outcome_plr = NA
+
+data_long %<>% 
+  mutate(outcome_plr = ifelse(competing_plr == 1, NA, outcome_plr))
+
+# 1. Cancer Ever vs. Never ---------------------------------------------------
+
+
+# 1.1. IPTW (cancer ever vs. never) ------------------------------------------
+
+# cancer_den <-
+#   glm(
+#     cancer_v ~ bs(age_0) + sex + education + apoe4 + as.factor(smoke1),
+#     data = data_wide,
+#     family = binomial
+#   )
+# 
+# summary(cancer_den)
+# 
[…]

### 2021-04-15 02_R/04b_auxiliary_fx.R

@@ -99,10 +99,15 @@ risks_km <- function(model){
     pivot_wider(names_from = strata, values_from = estimate) %>% 
     mutate(rd = .[[2]] - .[[1]],
            rr = .[[2]] / .[[1]]) %>% 
-    mutate_at(c(1:3), ~.*100)} 
+    mutate_at(c(1:3), ~.*100) %>% 
+    mutate_at(c(1:3), round, 1) %>% 
+    mutate(rr = round(rr, 2))} 
 
 tidy_hr <- function(model) {
  model %>% 
   broom::tidy(exponentiate = TRUE, conf.int = TRUE) %>%
-    select(term, estimate, contains("conf"))
+    select(term, estimate, contains("conf")) %>% 
+    mutate_at(c(2:4), round, 2) %>% 
+    select(-1, hr = estimate) %>% 
+    mutate(hr = paste0(hr, " (", conf.low, ", ", conf.high, ")"))
 }

### 2021-04-15 02_R/05b_plot_causes_death.R

@@ -3,7 +3,7 @@ library(tidyverse)
 
 ggthemes::scale_fill_colorblind()
 
-wide_data <- import(here::here("01_data", "clean_data", "wide_noltfu.RData"))
+wide_data <- import(here::here("01_data", "clean_data", "wide_after_truncation.RData"))
 
 cases_death <- import(here::here("01_data", "raw_data", "causes_death.RData"))
 
@@ -17,15 +17,15 @@ data <- wide_data %>%
 #   gt::gt()
 
 count_cancer_death <- data %>% 
-  filter(cohort == 1, dementia == 2) %>%
-  group_by(cancer) %>% 
+  filter(competing_plr == 1) %>% 
+  group_by(cancer_v) %>% 
   count(category) %>% 
   mutate(prop = 100*(n/sum(n))) 
 
 causes_death <- count_cancer_death %>% 
   filter(category != "Alive") %>% 
   mutate(category = str_to_title(category)) %>% 
-  mutate(cancer = ifelse(cancer == 0, "No cancer", "Incident cancer"),
+  mutate(cancer = ifelse(cancer_v == 0, "No cancer", "Incident cancer"),
     category = as_factor(category),
          category = fct_reorder(category, prop),
          cancer = as_factor(cancer)) %>% 
@@ -45,4 +45,33 @@ causes_death <- count_cancer_death %>%
     strip.background = element_rect(fill=NA),
     axis.text=element_text(size=10),
     axis.title=element_text(size=10))
-    
+
+count_dementia_death <- data %>% 
+  filter(competing_plr == 0) %>% 
+  group_by(outcome_plr) %>% 
+  count(category) %>% 
+  mutate(prop = 100*(n/sum(n))) 
+
+causes_death_dementia <- count_dementia_death %>% 
+  # filter(category != "Alive") %>% 
+  mutate(category = str_to_title(category)) %>% 
+  mutate(outcome_plr = ifelse(outcome_plr == 0, "Had no dementia diagnosis", "Had dementia diagnosis"),
+         category = as_factor(category),
+         category = fct_reorder(category, prop),
+         outcome_plr = as_factor(outcome_plr)) %>% 
+  ggplot(aes(outcome_plr, prop, fill = category)) +
+  geom_col() +
+  ggthemes::scale_fill_tableau() +
+  coord_flip() + 
+  theme_minimal() +
+  labs(y = "% of participants that died",
+       x = NULL,
+       fill = NULL) +
+  theme(legend.position = "bottom",
[…]

### 2021-04-16 04_reports/manuscript.Rmd
updated manuscript, adding drugs to the picture

@@
 ```{r, echo =  FALSE, message = FALSE, warning = FALSE}
-knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.align="center")
+knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
 library(survival)
@@
 
-There is significant interest in understanding the association between cancer and dementia.
-Many observational studies have consistently determined that individuals with a history or incident cancer have a lower risk of developing of Alzheimer´s disease or related dementias (ADRD) when compared to individuals with no history of cancer.
-These findings have motivated the identification of molecular and genetic mechanisms that explain this association.
-Specific molecular mechanisms, such as Pin1 enzyme over-expression promotes tumorigenesis, while its down-regulation is attributed to mechanisms that contribute to Alzheimer’s Disease.
-These findings suggest new opportunities to repurpose cancer chemotherapeutics for Alzheimer´s Disease.
+Many observational studies have consistently found that individuals with cancer have a lower risk of developing of Alzheimer´s disease or related dementias (ADRD) when compared to individuals with no history of cancer [1–4].
+These findings have motivated substantial research toward mechanistic explanations, including searching for and hypothesizing that molecular and genetic mechanisms may explain this association [5–12].
+These research inquiries inevitably lead to discussions of repurposing or augmenting current cancer chemotherapeutics for ADRD13.
 
-Nevertheless, observational studies are susceptible to several bias, and researchers have raised concerns related to the competing event of death as well as unmeasured confounding and ascertainment error  that could drive these results.
-Though, to understand sources of bias the causal question should be explicitly  specified.
-This particular question has at least two core challenges.
-First, defining the underlying biological mechanism we would ideally intervene upon and defining cancer diagnosis as the proxy for that mechanism.
-For example, researchers refer to biological mechanisms which onset may be prior to cancer diagnosis and to the effect of cancer treatment, both relate to different causal questions and would be translated into different study design and analytic choices.
-Second, defining the causal question (or estimand) of interest when the competing event of death is interest.
-Previous studies alluded to choosing a particular summary measure or modeling approach to account for death (incidence rate ratios (ref), hazard ratios (ref), multistate models (ref)).
-However, when competing events are present, there is more than one causal question that can be framed and each question has different interpretations and relies on different assumptions.
+Nevertheless, inferring any treatment or mechanistic effects from the observed cancer-ADRD inverse association is not straightforward, and researchers have raised concerns related to the competing event of death, unmeasured confounding, and ascertainment error that could explain these results [9,14].
+However, understanding these or other sources of bias first requires that we make explicit the causal question.
+Moreover, making explicit the causal question ensures that the research addresses a question that is relevant to clinical decision-making or population health .
 
-With these considerations in mind, we outline a case study focused on the question: _What is the direct effect of Pin1 over-expression at late-midlife in the risk of ADRD after 20 years of follow-up compared to having regulated expression of Pin1._ To identify this effect with real-world data, we illustrate the complexity of the research question by building progressively a directed acyclic graph (DAG) which help us connect the causal question to the observable data and the assumptions we rely on to identify the effect.
+To illustrate the complexities of inferring hypothetical or available treatments’ effects on ADRD from the observed cancer-ADRD association, we consider the case study of focusing on Pin1 enzyme.
+Previous studies have shown that Pin1 enzyme over-expression promotes tumorigenesis, while its down-regulation is attributed to mechanisms that contribute to Alzheimer’s Disease [11,12,15].
+If there was a drug that increases Pin1 expression specifically in brain tissue in hopes of preventing dementia, we could pose the question as: _What is the direct effect of this drug in the risk of ADRD after 5 years  of follow-up compared to standard treatments?_
+
+To explore how we might learn about this effect using real-world data on cancer and ADRD, we progressively build a causal directed acyclic graph (DAG) which help us connect the causal question to the observable data and the assumptions we rely on to identify the effect.
 We exemplify different scenarios with data collected from the Rotterdam Study, a population-based cohort study.
@@
 
-We begin by illustrating a hypothetical scenario through the causal graph of Figure 1, that depicts a situation where Pin1, represented as $P$, was measured in all participants in late midlife (for example between age 50 and 60).
+Since there is no drug developed at the moment that targets Pin1, at best we can use observational data on Pin1 expression measurements.
+We can imagine an scenario where a biomarker test is developed to measure Pin1 and we measure this biomarker from stored blood samples in a population based-cohort that is recruited in all participants in late midlife (for example between age 50 and 60).
 Based on the measurement of Pin1, participant's exposure is classified as 1) Over-expressed Pin1 or 2) Regulated Pin1.
-Although Pin1 expression acts as a biomarker and could be considered an ill-defined intervention, we can allude that sometime in the future people with over-expressed Pin1 could follow a specific treatment that lowers Pin1 under a specified threshold.
-Furthermore, through out the follow-up each participant followed a strict monitoring for ADRD onset, represented as $Y_{20}$, and no participant was lost to follow-up nor died during the study period (no competing events present).
+This hypothetical scenario is illustrated in the causal graph of Figure 1, that depicts a situation where Pin1 is represented as $P$.
+In this cohort each participant followed a strict monitoring for ADRD onset for twenty years, represented as $Y_{20}$, and no participant was lost to follow-up.
 
-Since Pin1 expression was not a randomized intervention, we consider a set of measured covariates $L$ for which conditional exchangeability holds.
-Previous studies have described age, sex, educational level and race/ethnicity as the minimal adjusting set of covariates (Ospina).
+Since the biomarker Pin1 is measured within an observational study, we have to consider a set of measured covariates $L$ for which conditional exchangeability holds.
+Previous studies have described age, sex, educational level and race/ethnicity as the minimal adjusting set of covariates)[4].
 However, environmental and behavioral factors such as smoking, which are known to cause microenvironmental changes such as inflammation and changes in tissue remodelling, may translate into Pin1 over-expression and are also related to the development of ADRD.
 Therefore, had we measured all $L$ we could identify the effect of $P$ in $Y_{20}$ by conditioning on $L$.
+For simplicity, we treat Pin1 expression as a point intervention and we assume we know the time-ordering of $L$ (that is we assume $L$ happens prior to $P$.
+In reality we would assume that $P$ changes over time and is affected by time-varying $L$, which produces treatment-confounder feedback loops, thus to emulate a target trial we would ideally measure $L$ and $P$ repeatedly over follow-up [16].
 
@@
 
-Unfortunately, Pin1 expression measurement is not available at the moment in real-world settings, we can only rely on a proxy of this exposure.
-Since Pin1 over-expression is present in tumors and tumors are only measured through screening and diagnosis, we considered cancer diagnosis as the proxy for Pin1 over-expression, as most observational studies have defined (all refs).
+Unfortunately, Pin1 expression measurement is not available at the moment in real-world settings, thus we can only rely on a proxy of this exposure.
+Since Pin1 over-expression is present in tumors and tumors are only measured through screening and diagnosis, we considered cancer diagnosis as the proxy for Pin1 over-expression, as most observational studies have defined [4,17–24].
 We depict this feature in Figure 2, where $P^*$ represents _incident cancer diagnosis_, the measured proxy of $P$.
@@
 \node [above of = p, yshift = -1cm](pstar) {$P^*$};
-\node [right of = p] (y) {$Y_{85}$};
+\node [right of = p] (y) {$Y_{20}$};
 \draw[->, violet] (p) -- (pstar);
@@
 This means that a participant within a prospective cohort study contributes to the "regulated Pin1" since study entry up to the time of cancer diagnosis and later on to the "Over-expressed Pin1" arm.
-Other studies have included participants with cancer diagnosis at the time of the diagnosis (for example from cancer registries such as SEER) and matched participants by age.
+Other studies have included participants with cancer diagnosis at the time of the diagnosis (for example from cancer registries[19,23,25,26]) and matched participants by age.
 In both cases we must remember that the the main interest is on the unmeasured $P$ thus we should only adjust for covariates prior to $P$ and be careful to adjust for post-baseline covariates of $P$ or mediators between $P$ and $P^*$.
@@
 $D = 0$ between $P$ and $P^*$, this illustrates that to observe $P^*$, we condition (box around $D = 0$) on surviving long enough to have a cancer diagnosis.
-Several risk factors that increase the risk of cancer might also cause death prior to cancer diagnosis, for example smoking may cause lung cancer and chronic obstructive pulmonary disease (a leading cause of death in this age group).
+Several risk factors that increase the risk of cancer might also cause death prior to cancer diagnosis, for example smoking may cause lung cancer and chronic obstructive pulmonary disease (a leading cause of death in this age group[27]).
 Therefore, to isolate the effect of $P$ in $P^*$ (the violet arrow), we need to block all shared causes between $P$ and $D$ and between $D$ and $P*$ both represented in Figure 3 as $C_1$ and $C_2$.
-In this way we assume a hypothetical scenario in which we could prevent death prior to cancer diagnosis, by conditioning on a rich set of covariates.
+In this way we assume a hypothetical scenario in which we could prevent death prior to cancer diagnosis, by conditioning on a rich set of covariates[28].
 This assumption must hold regardless of whether we use incident cancer as a time-varying exposure, or match cancer patients to participants free of cancer by age.
+Furthermore, we note that this is only one of issues with considering cancer diagnosis as the proxy for Pin1 expression, in terms of measurement.
+We could add more complexity by assigning measurement error to cancer and dementia diagnosis and unmeasured shared causes that affect both diseases, such as physician’s recommendations and type of health insurance.
 
@@
 
-Up to this point we have outlined the underlying assumptions of using incident cancer diagnosis as a proxy for Pin1 expression.
+Up to this point we have outlined certain underlying assumptions of using incident cancer diagnosis as a proxy for Pin1 expression.
 We continue outlining the additional assumptions related to death as a competing event of ADRD.
@@
 
-As illustrated in previous figures, we are interested in the effect of Pin1 in late midlife in ADRD risk after 20 years of follow-up.
-Nevertheless, mortality is substantial and increases steeply in late life, and a leading cause of death is cancer.
-This means that in order to measure ADRD over follow-up individuals have to had survive to cancer and to other leading causes of death.
+Mortality increases steeply in late life, which means that even in the setting of the randomized controlled trial, in order to measure ADRD over follow-up individuals have to survive long enough to have a diagnosis.
 For this reason death is a competing event of ADRD because if a participant dies prior to ADRD diagnosis, death prevents from observing ADRD at future time-points.
+In our setting, this becomes a major challenge, since we are measuring cancer diagnosis as the proxy for Pin1, and cancer is a leading cause of mortality in midlife and late life29.
 We can visualize this data feature in the causal diagram in Figure 4.
-In this DAG we include $D_{15}$ as a representation of death at 15 years of follow-up, an arrow between $P$ and $D_{15}$ since $P$ may increase the mortality risk irrespective of cancer diagnosis.
+In this DAG we include $D_{15}$ as a representation of death at 15 years of follow-up, an arrow between $P$ and $D_{15}$ since $P$ may increase the mortality risk.
 We also include an arrow between $P^*$ and $D_{15}$ since cancer diagnosis, and subsequent treatment (or lack of treatment) may have an effect on death.
@@
 In this DAG we observe that, had we measured and adjusted for $L$, we could estimate the total effect of $P$ in $Y_{20}$ without further assumptions.
-However the total effect includes all pathways between $P$ and $Y_{20}$.
+However the total effect includes all pathways between $P$ and $Y_{20}$[28].
 This means that if Pin1 has an effect on mortality through cancer or other mechanisms, we will observe a protective effect of Pin1 over-expression in ADRD, partially or fully mediated through death.
-For this reason methods that are used to approximate a total effect, such as the Aalen-Johansen estimator (underlying estimator for multistate models) or subdistribution hazard models, may not be useful in this setting.
-Since the question of interest is focused on the direct effect of $P$ in $Y_{20}$ as in Figure 4 (violet arrows), there are several causal questions (estimands) to represent this direct effect, such as: the controlled direct effect (CDE), the survivor average causal effect (SACE) and the natural separable direct effect.
-In this section we discuss the controlled direct effect as the causal question of interest since it translates to frequently used methods in this literature (such as Kaplan-Meier estimator and Cox-proportional hazard model).
-
-THe CDE represents the effect of Pin1 in ADRD in a setting where we could have prevented death over the study period.
-It relies on the assumption that we have measured all $C$ to block the pathway $Y20 \leftarrow C  \rightarrow D_{15} \rightarrow P \rightarrow P*$.
-This assumption is defined as the independent censoring assumption conditional on covariates.
-In this setting death is treated as a censoring event and it can be interpreted as those who died would have the same risk of developing dementia if prevented from dying than those who remained alive and free of dementia at a given time point, conditional on shared causes of dementia and death.
+Since the question of interest is focused on the direct effect of $P$ in $Y_{20}$ as in Figure 4 (violet arrows) we need to conceptualize the different mechanisms through which $P$ could affect $D_{15}$ and $Y_{20}$.
+With this in mind we can conceive different causal questions (estimands) to represent this direct effect , such as the controlled direct effect (CDE)  and the natural separable direct effect.
+In this section we discuss the controlled direct effect as the causal question of interest since it translates to frequently used methods in this literature (such as Kaplan-Meier estimator and Cox-proportional hazard model) and leave the separable direct effects question for discussion.
 
@@
 
+The CDE represents the effect of Pin1 over-expression in ADRD in a setting where we could have prevented death over the study period.
+It relies on the assumption that we have measured all $C$ to block the pathway $Y20 \leftarrow C  \rightarrow D_{15} \rightarrow P \rightarrow P*$.
+This assumption is defined as the independent censoring assumption conditional on covariates.
+In this setting death is treated as a censoring event and it can be interpreted as those who died would have the same risk of developing dementia if prevented from dying than those who remained alive and free of dementia at a given time point, conditional on shared causes of dementia and death.
+
+Therefore, if we combine the challenges in section two related to cancer diagnosis as a proxy for Pin1 over-expression, and having death as a competing event of ADRD we observe the complexity of the DAG in Figure 5.
+Though this is yet a simplified version since we omitted time varying $P*$, $L$ and $C's$.
+
+```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = " Direct effect of Pin1 in the risk of ADRD at 20 years of follow-up, with cancer diagnosis as proxy of Pin1, considering time (and death) bewteen P and P*"}
+\usetikzlibrary{arrows}
+\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
+\node (p) {$P$};
+\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$L$};
+\node [right of = p] (dp) {$D$};
+\node [right of = dp] (pstar) {$P*$};
+\node [right of = pstar] (d) {$D_{15}$};
+\node [right of = d] (y) {$Y_{20}$};
+\node [above of = dp, yshift = -1cm, xshift = +1cm](c) {$C_1$};
+\node [above of = d, yshift = -1cm, xshift = +1cm](cd) {$C_2$};
+\draw[->] (p) -- (dp);
+\draw[->] (dp) -- (pstar);
+\draw[->, violet] (p) to [out=-45,in=-135, looseness=0.3] (pstar);
+\draw[->] (pstar) -- (d);
+\draw[->] (p) to [out=-45,in=-135, looseness=0.3] (d);
+\draw[->, violet] (p) to [out=-45,in=-135, looseness=0.3] (y);
+\draw[->] (d) -- (y);
+\draw[->] (l) -- (p);
+\draw[->] (l) to [out=0,in=270, looseness=0.4] (y);
+\draw[->] (c) -- (dp);
+\draw[->] (c) -- (pstar);
+\draw[->] (cd) -- (d);
+\draw[->] (cd) -- (y);
+\end{tikzpicture}
+```
+
 ## 4. Application to the Rotterdam Study
 
-In this section we illustrate how previous DAGS may reflect on the analysis performed to unveil the effect of Pin1 in all cause-dementia.
+In this section we conduct an analysis of the cancer-ADRD association that is structured to the best of the available data’s abilities to inform the possible effect of Pin1 on all cause-dementia .
 We use data collected in the Rotterdam Study, a population-based prospective cohort study among persons living in the Ommoord district in Rotterdam, the Netherlands.
@@
 In contrast, among participants free of cancer diagnosis over follow-up, `r no_cancer_dem` were diagnosed with dementia and `r no_cancer_death` died over follow-up, `r no_cancer_alive` were alive at the end of follow-up.
+Figure x.
+shows the magnitude of the proportion of people who died over time.
 
@@
 
+[should include some final comments on this being interpreted as back to the original question about Pin1, rather than being results about the particular estimates from the models too, but I think we need to discuss the analyses first.]
+
 ## 5. Discussion (brainstorming points)
@@
 If the question was instead related to different cancer treatments it would require a different design and definition of confounders.
+And if we consider treatments given at a different time in life too.
+And how we want to think about which effect is of interest given competing events.
 As opposed to Ospina's paper that says this:
@@
 We first need to pick an estimand.
-If we are interested in the CDE we rely a strong assumptions.
+If we are interested in the CDE or any closely-related direct effect kind of question?
+we rely a strong assumptions.
 But to consider that unconditional independecy makes no sense.
@@
 
-- We could have change Pin1 to other molecular mechanism.
+- We could have change Pin1 to other molecular mechanism (MAYBE).
 This also extends to other questions that study the effect of one disease in the risk of another disease to understand the biological mechanisms behind the.
+
+\newpage
+
+### References
+
+1.	Ma, L. L. et al. Association between cancer and Alzheimer’s disease: Systematic review and meta-analysis. J. Alzheimer’s Dis. 42, 565–573 (2014).
+2.	Hanson, H.
+A., Horn, K.
+P., Rasmussen, K.
+M., Hoffman, J.
+M.
+& Smith, K.
+R.
+Is Cancer Protective for Subsequent Alzheimer’s Disease Risk?
+Evidence from the Utah Population Database.
+Journals Gerontol.
+- Ser.
+B Psychol.
+Sci.
+Soc.
+Sci.
+72, 1032–1043 (2017).
+3.	van der Willik, K. D., Schagen, S. B. & Ikram, M. A. Cancer and dementia: Two sides of the same coin? Eur. J. Clin. Invest. 48, 1–12 (2018).
+4.	Ospina-Romero, M. et al. Association Between Alzheimer Disease and Cancer With Evaluation of Study Biases: A Systematic Review and Meta-analysis. JAMA Netw. open 3, e2025515 (2020).
+5.	Behrens, M., Lendon, C. & Roe, C. A Common Biological Mechanism in Cancer and Alzheimers Disease? Curr. Alzheimer Res. 6, 196–204 (2009).
+6.	Harris, R. A., Tindale, L. & Cumming, R. C. Age-dependent metabolic dysregulation in cancer and Alzheimer’s disease. Biogerontology 15, 559–577 (2014).
+7.	Nudelman, K. N. H., McDonald, B. C., Lahiri, D. K. & Saykin, A. J. Biological Hallmarks of Cancer in Alzheimer’s Disease. Mol. Neurobiol. 56, 7173–7187 (2019).
+8.	Papin, S. & Paganetti, P. Emerging evidences for an implication of the neurodegeneration-associated protein tau in cancer. Brain Sci. 10, 1–24 (2020).
+9.	Driver, J. A. Inverse association between cancer and neurodegenerative disease: review of the epidemiologic and biological evidence. Biogerontology 15, 547–557 (2014).
+10.	Olson, B. & Marks, D. L. Pretreatment cancer-related cognitive impairment—mechanisms and outlook. Cancers (Basel). 11, 1–18 (2019).
+11.	Li, J. et al. Roles of peptidyl-prolyl isomerase Pin1 in disease pathogenesis. Theranostics 11, 3348–3358 (2021).
+12.	Driver, J.
+A., Zhou, X.
+Z.
+& Lu, K.
+P.
+Pin1 dysregulation helps to explain the inverse association between cancer and Alzheimer’s disease.
+Biochim.
+Biophys.
+Acta - Gen.
+Subj.
+1850, 2069–2076 (2015).
+13.	Snyder, H.
+M.
+et al.
+Exploring the nexus of Alzheimer’s disease and related dementias with cancer and cancer therapies: A convening of the Alzheimer’s Association & Alzheimer’s Drug Discovery Foundation.
+Alzheimer’s Dement.
+13, 267–273 (2017).
+14.	Ganguli, M. Cancer and dementia. Alzheimer Dis. Assoc. Disord. 29, 177–182 (2015).
+15.	Angelucci, F. & Hort, J. Prolyl isomerase Pin1 and neurotrophins: A loop that may determine the fate of cells in cancer and neurodegeneration. Ther. Adv. Med. Oncol. 9, 59–62 (2017).
+16.	Hernán, M. A. & Robins, J. M. Causal Inference: What If (Harvard book) - Last version. (2019).
+17.	Musicco, M. et al. Inverse occurrence of cancer and Alzheimer disease: A population-based incidence study. Neurology 81, 322–328 (2013).
+18.	Aiello Bowles, E. J. et al. Risk of Alzheimer’s disease or dementia following a cancer diagnosis. PLoS One 12, 1–11 (2017).
+19.	Ording, A. G. et al. Cancer and risk of Alzheimer’s disease: Small association in a nationwide cohort study. Alzheimer’s Dement. 16, 953–964 (2020).
+20.	Sun, M., Wang, Y., Sundquist, J., Sundquist, K. & Ji, J. The Association Between Cancer and Dementia: A National Cohort Study in Sweden. Front. Oncol. 10, (2020).
+21.	Frain, L. et al. Association of cancer and Alzheimer’s disease risk in a national cohort of veterans. Alzheimer’s Dement. 13, 1364–1370 (2017).
+22.	Schmidt, S. A. J., Ording, A. G., Horváth-Puhó, E., Sørensen, H. T. & Henderson, V. W. Non-melanoma skin cancer and risk of Alzheimer’s disease and all-cause dementia. PLoS One 12, 1–13 (2017).
+23.	Freedman, D. M. et al. Associations between cancer and Alzheimer’s disease in a U.S. Medicare population. Cancer Med. 5, 2965–2976 (2016).
+24.	Driver, J. A. et al. Inverse association between cancer and Alzheimer’s disease: Results from the Framingham Heart Study. BMJ 344, 19 (2012).
+25.	Bolch, C.
+A.
+et al.
+Inverse probability of treatment-weighted competing risks analysis: An application on long-term risk of urinary adverse events after prostate cancer treatments.
+BMC Med.
+Res.
+Methodol.
+(2017) doi:10.1186/s12874-017-0367-8.
+26.	Mezencev, R. & Chernoff, Y. O. Risk of alzheimer’s disease in cancer patients: Analysis of mortality data from the US SEER population-based registries. Cancers (Basel). 12, (2020).
+27.	Soriano, J.
+B.
+et al.
+Prevalence and attributable health burden of chronic respiratory diseases, 1990–2017: a systematic analysis for the Global Burden of Disease Study 2017.
+Lancet Respir.
+Med.
+8, 585–596 (2020).
+28.	Young, J. G., Stensrud, M. J., Tchetgen, E. J. T. & Hernán, M. A. A causal framework for classical statistical estimands in failure time settings with competing events. Stat. Med. 1, 1–38 (2020).
+29.	Bennett, J. E. et al. NCD Countdown 2030: worldwide trends in non-communicable disease mortality and progress towards Sustainable Development Goal target 3.4. Lancet 392, 1072–1088 (2018).
 
@@
   kableExtra::kable_styling(latex_options = c("striped", "hold_position"))
+
 ```

### 2021-04-16 02_R/04_direct_effect.R
saving results

@@ -1,7 +1,5 @@
-library(ggplot2)
 library(tidyverse)
 library(rio)
-library(wesanderson)
 library(survival)
 library(survminer)
 library(broom)
@@ -155,9 +153,15 @@ cancer_ever_cox_crude <-
   coxph(Surv(t2dem_efu, dementia_efu == 1) ~ cancer_v, data_wide)
 
 hr_1a <- cancer_ever_cox_crude %>% 
-  tidy_hr()
+  tidy_hr() %>% 
+  mutate(
+    model = "Unadjusted"
+  )
 
-rd_1a <- risks_km(cancer_ever_km_crude)
+rd_1a <- risks_km(cancer_ever_km_crude) %>% 
+  mutate(
+    model = "Unadjusted"
+  )
 
 
 # 1.b. KM Independent censoring -----------------------------------------------
@@ -172,9 +176,14 @@ cancer_ever_cox <-
   coxph(Surv(t2dem_efu, dementia_efu == 1) ~ cancer_v, data_wide, weights = cancer_weight_t, cluster = id)
 
 hr_1b <- cancer_ever_cox %>% 
-  tidy_hr()
+  tidy_hr() %>% mutate(
+    model = "IPTW"
+  )
 
-rr_1b <- risks_km(cancer_ever_km)
+rd_1b <- risks_km(cancer_ever_km) %>% 
+  mutate(
+    model = "IPTW"
+  )
 
 # 1.c. KM. IPCW for censoring -----------------------------------------------
 
@@ -188,10 +197,19 @@ cancer_ever_cox_ipcw <-
   coxph(Surv(t2dem_efu, dementia_efu == 1) ~ cancer_v, data_wide, weights = baseline_wt, cluster = id)
 
 hr_1c <- cancer_ever_cox_ipcw %>% 
-  tidy_hr()
+  tidy_hr() %>% 
+  mutate(
+    model = "IPTW + IPCW"
+  )
 
-rr_1c <- risks_km(cancer_ever_km_ipcw)
+rd_1c <- risks_km(cancer_ever_km_ipcw) %>% 
+  mutate(
+    model = "IPTW + IPCW"
+  )
 
+bind_rows(hr_1a, hr_1b, hr_1c)
+
+bind_rows(rd_1a, rd_1b, rd_1c)
 
 # 2. Time-varying cancer -----------------------------------------------------
 

### 2021-04-20 06_presentations/210421_cqm/index.Rmd
updating presentation for cqm

---
title: "‘Dissecting the causal question underlying the association between cancer and dementia"
subtitle: ""  
author: 
  - "L. Paloma Rojas-Saunero"
date: '`r Sys.Date()`'
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    seal: false
    nature:
      highlightStyle: github
      highlightLines: true
      highlightLanguage: ["r", "css"]
      countIncrementalSlides: FALSE
      ratio: "16:9"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, }
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(collapse = TRUE,
                      fig.retina = 3,
                      warning = FALSE,
                      message = FALSE, cache = TRUE)
library(WeightIt)
library(cobalt)
library(splines)

source(here::here("02_R", "05_plots.R"))
source(here::here("02_R", "05b_plot_causes_death.R"))
table_results <- rio::import(here::here("02_R", "table_results.csv"))
table_results_bounds <- rio::import(here::here("02_R", "table_results_bounds.csv"))

```

```{r xaringan-themer, include=FALSE}
library(xaringanthemer)
style_mono_accent(
  base_color = "#011A5E",
  header_font_google = google_font("Fira Sans", "500"),
  table_row_even_background_color = "#FFFFFF",
  table_border_color = "#011A5E",
  text_font_google   = google_font("Lato", "400", "400i"),
  text_font_size = "25px"
)
```

background-image: url(./figs/erasmus.png)
background-position: 5% 95%
background-size: 25%, 15%
class: center, middle

## Dissecting the causal question underlying the association between cancer and dementia

<br> <br>

.right[L. Paloma Rojas-Saunero, MD

**Epidemiology department**

**Causal inference and Neuroepi group**]

---

## Introduction

.pull-left[

Observational studies have found a protective effect of cancer diagnosis in the risk of Alzheimer´s disease or related dementias (ADRD).

These studies lead to discussions of repurposing or augmenting current cancer chemotherapeutics for ADRD
]

--

.pull-right[ 
![](figs/forest_plot.PNG)
]

.footnote[Ospina-Romero M. JAMA Network Open. 2020]

---

## Cancer diagnosis as an exposure

.pull-left[

- Molecular mechanisms

- Cancer treatments

- Shared factors

- Unmeasured confounding, measurement error and survival bias may explain this association

]

.pull-right[

![](figs/causal_risks.PNG)
]

.footnote[
L. Frain et al. Alzheimer’s & Dementia. 2017.
]

---
class: middle, center

## How can we conclude on causal pathways if we haven't measured them?

---

## Back to basics

.pull-left[
![](figs/pin_knockout.PNG) 
]

.pull-right[

Driver J. et al. Biochimica et Biophysica Acta. 2015.

![](figs/tment_pin1.PNG)

]

---

## Aims

---

### Overview of cancer diagnosis as a proxy for Pin1 enzime

.pull-left[
If there was a drug that targets Pin1 $(D)$ we would perform an RCT to assess the effect on dementia $(Y_{t})$.

We can imagine a setting where middle-aged participants with signs of cognitive decline take this drug and are followed over time.
]

--

<br><br>

.pull-right[
```{tikz, fig.ext = 'png', echo = FALSE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
\node (p) {$D$};
\node [right of = p] (y) {$Y_{t}$};
\draw[->] (p) -- (y);
\end{tikzpicture}
```
]

---
class: middle

.pull-left[
- The drug is not available yet, but we have better chances of having Pin1 expression as a biomarker in a closer future.

- This biomarker could be measured in stored blood samples in a cohort study

- We can set a threshold for _Pin1 over-expression_ = $P$

- Since this is within an observational setting we will have to address confounding $(L)$

]

.pull-right[
```{tikz, fig.ext = 'png', echo = FALSE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
\node (p) {$P$};
\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$L$};
\node [right of = p] (y) {$Y_{t}$};
\draw[->] (l) -- (p);
\draw[->, violet] (p) -- (y);
\draw[->] (l) -- (y);
\end{tikzpicture}
```
]

---

class: middle

.pull-left[

- Pin1 over-expression stimulates tumor growth

- Tumor growth leads to cancer diagnosis

- Cancer diagnosis $(P^*)$ represents a proxy for $P$

- We are assuming that the captured effect of $P^*$ and $Y_{t}$ is only through $P$
]

.pull-right[
```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = ""}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
\node (p) {$P$};
\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$L$};
\node [above of = p, yshift = -1cm](pstar) {$P^*$};
\node [right of = p] (y) {$Y_{t}$};
\draw[->, violet] (p) -- (pstar);
\draw[->, violet] (p) -- (y);
\draw[->] (l) -- (p);
\draw[->] (l) -- (y);

\end{tikzpicture}
```
]

???

Not everyone who had Pin1 over-expression will be diagnosed with cancer by late-midlife, but in late-life.

---
class:middle

.pull-left[
Some people will die before having a cancer diagnosis

]

.pull-right[
```{tikz, fig.ext = 'png', echo = FALSE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
\node (p) {$P_0$};
\node [right of = p] (d) {$D_{k} = 0$};
\node [right of = d] (pstar) {$P^*_{k+1}$};
\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$C_1$};
\node [below of = d, yshift = 1cm, xshift = 1cm](c) {$C_2$};
\draw[->] (p) -- (d);
\draw[->, violet] (p) to [out=45,in=135, looseness=0.4] (pstar);
\draw[->] (d) -- (pstar);
\draw[->] (l) -- (p);
\draw[->] (l) -- (d);
\draw[->] (c) -- (pstar);
\draw[->] (c) -- (d);
\end{tikzpicture}
```
]

---
class: middle, center

## Overview of the causal question and assumptions related to death as a competing event of ADRD

---
class: middle

.pull-left[

]

.pull-right[
```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = ""}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
\node (p) {$P$};
\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$L$};
\node [above of = p, yshift = -1cm](pstar) {$P^*$};
\node [right of = p] (d) {$D_{t}$};
\node [right of = d] (y) {$Y_{t+1}$};
\draw[->, violet] (p) -- (pstar);
\draw[->] (p) -- (d);
\draw[->] (d) -- (y);
\draw[->] (l) -- (p);
\draw[->] (l) -- (y);
\draw[->, violet] (p) to [out=45,in=135, looseness=0.4] (y);
\end{tikzpicture}
```
]

---
class: middle

.pull-left[]

.pull-right[
```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = ""}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
\node (p) {$P$};
\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$L$};
\node [above of = p, yshift = -1cm](pstar) {$P^*$};
\node [right of = p] (d) {$D_{t}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = d, yshift = 1cm, xshift = +1cm](c) {$C$};
\draw[->, violet] (p) -- (pstar);
\draw[->] (pstar) -- (d);
\draw[->] (p) -- (d);
\draw[->] (d) -- (y);
\draw[->] (l) -- (p);
\draw[->] (l) -- (y);
\draw[->] (c) -- (d);
\draw[->] (c) -- (y);
\draw[->, violet] (p) to [out=45,in=135, looseness=0.4] (y);
\end{tikzpicture}
```

]

---

## Application to the Rotterdam Study

.pull-left[

- Participants between 60 and 70 years old at study entry
- Without history of cancer diagnosis and dementia
- Data on cancer diagnosis and dementia diagnosis is measured through follow-up and linkage with national registries.
- Date and cause of death was collected on a weekly basis via municipal population registries

]

.pull-right[
```{r, echo = FALSE}
plot_distribution +
    theme_xaringan()
```
]

---

### Methods

---

### Results

```{r, echo = FALSE}
kableExtra::kable(table_results, booktabs = T) %>%
  kableExtra::kable_styling(latex_options = c("striped", "hold_position"))
```

---

### Extreme scenarios of independency

```{r, echo = FALSE, fig.showtext=TRUE, fig.align='center', fig.width = 14}
table_results_bounds %>% 
  select(-5, -6) %>%
  pivot_longer(cols = 3:4,
               names_to = "cancer",
               values_to = "risk") %>%
  pivot_wider(names_from = c(model),
              values_from = risk) %>%
  ggplot(aes(
    y = Proxy,
    x = IPTW,
    xmin = `Lower bound`,
    xmax = `Upper bound`
  )) +
  geom_vline(xintercept = 1,
             color = "#E69F00",
             size = 1.2) +
  geom_pointrange(shape = 15,
                  color = "#0072B2") +
  geom_errorbar(width = 0.2, color = "#0072B2") + facet_wrap(. ~ cancer) +
  theme_xaringan()
```

---

```{r, echo = FALSE}
kableExtra::kable(table_results_bounds, booktabs = T) %>%
  kableExtra::kable_styling(latex_options = c("striped", "hold_position")) 
```

---

### Causes of death

```{r, echo = FALSE, fig.align='center', fig.width = 14}
causes_death + 
      theme_xaringan(text_font_size = 20,
                       title_font_size = 20) +
     theme(legend.position = "right")

```

### 2021-04-20 02_R/04_direct_effect_simplified.R

@@ -652,7 +652,7 @@ bounds_tv_cancer <- bind_rows(rd_2b, rd_2d, rd_2e) %>%
   arrange(rd) %>% 
   mutate(Proxy = "Time-varying cancer")
 
-bounds_t2vcancer <- bind_rows(rd_3a, rd_3c, rd_3d) %>% 
+bounds_t2vcancer <- bind_rows(rd_3b, rd_3c, rd_3d) %>% 
   select(model, everything()) %>% 
   arrange(rd) %>% 
   mutate(Proxy = "Time to cancer")

### 2021-04-20 02_R/05b_plot_causes_death.R

@@ -57,7 +57,7 @@ causes_death_dementia <- count_dementia_death %>%
   mutate(category = str_to_title(category)) %>% 
   mutate(outcome_plr = ifelse(outcome_plr == 0, "Had no dementia diagnosis", "Had dementia diagnosis"),
          category = as_factor(category),
-         category = fct_reorder(category, prop),
+         # category = fct_reorder(category, prop),
          outcome_plr = as_factor(outcome_plr)) %>% 
   ggplot(aes(outcome_plr, prop, fill = category)) +
   geom_col() +

### 2021-04-21 06_presentations/210421_cqm/index.Rmd
final presentation for cqp

@@
 options(htmltools.dir.version = FALSE)
-knitr::opts_chunk$set(collapse = TRUE,
-                      fig.retina = 3,
-                      warning = FALSE,
-                      message = FALSE, cache = TRUE)
+knitr::opts_chunk$set(collapse = TRUE, fig.retina = 3, warning = FALSE,
+                      message = FALSE, echo = FALSE, fig.showtext = TRUE, cache = TRUE)
 library(WeightIt)
@@
 library(splines)
+library(knitr)
+library(kableExtra)
 
+source(here::here("02_R", "04_direct_effect_simplified.R"))
 source(here::here("02_R", "05_plots.R"))
@@
 
-<br> <br>
+<br> <br> <br>
 
@@
 
-**Epidemiology department**
-
-**Causal inference and Neuroepi group**]
+**Epidemiology department**]
 
@@
 
-.pull-left[
+```{r, fig.align='center', out.width= "40%"}
+include_graphics("figs/causal_risks.PNG")
 
-- Molecular mechanisms
-
-- Cancer treatments
-
-- Shared factors
-
-- Unmeasured confounding, measurement error and survival bias may explain this association
-
-]
-
-.pull-right[
-
-![](figs/causal_risks.PNG)
-]
-
-.footnote[
-L. Frain et al. Alzheimer’s & Dementia. 2017.
-]
-
+```
+.footnote[L. Frain et al. Alzheimer’s & Dementia. 2017]
 ---
-class: middle, center
-
-## How can we conclude on causal pathways if we haven't measured them?
-
----
-
-## Back to basics
 
@@
 
-Driver J. et al. Biochimica et Biophysica Acta. 2015.
+<br><br><br>
 
@@
 
+.pull-right[
+.footnote[Driver J. et al. Biochimica et Biophysica Acta. 2015.]
+]
+---
+
+class: middle, center
+
+## How can we conclude on causal pathways with observational data?
+
 ---
@@
 
+- To progressively build a causal directed acyclic graph (DAG) which help us connect the causal question to the observable data and the assumptions we rely on to identify the effect.
+
+--
+
+- To exemplify different scenarios with real data collected from the Rotterdam Study.
+
+--
+
+- To describe the challenges and how they translate into the analytic decisions.
+
 ---
 
-### Overview of cancer diagnosis as a proxy for Pin1 enzime
+### Causal question
 
 .pull-left[
-If there was a drug that targets Pin1 $(D)$ we would perform an RCT to assess the effect on dementia $(Y_{t})$.
 
-We can imagine a setting where middle-aged participants with signs of cognitive decline take this drug and are followed over time.
+What is the effect of a drug targeted to Pin1 $(D)$ in the risk of ADRD at time *t* $(Y_{t})$  compared to standard treatments?
+
+We can imagine an RCT where middle-aged participants with signs of cognitive decline are recruited.
+
 ]
@@
 .pull-left[
-Some people will die before having a cancer diagnosis
+
+- In the cohort setting people will develop cancer over time
+
+- You need to survive long enough to have a cancer diagnosis $(D_k = 0)$
+
+]
+
+.pull-right[
+```{tikz, fig.ext = 'png', echo = FALSE}
+\usetikzlibrary{arrows}
+\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
+\node (p) {$P_0$};
+\node [right of = p] (d) {$D_{k} = 0$};
+\node [right of = d] (pstar) {$P^*_{k+1}$};
+\draw[->] (p) -- (d);
+\draw[->, violet] (p) to [out=45,in=135, looseness=0.4] (pstar);
+\draw[->] (d) -- (pstar);
+\end{tikzpicture}
+```
+]
+
+---
+
+class:middle
+
+.pull-left[
+
+- We need to isolate the effect of $P$ in $P^*$ and mimic a setting where over time those who died had would have had the same risk of developing cancer over time based on additional information $C_1, C_2$
 
@@
 
+- We are interested in the direct effect of cancer in dementia, not mediated through death
+
+- We can consider death as a censoring event that could be prevented
+
+- We rely on the independent censoring assumption
+
 ]
@@
 
-.pull-left[]
+.pull-left[
+
+- But we cannot consider $D_{t}$ and $Y_{t+1}$ as independent events unconditionally
+
+- We need information on time-varying factors that increase the risk of dying and having dementia over time $C$
+
+]
 
@@
 \draw[->, violet] (p) -- (pstar);
-\draw[->] (pstar) -- (d);
 \draw[->] (p) -- (d);
@@
 
+???
+In this setting death is treated as a censoring event and it can be interpreted as those who died would have the same risk of developing dementia if prevented from dying than those who remained alive and free of dementia at a given time point, conditional on shared causes of dementia and death
+
 ---
+class: middle
 
+.pull-left[
+
+- We also have to acknowledge that cancer diagnosis affects death through other mechanisms independent of $P$
+
+]
+
+.pull-right[
+```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = ""}
+\usetikzlibrary{arrows}
+\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
+\node (p) {$P$};
+\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$L$};
+\node [above of = p, yshift = -1cm](pstar) {$P^*$};
+\node [right of = p] (d) {$D_{t}$};
+\node [right of = d] (y) {$Y_{t+1}$};
+\node [below of = d, yshift = 1cm, xshift = +1cm](c) {$C$};
+\draw[->, violet] (p) -- (pstar);
+\draw[->] (p) -- (d);
+\draw[->] (pstar) -- (d);
+\draw[->] (d) -- (y);
+\draw[->] (l) -- (p);
+\draw[->] (l) -- (y);
+\draw[->] (c) -- (d);
+\draw[->] (c) -- (y);
+\draw[->, violet] (p) to [out=45,in=135, looseness=0.4] (y);
+\end{tikzpicture}
+```
+
+]
+
+---
+class: middle, center
+
+```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = ""}
+\usetikzlibrary{arrows}
+\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
+\node (p) {$P$};
+\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$L$};
+\node [right of = p] (dp) {$D_t$};
+\node [right of = dp] (pstar) {$P*_t$};
+\node [right of = pstar] (d) {$D_{t+1}$};
+\node [right of = d] (y) {$Y_{t+1}$};
+\node [above of = dp, yshift = -1cm, xshift = +1cm](c) {$C_1$};
+\node [above of = d, yshift = -1cm, xshift = +1cm](cd) {$C_2$};
+\draw[->] (p) -- (dp);
+\draw[->] (dp) -- (pstar);
+\draw[->, violet] (p) to [out=-45,in=-135, looseness=0.3] (pstar);
+\draw[->] (pstar) -- (d);
+\draw[->] (p) to [out=-45,in=-135, looseness=0.3] (d);
+\draw[->, violet] (p) to [out=-45,in=-135, looseness=0.3] (y);
+\draw[->] (d) -- (y);
+\draw[->] (l) -- (p);
+\draw[->] (l) to [out=0,in=270, looseness=0.4] (y);
+\draw[->] (c) -- (dp);
+\draw[->] (c) -- (pstar);
+\draw[->] (cd) -- (d);
+\draw[->] (cd) -- (y);
+\end{tikzpicture}
+```
+
+...all the pieces together...
+
+---
 ## Application to the Rotterdam Study
 
-.pull-left[
+We conduct an analysis of the cancer-ADRD association that is structured to the best of the available data’s abilities to inform the possible effect of Pin1 on all cause-dementia
+
+--
 
@@
 - Date and cause of death was collected on a weekly basis via municipal population registries
+- Data on covariates is measured repeatedly over visits and through linkage to EHR
 
-]
+---
 
-.pull-right[
+### Methods
+
+We considered 3 scenarios to define Pin1 over-expression:
+
+--
+
+- _"Cancer ever vs. never"_ as if Pin1 measurement and cancer diagnosis happened at the same time if looked retrospectively.
+
+--
+
+- Cancer as a _"time-varying"_ exposure independent of death.
+
+--
+
+- _"Time to cancer diagnosis"_ as the proxy for Pin1 over-expression.
+
+--
+
+We used inverse probability treatment weighting (IPTW) to address confounding.
+
+---
+
+### Methods
+
+- To compare the risk of dementia at 20 years of follow-up, in a setting were death could have been prevented, we compared the complement of a weighted Kaplan-Meier survival estimator in participants with incident cancer vs.
+no incident of cancer with time indexed in years.
+
+--
+
+- The weights in this case are time-varying by follow-up year, defined as a product of the time-fixed IPT weights above and a year-specific inverse probability of censoring by death weights (IPC).
+
+--
+
+- For an individual still alive in year _t_, the time _t_ IPC weight is the product of the inverse probability of surviving in each year prior to _t_, conditional on measured common causes of death and dementia.
+For an individual who has died by time t, the year t IPC weight is zero.
+
+--
+
+- We additionally calculated the direct effect considering death as an unconditional independent censoring event (as if there were no arrows from {C} to $D_{t}$ and $Y_{t+1}$) for an illustrative purpose.
+
+---
+
+### Methods
+
+Since the independent censoring assumption is untestable, we can consider extreme scenarios of independence and fit Peterson upper and lower bounds.
+
+--
+
+1) Total independence, that refers to an scenario were those who died would never develop dementia (lower bound).
+  + Aalen-Johansen estimator treating death as a competing event.
+
+--
+
+2) Complete dependency, that refers to an scenario where those who died would have dementia prior to death (upper bound). 
+  + The Kaplan Meier estimator for the combined outcome of dementia or death.
+
+---
+### Results
+
+.pull-left[
 ```{r, echo = FALSE}
@@
 
+--
+
+.pull-right[
+- 24% had cancer, the median age of cancer diagnosis was 73 (IQR: 69-77).
+
+- Among participants with cancer, 6% had dementia and 63% (n = 549) died.
+
+- Among participants free of cancer diagnosis, 14% had dementia and 23% died.
+]
 ---
+### Results
 
-### Methods
+*Weight assessment for cancer diagnosis*
+
+.center[
+```{r}
+love.plot(w.out1) +
+  theme_minimal() +
+  theme(legend.position = "bottom") +
+  scale_color_manual(values = c("#011A5E", "#e4a803"))
+
+```
+]
+
+---
+### Results
+
+*Weight assessment for death including baseline covariates*
+
+.center[
+```{r}
+love.plot(w.out_death) +
+  theme_minimal() +
+  theme(legend.position = "bottom") +
+  scale_color_manual(values = c("#011A5E", "#e4a803"))
+```
+]
+---
+### Cancer as: never vs. ever
+
+.pull-left[
+```{r}
+plot_1b
+```
+
+]
+
+--
+
+.pull-right[
+```{r}
+plot_1c
+```
+
+]
+
+---
+### Cancer as time-varying exposure
+
+.pull-left[
+```{r}
+plot_2b
+```
+
+]
+
+--
+
+.pull-right[
+```{r}
+plot_2c
+```
+
+]
 
@@
 
+### Time to cancer
+
+.pull-left[
+```{r}
+plot_3a
+```
+
+]
+
+--
+
+.pull-right[
+```{r}
+plot_3b
+```
+
+]
+---
 ### Results
 
+<br>
+
+.center[
 ```{r, echo = FALSE}
-kableExtra::kable(table_results, booktabs = T) %>%
-  kableExtra::kable_styling(latex_options = c("striped", "hold_position"))
+tb <- table_results %>%
+  select(-c(3, 4)) %>%
+  rename(RD = rd, RR = rr, `HR (95%CI)` = hr)
+
+tb %>% kbl() %>%
+kable_styling(bootstrap_options = c("striped", "hover"))
 ```
-
+]
 ---
@@
 
-```{r, echo = FALSE, fig.showtext=TRUE, fig.align='center', fig.width = 14}
+.pull-left[
+```{r, echo = FALSE}
 table_results_bounds %>% 
-  select(-5, -6) %>%
-  pivot_longer(cols = 3:4,
-               names_to = "cancer",
-               values_to = "risk") %>%
+  select(-3, -4, - 6) %>%
     pivot_wider(names_from = c(model),
-              values_from = risk) %>%
+              values_from = rd) %>%
   ggplot(aes(
     y = Proxy,
-    x = IPTW,
+    x = `IPTW + IPCW`,
+    xmin = `Lower bound`,
+    xmax = `Upper bound`
+  )) +
+  geom_vline(xintercept = 0,
+             color = "#E69F00",
+             size = 1.2) +
+  geom_pointrange(shape = 15,
+                  color = "#0072B2") +
+  geom_errorbar(width = 0.2, color = "#0072B2") +
+  labs(x = "Risk difference",
+       y = NULL) + 
+  theme_xaringan()
+```
+]
+
+--
+
+.pull-right[
+```{r, echo = FALSE}
+table_results_bounds %>% 
+  select(-3, -4, - 5) %>%
+    pivot_wider(names_from = c(model),
+              values_from = rr) %>%
+  ggplot(aes(
+    y = Proxy,
+    x = `IPTW + IPCW`,
     xmin = `Lower bound`,
@@
                   color = "#0072B2") +
-  geom_errorbar(width = 0.2, color = "#0072B2") + facet_wrap(. ~ cancer) +
+  geom_errorbar(width = 0.2, color = "#0072B2") +
+  labs(x = "Risk Ratio",
+       y = NULL) + 
+   scale_x_continuous(
+    trans = "log10",
+    # breaks = seq(0.7, 1.2, by = 0.1)
+  ) +
   theme_xaringan()
 ```
-
----
-
-```{r, echo = FALSE}
-kableExtra::kable(table_results_bounds, booktabs = T) %>%
-  kableExtra::kable_styling(latex_options = c("striped", "hold_position")) 
-```
+]
 
@@
 ```
+
+---
+
+### Thoughts so far...
+
+- There big gap between what we measure and analyze as exposures, and what we truly want to study.
+
+--
+
+- The process of connecting our real question to the data help us identify the structure that will later represent the analytic decisions and assumptions.
+
+--
+
+- When death is a competing event, prior to choosing the estimator we need to define what is the estimand of interest and be explicit about the assumptions and interpretations.
+
+--
+
+- Descriptive information, such as % of people who died in each arm should be always presented.
+
+---
+
+### Questions
+
+- How can I make my points clear for a wide audience?
+
+- What to add, what to remove?

### 2021-04-21 02_R/04_direct_effect_simplified.R

@@ -185,7 +185,7 @@ cancer_ever_km <-
     cluster = id
   )
 
-plot_km(cancer_ever_km, "Risk of dementia, ever vs. never") +
+plot_1b <- plot_km(cancer_ever_km, "Risk of dementia, ever vs. never") +
   labs(subtitle = "IPTW")
 
 cancer_ever_cox <-
@@ -212,8 +212,8 @@ cancer_ever_km_ipcw <-
     cluster = id
   )
 
-plot_km(cancer_ever_km_ipcw, "Direct effect") +
-  labs(subtitle = "With IPCW for death")
+plot_1c <- plot_km(cancer_ever_km_ipcw, "Risk of dementia, ever vs. never") +
+  labs(subtitle = "IPTW + IPCW")
 
 cancer_ever_cox_ipcw <-
   coxph(
@@ -375,7 +375,7 @@ km_tv_iptw <- survfit(Surv(
 
 # plot(km_iptw)
 
-plot_km(km_tv_iptw, "Risk of dementia, Time-varying cancer") + 
+plot_2b <- plot_km(km_tv_iptw, "Risk of dementia, Time-varying cancer") + 
   labs(subtitle = "IPTW")
 
 cox_tv_iptw <- coxph(Surv(
@@ -398,8 +398,8 @@ km_tv_ipcw <- survfit(Surv(
 
 # plot(km_ipcw)
 
-plot_km(km_tv_ipcw, "Risk of dementia, time-varying cancer") + 
-  labs(subtitle = "With IPTW, conditional censoring on time-v")
+plot_2c <- plot_km(km_tv_ipcw, "Risk of dementia, time-varying cancer") + 
+  labs(subtitle = "IPTW + IPCW")
 
 cox_tv_ipcw <- coxph(Surv(
   tstart, time2 = fuptime, event = outcome_plr) ~ cancer_v,
@@ -548,8 +548,8 @@ km_t2cancer_iptw <- survfit(Surv(
   tstart, time2 = fuptime, event = outcome_plr) ~ cancer_v,
   data = data_long, cluster = id, weights = sw_t2cancer_t)
 
-plot_km(km_t2cancer_iptw, "Risk of dementia") + 
-  labs(subtitle = "IPTW for time to cancer")
+plot_3a <- plot_km(km_t2cancer_iptw, "Risk of dementia, time to cancer") + 
+  labs(subtitle = "IPTW")
 
 
[…]

### 2021-04-23 06_presentations/ser2021/index.Rmd
SER presentation

--- 2021-04-21 06_presentations/210421_cqm/index.Rmd
@@
 ---
-title: "‘Dissecting the causal question underlying the association between cancer and dementia"
+title: "Toward a clearer understanding of the inverse association between cancer and dementia"
 subtitle: ""  
@@
   text_font_google   = google_font("Lato", "400", "400i"),
-  text_font_size = "25px"
+  text_font_size = "30px"
 )
@@
 
-background-image: url(./figs/erasmus.png)
-background-position: 5% 95%
+background-image: url(./figs/ser.jpg), url(./figs/erasmus.png)
+background-position: 5% 95%, 95% 5%
 background-size: 25%, 15%
@@
 
-## Dissecting the causal question underlying the association between cancer and dementia
+## Towards a clearer understanding of the inverse association between cancer and dementia
 
@@
 ---
-
 ## Introduction
 
-.pull-left[
+<br><br>
 
-Observational studies have found a protective effect of cancer diagnosis in the risk of Alzheimer´s disease or related dementias (ADRD).
-
-These studies lead to discussions of repurposing or augmenting current cancer chemotherapeutics for ADRD
-]
+- Observational studies have found an inverse association between cancer and the risk of Alzheimer´s disease or related dementias (ADRD).
 
@@
 
-.pull-right[ 
-![](figs/forest_plot.PNG)
-]
+- To understand the potential sources of bias we first need to define the causal question.
 
-.footnote[Ospina-Romero M. JAMA Network Open. 2020]
+--
 
----
-
-## Cancer diagnosis as an exposure
-
-```{r, fig.align='center', out.width= "40%"}
-include_graphics("figs/causal_risks.PNG")
-
-```
-.footnote[L. Frain et al. Alzheimer’s & Dementia. 2017]
----
-
-.pull-left[
-![](figs/pin_knockout.PNG) 
-]
-
-.pull-right[
-
-<br><br><br>
-
-![](figs/tment_pin1.PNG)
-
-]
-
-.pull-right[
-.footnote[Driver J. et al. Biochimica et Biophysica Acta. 2015.]
-]
----
-
-class: middle, center
-
-## How can we conclude on causal pathways with observational data?
+- Refining the underlying question behind cancer diagnosis as the exposure unveils some of these concerns.
 
@@
 
-- To progressively build a causal directed acyclic graph (DAG) which help us connect the causal question to the observable data and the assumptions we rely on to identify the effect.
-
---
-
-- To exemplify different scenarios with real data collected from the Rotterdam Study.
-
---
-
-- To describe the challenges and how they translate into the analytic decisions.
+To conceptualize and progressively build a causal directed acyclic graph to connect this particular causal question to the observable data and the assumptions we rely on to identify the effect.
 
 ---
-
-### Causal question
-
-.pull-left[
-
-What is the effect of a drug targeted to Pin1 $(D)$ in the risk of ADRD at time *t* $(Y_{t})$  compared to standard treatments?
-
-We can imagine an RCT where middle-aged participants with signs of cognitive decline are recruited.
-
-]
-
---
+### Pin1-targeting drug
 
@@
 
-.pull-right[
+.center[
 ```{tikz, fig.ext = 'png', echo = FALSE}
@@
 ```
-]
 
----
-class: middle
-
-.pull-left[
-- The drug is not available yet, but we have better chances of having Pin1 expression as a biomarker in a closer future.
-
-- This biomarker could be measured in stored blood samples in a cohort study
-
-- We can set a threshold for _Pin1 over-expression_ = $P$
-
-- Since this is within an observational setting we will have to address confounding $(L)$
+What is the effect of a Pin1 targeting drug $(D)$ in the risk of ADRD over time *t* $(Y_{t})$  compared to standard treatments?
 
@@
 
-.pull-right[
+???
+In hypothetical setting, if Pin1-targeting drug was developed, the best way to understand its effect on dementia risk would be to do a randomized trial in which we randomize eligible participants in late midlife to receive this drug or not, and closely monitor ADRD over follow-up.
+
+---
+### Pin1 expression as biomarker
+
+<br><br>
+
+.center[
 ```{tikz, fig.ext = 'png', echo = FALSE}
@@
 \begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
-\node (p) {$P$};
+\node (p) {$P_0$};
 \node [below of = p, yshift = 1cm, xshift = -1cm](l) {$L$};
@@
 ```
+
+$P_0$ = Pin1 over-expression; $L$ = shared factors between $P$ and $Y_t$
+
 ]
 
+???
+We might be closer to developing biomarker for Pin1 expression in the future.
+So we can imagine a situation in which Pin1 exposure is measured in a population based cohort study, using stored blood samples.
+Since this would be a observational setting, Pin1 expression and ADRD at time may share causes such as smoking, and that assessing the causal relationship requires adjusting for these confounders.
+
 ---
+### Cancer diagnosis as the proxy of Pin1 biomarker
 
-class: middle
+<br>
 
-.pull-left[
-
-- Pin1 over-expression stimulates tumor growth
-
-- Tumor growth leads to cancer diagnosis
-
-- Cancer diagnosis $(P^*)$ represents a proxy for $P$
-
-- We are assuming that the captured effect of $P^*$ and $Y_{t}$ is only through $P$
-]
-
-.pull-right[
+.center[
 ```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = ""}
@@
 ```
+
+$P_0$ = Pin1 over-expression; $P^*$ = Cancer diagnosis; $L$ = shared factors between $P$ and $Y_t$ = ADRD diagnosis
 ]
@@
 ???
-
-Not everyone who had Pin1 over-expression will be diagnosed with cancer by late-midlife, but in late-life.
+At this moment this biomarker is not available, but we know that Pin1 over-expression stimulates tumor growth and for this reason we use cancer diagnosis as the proxy for Pin1.
 
 ---
-class:middle
+### Zoom in the P and P* association
 
-.pull-left[
+<br>
 
-- In the cohort setting people will develop cancer over time
-
-- You need to survive long enough to have a cancer diagnosis $(D_k = 0)$
-
-]
-
-.pull-right[
+.center[
 ```{tikz, fig.ext = 'png', echo = FALSE}
@@
 ```
-]
 
----
+<br> <br>
 
-class:middle
-
-.pull-left[
-
-- We need to isolate the effect of $P$ in $P^*$ and mimic a setting where over time those who died had would have had the same risk of developing cancer over time based on additional information $C_1, C_2$
+$P_0$ = Pin1 over-expression; $D_{k} = 0$ = surviving by time $k$; $P^*_{k+1}$ = Cancer diagnosis by time $k+1$ ;
 
@@
 
-.pull-right[
+???
+
+---
+### Selection bias in the P and P* association
+
+<br>
+
+.center[
 ```{tikz, fig.ext = 'png', echo = FALSE}
@@
 ```
-]
 
----
-class: middle, center
+<br> <br>
 
-## Overview of the causal question and assumptions related to death as a competing event of ADRD
-
----
-class: middle
-
-.pull-left[
-
-- We are interested in the direct effect of cancer in dementia, not mediated through death
-
-- We can consider death as a censoring event that could be prevented
-
-- We rely on the independent censoring assumption
+$P_0$ = Pin1 over-expression; $D_{k} = 0$ = surviving by time $k$; $P^*_{k+1}$ = Cancer diagnosis by time $k+1$; $C_1$, $C_2$ shared factors such as smoking
 
@@
 
-.pull-right[
+???
+In some way we need to assume that those who died would have the same risk of having a cancer diagnosis had they not died 
+---
+### Direct effect vs. total effect, potential bias
+
+<br>
+
+.center[
 ```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = ""}
@@
 \node (p) {$P$};
-\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$L$};
-\node [above of = p, yshift = -1cm](pstar) {$P^*$};
-\node [right of = p] (d) {$D_{t}$};
-\node [right of = d] (y) {$Y_{t+1}$};
-\draw[->, violet] (p) -- (pstar);
-\draw[->] (p) -- (d);
-\draw[->] (d) -- (y);
-\draw[->] (l) -- (p);
-\draw[->] (l) -- (y);
-\draw[->, violet] (p) to [out=45,in=135, looseness=0.4] (y);
-\end{tikzpicture}
-```
-]
-
----
-class: middle
-
-.pull-left[
-
-- But we cannot consider $D_{t}$ and $Y_{t+1}$ as independent events unconditionally
-
-- We need information on time-varying factors that increase the risk of dying and having dementia over time $C$
-
-]
-
-.pull-right[
-```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = ""}
-\usetikzlibrary{arrows}
-\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
-\node (p) {$P$};
-\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$L$};
 \node [above of = p, yshift = -1cm](pstar) {$P^*$};
@@
 \draw[->] (p) -- (d);
-\draw[->] (d) -- (y);
-\draw[->] (l) -- (p);
-\draw[->] (l) -- (y);
+\draw[->, very thick] (d) -- (y);
 \draw[->] (c) -- (d);
@@
 ```
-
 ]
@@
 ---
-class: middle
+### All pieces together
 
-.pull-left[
-
-- We also have to acknowledge that cancer diagnosis affects death through other mechanisms independent of $P$
-
-]
-
-.pull-right[
-```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = ""}
-\usetikzlibrary{arrows}
-\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
-\node (p) {$P$};
-\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$L$};
-\node [above of = p, yshift = -1cm](pstar) {$P^*$};
-\node [right of = p] (d) {$D_{t}$};
-\node [right of = d] (y) {$Y_{t+1}$};
-\node [below of = d, yshift = 1cm, xshift = +1cm](c) {$C$};
-\draw[->, violet] (p) -- (pstar);
-\draw[->] (p) -- (d);
-\draw[->] (pstar) -- (d);
-\draw[->] (d) -- (y);
-\draw[->] (l) -- (p);
-\draw[->] (l) -- (y);
-\draw[->] (c) -- (d);
-\draw[->] (c) -- (y);
-\draw[->, violet] (p) to [out=45,in=135, looseness=0.4] (y);
-\end{tikzpicture}
-```
-
-]
-
----
-class: middle, center
-
+.center[
 ```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = ""}
@@
 
-...all the pieces together...
+<br><br>
+
+At best we can test the sharp null hypothesis
+]
 
@@
 
-We conduct an analysis of the cancer-ADRD association that is structured to the best of the available data’s abilities to inform the possible effect of Pin1 on all cause-dementia
+- Participants between 60 and 70 years old at study entry.
 
---
+- Data on cancer diagnosis, dementia diagnosis and death was collected from national registries and electronical clinical records.
 
-- Participants between 60 and 70 years old at study entry
-- Without history of cancer diagnosis and dementia
-- Data on cancer diagnosis and dementia diagnosis is measured through follow-up and linkage with national registries.
-- Date and cause of death was collected on a weekly basis via municipal population registries
-- Data on covariates is measured repeatedly over visits and through linkage to EHR
+Three definitions of the proxy for Pin1 over-expression:
+
+- _"Cancer ever vs. never"_
+
+- Cancer as a _"time-varying"_
+
+- _"Time to cancer diagnosis"_
 
@@
 
-We considered 3 scenarios to define Pin1 over-expression:
+- Inverse probability treatment weights to address confounding
+
+  + *Age, sex, education, smoking status*
 
@@
 
-- _"Cancer ever vs. never"_ as if Pin1 measurement and cancer diagnosis happened at the same time if looked retrospectively.
+- Inverse probability censoring weights for death to relax the independent censoring assumption
+
+  + *Age, sex, Apoe4, education and time-varying covariates: smoking, systolic blood pressure, BMI and incident cancer, heart disease, stroke and diabetes*
 
@@
 
-- Cancer as a _"time-varying"_ exposure independent of death.
-
---
-
-- _"Time to cancer diagnosis"_ as the proxy for Pin1 over-expression.
-
---
-
-We used inverse probability treatment weighting (IPTW) to address confounding.
-
----
-
-### Methods
-
-- To compare the risk of dementia at 20 years of follow-up, in a setting were death could have been prevented, we compared the complement of a weighted Kaplan-Meier survival estimator in participants with incident cancer vs.
-no incident of cancer with time indexed in years.
-
---
-
-- The weights in this case are time-varying by follow-up year, defined as a product of the time-fixed IPT weights above and a year-specific inverse probability of censoring by death weights (IPC).
-
---
-
-- For an individual still alive in year _t_, the time _t_ IPC weight is the product of the inverse probability of surviving in each year prior to _t_, conditional on measured common causes of death and dementia.
-For an individual who has died by time t, the year t IPC weight is zero.
-
---
-
-- We additionally calculated the direct effect considering death as an unconditional independent censoring event (as if there were no arrows from {C} to $D_{t}$ and $Y_{t+1}$) for an illustrative purpose.
+- We estimate the risk of dementia using a weighted Kaplan-Meier estimator.
 
@@
 ---
-### Results
-
-*Weight assessment for cancer diagnosis*
-
-.center[
-```{r}
-love.plot(w.out1) +
-  theme_minimal() +
-  theme(legend.position = "bottom") +
-  scale_color_manual(values = c("#011A5E", "#e4a803"))
-
-```
-]
-
----
-### Results
-
-*Weight assessment for death including baseline covariates*
-
-.center[
-```{r}
-love.plot(w.out_death) +
-  theme_minimal() +
-  theme(legend.position = "bottom") +
-  scale_color_manual(values = c("#011A5E", "#e4a803"))
-```
-]
----
 ### Cancer as: never vs. ever
@@
 plot_1c
-```
-
-]
-
----
-### Cancer as time-varying exposure
-
-.pull-left[
-```{r}
-plot_2b
-```
-
-]
-
---
-
-.pull-right[
-```{r}
-plot_2c
 ```
@@
 tb <- table_results %>%
+  filter(model != "Unadjusted") %>% 
   select(-c(3, 4)) %>%
@@
 
-### Extreme scenarios of independency
+### Extreme scenarios of independency between death and dementia
 
@@
 ---
-
-### Causes of death
-
-```{r, echo = FALSE, fig.align='center', fig.width = 14}
-causes_death + 
-      theme_xaringan(text_font_size = 20,
-                       title_font_size = 20) +
-     theme(legend.position = "right")
-
-```
-
----
-
-### Thoughts so far...
-
-- There big gap between what we measure and analyze as exposures, and what we truly want to study.
-
---
+### Conclusions
 
@@
 
+- Cancer as the proxy for a molecular mechanism is tricky...
+--
+
 - When death is a competing event, prior to choosing the estimator we need to define what is the estimand of interest and be explicit about the assumptions and interpretations.
 
---
-
-- Descriptive information, such as % of people who died in each arm should be always presented.
-
 ---
-
-### Questions
-
-- How can I make my points clear for a wide audience?
-
-- What to add, what to remove?

### 2021-04-27 02_R/04_direct_effect_simplified.R
bounds for ever vs. never, lower bound fx.

@@ -49,6 +49,9 @@ data_wide %<>%
   mutate(t2dem_efu =
            ifelse(t2dem >= 240, 240, t2dem))
 
+data_wide %<>%
+  mutate(combined_outcome_efu = ifelse(dementia_efu != 0, 1, 0))
+
 ## If competing_plr = 1, then outcome_plr = NA
 
 data_long %<>% 
@@ -59,6 +62,7 @@ data_long %<>%
 
 # 1.1. IPTW (cancer ever vs. never) ------------------------------------------
 
+# By hand
 # cancer_den <-
 #   glm(
 #     cancer_v ~ bs(age_0) + sex + education + apoe4 + as.factor(smoke1),
@@ -79,7 +83,7 @@ data_long %<>%
 #     w_cancer = ifelse(cancer_v == 1, p_num/p_denom, (1 - p_num)/(1- p_denom)))
 
 
-## Check standardized mean
+## Weightit package, makes nice loveplots
 
 w.out1 <-
   weightit(
@@ -234,6 +238,34 @@ results_ever_never <- bind_rows(hr_1a, hr_1b, hr_1c) %>%
   left_join(bind_rows(rd_1a, rd_1b, rd_1c)) %>% 
   mutate(Proxy = "Ever vs. Never") %>% 
   select(Proxy, model, starts_with("cancer"), rd, rr, hr) 
+
+
+# 1.d. Lower bound --------------------------------------------------------
+
+km_ever_ipcw_low_bound <- survfit(
+  Surv(
+    t2dem_efu,
+    event = as.factor(dementia_efu),
+    type = 'mstate'
+  ) ~ cancer_v,
+  data = data_wide,
+  weights = baseline_wt)
+
+rd_1d <- lower_bound(km_ever_ipcw_low_bound)
+
+# 1.e. Upper bound -------------------------------------------------------------
+
+km_ever_ipcw_up_bound <- survfit(
+  Surv(
+    t2dem_efu,
+    event = combined_outcome_efu,
+  ) ~ cancer_v,
+  data = data_wide,
+  weights = baseline_wt)
+
+rd_1e  <- km_ever_ipcw_up_bound %>% 
+  risks_km()  %>% 
+  mutate(model = "Upper bound")
 
 
 # 2. Time-varying cancer -----------------------------------------------------
@@ -432,19 +464,7 @@ km_tv_ipcw_low_bound <- survfit(
[…]

### 2021-04-27 02_R/04b_auxiliary_fx.R

@@ -111,3 +111,20 @@ tidy_hr <- function(model) {
     select(-1, hr = estimate) %>% 
     mutate(hr = paste0(hr, " (", conf.low, ", ", conf.high, ")"))
 }
+
+
+lower_bound <- function(data){
+  data  %>% 
+    broom::tidy() %>% 
+    filter(state == "1") %>%
+    group_by(strata) %>% 
+    slice(n()) %>% 
+    select(estimate, strata) %>% 
+    pivot_wider(names_from = strata, values_from = estimate) %>% 
+    mutate(rd = .[[2]] - .[[1]],
+           rr = .[[2]] / .[[1]]) %>% 
+    mutate_at(c(1:3), ~.*100) %>% 
+    mutate_at(c(1:3), round, 1) %>% 
+    mutate(rr = round(rr, 2)) %>% 
+    mutate(model = "Lower bound")
+}

### 2021-04-27 03_figs/tikz364831c2bf8.tex
ser presentation updates

\documentclass[tikz]{standalone}
\usepackage{amsmath}
\usetikzlibrary{matrix}
%% EXTRA_TIKZ_PREAMBLE_CODE %%
\begin{document}
%% TIKZ_CODE %%
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
\node (p) {$P$};
\node [below of = p, yshift = 1cm, xshift = -1cm, gray!50](l) {$L$};
\node [right of = p, gray!50] (dp) {$D_t$};
\node [right of = dp, gray!50] (pstar) {$P*_t$};
\node [right of = pstar, gray!50] (d) {$D_{t+1}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [above of = dp, yshift = -1cm, xshift = +1cm, gray!50](c) {$C_1$};
\node [above of = d, yshift = -1cm, xshift = +1cm, gray!50](cd) {$C_2$};
\draw[->, gray!50] (p) -- (dp);
\draw[->, gray!50] (dp) -- (pstar);
\draw[->, gray!50] (p) to [out=-45,in=-135, looseness=0.3] (pstar);
\draw[->, gray!50] (pstar) -- (d);
\draw[->, gray!50] (p) to [out=-45,in=-135, looseness=0.3] (d);
\draw[->] (p) to [out=-45,in=-135, looseness=0.3] (y);
\draw[->, gray!50] (d) -- (y);
\draw[->, gray!50] (l) -- (p);
\draw[->, gray!50] (l) to [out=0,in=270, looseness=0.4] (y);
\draw[->, gray!50] (c) -- (dp);
\draw[->, gray!50] (c) -- (pstar);
\draw[->, v] (cd) -- (d);
\draw[->, gray!50] (cd) -- (y);
\end{tikzpicture}
\end{document}

### 2021-04-27 06_presentations/ser2021/index.Rmd
ser presentation updates

@@
 
-```{r setup, include=FALSE, }
+```{r setup, include=FALSE}
 options(htmltools.dir.version = FALSE)
-knitr::opts_chunk$set(collapse = TRUE, fig.retina = 3, warning = FALSE,
-                      message = FALSE, echo = FALSE, fig.showtext = TRUE, cache = TRUE)
-library(WeightIt)
-library(cobalt)
+knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE, fig.showtext = TRUE, cache = TRUE)
+
 library(splines)
@@
 
-## Aims
-
-To conceptualize and progressively build a causal directed acyclic graph to connect this particular causal question to the observable data and the assumptions we rely on to identify the effect.
-
----
 ### Pin1-targeting drug
 
-<br><br>
-
-.center[
-```{tikz, fig.ext = 'png', echo = FALSE}
+```{tikz, echo = FALSE}
 \usetikzlibrary{arrows}
-\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
-\node (p) {$D$};
-\node [right of = p] (y) {$Y_{t}$};
-\draw[->] (p) -- (y);
+\begin{tikzpicture}[node distance=2cm]
+\node (p) {$P_0$};
+\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
+\node [right of = p, gray!30] (dp) {$D_t$};
+\node [right of = dp, gray!30] (pstar) {$P^*_t$};
+\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
+\node [right of = d] (y) {$Y_{t+1}$};
+\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
+\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
+\draw[->, gray!30] (p) -- (dp);
+\draw[->, gray!30] (l) -- (p);
+\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
+\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->, gray!30] (dp) -- (pstar);
+\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (pstar);
+\draw[->, gray!30] (pstar) -- (d);
+\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (d);
+\draw[->, gray!30] (d) -- (y);
+\draw[->, gray!30] (c) -- (dp);
+\draw[->, gray!30] (c) -- (pstar);
+\draw[->, gray!30] (cd) -- (d);
+\draw[->, gray!30] (cd) -- (y);
 \end{tikzpicture}
@@
 
-What is the effect of a Pin1 targeting drug $(D)$ in the risk of ADRD over time *t* $(Y_{t})$  compared to standard treatments?
-
-]
+$P_0$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis
 
@@
 ---
-### Pin1 expression as biomarker
+### Pin1 expression as biomarker (Problem 1: _Confounding_)
 
-<br><br>
-
-.center[
-```{tikz, fig.ext = 'png', echo = FALSE}
+```{tikz, echo = FALSE}
 \usetikzlibrary{arrows}
-\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
+\begin{tikzpicture}[node distance=2cm]
 \node (p) {$P_0$};
-\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$L$};
-\node [right of = p] (y) {$Y_{t}$};
+\node [above of = p, yshift = -1cm, xshift = 1cm](l) {$L$};
+\node [right of = p, gray!30] (dp) {$D_t$};
+\node [right of = dp, gray!30] (pstar) {$P^*_t$};
+\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
+\node [right of = d] (y) {$Y_{t+1}$};
+\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
+\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
+\draw[->, gray!30] (p) -- (dp);
 \draw[->] (l) -- (p);
-\draw[->, violet] (p) -- (y);
-\draw[->] (l) -- (y);
+\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
+\draw[->, semithick] (p) to [out=45,in=135, looseness=0.3] (y);
+\draw[->, gray!30] (dp) -- (pstar);
+\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (pstar);
+\draw[->, gray!30] (pstar) -- (d);
+\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (d);
+\draw[->, gray!30] (d) -- (y);
+\draw[->, gray!30] (c) -- (dp);
+\draw[->, gray!30] (c) -- (pstar);
+\draw[->, gray!30] (cd) -- (d);
+\draw[->, gray!30] (cd) -- (y);
 \end{tikzpicture}
@@
 
-$P_0$ = Pin1 over-expression; $L$ = shared factors between $P$ and $Y_t$
-
-]
+$P_0$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis; $L$ = Shared causes of $P_0$ and $Y_{t+1}$
 
@@
 ---
-### Cancer diagnosis as the proxy of Pin1 biomarker
+### Cancer diagnosis as the proxy of Pin1 biomarker (Problem 2: _Time zero_)
 
-<br>
-
-.center[
-```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = ""}
+```{tikz, echo = FALSE}
 \usetikzlibrary{arrows}
-\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
+\begin{tikzpicture}[node distance=2cm]
 \node (p) {$P_0$};
-\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$L$};
-\node [above of = p, yshift = -1cm](pstar) {$P^*$};
-\node [right of = p] (y) {$Y_{t}$};
-\draw[->, violet] (p) -- (pstar);
-\draw[->, violet] (p) -- (y);
-\draw[->] (l) -- (p);
-\draw[->] (l) -- (y);
-
+\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
+\node [right of = p, gray!30] (dp) {$D_t$};
+\node [right of = dp] (pstar) {$P^*_t$};
+\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
+\node [right of = d] (y) {$Y_{t+1}$};
+\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
+\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
+\draw[->, gray!30] (p) -- (dp);
+\draw[->, gray!30] (l) -- (p);
+\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
+\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->, gray!30] (dp) -- (pstar);
+\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
+\draw[->, gray!30] (pstar) -- (d);
+\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (d);
+\draw[->, gray!30] (d) -- (y);
+\draw[->, gray!30] (c) -- (dp);
+\draw[->, gray!30] (c) -- (pstar);
+\draw[->, gray!30] (cd) -- (d);
+\draw[->, gray!30] (cd) -- (y);
 \end{tikzpicture}
@@
 
-$P_0$ = Pin1 over-expression; $P^*$ = Cancer diagnosis; $L$ = shared factors between $P$ and $Y_t$ = ADRD diagnosis
-]
+$P_0$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis; $P^*_t$ = Cancer diagnosis
 
@@
 ---
-### Zoom in the P and P* association
+### Death prior to cancer diagnosis (Problem 3: _Survival bias_)
 
-<br>
-
-.center[
-```{tikz, fig.ext = 'png', echo = FALSE}
+```{tikz, echo = FALSE}
 \usetikzlibrary{arrows}
-\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
+\begin{tikzpicture}[node distance=2cm]
 \node (p) {$P_0$};
-\node [right of = p] (d) {$D_{k} = 0$};
-\node [right of = d] (pstar) {$P^*_{k+1}$};
-\draw[->] (p) -- (d);
-\draw[->, violet] (p) to [out=45,in=135, looseness=0.4] (pstar);
-\draw[->] (d) -- (pstar);
+\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
+\node [right of = p] (dp) {$D_t$};
+\node [right of = dp] (pstar) {$P^*_t$};
+\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
+\node [right of = d] (y) {$Y_{t+1}$};
+\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
+\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
+\draw[->] (p) -- (dp);
+\draw[->, gray!30] (l) -- (p);
+\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
+\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->] (dp) -- (pstar);
+\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
+\draw[->, gray!30] (pstar) -- (d);
+\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (d);
+\draw[->, gray!30] (d) -- (y);
+\draw[->, gray!30] (c) -- (dp);
+\draw[->, gray!30] (c) -- (pstar);
+\draw[->, gray!30] (cd) -- (d);
+\draw[->, gray!30] (cd) -- (y);
 \end{tikzpicture}
@@
 
-<br> <br>
-
-$P_0$ = Pin1 over-expression; $D_{k} = 0$ = surviving by time $k$; $P^*_{k+1}$ = Cancer diagnosis by time $k+1$ ;
-
-]
-
+$P_0$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis; $P^*_t$ = Cancer diagnosis; $D_t$ = Death prior to cancer diagnosis
 ???
@@
 ---
-### Selection bias in the P and P* association
+### Death prior to cancer diagnosis (Problem 3: _Survival bias_)
 
-<br>
-
-.center[
-```{tikz, fig.ext = 'png', echo = FALSE}
+```{tikz, echo = FALSE}
 \usetikzlibrary{arrows}
-\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
+\begin{tikzpicture}[node distance=2cm]
 \node (p) {$P_0$};
-\node [right of = p] (d) {$D_{k} = 0$};
-\node [right of = d] (pstar) {$P^*_{k+1}$};
-\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$C_1$};
-\node [below of = d, yshift = 1cm, xshift = 1cm](c) {$C_2$};
-\draw[->] (p) -- (d);
-\draw[->, violet] (p) to [out=45,in=135, looseness=0.4] (pstar);
-\draw[->] (d) -- (pstar);
-\draw[->] (l) -- (p);
-\draw[->] (l) -- (d);
+\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
+\node [right of = p] (dp) {$D_t$};
+\node [right of = dp] (pstar) {$P^*_t$};
+\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
+\node [right of = d] (y) {$Y_{t+1}$};
+\node [below of = dp, yshift = 1cm, xshift = 1cm](c) {$C_1$};
+\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
+\draw[->] (p) -- (dp);
+\draw[->, gray!30] (l) -- (p);
+\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
+\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->] (dp) -- (pstar);
+\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
+\draw[->, gray!30] (pstar) -- (d);
+\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (d);
+\draw[->, gray!30] (d) -- (y);
+\draw[->] (c) -- (dp);
 \draw[->] (c) -- (pstar);
-\draw[->] (c) -- (d);
+\draw[->, gray!30] (cd) -- (d);
+\draw[->, gray!30] (cd) -- (y);
 \end{tikzpicture}
@@
 
-<br> <br>
+$P_0$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis; $P^*_t$ = Cancer diagnosis; $D_t$ = Death prior to cancer diagnosis; $C_1$ = Shared causes of $D_t$ and $P^*$
+???
 
-$P_0$ = Pin1 over-expression; $D_{k} = 0$ = surviving by time $k$; $P^*_{k+1}$ = Cancer diagnosis by time $k+1$; $C_1$, $C_2$ shared factors such as smoking
+---
+### Death prior to dementia diagnosis (Problem 3: _Survival bias_)
 
-]
-
-???
-In some way we need to assume that those who died would have the same risk of having a cancer diagnosis had they not died 
----
-### Direct effect vs. total effect, potential bias
-
-<br>
-
-.center[
-```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = ""}
+```{tikz, echo = FALSE}
 \usetikzlibrary{arrows}
-\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
-\node (p) {$P$};
-\node [above of = p, yshift = -1cm](pstar) {$P^*$};
-\node [right of = p] (d) {$D_{t}$};
+\begin{tikzpicture}[node distance=2cm]
+\node (p) {$P_0$};
+\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
+\node [right of = p, gray!30, gray!30] (dp) {$D_t$};
+\node [right of = dp, gray!30] (pstar) {$P^*_t$};
+\node [right of = pstar] (d) {$D_{t+1}$};
 \node [right of = d] (y) {$Y_{t+1}$};
-\node [below of = d, yshift = 1cm, xshift = +1cm](c) {$C$};
-\draw[->, violet] (p) -- (pstar);
-\draw[->] (p) -- (d);
-\draw[->, very thick] (d) -- (y);
-\draw[->] (c) -- (d);
-\draw[->] (c) -- (y);
-\draw[->, violet] (p) to [out=45,in=135, looseness=0.4] (y);
+\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
+\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
+\draw[->, gray!30] (p) -- (dp);
+\draw[->, gray!30] (l) -- (p);
+\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
+\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->, gray!30] (dp) -- (pstar);
+\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (pstar);
+\draw[->, gray!30] (pstar) -- (d);
+\draw[->] (p) to [out=+45,in=+135, looseness=0.3] (d);
+\draw[->] (d) -- (y);
+\draw[->, gray!30] (c) -- (dp);
+\draw[->, gray!30] (c) -- (pstar);
+\draw[->, gray!30] (cd) -- (d);
+\draw[->, gray!30] (cd) -- (y);
 \end{tikzpicture}
 ```
-]
 
-???
-In this setting death is treated as a censoring event and it can be interpreted as those who died would have the same risk of developing dementia if prevented from dying than those who remained alive and free of dementia at a given time point, conditional on shared causes of dementia and death
+$P_0$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis; $D_{t+1}$ = Death prior to dementia diagnosis
 
 ---
-### All pieces together
+### Death prior to dementia diagnosis (Problem 3: _Survival bias_)
 
-.center[
-```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = ""}
+```{tikz, echo = FALSE}
 \usetikzlibrary{arrows}
-\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
-\node (p) {$P$};
-\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$L$};
-\node [right of = p] (dp) {$D_t$};
-\node [right of = dp] (pstar) {$P*_t$};
+\begin{tikzpicture}[node distance=2cm]
+\node (p) {$P_0$};
+\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
+\node [right of = p, gray!30, gray!30] (dp) {$D_t$};
+\node [right of = dp, gray!30] (pstar) {$P^*_t$};
 \node [right of = pstar] (d) {$D_{t+1}$};
 \node [right of = d] (y) {$Y_{t+1}$};
-\node [above of = dp, yshift = -1cm, xshift = +1cm](c) {$C_1$};
-\node [above of = d, yshift = -1cm, xshift = +1cm](cd) {$C_2$};
-\draw[->] (p) -- (dp);
-\draw[->] (dp) -- (pstar);
-\draw[->, violet] (p) to [out=-45,in=-135, looseness=0.3] (pstar);
-\draw[->] (pstar) -- (d);
-\draw[->] (p) to [out=-45,in=-135, looseness=0.3] (d);
-\draw[->, violet] (p) to [out=-45,in=-135, looseness=0.3] (y);
+\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
+\node [below of = d, yshift = 1cm, xshift = 1cm](cd) {$C_2$};
+\draw[->, gray!30] (p) -- (dp);
+\draw[->, gray!30] (l) -- (p);
+\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
+\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->, gray!30] (dp) -- (pstar);
+\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (pstar);
+\draw[->, gray!30] (pstar) -- (d);
+\draw[->] (p) to [out=+45,in=+135, looseness=0.3] (d);
 \draw[->] (d) -- (y);
-\draw[->] (l) -- (p);
-\draw[->] (l) to [out=0,in=270, looseness=0.4] (y);
-\draw[->] (c) -- (dp);
-\draw[->] (c) -- (pstar);
+\draw[->, gray!30] (c) -- (dp);
+\draw[->, gray!30] (c) -- (pstar);
 \draw[->] (cd) -- (d);
@@
 
-<br><br>
-
-At best we can test the sharp null hypothesis
-]
+$P_0$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis; $D_{t+1}$ = Death prior to dementia diagnosis; $C_2$ as shared causes of $D_{t+1}$ and $Y_{t+1}$
 
@@
 
-- Participants between 60 and 70 years old at study entry.
+- Participants between 60 and 70 years old at study entry, free of cancer and dementia diagnosis at baseline
+
+--
 
@@
 
-Three definitions of the proxy for Pin1 over-expression:
+--
 
-- _"Cancer ever vs. never"_
-
-- Cancer as a _"time-varying"_
-
-- _"Time to cancer diagnosis"_
+- Cancer as _"time-fixed ever vs. never"_ vs. cancer as a _time-varying proxy_
 
 ---
+## Application to the Rotterdam Study
 
-### Methods
+- Inverse probability treatment weights to address confounding (IPTW)
 
-- Inverse probability treatment weights to address confounding
-
-  + *Age, sex, education, smoking status*
+  + *Age, sex, Apoe4, education*
 
@@
 
-- Inverse probability censoring weights for death to relax the independent censoring assumption
+- Inverse probability censoring weights for death (IPCW)
 
@@
 
-- We estimate the risk of dementia using a weighted Kaplan-Meier estimator.
+- Risk of dementia is estimated with a weighted Kaplan-Meier estimator.
 
 ---
+### Peterson upper and lower bounds
 
-### Methods
-
-Since the independent censoring assumption is untestable, we can consider extreme scenarios of independence and fit Peterson upper and lower bounds.
-
+- **1) Lower Bound**: Total independence, that refers to an scenario were those who died would never develop dementia (Death as competing event)
 --
 
-1) Total independence, that refers to an scenario were those who died would never develop dementia (lower bound).
-  + Aalen-Johansen estimator treating death as a competing event.
-
---
-
-2) Complete dependency, that refers to an scenario where those who died would have dementia prior to death (upper bound). 
-  + The Kaplan Meier estimator for the combined outcome of dementia or death.
+- **2) Upper Bound**: Complete dependency, that refers to an scenario where those who died would have dementia prior to death (Composite outcome, death + dementia)
 
@@
 
-.pull-left[
-```{r, echo = FALSE}
+.center[
+```{r, echo = FALSE, fig.retina = 3}
 plot_distribution +
@@
 ```
+
+In the cancer arm 6% had dementia and 63% died, in the free of cancer arm 14% had dementia and 23% died.
+
 ]
 
---
-
-.pull-right[
-- 24% had cancer, the median age of cancer diagnosis was 73 (IQR: 69-77).
-
-- Among participants with cancer, 6% had dementia and 63% (n = 549) died.
-
-- Among participants free of cancer diagnosis, 14% had dementia and 23% died.
-]
 ---
-### Cancer as: never vs. ever
+### Cancer as time-fixed ever vs. never proxy
 
@@
 
-### Time to cancer
+### Cancer diagnosis as time-varying proxy
 
@@
 ```{r}
-plot_3a
+plot_2b
 ```
@@
 ```{r}
-plot_3b
+plot_2c
 ```
@@
   filter(model != "Unadjusted") %>%
-  select(-c(3, 4)) %>%
-  rename(RD = rd, RR = rr, `HR (95%CI)` = hr)
+  select(-c(3, 4, 5)) %>%
+  rename(RR = rr, `HR (95%CI)` = hr) %>% 
+  filter(Proxy != "Time to cancer")
 

### 2021-04-27 02_R/04c_bootstraps_ever.R
Bootstraps for ever vs. never

@@ -1,0 +1,171 @@
+source(here::here("02_R", "04b_auxiliary_fx.R"))
+
+data_wide <-
+  import(here::here("01_data", "clean_data", "wide_after_truncation.RData"))
+
+data_wide %<>%
+  mutate(dementia_efu =
+           case_when(outcome_plr == 1 ~ 1,
+                     competing_plr == 1 ~ 2,
+                     TRUE ~ 0))
+
+data_wide %<>%
+  mutate(t2dem_efu =
+           ifelse(t2dem >= 240, 240, t2dem))
+
+data_wide %<>%
+  mutate(combined_outcome_efu = ifelse(dementia_efu != 0, 1, 0))
+
+# Weighted KM function for bootstrap ------------------------------------------------------
+
+km_ever <- function(data, crude = TRUE, ipcw = FALSE) {
+  if (crude != FALSE) {
+    
+    model <-
+      survfit(Surv(t2dem_efu, dementia_efu == 1) ~ cancer_v, data)
+    
+    output <- risks_km(model) %>% 
+      mutate(
+        model = "Unadjusted"
+      )
+  }
+  
+  else {  
+    cancer_den <-
+      glm(
+        cancer_v ~ bs(age_0) + sex + education + apoe4 + as.factor(smoke1),
+        data = data,
+        family = binomial
+      )
+    
+    cancer_num <- glm(cancer_v ~ 1, data = data)
+    
+    data <- data %>%
+      mutate(
+        p_num = predict(cancer_num, type = "response"),
+        p_denom = predict(cancer_den, type = "response"),
+        w_cancer = ifelse(cancer_v == 1, p_num/p_denom, (1 - p_num)/(1- p_denom)))
+    
+    data %<>%
+      mutate(w_cancer_t = ifelse(
+        w_cancer > quantile(w_cancer, 0.99),
+        quantile(w_cancer, 0.99),
+        w_cancer
+      ))
+    
+    if (ipcw != FALSE) {
+      death_den <-
+        glm(
+          competing_plr ~ bs(age_0, 3) + sex + education + apoe4 + 
+            as.factor(smoke1) + ht1 + bs(sbp1, 3) + bs(bmi1,3) +
+            as.factor(diabetes_prev) +  cancer_v + cohort,
+          data = data,
+          family = binomial
+        )
+      
+      death_num <- glm(competing_plr ~ 1, data = data)
+      
+      data <- data %>%
+        mutate(
[…]

### 2021-04-27 02_R/04c_bootstraps_tv.R
time-var bootstraps

@@ -1,0 +1,199 @@
+library(tidyverse)
+library(rio)
+library(survival)
+library(survminer)
+library(broom)
+library(splines)
+library(magrittr)
+library(WeightIt)
+library(cobalt)
+
+# Import data -------------------------------------------------------------
+source(here::here("02_R", "04b_auxiliary_fx.R"))
+
+data_long <-
+  import(here::here("01_data", "clean_data", "dementia_long.RData"))
+
+data_wide <-
+  import(here::here("01_data", "clean_data", "wide_after_truncation.RData"))
+
+data_long %<>% 
+  group_by(id) %>% 
+  mutate(
+    fuptime = time - 1,
+    tstart = lag(fuptime),
+    tstart = ifelse(is.na(tstart), -1, tstart)) %>% 
+  ungroup()
+
+## If competing_plr = 1, then outcome_plr = NA
+
+data_long %<>% 
+  mutate(outcome_plr = ifelse(competing_plr == 1, NA, outcome_plr))
+
+## Merge with baseline
+smoke1 <- data_wide %>% select(id, smoke1, ht1, sbp1, bmi1)
+
+data_long <- data_long %>% 
+  left_join(smoke1)
+
+# Weighted KM function for bootstrap  -------------------------------------------------------
+
+
+km_tv <- function(data_long, crude = TRUE, ipcw = FALSE) {
+  if (crude != FALSE) {
+    
+    model <- survfit(Surv(
+        tstart, time2 = fuptime, event = outcome_plr) ~ cancer_v,
+        data = data_long)
+    
+    
+    output <- risks_km(model) %>% 
+      mutate(
+        model = "Unadjusted")
+  }
+  
+  else {  
+    cancer_den <-
+      glm(
+        cancer_v ~ bs(age_0, 3) + sex + education + as.factor(smoke1) + cohort,
+        data = data_long,
+        family = binomial
+      )
+    
+    cancer_num <- glm(cancer_v ~ 1 , data = data_long)
+    
+    data_long <- data_long %>%
+      mutate(
+        p_num = predict(cancer_num, type = "response"),
+        p_denom = predict(cancer_den, type = "response"))
+    
+    data_long <- data_long %>%
+      group_by(id) %>% 
+      mutate(
+        sw_cancer = ifelse(cancer_v == 1, p_num/p_denom, (1 - p_num)/(1- p_denom)),
+        sw_cancer = cumprod(sw_cancer),
[…]

### 2021-04-27 02_R/04b_auxiliary_fx.R
fix hr for graphs

@@ -108,8 +108,7 @@ tidy_hr <- function(model) {
   broom::tidy(exponentiate = TRUE, conf.int = TRUE) %>%
     select(term, estimate, contains("conf")) %>% 
     mutate_at(c(2:4), round, 2) %>% 
-    select(-1, hr = estimate) %>% 
-    mutate(hr = paste0(hr, " (", conf.low, ", ", conf.high, ")"))
+    select(-1, point = estimate)
 }
 
 

### 2021-04-27 06_presentations/ser2021/index.Rmd
updated results, slides, todo much wow

@@
 
-source(here::here("02_R", "04_direct_effect_simplified.R"))
+load(here::here("03_figs", "plots.RData"))
 source(here::here("02_R", "05_plots.R"))
-source(here::here("02_R", "05b_plot_causes_death.R"))
-table_results <- rio::import(here::here("02_R", "table_results.csv"))
-table_results_bounds <- rio::import(here::here("02_R", "table_results_bounds.csv"))
+hr_results <- rio::import(here::here("02_R", "hr_results.csv"))
+rd_results_ever <- rio::import(here::here("02_R", "rd_results_ever.csv"))
+rd_results_tv <- rio::import(here::here("02_R", "rd_results_tv.csv"))
+table_results_bounds <- rio::import(here::here("02_R", "table_results_bounds.csv")) %>% 
+  select(1,2,rr) %>% 
+  filter(Proxy != "Time to cancer",
+         model != "IPTW + IPCW") %>% 
+  pivot_wider(names_from = model, values_from = rr)
 
+rr <- rd_results_ever %>%
+  bind_rows(rd_results_tv) %>% 
+  select(Proxy, model, rr, conf) %>% 
+  pivot_wider(names_from = conf,
+              values_from = rr) %>% 
+  mutate(contrast = "Risk Ratio") %>% 
+  rename(conf.high = conf.higher) %>% 
+  mutate(Proxy = str_replace(Proxy, "Time-varing", "Time-varying"))
+
+results <- hr_results %>% 
+  mutate(contrast = "Hazard Ratio") %>%
+  filter(Proxy != "Time to cancer") %>%
+  mutate(Proxy = ifelse(
+    Proxy == "Ever vs. Never", "Ever vs. never", "Time-varying")) %>% 
+  bind_rows(rr) %>% 
+  mutate(Proxy = str_to_sentence(Proxy))
 ```
@@
 ---
-## Introduction
+### Introduction
 
@@
 
-- Observational studies have found an inverse association between cancer and the risk of Alzheimer´s disease or related dementias (ADRD).
+- Observational studies have found an inverse association between cancer and dementia risk.
 
@@
 
-- To understand the potential sources of bias we first need to define the causal question.
+- Several genetic and molecular mechanisms have been proposed.
 
@@
 
-- Refining the underlying question behind cancer diagnosis as the exposure unveils some of these concerns.
+- Researchers have raised concerns related to measurement error, unmeasured confounding and selection bias.
 
@@
 
-### Pin1-targeting drug
+### Introduction
 
@@
 
-$P_0$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis
-
-???
-In hypothetical setting, if Pin1-targeting drug was developed, the best way to understand its effect on dementia risk would be to do a randomized trial in which we randomize eligible participants in late midlife to receive this drug or not, and closely monitor ADRD over follow-up.
+$P_0$ = Pin1 targeted-drug; $Y_{t+1}$ = Dementia diagnosis
 
@@
 ---
-### Cancer diagnosis as the proxy of Pin1 biomarker (Problem 2: _Time zero_)
+### Cancer diagnosis as the proxy of Pin1 (Problem 2: _Time zero_)
 
@@
 
-- Data on cancer diagnosis, dementia diagnosis and death was collected from national registries and electronical clinical records.
+- Data on cancer diagnosis, dementia diagnosis and death was collected from national registries and clinical records.
 
@@
 
-- Cancer as _"time-fixed ever vs. never"_ vs. cancer as a _time-varying proxy_
+- Cancer:
+
+  + _Time-fixed ever vs. never_
+
+  + _Time-varying_
 
@@
 
-- Risk of dementia is estimated with a weighted Kaplan-Meier estimator.
+- Risk of dementia and risk ratios are estimated with a weighted Kaplan-Meier estimator.
+
+--
+
+- Hazard ratios are estimated with a weighted Cox Proportional-Hazards Model for comparative purposes.
 
 ---
-### Peterson upper and lower bounds
+### Bounds
+
+<br><br>
 
 - **1) Lower Bound**: Total independence, that refers to an scenario were those who died would never develop dementia (Death as competing event)
+
 --
@@
 .center[
-```{r, echo = FALSE, fig.retina = 3}
+```{r, echo = FALSE, fig.width= 14, fig.retina=3}
 plot_distribution +
-    theme_xaringan()
+  theme(legend.position = "right",
+        title = element_text(face = "bold", size = rel(1.5), color = "#011A5E"),
+        axis.title.x = element_text(face = "bold", size = rel(1.5), color = "#011A5E")) +
+  labs(title = "Proportion of participants in each status")
 ```
-
-In the cancer arm 6% had dementia and 63% died, in the free of cancer arm 14% had dementia and 23% died.
-
 ]
@@
 ---
-### Cancer as time-fixed ever vs. never proxy
+
+### Results
 
 .pull-left[
-```{r}
-plot_1b
+```{r, echo = FALSE, fig.retina = 3}
+temp <-
+  table_results_bounds %>% slice(1) %>% mutate(model = "IPTW + IPCW", contrast = "Risk Ratio") %>% 
+  rename(conf.low = `Lower bound`, conf.high = `Upper bound`)
+
+results %>%
+  filter(Proxy == "Ever vs. never") %>%
+  mutate(model = fct_relevel(model, c("Unadjusted", "IPTW", "IPTW + IPCW"))) %>%
+  ggplot(aes(
+    y = contrast,
+    x = point,
+    xmin = conf.low,
+    xmax = conf.high
+  )) +
+  geom_vline(xintercept = 1,
+             color = "#E69F00",
+             size = 1.2) +
+  geom_pointrange(aes(shape = contrast), color = "#0072B2") +
+  geom_errorbar(aes(group = contrast),
+                width = 0.2,
+                color = "#0072B2") +
+  geom_point(
+    data = temp,
+    aes(x = conf.low),
+    shape = 18,
+    color = "#011A5E",
+    size = 3
+  ) +
+  geom_point(
+    data = temp,
+    aes(x = conf.high),
+    shape = 18,
+    color = "#011A5E",
+    size = 3
+  ) +
+  facet_grid(factor(model, levels = c("Unadjusted", "IPTW", "IPTW + IPCW")) ~
+                 ., switch = "y") +
+  # facet_grid(model ~., switch = "y", margins = FALSE) +
+  scale_x_continuous(trans = "log10") +
+  labs(
+    x = NULL,
+    y = NULL,
+    group = NULL,
+    shape = NULL,
+    title = "Cancer as ever vs. never",
+    subtitle = "Risk ratios \U25B2, Hazard ratios \U25CF, Bounds \U25C6" ) +
+  theme(
+    panel.spacing = unit(0, "lines"),
+    panel.background = element_blank(),
+    axis.text.y = element_blank(),
+    axis.ticks.y = element_blank(),
+    strip.text.y = element_text(
+      face = "bold",
+      size = rel(1.5),
+      color = "#011A5E"
+    ),
+    axis.line.x = element_line(color = "#011A5E"),
+    axis.text.x = element_text(size = rel(1.5), color = "#011A5E"),
+    strip.background = element_blank(),
+    plot.title = element_text(face = "bold", size = rel(2), color = "#011A5E"),
+    plot.subtitle = element_text(face = "bold", size = rel(1.5), color = "#011A5E", hjust = 0.05),
+    legend.position = "none")
+
 ```
@@
 .pull-right[
-```{r}
-plot_1c
-```
+```{r, echo = FALSE, fig.retina = 3}
+temp <-
+  table_results_bounds %>% slice(2) %>% mutate(model = "IPTW + IPCW", contrast = "Risk Ratio") %>% 
+  rename(conf.low = `Lower bound`, conf.high = `Upper bound`)
 
-]
-
----
-
-### Cancer diagnosis as time-varying proxy
-
-.pull-left[
-```{r}
-plot_2b
-```
-
-]
-
---
-
-.pull-right[
-```{r}
-plot_2c
-```
-
-]
----
-### Results
-
-<br>
-
-.center[
-```{r, echo = FALSE}
-tb <- table_results %>%
-  filter(model != "Unadjusted") %>%
-  select(-c(3, 4, 5)) %>%
-  rename(RR = rr, `HR (95%CI)` = hr) %>% 
-  filter(Proxy != "Time to cancer")
-
-tb %>% kbl() %>%
-kable_styling(bootstrap_options = c("striped", "hover"))
-```
-]
----
-
-### Extreme scenarios of independency between death and dementia
-
-.pull-left[
-```{r, echo = FALSE}
-table_results_bounds %>%
-  select(-3, -4, - 6) %>%
-    pivot_wider(names_from = c(model),
-              values_from = rd) %>%
+results %>%
+  filter(Proxy == "Time-varying") %>%
+  mutate(model = fct_relevel(model, c("Unadjusted", "IPTW", "IPTW + IPCW"))) %>%
   ggplot(aes(
-    y = Proxy,
-    x = `IPTW + IPCW`,
-    xmin = `Lower bound`,
-    xmax = `Upper bound`
-  )) +
-  geom_vline(xintercept = 0,
-             color = "#E69F00",
-             size = 1.2) +
-  geom_pointrange(shape = 15,
-                  color = "#0072B2") +
-  geom_errorbar(width = 0.2, color = "#0072B2") +
-  labs(x = "Risk difference",
-       y = NULL) +
-  theme_xaringan()
-```
-]
-
---
-
-.pull-right[
-```{r, echo = FALSE}
-table_results_bounds %>%
-  select(-3, -4, - 5) %>%
-    pivot_wider(names_from = c(model),
-              values_from = rr) %>%
-  ggplot(aes(
-    y = Proxy,
-    x = `IPTW + IPCW`,
-    xmin = `Lower bound`,
-    xmax = `Upper bound`
+    y = contrast,
+    x = point,
+    xmin = conf.low,
+    xmax = conf.high
   )) +
@@
              size = 1.2) +
-  geom_pointrange(shape = 15,
+  geom_pointrange(aes(shape = contrast), color = "#0072B2") +
+  geom_errorbar(aes(group = contrast),
+                width = 0.2,
                 color = "#0072B2") +
-  geom_errorbar(width = 0.2, color = "#0072B2") +
-  labs(x = "Risk Ratio",
-       y = NULL) +
-   scale_x_continuous(
-    trans = "log10",
-    # breaks = seq(0.7, 1.2, by = 0.1)
+  geom_point(
+    data = temp,
+    aes(x = conf.low),
+    shape = 18,
+    color = "#011A5E",
+    size = 3
   ) +
-  theme_xaringan()
+  geom_point(
+    data = temp,
+    aes(x = conf.high),
+    shape = 18,
+    color = "#011A5E",
+    size = 3
+  ) +
+  facet_grid(factor(model, levels = c("Unadjusted", "IPTW", "IPTW + IPCW")) ~
+                 ., switch = "y") +
+  scale_x_continuous(trans = "log10") +
+  labs(
+    x = NULL,
+    y = NULL,
+    group = NULL,
+    shape = NULL,
+    title = "Time-varying cancer",
+    subtitle = "Risk ratios \U25B2, Hazard ratios \U25CF, Bounds \U25C6" ) +
+  theme(
+    panel.spacing = unit(0, "lines"),
+    panel.background = element_blank(),
+    axis.text.y = element_blank(),
+    axis.ticks.y = element_blank(),
+    strip.text.y = element_text(
+      face = "bold",
+      size = rel(1.5),
+      color = "#011A5E"
+    ),
+    axis.text.x = element_text(size = rel(1.5), color = "#011A5E"),
+    axis.line.x = element_line(color = "#011A5E"),
+    strip.background = element_blank(),
+    plot.title = element_text(face = "bold", size = rel(2), color = "#011A5E"),
+    plot.subtitle = element_text(face = "bold", size = rel(1.5), color = "#011A5E", hjust = 0.05),
+    legend.position = "none")
 ```
+
 ]

### 2021-04-27 02_R/04_direct_effect_simplified.R

@@ -234,10 +234,8 @@ hr_1c <- cancer_ever_cox_ipcw %>%
 rd_1c <- risks_km(cancer_ever_km_ipcw) %>%
   mutate(model = "IPTW + IPCW")
 
-results_ever_never <- bind_rows(hr_1a, hr_1b, hr_1c) %>% 
-  left_join(bind_rows(rd_1a, rd_1b, rd_1c)) %>% 
-  mutate(Proxy = "Ever vs. Never") %>% 
-  select(Proxy, model, starts_with("cancer"), rd, rr, hr) 
+hr_ever_never <- bind_rows(hr_1a, hr_1b, hr_1c) %>% 
+  mutate(Proxy = "Ever vs. Never") 
 
 
 # 1.d. Lower bound --------------------------------------------------------
@@ -444,10 +442,8 @@ hr_2c <- cox_tv_ipcw %>%
 rd_2c <- risks_km(km_tv_ipcw) %>% 
     mutate(model = "IPTW + IPCW")
 
-results_tv_cancer <- bind_rows(hr_2a, hr_2b, hr_2c) %>% 
-  left_join(bind_rows(rd_2a, rd_2b, rd_2c)) %>% 
-  mutate(Proxy = "Time-varying cancer") %>% 
-  select(Proxy, model, starts_with("cancer"), rd, rr, hr)
+hr_timevar <- bind_rows(hr_2a, hr_2b, hr_2c) %>% 
+  mutate(Proxy = "Time-varying cancer")
 
 #  2.d. Lower bound ------------------------------------------------------------
 
@@ -603,10 +599,9 @@ hr_3b <- cox_t2cancer_ipcw %>%
 rd_3b <- risks_km(km_t2cancer_ipcw) %>% 
   mutate(model = "IPTW + IPCW")
 
-results_t2cancer <- bind_rows(hr_3a, hr_3b) %>% 
-  left_join(bind_rows(rd_3a, rd_3b)) %>% 
-  mutate(Proxy = "Time to cancer") %>% 
-  select(Proxy, model, starts_with("cancer"), rd, rr, hr)
+hr_t2cancer <- bind_rows(hr_3a, hr_3b) %>% 
+  mutate(Proxy = "Time to cancer") 
+
 
 # 3.c. Lower bound ------------------------------------------------------------------
 
@@ -647,9 +642,9 @@ rd_3d  <- km_t2cancer_ipcw_up_bound %>%
 
 # 4. All results ----------------------------------------------------------
 
[…]

### 2021-04-27 02_R/04c_bootstraps_tv.R

@@ -177,11 +177,13 @@ risks_boot_long <- function(data_long, n, seed, crude = TRUE, ipcw = FALSE){
 # Results -----------------------------------------------------------------
 
 point_km_crude <- km_tv(data_long, crude = TRUE)
+point_km_crude$conf <- "point"
 
 point_km_iptw <- km_tv(data_long, crude = FALSE, ipcw = FALSE)
+point_km_iptw$conf <- "point"
 
 point_km_ipcw <- km_tv(data_long, crude = FALSE, ipcw = TRUE)
-
+point_km_ipcw$conf <- "point"
 
 boots_km_crude <- risks_boot_long(data_long, seed = 123, n = 100, crude = TRUE, ipcw = FALSE)
 
@@ -197,3 +199,9 @@ boots_km_ipcw <- risks_boot_long(data_long, seed = 123, n = 100, crude = FALSE, 
 boots_km_ipcw$conf <- c("conf.low", "conf.higher")
 boots_km_ipcw$model <- "IPTW + IPCW"
 
+
+rd_results_tv <- bind_rows(point_km_crude, point_km_iptw, point_km_ipcw) %>%
+  bind_rows(boots_km_crude, boots_km_iptw, boots_km_ipcw) %>%
+  mutate(Proxy = "Time-varing")
+
+export(rd_results_tv, here::here("02_R", "rd_results_tv.csv"))

### 2021-04-28 06_presentations/ser2021/index.Rmd
presentation almost done

@@
 
+<br>
+
 ## Towards a clearer understanding of the inverse association between cancer and dementia
@@
 
-.right[L. Paloma Rojas-Saunero, MD
+.right[
 
-**Epidemiology department**]
+L. Paloma Rojas-Saunero, MD
+
+**Epidemiology department**
+
+`r icons::fontawesome$brands$twitter` @palolili23
+
+]
 
@@
 ---
-## Application to the Rotterdam Study
+### Application to the Rotterdam Study
 
@@
 
-- The process of connecting our real question to the data help us identify the structure that will later represent the analytic decisions and assumptions.
+- Defining the underlying mechanism behind cancer diagnosis helps in multiple way.
+
+  + Helps identify sources of bias and guides analysis
+
+  + Opens conversations about future data collection
 
@@
 
-- Cancer as the proxy for a molecular mechanism is tricky...
---
+- Since death acts as a competing event, we first need to define the estimand of interest.
 
-- When death is a competing event, prior to choosing the estimator we need to define what is the estimand of interest and be explicit about the assumptions and interpretations.
+  + The direct effect relies on the independent censoring assumption
+
+  + Interpretation implies preventing death.
+
+  + The separable effects provides a clearer intuition to address this question.
 
 ---
+class: middle
+
+### Acknowledgments
+
+- Sonja A. Swanson
+
+- Kimberly van der Willik
+
+- Arfan M. Ikram
+
+---
+class:middle
+
+### Gracias!   Thank you!
+
+`r icons::fontawesome("paper-plane", style = "solid")` </i>&nbsp; l.rojassaunero@erasmusmc.nl </a> <br>
+
+`r icons::fontawesome$brands$twitter` @palolili23
+
+`r icons::fontawesome$brands$github` @palolili23

### 2021-04-29 06_presentations/ser2021/index.Rmd
final slides

@@
 
+???
+Hello, my name is Paloma Rojas-Saunero, I am a PhD candidate at Erasmus MC and today I will present my work called "towards a clearer understanding of the inverse association between cancer and dementia".
+
 ---
@@
 
-- Observational studies have found an inverse association between cancer and dementia risk.
+- Observational studies have found an inverse association between cancer and dementia.
 
@@
 - Researchers have raised concerns related to measurement error, unmeasured confounding and selection bias.
+
+???
+
+...with the intention of identifying drug targets to prevent dementia.
+However, researchers have raised concerns related to measurement error, unmeasured confounding and selection bias.
+
+But to understand the potential sources and magnitude of bias, we first need to define the causal question which implies defining what is the mechanism of interest behind cancer diagnosis as an exposure.
 
@@
 
+???
+
+For example, in this graph we defined a drug target for Pin1. This is an enzyme that is over-expressed in tumor development, but it's depletion is associated to neurodegeneration.
+
+We can imagine that someday in the future, there will be a drug that targets Pin1 to delay or prevent dementia, and the effectiveness of this drug would be assessed in an RCT.
+
+Since this drug is not available yet, we can at least imagine that Pin1 expression could become a biomarker in a closer future.
+Once this is available we could measure Pin1 expression in stored blood samples in population based cohort studies.
+
 ---
@@
 ???
-We might be closer to developing biomarker for Pin1 expression in the future.
-So we can imagine a situation in which Pin1 exposure is measured in a population based cohort study, using stored blood samples.
-Since this would be a observational setting, Pin1 expression and ADRD at time may share causes such as smoking, and that assessing the causal relationship requires adjusting for these confounders.
+So in this graph we changed P0 to pin1 overexpression.
+Since this study would be performed within an observational study, Pin1 expression and dementia may share risk factors such as behavioural or environmental, represented as L.
+Thus to identify the causal relationship we will have to adjust for all L.
+
+Now, since Pin1 is not available we defined a proxy for it.
 
@@
 ???
-At this moment this biomarker is not available, but we know that Pin1 over-expression stimulates tumor growth and for this reason we use cancer diagnosis as the proxy for Pin1.
+Given that Pin1-expression is present in tumors, and tumors are only measured through diagnosis, cancer diagnosis may act as the proxy for Pin1 over-expression, which is represented as Pstar.
+This would mean that if we study the association between pstar and y in the observed data, we assume that the captured effect is only through P.
+The challenge here is that there might be a gap in time between our intended exposure and cancer diagnosis.
+So we need to be careful to adjust for post-baseline covariates of P or mediators between P and Pstar.
 
@@
 $P_0$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis; $P^*_t$ = Cancer diagnosis; $D_t$ = Death prior to cancer diagnosis
+
 ???
+One of the mediators between P and Pstar is death prior to cancer diagnosis.
+Since individuals are at risk of dying from other causes (such as cardiovascular death), we can only measure Pstar in the subset of individuals who have survived long enough to have a cancer diagnosis.
 
@@
 $P_0$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis; $P^*_t$ = Cancer diagnosis; $D_t$ = Death prior to cancer diagnosis; $C_1$ = Shared causes of $D_t$ and $P^*$
+
 ???
+
+And since several risk factors that increase the risk of cancer might also cause death prior to cancer diagnosis, we need to block this backdoor pathways by adjusting for C.
 
@@
 
+???
+Even if we could have measured Pin1 or a drug in everybody at the same time, death is going to play the rol of a competing event, death prior to dementia diagnosis prevents us from observing dementia.
+This is key because it puts us in a situation where we have to choose an estimand.
+In this case a total effect that includes all causal pathways, including the ones mediated through dead would not be of interest, since one of the leading causes of death is cancer we would see an inverse association.
+
 ---
@@
 
+???
+
+Besides, dementia and death share risk factors (represented as C2).
+So if we are interested in the direct effect of Pin1 in dementia, which means that we would treat death as a censoring event, we rely in the independent censoring assumption that indicates that dementia and death are independent if we condition on C.
+
 ---
@@
 - Participants between 60 and 70 years old at study entry, free of cancer and dementia diagnosis at baseline
+
+???
+To show how this causal structure and assumptions connect to the analytic decisions, we use data from the rotterdam study (a population based cohort with over 25 years of follow up).
 
@@
   + _Time-varying_
+
+???
+We defined Cancer as a proxy in two ways.
 
@@
 --
+  + *Lower Bound*: Total independence, that refers to an scenario were those who died would never develop dementia.
+
+  + *Upper Bound*: Complete dependency, that refers to an scenario where those who died would have dementia prior to death.
+
+---
+class: middle
 
@@
 - Hazard ratios are estimated with a weighted Cox Proportional-Hazards Model for comparative purposes.
-
----
-### Bounds
-
-<br><br>
-
-- **1) Lower Bound**: Total independence, that refers to an scenario were those who died would never develop dementia (Death as competing event)
-
---
-
-- **2) Upper Bound**: Complete dependency, that refers to an scenario where those who died would have dementia prior to death (Composite outcome, death + dementia)
 
@@
 ]
+
+???
+If we look at the proportion of participants who was in each status over follow up, we observe that dementia diagnosis becomes more frequent after age of 80, while cancer diagnosis distribution is more homogenous across life span.
+In yellow we observe the amount of participants who died over followup, in larger proportion compared to dementia.
+Last, in pink we have participants who had cancer and dementia, which represents a 6% from all participants.
 
@@
 
+???
+First I will show all results related to cancer as ever vs. never.
+In this graph I present three situations, unadjusted, after weighting for confounding, and after weighting for confounding and censoring for death.
+In each case we have risk ratios and hazard ratios.
+We observe the inverse association in these two settings, and after including weights for censoring it shifts towards above the null. 
+Lets recall that the model with censoring weights is our best effort to achieve the conditional independence, but if we observe the bounds that represent extreme scenarios, these bounds cross the null.
+
 --
@@
 
+???
+When we define cancer as a time-varying exposure, we observe a shift towards the null prior to adjustment. An estimates and confidence shift above one by including the weights.
+And again, in our best scenario to achieve conditional independence, if we look at the bounds of extreme independency and dependency, they are far apart in both sides of the null.
+
+Please note that in all cases, we would interpret this results as if death could have been prevented.
+
 ---

### 2021-04-29 06_presentations/ser2021/index.Rmd
retoques finales

@@
 
-- Participants between 60 and 70 years old at study entry, free of cancer and dementia diagnosis at baseline
+- Participants between 60 and 70 years old at study entry, free of cancer and dementia diagnosis at baseline.
 
@@
 
-- Defining the underlying mechanism behind cancer diagnosis helps in multiple way.
+- Defining the underlying mechanism behind cancer diagnosis helps in multiple ways.
 
-  + Helps identify sources of bias and guides analysis
+  + Helps identify sources of bias and guides analysis.
 
-  + Opens conversations about future data collection
+  + Opens conversations about future data collection.
 
@@
 
-  + The direct effect relies on the independent censoring assumption
+  + The direct effect relies on the independent censoring assumption.
 
@@
 
-- Arfan M. Ikram
+- M. Arfan Ikram
 

### 2021-04-29 02_R/04_direct_effect.R
changes

@@ -1,442 +1,0 @@
-library(tidyverse)
-library(rio)
-library(survival)
-library(survminer)
-library(broom)
-library(splines)
-library(magrittr)
-library(WeightIt)
-library(cobalt)
-
-# Import data -------------------------------------------------------------
-source(here::here("02_R", "04b_auxiliary_fx.R"))
-
-data_long <-
-  import(here::here("01_data", "clean_data", "dementia_long.RData"))
-
-data_wide <-
-  import(here::here("01_data", "clean_data", "wide_after_truncation.RData"))
-
-data_long %<>% 
-  group_by(id) %>% 
-  mutate(
-    fuptime = time - 1,
-    tstart = lag(fuptime),
-    tstart = ifelse(is.na(tstart), -1, tstart)) %>% 
-  ungroup()
-
-## idea of how numbers look like over follow-up
-
-data_long %<>% 
-  mutate(both_outcomes = 
-           case_when(outcome_plr == 1 ~ 1,
-                     competing_plr == 1 ~2,
-                     TRUE ~ 0))
-
-data_wide %<>%
-  mutate(dementia_efu =
-           case_when(outcome_plr == 1 ~ 1,
-                     competing_plr == 1 ~ 2,
-                     TRUE ~ 0))
-
-data_wide %<>%
-  mutate(t2dem_efu =
-           ifelse(t2dem >= 240, 240, t2dem))
-
-
-# 1. Cancer Ever vs. Never ---------------------------------------------------
-
-
-# 1.1. IPTW (cancer ever vs. never) ------------------------------------------
-
-# cancer_den <-
-#   glm(
-#     cancer_v ~ bs(age_0) + sex + education + apoe4 + as.factor(smoke1),
-#     data = data_wide,
-#     family = binomial
-#   )
-# 
-# summary(cancer_den)
-# 
-# cancer_num <- glm(cancer_v ~ 1, data = data_wide)
-# 
-# summary(cancer_num)
-# 
-# data_wide <- data_wide %>%
-#   mutate(
-#     p_num = predict(cancer_num, type = "response"),
-#     p_denom = predict(cancer_den, type = "response"),
-#     w_cancer = ifelse(cancer_v == 1, p_num/p_denom, (1 - p_num)/(1- p_denom)))
-
-
-## Check standardized mean
-
-w.out1 <-
-  weightit(
[…]

### 2021-04-29 02_R/04_direct_effect_simplified.R

@@ -116,45 +116,43 @@ summary(data_wide$cancer_weight_t)
 
 # 1.2. IPCW for death, based on baseline covariates ----------------------------
 
-w.out_death <-
-  weightit(
+## Death weights were once death = 1
+
+death_den <-
+  glm(
     competing_plr ~ bs(age_0, 3) + sex + education + apoe4 + 
       as.factor(smoke1) + ht1 + bs(sbp1, 3) + bs(bmi1,3) +
       as.factor(diabetes_prev) +  cancer_v + cohort,
     data = data_wide,
-    stabilize = TRUE,
-    estimand = "ATE",
-    method = "ps"
+    family = binomial
   )
 
-w.out_death
+death_num <-
+  glm(
+    competing_plr ~ 1,
+    data = data_wide,
+    family = binomial
+  )
 
-data_wide <- data_wide %>% 
-  mutate(death_weight = w.out_death$weights)
-
-love.plot(w.out_death) +
-  theme_minimal() +
-  theme(legend.position = "bottom") +
-  scale_color_manual(values = c("#011A5E", "#e4a803"))
+data_wide %<>% mutate(
+  p_death_den = predict(object = death_den, type = "response"),
+  p_death_num = predict(object = death_num, type = "response"),
+  sw_death = if_else(competing_plr == 0, (1-p_death_num)/(1-p_death_den), 1))
 
 data_wide %<>%
-  mutate(death_weight_t = ifelse(
-    death_weight > quantile(death_weight, 0.99),
-    quantile(death_weight, 0.99),
-    death_weight
+  mutate(sw_death_t = ifelse(
+    sw_death > quantile(sw_death, 0.99),
+    quantile(sw_death, 0.99),
+    sw_death
   ))
 
-
-summary(data_wide$death_weight)
-summary(data_wide$death_weight_t)
 
 ## Multiply weights
 
 data_wide %<>%
   mutate(
-    baseline_w = cancer_weight * death_weight,
-    baseline_wt = cancer_weight_t * death_weight_t)
-
+    baseline_w = cancer_weight * sw_death,
+    baseline_wt = cancer_weight_t * sw_death_t)
 
 # 1.a. KM Unadjusted ------------------------------------------------------
 
@@ -247,7 +245,7 @@ km_ever_ipcw_low_bound <- survfit(
[…]

### 2021-04-29 02_R/04c_bootstraps_ever.R

@@ -69,7 +69,7 @@ km_ever <- function(data, crude = TRUE, ipcw = FALSE) {
         mutate(
           dp_num = predict(death_num, type = "response"),
           dp_denom = predict(death_den, type = "response"),
-          w_death = ifelse(competing_plr == 1, dp_num/dp_denom, (1 - dp_num)/(1- dp_denom)))
+          w_death = if_else(competing_plr == 0, (1-dp_num)/(1-dp_denom), 1))
       
       data %<>%
         mutate(w_death_t = ifelse(

### 2021-04-29 02_R/04c_bootstraps_tv.R

@@ -71,9 +71,10 @@ km_tv <- function(data_long, crude = TRUE, ipcw = FALSE) {
       group_by(id) %>% 
       mutate(
         sw_cancer = ifelse(cancer_v == 1, p_num/p_denom, (1 - p_num)/(1- p_denom)),
-        sw_cancer = cumprod(sw_cancer),
+        # sw_cancer = cumprod(sw_cancer),
         w_cancer = ifelse(cancer_v == 1, 1/p_denom, 1/(1- p_denom)),
-        w_cancer = cumprod(w_cancer)) %>% 
+        # w_cancer = cumprod(w_cancer)
+        ) %>% 
       ungroup()
     
     data_long %<>% 

### 2021-04-29 02_R/04x_extra_models.R

@@ -1,0 +1,54 @@
+w.out1 <-
+  weightit(
+    cancer_v ~ bs(age_0, 3) + sex + education + apoe4 + as.factor(smoke1) +
+      bs(sbp1, 3) + bs(bmi1, 3) + ht1 + as.factor(diabetes_prev) + cohort,
+    data = data_wide,
+    stabilize = TRUE,
+    estimand = "ATE",
+    method = "ps"
+  )
+w.out1
+
+# 1.2. IPCW for death, based on baseline covariates ----------------------------
+
+w.out_death <-
+  weightit(
+    competing_plr ~ bs(age_0, 3) + sex + education + apoe4 + 
+      as.factor(smoke1) + ht1 + bs(sbp1, 3) + bs(bmi1,3) +
+      as.factor(diabetes_prev) +  cancer_v + cohort,
+    data = data_wide,
+    stabilize = TRUE,
+    estimand = "ATE",
+    method = "ps"
+  )
+
+
+# 2. Time-varying cancer -----------------------------------------------------
+
+cancer_den <-
+  glm(
+    cancer_v ~ bs(age_0, 3) + sex + education + apoe4 + as.factor(smoke1) + ht1 + 
+      bs(sbp1, 3) + bs(bmi1, 3) + as.factor(diabetes_prev) + cohort,
+    data = data_long,
+    family = binomial
+  )
+
+
+# 2.2. IPCW. Weights on death - dementia ---------------------------------------------
+
+death_den <- glm(
+  competing_plr ~ cancer_v + bs(time, 3) + bs(age_0, 3) + sex + education + apoe4 +
+    as.factor(smoke1) + ht1 + bs(sbp1, 3) + bs(bmi1, 3) + as.factor(diabetes_prev) +
+    as.factor(smoke) + bs(sbp, 3) + bs(bmi, 3) + ht + ht_drug + hd_v + stroke_v + diab_v + cohort,
+  data = data_long,
+  family = quasibinomial
+)
+
+death_num <-
+  glm(
+    competing_plr ~ cancer_v + bs(time, 3) + bs(age_0, 3) + sex + education + apoe4 +
+      as.factor(smoke1) + ht1 + bs(sbp1, 3) + bs(bmi1, 3) + as.factor(diabetes_prev) + cohort,
+    data = data_long,
+    family = binomial
+  )
+

### 2021-04-30 04_reports/dags.Rmd
updates on manuscript

--- 2021-04-28 06_presentations/ser2021/index.Rmd
@@
 ---
-title: "Toward a clearer understanding of the inverse association between cancer and dementia"
-subtitle: ""  
-author: 
-  - "L. Paloma Rojas-Saunero"
-date: '`r Sys.Date()`'
+title: "Dags"
 output:
-  xaringan::moon_reader:
-    lib_dir: libs
-    css: xaringan-themer.css
-    seal: false
-    nature:
-      highlightStyle: github
-      highlightLines: true
-      highlightLanguage: ["r", "css"]
-      countIncrementalSlides: FALSE
-      ratio: "16:9"
-editor_options: 
-  chunk_output_type: console
+  xaringan::moon_reader
 ---
 
-```{r setup, include=FALSE}
-options(htmltools.dir.version = FALSE)
-knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE, fig.showtext = TRUE, cache = TRUE)
-
-library(splines)
-library(knitr)
-library(kableExtra)
-
-load(here::here("03_figs", "plots.RData"))
-source(here::here("02_R", "05_plots.R"))
-hr_results <- rio::import(here::here("02_R", "hr_results.csv"))
-rd_results_ever <- rio::import(here::here("02_R", "rd_results_ever.csv"))
-rd_results_tv <- rio::import(here::here("02_R", "rd_results_tv.csv"))
-table_results_bounds <- rio::import(here::here("02_R", "table_results_bounds.csv")) %>% 
-  select(1,2,rr) %>% 
-  filter(Proxy != "Time to cancer",
-         model != "IPTW + IPCW") %>% 
-  pivot_wider(names_from = model, values_from = rr)
-
-rr <- rd_results_ever %>%
-  bind_rows(rd_results_tv) %>% 
-  select(Proxy, model, rr, conf) %>% 
-  pivot_wider(names_from = conf,
-              values_from = rr) %>% 
-  mutate(contrast = "Risk Ratio") %>% 
-  rename(conf.high = conf.higher) %>% 
-  mutate(Proxy = str_replace(Proxy, "Time-varing", "Time-varying"))
-
-results <- hr_results %>% 
-  mutate(contrast = "Hazard Ratio") %>%
-  filter(Proxy != "Time to cancer") %>%
-  mutate(Proxy = ifelse(
-    Proxy == "Ever vs. Never", "Ever vs. never", "Time-varying")) %>% 
-  bind_rows(rr) %>% 
-  mutate(Proxy = str_to_sentence(Proxy))
-```
-
-```{r xaringan-themer, include=FALSE}
-library(xaringanthemer)
-style_mono_accent(
-  base_color = "#011A5E",
-  header_font_google = google_font("Fira Sans", "500"),
-  table_row_even_background_color = "#FFFFFF",
-  table_border_color = "#011A5E",
-  text_font_google   = google_font("Lato", "400", "400i"),
-  text_font_size = "30px"
-)
-```
-
-background-image: url(./figs/ser.jpg), url(./figs/erasmus.png)
-background-position: 5% 95%, 95% 5%
-background-size: 25%, 15%
-class: center, middle
-
-<br>
-
-## Towards a clearer understanding of the inverse association between cancer and dementia
-
-<br> <br> <br>
-
-.right[
-
-L. Paloma Rojas-Saunero, MD
-
-**Epidemiology department**
-
-`r icons::fontawesome$brands$twitter` @palolili23
-
-]
-
----
-### Introduction
-
-<br><br>
-
-- Observational studies have found an inverse association between cancer and dementia risk.
-
---
-
-- Several genetic and molecular mechanisms have been proposed.
-
---
-
-- Researchers have raised concerns related to measurement error, unmeasured confounding and selection bias.
-
----
-
-### Introduction
-
-```{tikz, echo = FALSE}
-\usetikzlibrary{arrows}
-\begin{tikzpicture}[node distance=2cm]
-\node (p) {$P_0$};
-\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
-\node [right of = p, gray!30] (dp) {$D_t$};
-\node [right of = dp, gray!30] (pstar) {$P^*_t$};
-\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
-\node [right of = d] (y) {$Y_{t+1}$};
-\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
-\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
-\draw[->, gray!30] (p) -- (dp);
-\draw[->, gray!30] (l) -- (p);
-\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
-\draw[->, gray!30] (dp) -- (pstar);
-\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (pstar);
-\draw[->, gray!30] (pstar) -- (d);
-\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (d);
-\draw[->, gray!30] (d) -- (y);
-\draw[->, gray!30] (c) -- (dp);
-\draw[->, gray!30] (c) -- (pstar);
-\draw[->, gray!30] (cd) -- (d);
-\draw[->, gray!30] (cd) -- (y);
-\end{tikzpicture}
-```
-
-$P_0$ = Pin1 targeted-drug; $Y_{t+1}$ = Dementia diagnosis
-
----
-### Pin1 expression as biomarker (Problem 1: _Confounding_)
-
-```{tikz, echo = FALSE}
+```{tikz, echo = FALSE, fig.cap = "Pin1 as biomarker and risk of dementia"}
 \usetikzlibrary{arrows}
@@
 
-$P_0$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis; $L$ = Shared causes of $P_0$ and $Y_{t+1}$
+---
 
-???
-We might be closer to developing biomarker for Pin1 expression in the future.
-So we can imagine a situation in which Pin1 exposure is measured in a population based cohort study, using stored blood samples.
-Since this would be a observational setting, Pin1 expression and ADRD at time may share causes such as smoking, and that assessing the causal relationship requires adjusting for these confounders.
-
----
-### Cancer diagnosis as the proxy of Pin1 (Problem 2: _Time zero_)
-
-```{tikz, echo = FALSE}
+```{tikz, echo = FALSE, fig.cap = "Cancer diagnosis as proxy for Pin1 expression"}
 \usetikzlibrary{arrows}
@@
 
-$P_0$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis; $P^*_t$ = Cancer diagnosis
+---
 
-???
-At this moment this biomarker is not available, but we know that Pin1 over-expression stimulates tumor growth and for this reason we use cancer diagnosis as the proxy for Pin1.
+```{tikz, echo = FALSE, fig.cap = "Survival bias when treating cancer diagnosis as the proxy for Pin1 expression"}
 
----
-### Death prior to cancer diagnosis (Problem 3: _Survival bias_)
-
-```{tikz, echo = FALSE}
-\usetikzlibrary{arrows}
-\begin{tikzpicture}[node distance=2cm]
-\node (p) {$P_0$};
-\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
-\node [right of = p] (dp) {$D_t$};
-\node [right of = dp] (pstar) {$P^*_t$};
-\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
-\node [right of = d] (y) {$Y_{t+1}$};
-\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
-\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
-\draw[->] (p) -- (dp);
-\draw[->, gray!30] (l) -- (p);
-\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
-\draw[->] (dp) -- (pstar);
-\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
-\draw[->, gray!30] (pstar) -- (d);
-\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (d);
-\draw[->, gray!30] (d) -- (y);
-\draw[->, gray!30] (c) -- (dp);
-\draw[->, gray!30] (c) -- (pstar);
-\draw[->, gray!30] (cd) -- (d);
-\draw[->, gray!30] (cd) -- (y);
-\end{tikzpicture}
-```
-
-$P_0$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis; $P^*_t$ = Cancer diagnosis; $D_t$ = Death prior to cancer diagnosis
-???
-
----
-### Death prior to cancer diagnosis (Problem 3: _Survival bias_)
-
-```{tikz, echo = FALSE}
 \usetikzlibrary{arrows}
@@
 \draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (d);
-\draw[->, gray!30] (d) -- (y);
+\draw[->] (d) -- (y);
 \draw[->] (c) -- (dp);
@@
 
-$P_0$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis; $P^*_t$ = Cancer diagnosis; $D_t$ = Death prior to cancer diagnosis; $C_1$ = Shared causes of $D_t$ and $P^*$
-???
+---
 
----
-### Death prior to dementia diagnosis (Problem 3: _Survival bias_)
-
-```{tikz, echo = FALSE}
+```{tikz, echo = FALSE, fig.cap = "Death as a competing event for dementia over follow-up"}
 \usetikzlibrary{arrows}
@@
 \draw[->, gray!30] (pstar) -- (d);
-\draw[->] (p) to [out=+45,in=+135, looseness=0.3] (d);
+\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (d);
 \draw[->] (d) -- (y);
@@
 
-$P_0$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis; $D_{t+1}$ = Death prior to dementia diagnosis
+---
 
----
-### Death prior to dementia diagnosis (Problem 3: _Survival bias_)
-
-```{tikz, echo = FALSE}
+```{tikz, echo = FALSE, fig.cap = "Direct effect"}
 \usetikzlibrary{arrows}
@@
 \node [right of = p, gray!30, gray!30] (dp) {$D_t$};
-\node [right of = dp, gray!30] (pstar) {$P^*_t$};
+\node [right of = dp] (pstar) {$P^*_t$};
 \node [right of = pstar] (d) {$D_{t+1}$};
@@
 \draw[->, gray!30] (dp) -- (pstar);
-\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (pstar);
-\draw[->, gray!30] (pstar) -- (d);
+\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
+\draw[->] (pstar) -- (d);
 \draw[->] (p) to [out=+45,in=+135, looseness=0.3] (d);
@@
 
-$P_0$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis; $D_{t+1}$ = Death prior to dementia diagnosis; $C_2$ as shared causes of $D_{t+1}$ and $Y_{t+1}$
-
----
-### Application to the Rotterdam Study
-
-- Participants between 60 and 70 years old at study entry, free of cancer and dementia diagnosis at baseline
-
---
-
-- Data on cancer diagnosis, dementia diagnosis and death was collected from national registries and clinical records.
-
---
-
-- Cancer:
-
-  + _Time-fixed ever vs. never_
-
-  + _Time-varying_
-
----
-### Application to the Rotterdam Study
-
-- Inverse probability treatment weights to address confounding (IPTW)
-
-  + *Age, sex, Apoe4, education*
-
---
-
-- Inverse probability censoring weights for death (IPCW)
-
-  + *Age, sex, Apoe4, education and time-varying covariates: smoking, systolic blood pressure, BMI and incident cancer, heart disease, stroke and diabetes*
-
---
-
-- Risk of dementia and risk ratios are estimated with a weighted Kaplan-Meier estimator.
-
---
-
-- Hazard ratios are estimated with a weighted Cox Proportional-Hazards Model for comparative purposes.
-
----
-### Bounds
-
-<br><br>
-
-- **1) Lower Bound**: Total independence, that refers to an scenario were those who died would never develop dementia (Death as competing event)
-
---
-
-- **2) Upper Bound**: Complete dependency, that refers to an scenario where those who died would have dementia prior to death (Composite outcome, death + dementia)
-
----
-### Results
-
-.center[
-```{r, echo = FALSE, fig.width= 14, fig.retina=3}
-plot_distribution +
-  theme(legend.position = "right",
-        title = element_text(face = "bold", size = rel(1.5), color = "#011A5E"),
-        axis.title.x = element_text(face = "bold", size = rel(1.5), color = "#011A5E")) +
-  labs(title = "Proportion of participants in each status")
-```
-]
-
 ---
 
-### Results
-
-.pull-left[
-```{r, echo = FALSE, fig.retina = 3}
-temp <-
-  table_results_bounds %>% slice(1) %>% mutate(model = "IPTW + IPCW", contrast = "Risk Ratio") %>% 
-  rename(conf.low = `Lower bound`, conf.high = `Upper bound`)
-
-results %>%
-  filter(Proxy == "Ever vs. never") %>%
-  mutate(model = fct_relevel(model, c("Unadjusted", "IPTW", "IPTW + IPCW"))) %>%
-  ggplot(aes(
-    y = contrast,
-    x = point,
-    xmin = conf.low,
-    xmax = conf.high
-  )) +
-  geom_vline(xintercept = 1,
-             color = "#E69F00",
-             size = 1.2) +
-  geom_pointrange(aes(shape = contrast), color = "#0072B2") +
-  geom_errorbar(aes(group = contrast),
-                width = 0.2,
-                color = "#0072B2") +
-  geom_point(
-    data = temp,
-    aes(x = conf.low),
-    shape = 18,
-    color = "#011A5E",
-    size = 3
-  ) +
-  geom_point(
-    data = temp,
-    aes(x = conf.high),
-    shape = 18,
-    color = "#011A5E",
-    size = 3
-  ) +
-  facet_grid(factor(model, levels = c("Unadjusted", "IPTW", "IPTW + IPCW")) ~
-                 ., switch = "y") +
-  # facet_grid(model ~., switch = "y", margins = FALSE) +
-  scale_x_continuous(trans = "log10") +
-  labs(
-    x = NULL,
-    y = NULL,
-    group = NULL,
-    shape = NULL,
-    title = "Cancer as ever vs. never",
-    subtitle = "Risk ratios \U25B2, Hazard ratios \U25CF, Bounds \U25C6" ) +
-  theme(
-    panel.spacing = unit(0, "lines"),
-    panel.background = element_blank(),
-    axis.text.y = element_blank(),
-    axis.ticks.y = element_blank(),
-    strip.text.y = element_text(
-      face = "bold",
-      size = rel(1.5),
-      color = "#011A5E"
-    ),
-    axis.line.x = element_line(color = "#011A5E"),
-    axis.text.x = element_text(size = rel(1.5), color = "#011A5E"),
-    strip.background = element_blank(),
-    plot.title = element_text(face = "bold", size = rel(2), color = "#011A5E"),
-    plot.subtitle = element_text(face = "bold", size = rel(1.5), color = "#011A5E", hjust = 0.05),
-    legend.position = "none")
-
+```{tikz, echo = FALSE, fig.cap = "Direct effect"}
+\usetikzlibrary{arrows}
+\begin{tikzpicture}[node distance=2cm]
+\node (p) {$P_0$};
+\node [above of = p, yshift = -1cm, xshift = 1cm](l) {$L$};
+\node [right of = p] (dp) {$D_t$};
+\node [right of = dp] (pstar) {$P^*_t$};
+\node [right of = pstar] (d) {$D_{t+1}$};
+\node [right of = d] (y) {$Y_{t+1}$};
+\node [below of = dp, yshift = 1cm, xshift = 1cm](c) {$C_1$};
+\node [below of = d, yshift = 1cm, xshift = 1cm](cd) {$C_2$};
+\draw[->] (p) -- (dp);
+\draw[->] (l) -- (p);
+\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
+\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->] (dp) -- (pstar);
+\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
+\draw[->] (pstar) -- (d);
+\draw[->] (p) to [out=+45,in=+135, looseness=0.3] (d);
+\draw[->] (d) -- (y);
+\draw[->] (c) -- (dp);
+\draw[->] (c) -- (pstar);
+\draw[->] (cd) -- (d);
+\draw[->] (cd) -- (y);
+\end{tikzpicture}
 ```
-
-]
-
---
-
-.pull-right[
-```{r, echo = FALSE, fig.retina = 3}
-temp <-
-  table_results_bounds %>% slice(2) %>% mutate(model = "IPTW + IPCW", contrast = "Risk Ratio") %>% 
-  rename(conf.low = `Lower bound`, conf.high = `Upper bound`)
-
-results %>%
-  filter(Proxy == "Time-varying") %>%
-  mutate(model = fct_relevel(model, c("Unadjusted", "IPTW", "IPTW + IPCW"))) %>%
-  ggplot(aes(
-    y = contrast,
-    x = point,
-    xmin = conf.low,
-    xmax = conf.high
-  )) +
-  geom_vline(xintercept = 1,
-             color = "#E69F00",
-             size = 1.2) +
-  geom_pointrange(aes(shape = contrast), color = "#0072B2") +
-  geom_errorbar(aes(group = contrast),
-                width = 0.2,
-                color = "#0072B2") +
-  geom_point(
-    data = temp,
-    aes(x = conf.low),
-    shape = 18,
-    color = "#011A5E",
-    size = 3
-  ) +
-  geom_point(
-    data = temp,
-    aes(x = conf.high),
-    shape = 18,
-    color = "#011A5E",
-    size = 3
-  ) +
-  facet_grid(factor(model, levels = c("Unadjusted", "IPTW", "IPTW + IPCW")) ~
-                 ., switch = "y") +
-  scale_x_continuous(trans = "log10") +
-  labs(
-    x = NULL,
-    y = NULL,
-    group = NULL,
-    shape = NULL,
-    title = "Time-varying cancer",
-    subtitle = "Risk ratios \U25B2, Hazard ratios \U25CF, Bounds \U25C6" ) +
-  theme(
-    panel.spacing = unit(0, "lines"),
-    panel.background = element_blank(),
-    axis.text.y = element_blank(),
-    axis.ticks.y = element_blank(),
-    strip.text.y = element_text(
-      face = "bold",
-      size = rel(1.5),
-      color = "#011A5E"
-    ),
-    axis.text.x = element_text(size = rel(1.5), color = "#011A5E"),
-    axis.line.x = element_line(color = "#011A5E"),
-    strip.background = element_blank(),
-    plot.title = element_text(face = "bold", size = rel(2), color = "#011A5E"),
-    plot.subtitle = element_text(face = "bold", size = rel(1.5), color = "#011A5E", hjust = 0.05),
-    legend.position = "none")
-```
-
-]
-
----
-### Conclusions
-
-- Defining the underlying mechanism behind cancer diagnosis helps in multiple way.
-
-  + Helps identify sources of bias and guides analysis
-
-  + Opens conversations about future data collection
-
---
-
-- Since death acts as a competing event, we first need to define the estimand of interest.
-
-  + The direct effect relies on the independent censoring assumption
-
-  + Interpretation implies preventing death.
-
-  + The separable effects provides a clearer intuition to address this question.
-
----
-class: middle
-
-### Acknowledgments
-
-- Sonja A. Swanson
-
-- Kimberly van der Willik
-
-- Arfan M. Ikram
-
----
-class:middle
-
-### Gracias!   Thank you!
-
-`r icons::fontawesome("paper-plane", style = "solid")` </i>&nbsp; l.rojassaunero@erasmusmc.nl </a> <br>
-
-`r icons::fontawesome$brands$twitter` @palolili23
-
-`r icons::fontawesome$brands$github` @palolili23

### 2021-04-30 04_reports/manuscript.Rmd
updates on manuscript

--- 2021-04-12 04_reports/manuscript.Rmd
@@
 output: [pdf_document]
-
+editor_options: 
+  chunk_output_type: console
 ---
 ```{r, echo =  FALSE, message = FALSE, warning = FALSE}
-knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.align="center")
+knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
 library(survival)
@@
 library(tableone)
-source(here::here("02_R", "03_descriptives.R"))
-source(here::here("02_R", "05_plots.R"))
-source(here::here("02_R", "05b_plot_causes_death.R"))
+library(magrittr)
+library(kableExtra)
+# source(here::here("02_R", "03_descriptives.R"))
+# source(here::here("02_R", "05_plots.R"))
+# source(here::here("02_R", "05b_plot_causes_death.R"))
+# table_results <- rio::import(here::here("02_R", "table_results.csv"))
+# table_results_bounds <- rio::import(here::here("02_R", "table_results_bounds.csv"))
+# models <- read_csv(here::here("04_reports", "models.csv")) %>% select(-6)
 ```
@@
 
-There is significant interest in understanding the association between cancer and dementia1.
-Many observational studies have consistently determined that individuals with a history or incident cancer have a lower risk of developing of Alzheimer´s disease or related dementias (ADRD) when compared to individuals with no history of cancer.
-These findings have motivated the identification of molecular and genetic mechanisms that explain this association.
-Specific molecular mechanisms, such as Pin1 enzyme over-expression promotes tumorigenesis, while it’s down-regulation is attributed to mechanisms that contribute to Alzheimer’s Disease.
-These findings are suggest new opportunities to repurpose cancer chemotherapeutics for Alzheimer´s Disease.
+Many observational studies have consistently found that individuals with cancer have a lower risk of developing of Alzheimer´s disease or related dementias (ADRD) when compared to individuals with no history of cancer1–4.
+These findings have motivated substantial research toward mechanistic explanations, including searching for and hypothesizing that molecular and genetic mechanisms may explain this association5–12.
+These research inquiries inevitably lead to discussions of repurposing or augmenting current cancer chemotherapeutics for ADRD13.
 
-Nevertheless, observational studies are susceptible to several bias, and researchers have raised concerns related to the competing event of death as well as unmeasured confounding and ascertainment error  that could drive these results.
-Though, to understand sources of bias the causal question should be explicitly  specified.
-This particular question has at least two core challenges.
-First, defining the underlying biological mechanism we would ideally intervene upon and defining cancer diagnosis as the proxy for that mechanism.
-For example, researchers refer to biological mechanisms which onset may be prior to cancer diagnosis and to the effect of cancer treatment, both relate to different causal questions and would be translated into different study design and analytic choices.
-Second, defining the causal question (or estimand) of interest when the competing event of death is interest.
-Previous studies alluded to choosing a particular summary measure or modeling approach to account for death (incidence rate ratios (ref), hazard ratios (ref), multistate models (ref)).
-However, when competing events are present, there is more than one causal question that can be framed and each question has different interpretations and relies on different assumptions.
+Nevertheless, inferring any treatment or mechanistic effects from the observed cancer-ADRD inverse association is not straightforward.
+Researchers have raised concerns related to the competing event of death, unmeasured confounding, and ascertainment error that could explain these results9,14.
+However, understanding these or other sources of bias first requires that we make explicit the causal question.
+Moreover, making explicit the causal question is one step toward tying a research study to a question that is relevant to decision-making.
 
-With these considerations in mind, we outline a case study focused on the question: _What is the direct effect of Pin1 over-expression at late-midlife in the risk of ADRD after 20 years of follow-up compared to having regulated expression of Pin1._ To identify this effect with real-world data, we illustrate the complexity of the research question by building progressively a directed acyclic graph (DAG) which help us connect the causal question to the observable data and the assumptions we rely on to identify the effect.
+To illustrate the complexities of inferring hypothetical or available treatments’ effects on ADRD from the observed cancer-ADRD association, we focus on the   Pin1 enzyme.
+Previous animal studies have shown that the Pin1 enzyme over-expression promotes tumorigenesis, while its down-regulation is attributed to mechanisms that contribute to neurodegeneration and amyloid deposition11,12,15.
+If we one day could develop a drug that increases Pin1 expression specifically in brain tissue in hopes of preventing dementia, we could pose the question as: _What is the direct effect of this Pin1-targeting drug on the risk of ADRD over time compared to standard treatments?_
+
+To explore how we might learn about this effect using real-world data on cancer and ADRD, we progressively build a causal directed acyclic graph to connect this particular causal question to the observable data and the assumptions we rely on to identify the effect.
 We exemplify different scenarios with data collected from the Rotterdam Study, a population-based cohort study.
@@
 
-## 2. Overview of the causal structure and assumptions related to cancer diagnosis as a proxy for Pin1
+## 2. Overview of the causal structure
 
-We begin by illustrating a hypothetical scenario through the causal graph of Figure 1, that depicts a situation where Pin1, represented as $P$, was measured in all participants in late midlife (for example between age 50 and 60).
-Based on the measurement of Pin1, participant's exposure is classified as 1) Over-expressed Pin1 or 2) Regulated Pin1.
-Although Pin1 expression acts as a biomarker and is not a well-defined intervention, we can allude that sometime in the future people with over-expressed Pin1 could follow a specific treatment that lowers Pin1 under a specified threshold.
-Furthermore, through out the follow-up each participant followed a strict monitoring for ADRD onset, represented as $Y_{20}$, and no participant was lost to follow-up nor died during the study period (no competing events present).
+If this hypothetical Pin1-targeting drug was developed, the best way to understand its effect on dementia risk would be to do a well-conducted randomized trial in which we randomize eligible participants in late midlife (e.g., ages 50-60 years) to receive this drug or not, and closely monitor ADRD over a lengthy follow-up.
 
-Since Pin1 expression was not a randomized intervention, we consider a set of measured covariates $L$ for which conditional exchangeability holds.
+Since there is no drug currently available that targets Pin1, at best we can use observational data on Pin 1 expression measurements.
+For example, suppose that a biomarker test was available to measure Pin1 and we measured this biomarker from (stored) baseline blood samples in a population based-cohort that recruited participants in late midlife.
+Since the biomarker Pin1 is measured within an observational study, confounding can explain an observed association between it and ADRD.
+In Figure 1, we show that, Pin1 expression ($P_0$) and ADRD at time $t+1$ ($Y_{t+1}$) may share causes $L$, and that assessing the causal relationship requires adjusting for these confounders $L$.
 Previous studies have described age, sex, educational level and race/ethnicity as the minimal adjusting set of covariates (Ospina).
-However, environmental and behavioral factors such as smoking, alcohol intake, or chronic diseases which are known to cause microenvironmental changes such as inflammation and changes in tissue remodelling, may translate into Pin1 over-expression and are also related to the development of ADRD.
-Therefore, had we measured all $L$ we could identify the effect of $P$ in $Y_{20}$ by conditioning on $L$.
+However, environmental and behavioral factors such as smoking, which are known to cause microenvironmental changes such as inflammation and changes in tissue remodeling, may translate into Pin1 over-expression and are also related to the development of ADRD.
 
-```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = "Direct effect of Pin1 in ADRD"}
+For simplicity, we treat Pin1 expression as a point intervention and we fix the time-ordering of covariates (that is, we assume $L$ happens prior to $P_0$).
+In reality, it is possible that Pin1 expression changes over time and is affected by time-varying confounders (like smoking), which produces treatment-confounder feedback loops.
+Addressing such time-varying confounding would require repeated measurements of  $L$ and $P_0$16.
+
+```{tikz, echo = FALSE, fig.cap = "Pin1 as biomarker and risk of dementia"}
 \usetikzlibrary{arrows}
-\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
-\node (p) {$P$};
-\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$L$};
-\node [right of = p] (y) {$Y_{20}$};
+\begin{tikzpicture}[node distance=2cm]
+\node (p) {$P_0$};
+\node [above of = p, yshift = -1cm, xshift = 1cm](l) {$L$};
+\node [right of = p, gray!30] (dp) {$D_t$};
+\node [right of = dp, gray!30] (pstar) {$P^*_t$};
+\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
+\node [right of = d] (y) {$Y_{t+1}$};
+\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
+\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
+\draw[->, gray!30] (p) -- (dp);
 \draw[->] (l) -- (p);
-\draw[->, violet] (p) -- (y);
-\draw[->] (l) -- (y);
+\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
+\draw[->, semithick] (p) to [out=45,in=135, looseness=0.3] (y);
+\draw[->, gray!30] (dp) -- (pstar);
+\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (pstar);
+\draw[->, gray!30] (pstar) -- (d);
+\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (d);
+\draw[->, gray!30] (d) -- (y);
+\draw[->, gray!30] (c) -- (dp);
+\draw[->, gray!30] (c) -- (pstar);
+\draw[->, gray!30] (cd) -- (d);
+\draw[->, gray!30] (cd) -- (y);
 \end{tikzpicture}
@@
 
-Unfortunately, Pin1 expression measurement is not available at the moment in real-world settings, we can only rely on a proxy of this exposure.
-Since Pin1 over-expression is present in tumors and tumors are only measured through screening and diagnosis, we considered cancer diagnosis as the proxy for Pin1 over-expression, as most observational studies have defined (all refs).
-We depict this feature in Figure 2, where $P^*$ represents _incident cancer diagnosis_, the measured proxy of $P$.
-In this DAG we colored the path $P^* \leftarrow P \rightarrow Y_{85}$ because although we are measuring the association between $P^*$ and $Y_{20}$ in the observed data, we are assuming that the captured effect is only through $P$.
+However, unfortunately, in current studies we do not even have a single measurement of Pin1 expression, let alone repeated measurements.
+Thus, we can only rely on a proxy of this exposure.
+Since Pin1 over-expression is present in tumors, and tumors are only measured through screening and diagnosis, we considered cancer diagnosis as the proxy for Pin1 over-expression4,17–24.
+We depict this feature in Figure 2, where $P^*$ represents _incident cancer diagnosis_, the measured proxy of $P_0$.
+In this causal graph  we colored the path $P^* \leftarrow P \leftarrow Y_{t+1}$ because although we are measuring the association between $P^*$ and $Y_{t+1}$ in the observed data, we are assuming that the captured effect is only through $P_0$.
 
-```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = "Direct effect of Pin1 in ADRD, with cancer diagnosis as proxy of Pin1"}
+```{tikz, echo = FALSE, fig.cap = "Cancer diagnosis as proxy for Pin1 expression"}
 \usetikzlibrary{arrows}
-\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
-\node (p) {$P$};
-\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$L$};
-\node [above of = p, yshift = -1cm](pstar) {$P^*$};
-\node [right of = p] (y) {$Y_{85}$};
-\draw[->, violet] (p) -- (pstar);
-\draw[->, violet] (p) -- (y);
-\draw[->] (l) -- (p);
-\draw[->] (l) -- (y);
-
+\begin{tikzpicture}[node distance=2cm]
+\node (p) {$P_0$};
+\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
+\node [right of = p, gray!30] (dp) {$D_t$};
+\node [right of = dp] (pstar) {$P^*_t$};
+\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
+\node [right of = d] (y) {$Y_{t+1}$};
+\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
+\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
+\draw[->, gray!30] (p) -- (dp);
+\draw[->, gray!30] (l) -- (p);
+\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
+\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->, gray!30] (dp) -- (pstar);
+\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
+\draw[->, gray!30] (pstar) -- (d);
+\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (d);
+\draw[->, gray!30] (d) -- (y);
+\draw[->, gray!30] (c) -- (dp);
+\draw[->, gray!30] (c) -- (pstar);
+\draw[->, gray!30] (cd) -- (d);
+\draw[->, gray!30] (cd) -- (y);
 \end{tikzpicture}
@@
 For this reason prospective cohort studies have considered cancer diagnosis as a time-varying or time-dependent exposure.
-This means that a participant within a prospective cohort study contributes to the "regulated Pin1" since study entry up to the time of cancer diagnosis and later on to the "Over-expressed Pin1" arm.
-Other studies have included participants with cancer diagnosis at the time of the diagnosis (for example from cancer registries such as SEER) and matched participants by age.
-In both cases we must remember that the the main interest is on the unmeasured $P$ and we should only adjust for covariates prior to $P$, and be careful to adjust for post-baseline covariates of $P$ or mediators between $P$ and $P^*$.
+This means that a participant within a prospective cohort study contributes to the "regulated Pin1" arm since study entry up to the time of cancer diagnosis and later on to the "Over-expressed Pin1" arm.
+Other studies have included participants with cancer diagnosis at the time of the diagnosis (for example from cancer registries19,23,25,26) and matched participants by age.
+In both cases we must remember that the main interest is on the unmeasured $P_0$ thus we should only adjust for covariates prior to $P_0$ and be careful to adjust for post-baseline covariates of $P_0$ or mediators between $P_0$ and $P^*$.
 
-One of the mediators between $P$ and $P^*$ is death prior to cancer diagnosis.
-Since individuals are at risk of dying from other causes (such as cardivascular death), we can only measure $P^*$ in the subset of individuals who have survived long enough to have a cancer diagnosis.
-We zoom in the relationship of $P \rightarrow P^*$ in Figure 3.
-In this DAG we include
-$D = 0$ between $P$ and $P^*$, this illustrates that to observe $P^*$, we condition (box around $D = 0$) on surviving long enough to have a cancer diagnosis.
-Several risk factors that increase the risk of cancer might also cause death prior to cancer diagnosis, for example smoking may cause lung cancer and chronic obstructive pulmonary disease (a leading cause of death in this age group).
-Therefore, to isolate the effect of $P$ in $P^*$ (the violet arrow), we need to block all shared causes between $P$ and $D$ and between $D$ and $P*$ both represented in Figure 3 as $C_1$ and $C_2$.
-In this way we assume a hypothetical scenario in which we could prevent death prior to cancer diagnosis, by conditioning on a rich set of covariates.
+One of the mediators between $P_0$ and $P^*$ is death prior to cancer diagnosis, represented as $D_t$ in Figure 3.
+Since individuals are at risk of dying from other causes (such as cardiovascular death), we can only measure $P^*$ in the subset of individuals who have survived long enough to have a cancer diagnosis.
+Several risk factors that increase the risk of cancer might also cause death prior to cancer diagnosis, for example smoking may cause lung cancer and chronic obstructive pulmonary disease (a leading cause of death in this age group27).
+Therefore, we can only isolate the effect of $P_0$ in $P^*$ if we condition on $C_1$ and block the backdoor pathway that is open by conditioning on $D_t$.
+In this way we assume a hypothetical scenario in which we could prevent death prior to cancer diagnosis, by conditioning on a rich set of covariates28.
 This assumption must hold regardless of whether we use incident cancer as a time-varying exposure, or match cancer patients to participants free of cancer by age.
 
-```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = "Zoom in the association between P and P star"}
+Furthermore, we note that this is only one of the issues with considering cancer diagnosis as the proxy for Pin1 expression in terms of information bias(ref).
+We could add more complexity by considering further shared causes of cancer and ADRD diagnosis, or further mediators between Pin1 expression and cancer diagnosis, including but not limited to: screening guidelines, type of healthcare coverage, health-seeking behaviors, and healthcare availability.
+
+```{tikz, echo = FALSE, fig.cap = "Survival bias when treating cancer diagnosis as the proxy for Pin1 expression"}
+
 \usetikzlibrary{arrows}
-\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
-\node (p) {$P$};
-\node [draw, right of = p] (d) {$D = 0$};
-\node [right of = d] (pstar) {$P^*$};
-\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$C_1$};
-\node [below of = d, yshift = 1cm, xshift = 1cm](c) {$C_2$};
-\draw[->] (p) -- (d);
-\draw[->, violet] (p) to [out=45,in=135, looseness=0.4] (pstar);
-\draw[->] (d) -- (pstar);
-\draw[->] (l) -- (p);
-\draw[->] (l) -- (d);
+\begin{tikzpicture}[node distance=2cm]
+\node (p) {$P_0$};
+\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
+\node [right of = p] (dp) {$D_t$};
+\node [right of = dp] (pstar) {$P^*_t$};
+\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
+\node [right of = d] (y) {$Y_{t+1}$};
+\node [below of = dp, yshift = 1cm, xshift = 1cm](c) {$C_1$};
+\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
+\draw[->] (p) -- (dp);
+\draw[->, gray!30] (l) -- (p);
+\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
+\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->] (dp) -- (pstar);
+\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
+\draw[->, gray!30] (pstar) -- (d);
+\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (d);
+\draw[->] (d) -- (y);
+\draw[->] (c) -- (dp);
 \draw[->] (c) -- (pstar);
-\draw[->] (c) -- (d);
+\draw[->, gray!30] (cd) -- (d);
+\draw[->, gray!30] (cd) -- (y);
 \end{tikzpicture}
@@
 
-Up to this point we have outlined the underlying assumptions of using incident cancer diagnosis as a proxy for Pin1 expression.
-We continue outlining the additional assumptions related to death as a competing event of ADRD.
+Up to this point we have outlined certain underlying assumptions of using incident cancer diagnosis as a proxy for Pin1 expression.
+However, as mortality increases steeply in late life, even in the setting of the ideal randomized trial of a hypothetical Pin1-targeting drug, we can only measure ADRD over follow-up in the individuals who survive long enough to have a diagnosis.
+For this reason, death is a competing event of ADRD because if a participant dies prior to ADRD diagnosis, death prevents observing ADRD at future time-points.
+As such, to identify the direct effect of $P_0$ in $Y_{t+1}$ we have to first assess if the drug has an any side effects that could increase the risk of death in any way.
+If the drug has no systemic side-effects, such as that there is no arrow between $P_0$ and $D_{t+1}$, such as observed in Figure 4.
 
-## 3. Overview of the causal question and assumptions related to death as a competing event of ADRD
+```{tikz, echo = FALSE, fig.cap = "Death as a competing event for dementia over follow-up"}
+\usetikzlibrary{arrows}
+\begin{tikzpicture}[node distance=2cm]
+\node (p) {$P_0$};
+\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
+\node [right of = p, gray!30, gray!30] (dp) {$D_t$};
+\node [right of = dp, gray!30] (pstar) {$P^*_t$};
+\node [right of = pstar] (d) {$D_{t+1}$};
+\node [right of = d] (y) {$Y_{t+1}$};
+\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
+\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
+\draw[->, gray!30] (p) -- (dp);
+\draw[->, gray!30] (l) -- (p);
+\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
+\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->, gray!30] (dp) -- (pstar);
+\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (pstar);
+\draw[->, gray!30] (pstar) -- (d);
+\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (d);
+\draw[->] (d) -- (y);
+\draw[->, gray!30] (c) -- (dp);
+\draw[->, gray!30] (c) -- (pstar);
+\draw[->, gray!30] (cd) -- (d);
+\draw[->, gray!30] (cd) -- (y);
+\end{tikzpicture}
+```
 
-As illustrated in previous figures, we are interested in the effect of Pin1 in late midlife in ADRD risk after 20 years of follow-up.
-Nevertheless, mortality is substantial and increases steeply in late life, and a leading cause of death is cancer.
-This means that in order to measure ADRD over follow-up individuals have to had survive to cancer and to other leading causes of death.
-For this reason death is a competing event of ADRD because if a participant dies prior to ADRD diagnosis, death prevents from observing ADRD at future time-points.
-We can visualize this data feature in the causal diagram in Figure 4.
-In this DAG we include $D_{15}$ as a representation of death at 15 years of follow-up, an arrow between $P$ and $D_{15}$ since $P$ may increase the mortality risk irrespective of cancer diagnosis.
-We also include an arrow between $P^*$ and $D_{15}$ since cancer diagnosis, and subsequent treatment (or lack of treatment) may have an effect on death.
-Last, the arrow from $D_{15}$ and $Y_{20}$ represents the key feature of a competing events data structure: an individual who dies at 15 years of follow-up cannot subsequently develop ADRD, and since $D_{15}$ and $Y_{20}$ are events related to aging, $C$ represent the shared causes of both events.
+However, if the drug has any side-effects we have to decide if either we want to estimate a total effect (that is, the effect on ADRD even if partly mediated by the effect on death) or a direct effect (that is, the effect directly on ADRD not mediated by death) (ref).
+We face this challenge in our observational data setting, since we are measuring cancer diagnosis as the proxy for Pin1 over-expression, and cancer is a leading cause of death29.
+We can visualize this data feature in the causal diagram in Figure 5.
+In this causal diagram we include $D_{t+1}$ as a representation of death over follow-up, an arrow between $P_0$ and $D_{t+1}$ since $P_0$ may increase the mortality risk.
+We also include an arrow between $P^*_t$ and $D_{t+1}$ since cancer diagnosis, and subsequent treatment (or lack of treatment) may have an effect on death.
+Last, the arrow from $D_{t+1}$ and $Y_{t+1}$ represents the key feature of a competing events data structure: an individual who dies over follow-up (prior to dementia diagnosis) cannot subsequently develop ADRD, and since $D_{t+1}$ and $Y_{t+1}$ are events related to aging, $C_2$ represent the shared causes of both events.
 
-In this DAG we observe that, had we measured and adjusted for $L$, we could estimate the total effect of $P$ in $Y_{20}$ without further assumptions.
-However the total effect includes all pathways between $P$ and $Y_{20}$.
+In this causal diagram we observe that, had we measured and adjusted for $L$, we could estimate the total effect of $P_0$ in $Y_{t+1}$ without further assumptions.
+However the total effect includes all pathways between $P_0$ and $Y_{20}$28.
 This means that if Pin1 has an effect on mortality through cancer or other mechanisms, we will observe a protective effect of Pin1 over-expression in ADRD, partially or fully mediated through death.
-For this reason estimators that are used to approximate a total effect (JY2020), such as the Aalen-Johansen estimator (underlying estimator for multistate models) or subdistribution hazard models, may not be useful in this setting.
-Since the question of interest is focused on the direct effect of $P$ in $Y_{20}$ as in Figure 4 (violet arrows), there are several causal questions (estimands) to represent this direct effect, such as: the controlled direct effect (CDE), the survivor average causal effect (SACE) and the natural separable direct effect.
-In this section we discuss the controlled direct effect as the causal question of interest since it translates to frequently used estimators in this literature (such as Kaplan-Meier estimator and Cox-proportional hazard model).
+Since the question of interest is focused on the direct effect of $P_0$ in $Y_{t+1}$ as in Figure 4 (violet arrows) we need to conceptualize the different mechanisms through which $P_0$ could affect $D_{t+1}$ and $Y_{20}$.
+With this in mind we can conceive different causal questions (estimands) to represent this direct effect , such as the controlled direct effect (CDE) and the natural separable direct effect.
+In this section we discuss the controlled direct effect as the causal question of interest since it translates to frequently used methods in this literature (such as Kaplan-Meier estimator and Cox-proportional hazard model) and leave the separable direct effects question for discussion.
 
-THe CDE represents the effect of Pin1 in ADRD in a setting where we could have prevented death over the study period.
-It relies on the assumption that we have measured all $C$ to block the pathway $Y85 \leftarrow C  \rightarrow D_{80} \rightarrow P \rightarrow P*$.
+```{tikz, echo = FALSE, fig.cap = "Direct effect"}
+\usetikzlibrary{arrows}
+\begin{tikzpicture}[node distance=2cm]
+\node (p) {$P_0$};
+\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
+\node [right of = p, gray!30, gray!30] (dp) {$D_t$};
+\node [right of = dp] (pstar) {$P^*_t$};
+\node [right of = pstar] (d) {$D_{t+1}$};
+\node [right of = d] (y) {$Y_{t+1}$};
+\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
+\node [below of = d, yshift = 1cm, xshift = 1cm](cd) {$C_2$};
+\draw[->, gray!30] (p) -- (dp);
+\draw[->, gray!30] (l) -- (p);
+\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
+\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->, gray!30] (dp) -- (pstar);
+\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
+\draw[->] (pstar) -- (d);
+\draw[->] (p) to [out=+45,in=+135, looseness=0.3] (d);
+\draw[->] (d) -- (y);
+\draw[->, gray!30] (c) -- (dp);
+\draw[->, gray!30] (c) -- (pstar);
+\draw[->] (cd) -- (d);
+\draw[->] (cd) -- (y);
+\end{tikzpicture}
+```
+
+The CDE represents the effect of Pin1 over-expression in ADRD in a setting where we could have prevented death over the study period.
+It relies on the assumption that we have measured all $C$ to block the pathway $Y20 \leftarrow C  \rightarrow D_{15} \rightarrow P \rightarrow P*$.
 This assumption is defined as the independent censoring assumption conditional on covariates.
@@
 
-```{tikz, fig.ext = 'png', echo = FALSE, fig.cap = "Effect of Pin1 in the risk of ADRD by age 85, with cancer diagnosis as proxy of Pin1"}
+Therefore, if we combine the challenges in section two related to cancer diagnosis as a proxy for Pin1 over-expression, and having death as a competing event of ADRD we observe the complexity of the DAG in Figure 5.
+Though this is yet a simplified version since we omitted time varying $P*$, $L$ and $C's$.
+
+```{tikz, echo = FALSE, fig.cap = "Direct effect"}
 \usetikzlibrary{arrows}
-\begin{tikzpicture}[node distance=2cm, auto,>=latex, scale = 0.5]
-\node (p) {$P$};
-\node [below of = p, yshift = 1cm, xshift = -1cm](l) {$L$};
-\node [above of = p, yshift = -1cm](pstar) {$P^*$};
-\node [draw, right of = p] (d) {$D_{80}$};
-\node [right of = d] (y) {$Y_{85}$};
-\node [below of = d, yshift = 1cm, xshift = +1cm](c) {$C$};
-\draw[->, violet] (p) -- (pstar);
+\begin{tikzpicture}[node distance=2cm]
+\node (p) {$P_0$};
+\node [above of = p, yshift = -1cm, xshift = 1cm](l) {$L$};
+\node [right of = p] (dp) {$D_t$};
+\node [right of = dp] (pstar) {$P^*_t$};
+\node [right of = pstar] (d) {$D_{t+1}$};
+\node [right of = d] (y) {$Y_{t+1}$};
+\node [below of = dp, yshift = 1cm, xshift = 1cm](c) {$C_1$};
+\node [below of = d, yshift = 1cm, xshift = 1cm](cd) {$C_2$};
+\draw[->] (p) -- (dp);
+\draw[->] (l) -- (p);
+\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
+\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->] (dp) -- (pstar);
+\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
 \draw[->] (pstar) -- (d);
-\draw[->] (p) -- (d);
+\draw[->] (p) to [out=+45,in=+135, looseness=0.3] (d);
 \draw[->] (d) -- (y);
-\draw[->] (l) -- (p);
-\draw[->] (l) -- (y);
-\draw[->] (c) -- (d);
-\draw[->] (c) -- (y);
-\draw[->, violet] (p) to [out=45,in=135, looseness=0.4] (y);
+\draw[->] (c) -- (dp);
+\draw[->] (c) -- (pstar);
+\draw[->] (cd) -- (d);
+\draw[->] (cd) -- (y);
 \end{tikzpicture}
 ```
-
-Since the conditional independent censoring assumption is untestable, when there are no other mechanisms of censoring, we can explore the extreme scenarios and estimate the bounds.
-
-## 4. Motivating example with the RS data
-
-In this section we illustrate how previous DAGS may reflect on the analysis performed to unveil the effect of Pin1 in all cause-dementia.
-We use data collected in the Rotterdam Study, a population-based prospective cohort study among persons living in the Ommoord district in Rotterdam, the Netherlands.
-Recruitment and initial assessments were held between 1990 and 1993; it was later extended between 2000 and 2001 consisting of individuals who had reached the age of 55 years or who had moved into the study area.
-Participants from first subcohort had follow-up visits between 1993-1995, 1997-1999, 2002-2005, and 2008-2010, second subcohort had follow-up visits between 2004 and 2005, and between 2011 and 2012.
-All participants had data on incident cancer diagnosis and incident dementia diagnosis through follow-up, collected from medical records of general practitioners (including hospital discharge letters) and through linkage with national registries.
-Date and cause of death was collected on a weekly basis via municipal population registries.
-Data on clinical outcomes was available until 2015.
-
-To match the analysis to our initial question, we considered as inclusion criteria being between 60 and 70 years old at study entry, without history of cancer diagnosis, free of cognitive decline or with previous history of dementia.
-Out of 10998 persons who participated at study entry, `r total_n` were considered eligible.
-Time to cancer diagnosis, time to dementia diagnosis and death status was measured for all participants.
-All participants were followed from study entry until dementia diagnosis, death or 20 years after their individual baseline date, whichever occurred first.
-
-### 4.1. Methods
-
-First we illustrate the most simple scenario (as Figure 1), were Pin1 over-expression is defined as "cancer ever vs.
-never" as if Pin1 and cancer diagnosis happened at the same time if looked retrospectively.
-Second, we consider cancer as a time-varying exposure independent of death.
-This can be represented as follows, at five years from study entry the observed unadjusted risk of cancer was `r cancer_5y` and the risk of death was `r death_5y` for the entire cohort, thus we assume that those who died would have the same risk of cancer had they not died, unconditionally, over follow-up.
-Third, we consider time to cancer as the
-
-To address confounding we fit inverse probability treatment weights including age at study entry, sex, educational attainment, smoking status and cohort as covariates.
-
-To identify the controlled direct effect we considered death as an independent censoring mechanism conditional on time-varying shared causes of death and dementia.
-
-The difference in dementia risk considering cancer as a time-varying exposure independent of death.
-For example.
-at five years from follow-up the risk of cancer was `r cancer_5y` and the risk of death was `r death_5y`, thus we assume that those who died would have the same risk of cancer had they not died, unconditionally.
-
-### 4.2. Results
-
-Participants had a mean age of `r mean_age`, and `r women_prop` were women.
-Further details on participants are presented in Table 1.
-Over follow up, `r yes_cancer` developed cancer and `r no_cancer` remain free of cancer diagnosis, the median age of cancer diagnosis was `r cancer_time`.
-From the total sample, `r dem_total` had dementia over follow-up and median time to dementia was `r dem_time`.
-Among participants with incident cancer, `r cancer_dem` had dementia diagnosis and `r cancer_death` died over follow-up, `r cancer_alive` remain alive at 20 years since study entry.
-In contrast, among participants free of cancer diagnosis over follow-up, `r no_cancer_dem` were diagnosed with dementia and `r no_cancer_death` died over follow-up, `r no_cancer_alive` were alive at the end of follow-up.
-
-## 5. Discussion
-
-We need to figure out what the key aging-dependent changes are and how to modulate these factors safely
-
-Causes such as aging, smoking, irradiation etc are causing micro environmental changes, like inflamation and changes in tissue remodelling. This promotes selection for adaptive mutations.
-
-Efforts to prevent and treat cancer should converge with similar efforts to prevent other aging- associated diseases.
-
-We need better means to
-
-Youthful tissues
-
-Knowing the cause of death provides information about the direction of missclassification.
-In patients with cancer we observed a slight % that died with dementia as a primary cause.
-On the other hand, among individuals free of cancer, we observed % of individuals who died with cancer as a primary cause.
-
-In cancer you are afraid of missing dementia
-in cancer free you are afraid of missing cancer dx
-
-Make a table of definition of weights based on cancer def. and
-
-Explicitly outlining the estimands and the assumptions that connect the causal question to the observed data provide an opportunity to improve the design of observational studies and the interpretation of their findings, plus better insight of potential sources of bias.
-This is a crucial since these studies are providing insights that are guiding other fields of research in the area, from bench science to biostatistics and epidemiological methods15.
-To guide future observational and methodologic studies, our aim is to outline the potential estimands of interest that represent the underlying causal questions behind the association between cancer and dementia
-
-\newpage
-
-## Tables
-
-### Table 1: Descriptive characteristics of individuals who had a cancer diagnosis and of those free of cancer diagnosis over follow up.
-
-```{r}
-tableone_exp <-
-  print(
-    tableone,
-    explain = TRUE,
-    test = FALSE,
-    quote = FALSE,
-    noSpaces = TRUE,
-    printToggle = FALSE
-  )
-
-kableExtra::kable(tableone_exp, booktabs = T) %>% 
-  kableExtra::kable_styling(latex_options = c("striped", "hold_position"))
-
-```
-
-\newpage
-
-### Table 2. Risk difference and risk ratio for the risk of dementia
-
-\newpage
-
-## Figures
-
-### Figure 1: Distribution of participants under each health status, by age over follow-up
-
-```{r}
-plot_distribution
-```
-
-### Figure 2. Causes of death for participants with and without incident cancer
-
-```{r}
-causes_death
-```

### 2021-05-05 04_reports/manuscript.Rmd
updated manuscript and dags

@@
 date: ""
-output: [pdf_document]
+output: [html_document]
 editor_options: 
@@
 \node (p) {$P_0$};
-\node [above of = p, yshift = -1cm, xshift = 1cm](l) {$L$};
+\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
 \node [right of = p, gray!30] (dp) {$D_t$};
@@
 \node [right of = d] (y) {$Y_{t+1}$};
-\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
-\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
-\draw[->, gray!30] (p) -- (dp);
+\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C$};
 \draw[->] (l) -- (p);
 \draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->, semithick] (p) to [out=45,in=135, looseness=0.3] (y);
 \draw[->, gray!30] (dp) -- (pstar);
@@
 \draw[->, gray!30] (pstar) -- (d);
-\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (d);
 \draw[->, gray!30] (d) -- (y);
-\draw[->, gray!30] (c) -- (dp);
-\draw[->, gray!30] (c) -- (pstar);
 \draw[->, gray!30] (cd) -- (d);
 \draw[->, gray!30] (cd) -- (y);
+\draw[->, semithick] (p) to [out=45,in=135, looseness=0.2] (y);
 \end{tikzpicture}
@@
 \node (p) {$P_0$};
-\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
+\node [above of = p, yshift = -0.8cm, xshift = -1cm, gray!30](l) {$L$};
 \node [right of = p, gray!30] (dp) {$D_t$};
@@
 \node [right of = d] (y) {$Y_{t+1}$};
-\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
-\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
-\draw[->, gray!30] (p) -- (dp);
+\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C$};
 \draw[->, gray!30] (l) -- (p);
 \draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
 \draw[->, gray!30] (dp) -- (pstar);
@@
 \draw[->, gray!30] (pstar) -- (d);
-\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (d);
 \draw[->, gray!30] (d) -- (y);
-\draw[->, gray!30] (c) -- (dp);
-\draw[->, gray!30] (c) -- (pstar);
 \draw[->, gray!30] (cd) -- (d);
 \draw[->, gray!30] (cd) -- (y);
+\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
 \end{tikzpicture}
@@
 \node (p) {$P_0$};
-\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
+\node [above of = p, yshift = -0.8cm, xshift = -1cm, gray!30](l) {$L$};
 \node [right of = p] (dp) {$D_t$};
@@
 \node [right of = d] (y) {$Y_{t+1}$};
-\node [below of = dp, yshift = 1cm, xshift = 1cm](c) {$C_1$};
-\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
-\draw[->] (p) -- (dp);
+\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C$};
 \draw[->, gray!30] (l) -- (p);
 \draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
 \draw[->] (dp) -- (pstar);
@@
 \draw[->, gray!30] (pstar) -- (d);
-\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (d);
 \draw[->] (d) -- (y);
-\draw[->] (c) -- (dp);
-\draw[->] (c) -- (pstar);
 \draw[->, gray!30] (cd) -- (d);
 \draw[->, gray!30] (cd) -- (y);
+\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
 \end{tikzpicture}
@@
 \node (p) {$P_0$};
-\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
+\node [above of = p, yshift = -0.8cm, xshift = -1cm, gray!30](l) {$L$};
 \node [right of = p, gray!30, gray!30] (dp) {$D_t$};
@@
 \node [right of = d] (y) {$Y_{t+1}$};
-\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
-\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
-\draw[->, gray!30] (p) -- (dp);
+\node [below of = d, yshift = 1cm, xshift = 1cm,](cd) {$C$};
 \draw[->, gray!30] (l) -- (p);
 \draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
 \draw[->, gray!30] (dp) -- (pstar);
@@
 \draw[->, gray!30] (pstar) -- (d);
-\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (d);
 \draw[->] (d) -- (y);
-\draw[->, gray!30] (c) -- (dp);
-\draw[->, gray!30] (c) -- (pstar);
-\draw[->, gray!30] (cd) -- (d);
-\draw[->, gray!30] (cd) -- (y);
+\draw[->] (cd) -- (d);
+\draw[->] (cd) -- (y);
+\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
 \end{tikzpicture}
@@
 We also include an arrow between $P^*_t$ and $D_{t+1}$ since cancer diagnosis, and subsequent treatment (or lack of treatment) may have an effect on death.
-Last, the arrow from $D_{t+1}$ and $Y_{t+1}$ represents the key feature of a competing events data structure: an individual who dies over follow-up (prior to dementia diagnosis) cannot subsequently develop ADRD, and since $D_{t+1}$ and $Y_{t+1}$ are events related to aging, $C_2$ represent the shared causes of both events.
+Last, the arrow from $D_{t+1}$ and $Y_{t+1}$ represents the key feature of a competing events data structure: an individual who dies over follow-up (prior to dementia diagnosis) cannot subsequently develop ADRD, and since $D_{t+1}$ and $Y_{t+1}$ are events related to aging, $C$ represent the shared causes of both events.
 
@@
 \node [right of = d] (y) {$Y_{t+1}$};
-\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
-\node [below of = d, yshift = 1cm, xshift = 1cm](cd) {$C_2$};
-\draw[->, gray!30] (p) -- (dp);
+\node [below of = d, yshift = 1cm, xshift = 1cm](cd) {$C$};
 \draw[->, gray!30] (l) -- (p);
 \draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
 \draw[->, gray!30] (dp) -- (pstar);
@@
 \draw[->] (pstar) -- (d);
-\draw[->] (p) to [out=+45,in=+135, looseness=0.3] (d);
 \draw[->] (d) -- (y);
-\draw[->, gray!30] (c) -- (dp);
-\draw[->, gray!30] (c) -- (pstar);
 \draw[->] (cd) -- (d);
 \draw[->] (cd) -- (y);
+\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
 \end{tikzpicture}
@@
 \node (p) {$P_0$};
-\node [above of = p, yshift = -1cm, xshift = 1cm](l) {$L$};
+\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
 \node [right of = p] (dp) {$D_t$};
@@
 \node [right of = d] (y) {$Y_{t+1}$};
-\node [below of = dp, yshift = 1cm, xshift = 1cm](c) {$C_1$};
-\node [below of = d, yshift = 1cm, xshift = 1cm](cd) {$C_2$};
-\draw[->] (p) -- (dp);
+\node [below of = d, yshift = 1cm, xshift = 1cm](cd) {$C$};
 \draw[->] (l) -- (p);
 \draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
 \draw[->] (dp) -- (pstar);
@@
 \draw[->] (pstar) -- (d);
-\draw[->] (p) to [out=+45,in=+135, looseness=0.3] (d);
 \draw[->] (d) -- (y);
-\draw[->] (c) -- (dp);
-\draw[->] (c) -- (pstar);
 \draw[->] (cd) -- (d);
 \draw[->] (cd) -- (y);
+\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
 \end{tikzpicture}

### 2021-05-31 04_reports/manuscript.Rmd
updates and more updates
deleted old stuff, added refs, add bibtex

@@
 ---
-title: "Dissecting the causal question underlying the association between cancer and ADRD "
+title: "Dissecting the causal question underlying the association between cancer and ADRD: an opportunity to understand sources of bias"
 author: "L. Paloma Rojas-Saunero"
@@
   chunk_output_type: console
+bibliography: cancer_dementia.bib
 ---
@@
 
-Many observational studies have consistently found that individuals with cancer have a lower risk of developing of Alzheimer´s disease or related dementias (ADRD) when compared to individuals with no history of cancer1–4.
-These findings have motivated substantial research toward mechanistic explanations, including searching for and hypothesizing that molecular and genetic mechanisms may explain this association5–12.
-These research inquiries inevitably lead to discussions of repurposing or augmenting current cancer chemotherapeutics for ADRD13.
+Many observational studies have consistently found that individuals with cancer have a lower risk of developing of Alzheimer´s disease or related dementias (ADRD) when compared to individuals with no history of cancer [@ma2014;@hanson2016;@vanderwillik2018;@ospina2020] .
+These findings have motivated substantial research toward mechanistic explanations, including searching for and hypothesizing that molecular and genetic mechanisms may explain this association[@behrens2009;@harris2014;@nudelman2019;@Papin2020;@driverbiogeront2014;@olson2019;@li2021;@driverpin12015].
+These research inquiries inevitably lead to discussions of repurposing or augmenting current cancer chemotherapeutics for ADRD [@Snyder2017].
 
 Nevertheless, inferring any treatment or mechanistic effects from the observed cancer-ADRD inverse association is not straightforward.
-Researchers have raised concerns related to the competing event of death, unmeasured confounding, and ascertainment error that could explain these results9,14.
-However, understanding these or other sources of bias first requires that we make explicit the causal question.
-Moreover, making explicit the causal question is one step toward tying a research study to a question that is relevant to decision-making.
+Researchers have raised concerns related to the competing event of death, unmeasured confounding, and ascertainment error that could explain these results[@driverbiogeront2014;@ganguli2015].
+However, understanding these or other sources of bias first requires making explicit the causal question.
+Moreover, making explicit the causal question is one step toward tying a research study to a question that is relevant to decision-making[@didelez2016;@labrecque2017].
 
-To illustrate the complexities of inferring hypothetical or available treatments’ effects on ADRD from the observed cancer-ADRD association, we focus on the   Pin1 enzyme.
-Previous animal studies have shown that the Pin1 enzyme over-expression promotes tumorigenesis, while its down-regulation is attributed to mechanisms that contribute to neurodegeneration and amyloid deposition11,12,15.
-If we one day could develop a drug that increases Pin1 expression specifically in brain tissue in hopes of preventing dementia, we could pose the question as: _What is the direct effect of this Pin1-targeting drug on the risk of ADRD over time compared to standard treatments?_
+To illustrate the complexities of inferring hypothetical or available treatments’ effects on ADRD from the observed cancer-ADRD association, we focus on a specific question conceptualizing the  Pin1 enzyme as the target of intervention.
+Previous animal studies have shown that Pin1 enzyme over-expression promotes tumorigenesis, while its down-regulation is attributed to mechanisms that contribute to neurodegeneration and amyloid deposition[@driverpin12015;@angelucci2017;@li2021].
+If we one day could develop a drug that increases Pin1 expression specifically in brain tissue in hopes of preventing dementia, we could pose the question as: _What is the effect of this Pin1-targeting drug on the risk of ADRD over time compared to standard treatments?_
 
-To explore how we might learn about this effect using real-world data on cancer and ADRD, we progressively build a causal directed acyclic graph to connect this particular causal question to the observable data and the assumptions we rely on to identify the effect.
+To explore how we might learn about this effect using real-world data on cancer and ADRD, we progressively build a causal directed acyclic graph to connect this particular causal question to the observable data and the assumptions we rely on to study the effect.
 We exemplify different scenarios with data collected from the Rotterdam Study, a population-based cohort study.
@@
 
-If this hypothetical Pin1-targeting drug was developed, the best way to understand its effect on dementia risk would be to do a well-conducted randomized trial in which we randomize eligible participants in late midlife (e.g., ages 50-60 years) to receive this drug or not, and closely monitor ADRD over a lengthy follow-up.
+If this hypothetical Pin1-targeting drug was developed, the best way to understand its effect on dementia risk would be to have a well-conducted randomized trial in which we randomize eligible participants in late midlife (e.g., ages 50-60 years) to receive this drug or not, and closely monitor ADRD over a lengthy follow-up.
+Since this drug is not currently available, at best we can use observational data on Pin1 expression measurements.
+For example, suppose that a biomarker test was available to measure Pin1 and we measured this biomarker from (stored) baseline blood samples in a population based-cohort that recruited participants in late midlife.
+Although this means we could measure the biomarker at a particular time in everyone, defined by the cohort study design, an unanswered question that would remain is related to the best time to measure the biomarker in an applied setting.
+This question is beyond the scope of this exercise and for didactic purpose we may consider that the study entrance (and subsequent data collection) date was part of a screening process.
 
-Since there is no drug currently available that targets Pin1, at best we can use observational data on Pin 1 expression measurements.
-For example, suppose that a biomarker test was available to measure Pin1 and we measured this biomarker from (stored) baseline blood samples in a population based-cohort that recruited participants in late midlife.
-Since the biomarker Pin1 is measured within an observational study, confounding can explain an observed association between it and ADRD.
-In Figure 1, we show that, Pin1 expression ($P_0$) and ADRD at time $t+1$ ($Y_{t+1}$) may share causes $L$, and that assessing the causal relationship requires adjusting for these confounders $L$.
-Previous studies have described age, sex, educational level and race/ethnicity as the minimal adjusting set of covariates (Ospina).
-However, environmental and behavioral factors such as smoking, which are known to cause microenvironmental changes such as inflammation and changes in tissue remodeling, may translate into Pin1 over-expression and are also related to the development of ADRD.
-
-For simplicity, we treat Pin1 expression as a point intervention and we fix the time-ordering of covariates (that is, we assume $L$ happens prior to $P_0$).
-In reality, it is possible that Pin1 expression changes over time and is affected by time-varying confounders (like smoking), which produces treatment-confounder feedback loops.
-Addressing such time-varying confounding would require repeated measurements of  $L$ and $P_0$16.
+In this observational study, confounding could explain the observed association between Pin1 and ADRD.
+In the directed acyclic graph [@whatif2020] of Figure 1, we present Pin1 expression as $P$ and ADRD diagnosis over time as $Y_{t+1}$.
+Both nodes may share causes $L$, and to assess the causal relationship between $P$ and $Y_{t+1}$ we would require adjusting for $L$.
+Previous studies have described age, sex, educational level and race/ethnicity as the minimal adjusting set of covariates[@ospina2020].
+However, environmental and behavioral factors such as smoking, which are known to cause microenvironmental changes such as inflammation and changes in tissue remodeling, may translate into Pin1 over-expression and are also related to the development of ADRD[ref].
 
@@
 \begin{tikzpicture}[node distance=2cm]
-\node (p) {$P_0$};
+\node (p) {$P$};
 \node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
-\node [right of = p, gray!30] (dp) {$D_t$};
-\node [right of = dp, gray!30] (pstar) {$P^*_t$};
-\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
-\node [right of = d] (y) {$Y_{t+1}$};
-\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C$};
+\node [gray!30, right of = p] (pstar) {$P^*_t$};
+\node [gray!30, right of = pstar] (dp) {$D_t$};
+\node [gray!30, right of = dp] (pstar2) {$P^*_{t+1}$};
+\node [gray!30, right of = pstar2] (dp2) {$D_{t+1}$};
+\node [right of = dp2] (y) {$Y_{t+1}$};
+\node [gray!30, below of = dp2, yshift = 1cm](cd) {$C$};
 \draw[->] (l) -- (p);
+\draw[->, gray!30] (l) to (dp);
+\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (dp2);
 \draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->, gray!30] (dp) -- (pstar);
-\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (pstar);
-\draw[->, gray!30] (pstar) -- (d);
-\draw[->, gray!30] (d) -- (y);
-\draw[->, gray!30] (cd) -- (d);
+\draw[->, gray!30] (p) to (pstar);
+\draw[->, gray!30] (pstar) -- (dp);
+\draw[->, gray!30] (dp) -- (pstar2);
+\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (pstar2);
+\draw[->, gray!30] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
+\draw[->, gray!30] (dp2) -- (y);
+\draw[->, gray!30] (pstar2) -- (dp2);
+\draw[->, gray!30] (cd) -- (dp);
+\draw[->, gray!30] (cd) -- (dp2);
 \draw[->, gray!30] (cd) -- (y);
-\draw[->, semithick] (p) to [out=45,in=135, looseness=0.2] (y);
+\draw[->, gray!30, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
 \end{tikzpicture}
@@
 
-However, unfortunately, in current studies we do not even have a single measurement of Pin1 expression, let alone repeated measurements.
-Thus, we can only rely on a proxy of this exposure.
-Since Pin1 over-expression is present in tumors, and tumors are only measured through screening and diagnosis, we considered cancer diagnosis as the proxy for Pin1 over-expression4,17–24.
-We depict this feature in Figure 2, where $P^*$ represents _incident cancer diagnosis_, the measured proxy of $P_0$.
-In this causal graph  we colored the path $P^* \leftarrow P \leftarrow Y_{t+1}$ because although we are measuring the association between $P^*$ and $Y_{t+1}$ in the observed data, we are assuming that the captured effect is only through $P_0$.
+Currently, Pin1 expression is not an available biomarker for population-based research, so instead we may have to measure a proxy to approximate this exposure.
+Because Pin1 over-expression is present in tumors, and tumors are only measured through screening and diagnosis, previous studies have considered Pin1 over-expression as the explanation behind the inverse association between cancer and ADRD, though Pin1 was not explicitly part of the research question[@ospina2020;@driver2012;@musicco2013;@freedman2016;@bowles2017;@frain2017;@schmidt2017;@sun2020;@ording2020].
+
+In a cohort setting, participants may have a cancer diagnosis at any point over follow-up.
+As such, we may consider cancer diagnosis as a time-varying process and to define cancer diagnosis as the proxy for Pin1 we may assume that Pin1 is over-expressed only by the time a person has the diagnosis, though this assumption may not represent the true biological process.
+We depict this feature in Figure 2, where $P^*_t$ and $P^*_{t+1}$ represent _cancer diagnosis_ over time, the measured proxy of $P$.
+Although we are measuring the association between cancer diagnosis over time and ADRD in the observed data, we are assuming that the captured effect is only through the pathway that involves $P$.
 
@@
 \begin{tikzpicture}[node distance=2cm]
-\node (p) {$P_0$};
-\node [above of = p, yshift = -0.8cm, xshift = -1cm, gray!30](l) {$L$};
-\node [right of = p, gray!30] (dp) {$D_t$};
-\node [right of = dp] (pstar) {$P^*_t$};
-\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
-\node [right of = d] (y) {$Y_{t+1}$};
-\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C$};
+\node (p) {$P$};
+\node [gray!30, above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
+\node [right of = p] (pstar) {$P^*_t$};
+\node [gray!30, right of = pstar] (dp) {$D_t$};
+\node [right of = dp] (pstar2) {$P^*_{t+1}$};
+\node [gray!30, right of = pstar2] (dp2) {$D_{t+1}$};
+\node [right of = dp2] (y) {$Y_{t+1}$};
+\node [gray!30, below of = dp2, yshift = 1cm](cd) {$C$};
 \draw[->, gray!30] (l) -- (p);
+\draw[->, gray!30] (l) to (dp);
+\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (dp2);
 \draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->, gray!30] (dp) -- (pstar);
-\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
-\draw[->, gray!30] (pstar) -- (d);
-\draw[->, gray!30] (d) -- (y);
-\draw[->, gray!30] (cd) -- (d);
+\draw[->] (p) to (pstar);
+\draw[->, gray!30] (pstar) -- (dp);
+\draw[->, gray!30] (dp) -- (pstar2);
+\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar2);
+\draw[->, gray!30] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
+\draw[->, gray!30] (dp2) -- (y);
+\draw[->, gray!30] (pstar2) -- (dp2);
+\draw[->, gray!30] (cd) -- (dp);
+\draw[->, gray!30] (cd) -- (dp2);
 \draw[->, gray!30] (cd) -- (y);
@@
 
-A major challenge related to assigning cancer diagnosis as the proxy of Pin1 is defining time zero.
-Not everyone who had Pin1 over-expression will be diagnosed with cancer by late-midlife, but in late-life.
-For this reason prospective cohort studies have considered cancer diagnosis as a time-varying or time-dependent exposure.
-This means that a participant within a prospective cohort study contributes to the "regulated Pin1" arm since study entry up to the time of cancer diagnosis and later on to the "Over-expressed Pin1" arm.
-Other studies have included participants with cancer diagnosis at the time of the diagnosis (for example from cancer registries19,23,25,26) and matched participants by age.
-In both cases we must remember that the main interest is on the unmeasured $P_0$ thus we should only adjust for covariates prior to $P_0$ and be careful to adjust for post-baseline covariates of $P_0$ or mediators between $P_0$ and $P^*$.
+Given the nature of an elderly cohort, cancer diagnosis will be measured in the subset of the population who survives other causes of death.
+In Figure 3 we represent death prior to cancer diagnosis as $D_t$, and the arrow between $D_t$ and $P^*_{t+1}$ represents a deterministic association such as that $P^*$ is only observed if $D_t$ is zero.
+The shared causes between $P$ and $D_t$ are also represented by $L$,  for simplicity we illustrate them with the same node, however, although some shared causes might affect $Y_{t+1}$ this representation would not restrict to those covariates only.
 
-One of the mediators between $P_0$ and $P^*$ is death prior to cancer diagnosis, represented as $D_t$ in Figure 3.
-Since individuals are at risk of dying from other causes (such as cardiovascular death), we can only measure $P^*$ in the subset of individuals who have survived long enough to have a cancer diagnosis.
-Several risk factors that increase the risk of cancer might also cause death prior to cancer diagnosis, for example smoking may cause lung cancer and chronic obstructive pulmonary disease (a leading cause of death in this age group27).
-Therefore, we can only isolate the effect of $P_0$ in $P^*$ if we condition on $C_1$ and block the backdoor pathway that is open by conditioning on $D_t$.
-In this way we assume a hypothetical scenario in which we could prevent death prior to cancer diagnosis, by conditioning on a rich set of covariates28.
-This assumption must hold regardless of whether we use incident cancer as a time-varying exposure, or match cancer patients to participants free of cancer by age.
+We assume that there is no effect of Pin1 over-expression in all-cause mortality prior to cancer diagnosis (no arrow between $P$ and $D_t$), however dying prior to cancer diagnosis may represent a selection mechanism that prevents generalizability or transportability of the effect measure to the entire target population [@hernan2017].
 
-Furthermore, we note that this is only one of the issues with considering cancer diagnosis as the proxy for Pin1 expression in terms of information bias(ref).
-We could add more complexity by considering further shared causes of cancer and ADRD diagnosis, or further mediators between Pin1 expression and cancer diagnosis, including but not limited to: screening guidelines, type of healthcare coverage, health-seeking behaviors, and healthcare availability.
+Furthermore, we could consider unmeasured and differential measurement error due to other shared causes of cancer and ADRD including but not limited to: screening guidelines, type of health care coverage, health-seeking behaviors, and health care availability.
 
-```{tikz, echo = FALSE, fig.cap = "Survival bias when treating cancer diagnosis as the proxy for Pin1 expression"}
+```{tikz, echo = FALSE, fig.cap = "Death prior to cancer diagnosis"}
 
@@
 \begin{tikzpicture}[node distance=2cm]
-\node (p) {$P_0$};
-\node [above of = p, yshift = -0.8cm, xshift = -1cm, gray!30](l) {$L$};
-\node [right of = p] (dp) {$D_t$};
-\node [right of = dp] (pstar) {$P^*_t$};
-\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
-\node [right of = d] (y) {$Y_{t+1}$};
-\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C$};
-\draw[->, gray!30] (l) -- (p);
-\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->] (dp) -- (pstar);
-\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
-\draw[->, gray!30] (pstar) -- (d);
-\draw[->] (d) -- (y);
-\draw[->, gray!30] (cd) -- (d);
+\node (p) {$P$};
+\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
+\node [right of = p] (pstar) {$P^*_t$};
+\node [right of = pstar] (dp) {$D_t$};
+\node [right of = dp] (pstar2) {$P^*_{t+1}$};
+\node [gray!30, right of = pstar2] (dp2) {$D_{t+1}$};
+\node [right of = dp2] (y) {$Y_{t+1}$};
+\node [gray!30, below of = dp2, yshift = 1cm](cd) {$C$};
+\draw[->] (l) -- (p);
+\draw[->] (l) to (dp);
+\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (dp2);
+\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
+\draw[->] (p) to (pstar);
+\draw[->, gray!30] (pstar) -- (dp);
+\draw[->] (dp) -- (pstar2);
+\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar2);
+\draw[->, gray!30] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
+\draw[->, gray!30] (dp2) -- (y);
+\draw[->, gray!30] (pstar2) -- (dp2);
+\draw[->, gray!30] (cd) -- (dp);
+\draw[->, gray!30] (cd) -- (dp2);
 \draw[->, gray!30] (cd) -- (y);
-\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->, gray!30, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
 \end{tikzpicture}
@@
 
-Up to this point we have outlined certain underlying assumptions of using incident cancer diagnosis as a proxy for Pin1 expression.
-However, as mortality increases steeply in late life, even in the setting of the ideal randomized trial of a hypothetical Pin1-targeting drug, we can only measure ADRD over follow-up in the individuals who survive long enough to have a diagnosis.
-For this reason, death is a competing event of ADRD because if a participant dies prior to ADRD diagnosis, death prevents observing ADRD at future time-points.
-As such, to identify the direct effect of $P_0$ in $Y_{t+1}$ we have to first assess if the drug has an any side effects that could increase the risk of death in any way.
-If the drug has no systemic side-effects, such as that there is no arrow between $P_0$ and $D_{t+1}$, such as observed in Figure 4.
+Even if survival prior to cancer diagnosis does not introduce bias over time, we can only measure ADRD over follow-up in the individuals who survive long enough to have an ADRD diagnosis, as represented in Figure 4.
+In this causal diagram we include one more node of death over time that comes after the exposure, represented as $D_{t+1}$, this acts as a competing event of $Y_{t+1}$ because if a participant dies prior to ADRD diagnosis, death prevents observing ADRD at future time-points.
+Furthermore, since $D_{t+1}$ and $Y_{t+1}$ are events related to aging and its consequences, $C$ represents the shared causes of both events such as cardiovascular conditions.
 
@@
 \begin{tikzpicture}[node distance=2cm]
-\node (p) {$P_0$};
-\node [above of = p, yshift = -0.8cm, xshift = -1cm, gray!30](l) {$L$};
-\node [right of = p, gray!30, gray!30] (dp) {$D_t$};
-\node [right of = dp, gray!30] (pstar) {$P^*_t$};
-\node [right of = pstar] (d) {$D_{t+1}$};
-\node [right of = d] (y) {$Y_{t+1}$};
-\node [below of = d, yshift = 1cm, xshift = 1cm,](cd) {$C$};
-\draw[->, gray!30] (l) -- (p);
-\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->, gray!30] (dp) -- (pstar);
-\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (pstar);
-\draw[->, gray!30] (pstar) -- (d);
-\draw[->] (d) -- (y);
-\draw[->] (cd) -- (d);
+\node (p) {$P$};
+\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
+\node [right of = p, gray!30] (pstar) {$P^*_t$};
+\node [right of = pstar] (dp) {$D_t$};
+\node [right of = dp, gray!30] (pstar2) {$P^*_{t+1}$};
+\node [right of = pstar2] (dp2) {$D_{t+1}$};
+\node [right of = dp2] (y) {$Y_{t+1}$};
+\node [below of = dp2, yshift = 1cm](cd) {$C$};
+\draw[->, gray!30] (pstar2) -- (dp2);
+\draw[->, gray!30] (p) to (pstar);
+\draw[->, gray!30] (pstar) -- (dp);
+\draw[->, gray!30] (dp) -- (pstar2);
+\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (pstar2);
+\draw[->, gray!30, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->] (l) -- (p);
+\draw[->] (l) to (dp);
+\draw[->] (l) to [out=0,in=90, looseness=0.4] (dp2);
+\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
+\draw[->] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
+\draw[->] (dp2) -- (y);
+\draw[->] (cd) -- (dp);
+\draw[->] (cd) -- (dp2);
 \draw[->] (cd) -- (y);
-\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
 \end{tikzpicture}
@@
 
-However, if the drug has any side-effects we have to decide if either we want to estimate a total effect (that is, the effect on ADRD even if partly mediated by the effect on death) or a direct effect (that is, the effect directly on ADRD not mediated by death) (ref).
-We face this challenge in our observational data setting, since we are measuring cancer diagnosis as the proxy for Pin1 over-expression, and cancer is a leading cause of death29.
-We can visualize this data feature in the causal diagram in Figure 5.
-In this causal diagram we include $D_{t+1}$ as a representation of death over follow-up, an arrow between $P_0$ and $D_{t+1}$ since $P_0$ may increase the mortality risk.
-We also include an arrow between $P^*_t$ and $D_{t+1}$ since cancer diagnosis, and subsequent treatment (or lack of treatment) may have an effect on death.
-Last, the arrow from $D_{t+1}$ and $Y_{t+1}$ represents the key feature of a competing events data structure: an individual who dies over follow-up (prior to dementia diagnosis) cannot subsequently develop ADRD, and since $D_{t+1}$ and $Y_{t+1}$ are events related to aging, $C$ represent the shared causes of both events.
+In the scenario where $P$ represented the targeted-drug for Pin1, and if this drug has no systemic beneficial or harmful side-effects such as that there is no arrow between $P$ and $D_{t+1}$, a total effect would quantify the effect of $P_0$ on $Y_{t+1}$ that does not include any pathway mediated through $D_{t+1}$[@young2020].
 
-In this causal diagram we observe that, had we measured and adjusted for $L$, we could estimate the total effect of $P_0$ in $Y_{t+1}$ without further assumptions.
-However the total effect includes all pathways between $P_0$ and $Y_{20}$28.
-This means that if Pin1 has an effect on mortality through cancer or other mechanisms, we will observe a protective effect of Pin1 over-expression in ADRD, partially or fully mediated through death.
-Since the question of interest is focused on the direct effect of $P_0$ in $Y_{t+1}$ as in Figure 4 (violet arrows) we need to conceptualize the different mechanisms through which $P_0$ could affect $D_{t+1}$ and $Y_{20}$.
-With this in mind we can conceive different causal questions (estimands) to represent this direct effect , such as the controlled direct effect (CDE) and the natural separable direct effect.
-In this section we discuss the controlled direct effect as the causal question of interest since it translates to frequently used methods in this literature (such as Kaplan-Meier estimator and Cox-proportional hazard model) and leave the separable direct effects question for discussion.
+However, in the context of cancer diagnosis we cannot rule out it's effect on death, as observed in Figure 5.
+In this case a total effect of $P$ in $Y_{t+1}$ would include the causal pathway mediated by the effect of cancer diagnosis in mortality, which may translate into an inverse association[29].
+To isolate the direct effect of $P$ in $Y_{t+1}$ through measurement of $P^*_t$ we have at least two alternatives.
+We can either imagine a causal question where we decompose the effect of cancer by the different mechanisms that affect dementia and death separately[@stensrud2020].
+Alternatively, we can define an scenario where death could have been prevented.
+The latter is defined as the controlled direct effect, where death is considered as a censoring event and it relies on the assumption that we have measured all $C$ to block the pathway $Y_{t+1} \leftarrow C  \rightarrow D_{t+1} \rightarrow P^*_t \rightarrow P_0$.
+Previous studies have defined death as a censoring event [ref], although failed to explicitly define how to relax the independent censoring assumption and while adjusting for confounders for dementia may represent partially the adjustment for the competing event of death [@young2020], adjusting for baseline covariates may not be sufficient.
+Furthermore interpretation regarding the prevention of death was not explicitly portrait.
 
@@
 \begin{tikzpicture}[node distance=2cm]
-\node (p) {$P_0$};
-\node [above of = p, yshift = -0.8cm, xshift = -1cm, gray!30](l) {$L$};
-\node [right of = p, gray!30, gray!30] (dp) {$D_t$};
-\node [right of = dp] (pstar) {$P^*_t$};
-\node [right of = pstar] (d) {$D_{t+1}$};
-\node [right of = d] (y) {$Y_{t+1}$};
-\node [below of = d, yshift = 1cm, xshift = 1cm](cd) {$C$};
+\node (p) {$P$};
+\node [gray!30, above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
+\node [gray!30, right of = p] (pstar) {$P^*_t$};
+\node [gray!30, right of = pstar] (dp) {$D_t$};
+\node [gray!30, right of = dp] (pstar2) {$P^*_{t+1}$};
+\node [gray!30, right of = pstar2] (dp2) {$D_{t+1}$};
+\node [gray!30, right of = dp2] (y) {$Y_{t+1}$};
+\node [gray!30, below of = dp2, yshift = 1cm](cd) {$C$};
 \draw[->, gray!30] (l) -- (p);
+\draw[->, gray!30] (l) to (dp);
+\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (dp2);
 \draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->, gray!30] (dp) -- (pstar);
-\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
-\draw[->] (pstar) -- (d);
-\draw[->] (d) -- (y);
-\draw[->] (cd) -- (d);
-\draw[->] (cd) -- (y);
-\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->, gray!30] (p) to (pstar);
+\draw[->, gray!30] (pstar) -- (dp);
+\draw[->, gray!30] (dp) -- (pstar2);
+\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (pstar2);
+\draw[->, gray!30] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
+\draw[->, gray!30] (dp2) -- (y);
+\draw[->, gray!30] (pstar2) -- (dp2);
+\draw[->, gray!30] (cd) -- (dp);
+\draw[->, gray!30] (cd) -- (dp2);
+\draw[->, gray!30] (cd) -- (y);
+\draw[->, gray!30, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
 \end{tikzpicture}
@@
 
-The CDE represents the effect of Pin1 over-expression in ADRD in a setting where we could have prevented death over the study period.
-It relies on the assumption that we have measured all $C$ to block the pathway $Y20 \leftarrow C  \rightarrow D_{15} \rightarrow P \rightarrow P*$.
-This assumption is defined as the independent censoring assumption conditional on covariates.
-In this setting death is treated as a censoring event and it can be interpreted as those who died would have the same risk of developing dementia if prevented from dying than those who remained alive and free of dementia at a given time point, conditional on shared causes of dementia and death.
+Finally, if we combine the challenges related to cancer diagnosis as a proxy for Pin1 over-expression, and having death as a competing event of ADRD we observe the complexity of the causal diagram in Figure 5.
+This As mentioned above, this is nevertheless a simplified version since we omitted the time varying nature of all nodes in the graph, and thus are ignoring how feedback loops between them further complicate interpretability and identificability.
+We also ignore further measurement error on $P^*_t$ and $Y_{t+1}$.
 
-Therefore, if we combine the challenges in section two related to cancer diagnosis as a proxy for Pin1 over-expression, and having death as a competing event of ADRD we observe the complexity of the DAG in Figure 5.
-Though this is yet a simplified version since we omitted time varying $P*$, $L$ and $C's$.
+We now turn to an application where we show how these challenges, tied to different assumptions lead to different analytic decisions, and we will show ways to mitigate or better understand them given current available data.
+Nevertheless, we acknowledge that given the large gap between $P_0$ and $P^*_t$, not only in terms of time between measurements and potential selection bias, but since cancer diagnosis represents a complex and heterogenous health condition that exceeds the representation of Pin1 expression, at best we may test the sharp null hypothesis rather than estimating the causal effect of Pin1.
 
@@
 \begin{tikzpicture}[node distance=2cm]
-\node (p) {$P_0$};
-\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
-\node [right of = p] (dp) {$D_t$};
-\node [right of = dp] (pstar) {$P^*_t$};
-\node [right of = pstar] (d) {$D_{t+1}$};
-\node [right of = d] (y) {$Y_{t+1}$};
-\node [below of = d, yshift = 1cm, xshift = 1cm](cd) {$C$};
-\draw[->] (l) -- (p);
-\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->] (dp) -- (pstar);
-\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
-\draw[->] (pstar) -- (d);
-\draw[->] (d) -- (y);
-\draw[->] (cd) -- (d);
-\draw[->] (cd) -- (y);
-\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
+\node (p) {$P$};
+\node [gray!30, above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
+\node [gray!30, right of = p] (pstar) {$P^*_t$};
+\node [gray!30, right of = pstar] (dp) {$D_t$};
+\node [gray!30, right of = dp] (pstar2) {$P^*_{t+1}$};
+\node [gray!30, right of = pstar2] (dp2) {$D_{t+1}$};
+\node [gray!30, right of = dp2] (y) {$Y_{t+1}$};
+\node [gray!30, below of = dp2, yshift = 1cm](cd) {$C$};
+\draw[->, gray!30] (l) -- (p);
+\draw[->, gray!30] (l) to (dp);
+\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (dp2);
+\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
+\draw[->, gray!30] (p) to (pstar);
+\draw[->, gray!30] (pstar) -- (dp);
+\draw[->, gray!30] (dp) -- (pstar2);
+\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (pstar2);
+\draw[->, gray!30] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
+\draw[->, gray!30] (dp2) -- (y);
+\draw[->, gray!30] (pstar2) -- (dp2);
+\draw[->, gray!30] (cd) -- (dp);
+\draw[->, gray!30] (cd) -- (dp2);
+\draw[->, gray!30] (cd) -- (y);
+\draw[->, gray!30, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
 \end{tikzpicture}

### 2021-06-03 04_reports/manuscript.Rmd
presentantion melodem plus updated dags in manuscript finally yooohooo

@@
 library(kableExtra)
-# source(here::here("02_R", "03_descriptives.R"))
-# source(here::here("02_R", "05_plots.R"))
-# source(here::here("02_R", "05b_plot_causes_death.R"))
+source(here::here("02_R", "03_descriptives.R"))
+source(here::here("02_R", "05_plots.R"))
+source(here::here("02_R", "05b_plot_causes_death.R"))
 # table_results <- rio::import(here::here("02_R", "table_results.csv"))
@@
 
-Many observational studies have consistently found that individuals with cancer have a lower risk of developing of Alzheimer´s disease or related dementias (ADRD) when compared to individuals with no history of cancer [@ma2014;@hanson2016;@vanderwillik2018;@ospina2020] .
+Many observational studies have consistently found that individuals with cancer have a lower risk of developing of Alzheimer´s disease or related dementias (ADRD) when compared to individuals with no history of cancer [@ma2014;@hanson2016;@vanderwillik2018;@ospina2020].
 These findings have motivated substantial research toward mechanistic explanations, including searching for and hypothesizing that molecular and genetic mechanisms may explain this association[@behrens2009;@harris2014;@nudelman2019;@Papin2020;@driverbiogeront2014;@olson2019;@li2021;@driverpin12015].
@@
 Moreover, making explicit the causal question is one step toward tying a research study to a question that is relevant to decision-making[@didelez2016;@labrecque2017].
-
 To illustrate the complexities of inferring hypothetical or available treatments’ effects on ADRD from the observed cancer-ADRD association, we focus on a specific question conceptualizing the  Pin1 enzyme as the target of intervention.
@@
 We describe the challenges and how they translate into the analytic decisions.
-Last, we discuss how information on mortality and cause of death can provides insight about the direction of some sources of bias.
+Last, we discuss how information on mortality and cause of death can provide insight about the direction of some sources of bias.
 
@@
 For example, suppose that a biomarker test was available to measure Pin1 and we measured this biomarker from (stored) baseline blood samples in a population based-cohort that recruited participants in late midlife.
-Although this means we could measure the biomarker at a particular time in everyone, defined by the cohort study design, an unanswered question that would remain is related to the best time to measure the biomarker in an applied setting.
-This question is beyond the scope of this exercise and for didactic purpose we may consider that the study entrance (and subsequent data collection) date was part of a screening process.
 
-In this observational study, confounding could explain the observed association between Pin1 and ADRD.
-In the directed acyclic graph [@whatif2020] of Figure 1, we present Pin1 expression as $P$ and ADRD diagnosis over time as $Y_{t+1}$.
-Both nodes may share causes $L$, and to assess the causal relationship between $P$ and $Y_{t+1}$ we would require adjusting for $L$.
+In the observational setting, confounding could explain the observed association between Pin1 and ADRD.
+In the directed acyclic graph [@whatif2020] of Figure 1, we present Pin1 expression measured at time $t$ as $P_{t-1}$ and ADRD diagnosis over time as $Y_{t+1}$.
+Both nodes may share causes $L$, and to assess the causal relationship between $P_{t-1}$ and $Y_{t+1}$ we would require adjusting for $L$.
 Previous studies have described age, sex, educational level and race/ethnicity as the minimal adjusting set of covariates[@ospina2020].
@@
 \begin{tikzpicture}[node distance=2cm]
-\node (p) {$P$};
+\node (p) {$P_{t-1}$};
 \node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
-\node [gray!30, right of = p] (pstar) {$P^*_t$};
-\node [gray!30, right of = pstar] (dp) {$D_t$};
-\node [gray!30, right of = dp] (pstar2) {$P^*_{t+1}$};
-\node [gray!30, right of = pstar2] (dp2) {$D_{t+1}$};
+\node [gray!30, below of = p, yshift = 1cm] (pstar) {$P^*_{t-1}$};
+\node [gray!30, right of = p] (dp) {$D_t$};
+\node [gray!30, right of = dp] (p2) {$P_t$};
+\node [gray!30, below of = p2, yshift = 1cm] (pstar2) {$P^*_t$};
+\node [gray!30, right of = p2] (dp2) {$D_{t+1}$};
 \node [right of = dp2] (y) {$Y_{t+1}$};
-\node [gray!30, below of = dp2, yshift = 1cm](cd) {$C$};
+\node [gray!30, below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$C$};
+\draw[->, gray!30] (p) to (pstar);
+\draw[->, gray!30] (p2) to (pstar2);
 \draw[->] (l) -- (p);
+\draw[->, gray!30] (l) -- (p2);
 \draw[->, gray!30] (l) to (dp);
@@
 \draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->, gray!30] (p) to (pstar);
 \draw[->, gray!30] (pstar) -- (dp);
-\draw[->, gray!30] (dp) -- (pstar2);
-\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (pstar2);
+\draw[->, gray!30] (pstar2) -- (dp2);
 \draw[->, gray!30] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
+\draw[->, gray!30] (dp) -- (p2);
 \draw[->, gray!30] (dp2) -- (y);
-\draw[->, gray!30] (pstar2) -- (dp2);
-\draw[->, gray!30] (cd) -- (dp);
 \draw[->, gray!30] (cd) -- (dp2);
 \draw[->, gray!30] (cd) -- (y);
-\draw[->, gray!30, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->, gray!30, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
 \end{tikzpicture}
@@
 
-Currently, Pin1 expression is not an available biomarker for population-based research, so instead we may have to measure a proxy to approximate this exposure.
-Because Pin1 over-expression is present in tumors, and tumors are only measured through screening and diagnosis, previous studies have considered Pin1 over-expression as the explanation behind the inverse association between cancer and ADRD, though Pin1 was not explicitly part of the research question[@ospina2020;@driver2012;@musicco2013;@freedman2016;@bowles2017;@frain2017;@schmidt2017;@sun2020;@ording2020].
-
-In a cohort setting, participants may have a cancer diagnosis at any point over follow-up.
-As such, we may consider cancer diagnosis as a time-varying process and to define cancer diagnosis as the proxy for Pin1 we may assume that Pin1 is over-expressed only by the time a person has the diagnosis, though this assumption may not represent the true biological process.
-We depict this feature in Figure 2, where $P^*_t$ and $P^*_{t+1}$ represent _cancer diagnosis_ over time, the measured proxy of $P$.
-Although we are measuring the association between cancer diagnosis over time and ADRD in the observed data, we are assuming that the captured effect is only through the pathway that involves $P$.
+Currently, Pin1 expression is not an available biomarker for population-based research, so at best we can only rely on a proxy for it.
+Because Pin1 over-expression is present in tumors, and tumors are only measured through diagnosis, previous studies have considered Pin1 over-expression as the explanation behind the inverse association between cancer diagnosis and ADRD, though Pin1 was not explicitly part of the research question[@ospina2020;@driver2012;@musicco2013;@freedman2016;@bowles2017;@frain2017;@schmidt2017;@sun2020;@ording2020].
+Unlike measuring Pin1 at the same time to all participants (thought this would not necessarily mean this would be the ideal time to measure it, we discuss this point further in the discussion section), cancer diagnosis is a time-varying process.
+We depict this feature in Figure 2, where $P^*_{t-1}$ and $P^*_t$ represent _cancer diagnosis_ over time, the measured proxy of $P_{t-1}$ and $P_t$ respectively.
+Although this means we would measure the association between cancer diagnosis over time and ADRD in the observed data, we are assuming that the captured effect is only through the pathway that involves Pin1 expression over time.
 
@@
 \begin{tikzpicture}[node distance=2cm]
-\node (p) {$P$};
-\node [gray!30, above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
-\node [right of = p] (pstar) {$P^*_t$};
-\node [gray!30, right of = pstar] (dp) {$D_t$};
-\node [right of = dp] (pstar2) {$P^*_{t+1}$};
-\node [gray!30, right of = pstar2] (dp2) {$D_{t+1}$};
+\node (p) {$P_{t-1}$};
+\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
+\node [below of = p, yshift = 1cm] (pstar) {$P^*_{t-1}$};
+\node [gray!30, right of = p] (dp) {$D_t$};
+\node [right of = dp] (p2) {$P_t$};
+\node [below of = p2, yshift = 1cm] (pstar2) {$P^*_t$};
+\node [gray!30, right of = p2] (dp2) {$D_{t+1}$};
 \node [right of = dp2] (y) {$Y_{t+1}$};
-\node [gray!30, below of = dp2, yshift = 1cm](cd) {$C$};
-\draw[->, gray!30] (l) -- (p);
+\node [gray!30, below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$C$};
+\draw[->] (p) to (pstar);
+\draw[->] (p2) to (pstar2);
+\draw[->] (l) -- (p);
+\draw[->] (l) -- (p2);
+\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
 \draw[->, gray!30] (l) to (dp);
 \draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (dp2);
-\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->] (p) to (pstar);
+\draw[->, gray!30] (dp) -- (p2);
 \draw[->, gray!30] (pstar) -- (dp);
-\draw[->, gray!30] (dp) -- (pstar2);
-\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar2);
+\draw[->, gray!30] (pstar2) -- (dp2);
 \draw[->, gray!30] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
 \draw[->, gray!30] (dp2) -- (y);
-\draw[->, gray!30] (pstar2) -- (dp2);
-\draw[->, gray!30] (cd) -- (dp);
 \draw[->, gray!30] (cd) -- (dp2);
 \draw[->, gray!30] (cd) -- (y);
+\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
+\end{tikzpicture}
+```
+
+A challenge that arises by defining cancer diagnosis as the proxy for Pin1 is defining the time zero, the time where follow-up and screening for ADRD would begin after having the exposure measured.
+The eligibility criteria to join the study may not align with the time cancer diagnosis is measured.
+This situation may introduce inmortal-time bias [@hernanInmortal2016].
+For example, a study performed using data from the Framingham study [@driver2012] defined the exposed group with cancer as those participants with prevalent or incident cancer diagnosis (alternatively defined as "ever cancer" [@hanson2016]).
+This meant that a participant who had cancer diagnosis over follow-up contributed all their person-time to the cancer arm, including the time prior to the cancer diagnosis.
+By defining the exposure after the time of inclusion to the cohort, only participants who remain alive and free of ADRD diagnosis over follow-up can be defined as "ever cancer" [@anderson1983, @hernanInmortal2016].
+
+This problem is avoided by recognizing the time-varying nature of cancer diagnosis.
+Several studies have considered cancer diagnosis as a time-dependent exposure.
+In this way, the time prior to cancer diagnosis is allocated to the non-exposed arm, and the time after cancer diagnosis to the exposed arm [@white2013; @hanson2016; @bowles2017].
+The price we pay with this approximation is that implicitly, this means that Pin1 would overexpress at the time of cancer diagnosis, which is biologically implausible.
+Moreover, cancer diagnosis will only be measured in the subset of participants who are alive over follow-up.
+Thus we included death prior to cancer diagnosis as $D_t$ and an arrow between $D_t$ and $P_t$ that represents a deterministic association such as that $P_t$ is only observed if $D_t$ is zero in Figure 3.
+In addition, we added an arrow between $L$ and $D_t$, since covariates such as smoking may affect Pin1 over-expression but also affect the risk of death due to other causes such as COPD.
+
+```{tikz, echo = FALSE, fig.cap = "Time-varying cancer diagnosis and inmortal-time bias"}
+\usetikzlibrary{arrows}
+\begin{tikzpicture}[node distance=2cm]
+\node (p) {$P_{t-1}$};
+\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
+\node [below of = p, yshift = 1cm] (pstar) {$P^*_{t-1}$};
+\node [right of = p] (dp) {$D_t$};
+\node [right of = dp] (p2) {$P_t$};
+\node [below of = p2, yshift = 1cm] (pstar2) {$P^*_t$};
+\node [gray!30, right of = p2] (dp2) {$D_{t+1}$};
+\node [right of = dp2] (y) {$Y_{t+1}$};
+\node [gray!30, below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$C$};
+\draw[->] (p) to (pstar);
+\draw[->] (p2) to (pstar2);
+\draw[->] (l) -- (p);
+\draw[->] (l) -- (p2);
+\draw[->] (l) to (dp);
+\draw[->] (dp) -- (p2);
+\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (dp2);
+\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
+\draw[->, gray!30] (pstar) -- (dp);
+\draw[->, gray!30] (pstar2) -- (dp2);
+\draw[->, gray!30] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
+\draw[->, gray!30] (dp2) -- (y);
+\draw[->, gray!30] (cd) -- (dp2);
+\draw[->, gray!30] (cd) -- (y);
+\draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
 \draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
@@
 
-Given the nature of an elderly cohort, cancer diagnosis will be measured in the subset of the population who survives other causes of death.
-In Figure 3 we represent death prior to cancer diagnosis as $D_t$, and the arrow between $D_t$ and $P^*_{t+1}$ represents a deterministic association such as that $P^*$ is only observed if $D_t$ is zero.
-The shared causes between $P$ and $D_t$ are also represented by $L$,  for simplicity we illustrate them with the same node, however, although some shared causes might affect $Y_{t+1}$ this representation would not restrict to those covariates only.
+We now move to the challenge of death as a competing event for ADRD, represented in Figure 5.
+For interpretability we exclude the time varying process of cancer diagnosis and focus on Pin1 (and cancer diagnosis) as it had been measured in all participants at $t$.
+Given that some participants may die prior to having a cancer diagnosis over follow-up, we can only have ADRD over follow-up in the individuals who survive long enough to have an ADRD diagnosis.
+In the causal diagram of Figure 4 we include a node for death after the exposure $P^*t$ has been measured, represented as $D_{t+1}$, this acts as a competing event of $Y_{t+1}$ because if a participant dies prior to ADRD diagnosis, death prevents observing ADRD at future time-points.
+Furthermore, since $D_{t+1}$ and $Y_{t+1}$ are events related to aging and its consequences, $C$ represents the shared causes of both events such as cardiovascular conditions.
+We also include an arrow between $L$ and $D_{t+1}$ following the argument discussed previously.
 
-We assume that there is no effect of Pin1 over-expression in all-cause mortality prior to cancer diagnosis (no arrow between $P$ and $D_t$), however dying prior to cancer diagnosis may represent a selection mechanism that prevents generalizability or transportability of the effect measure to the entire target population [@hernan2017].
+In the scenario where $P_t$ represented the targeted-drug for Pin1, and if this drug had no systemic beneficial or harmful side-effects such as that there is no arrow between $P$ and $D_{t+1}$, a total effect would quantify the effect of $P_t$ on $Y_{t+1}$ that does not include any pathway mediated through $D_{t+1}$[@young2020].
+However, in the context of cancer diagnosis as the proxy for Pin1 over-expression, we cannot rule out the effect of cancer diagnosis on death, represented as the arrow between $P^*_t$ and  $D_{t+1}$.
+In this setting, a total effect of $P_t$ in $Y_{t+1}$ would include the causal pathway mediated by the effect of cancer diagnosis in mortality, which may translate into an inverse association[@young2020].
 
-Furthermore, we could consider unmeasured and differential measurement error due to other shared causes of cancer and ADRD including but not limited to: screening guidelines, type of health care coverage, health-seeking behaviors, and health care availability.
+To isolate the direct effect of $P_t$ in $Y_{t+1}$ through measurement of $P^*_t$ we have at least two alternatives of causal questions we can ask.
+We can either imagine a causal question where we decompose the effect of cancer by the different mechanisms that affect dementia and death separately[@stensrud2020].
+Alternatively, we can define an scenario where death could have been prevented.
+The latter is defined as the controlled direct effect, where death is considered as a censoring event and it relies on the assumption that we have measured all $C$ to block the pathway $Y_{t+1} \leftarrow C  \rightarrow D_{t+1} \rightarrow P^*_t \rightarrow P_t$.
+Previous studies have defined death as a censoring event [ref], although failed to explicitly define how to relax the independent censoring assumption.
+Moreover, adjusting for confounders for dementia may represent partially the adjustment for the competing event of death [@young2020], adjusting for baseline covariates may not be sufficient.
 
-```{tikz, echo = FALSE, fig.cap = "Death prior to cancer diagnosis"}
-
+```{tikz, echo = FALSE, fig.cap = "Death as a competing event"}
 \usetikzlibrary{arrows}
 \begin{tikzpicture}[node distance=2cm]
-\node (p) {$P$};
+\node [gray!30](p) {$P_{t-1}$};
 \node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
-\node [right of = p] (pstar) {$P^*_t$};
-\node [right of = pstar] (dp) {$D_t$};
-\node [right of = dp] (pstar2) {$P^*_{t+1}$};
-\node [gray!30, right of = pstar2] (dp2) {$D_{t+1}$};
+\node [gray!30, below of = p, yshift = 1cm] (pstar) {$P^*_{t-1}$};
+\node [gray!30, right of = p] (dp) {$D_t$};
+\node [right of = dp] (p2) {$P_t$};
+\node [below of = p2, yshift = 1cm] (pstar2) {$P^*_t$};
+\node [right of = p2] (dp2) {$D_{t+1}$};
 \node [right of = dp2] (y) {$Y_{t+1}$};
-\node [gray!30, below of = dp2, yshift = 1cm](cd) {$C$};
-\draw[->] (l) -- (p);
-\draw[->] (l) to (dp);
-\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (dp2);
+\node [below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$C$};
+\draw[->, gray!30] (p) to (pstar);
+\draw[->, gray!30] (l) -- (p);
+\draw[->, gray!30] (l) to (dp);
+\draw[->, gray!30] (pstar) -- (dp);
+\draw[->, gray!30] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
+\draw[->, gray!30] (dp) -- (p2);
+\draw[->, semithick, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->] (p2) to (pstar2);
+\draw[->] (dp2) -- (y);
+\draw[->] (cd) -- (dp2);
+\draw[->] (cd) -- (y);
+\draw[->] (l) -- (p2);
+\draw[->] (l) to [out=0,in=90, looseness=0.4] (dp2);
 \draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->] (p) to (pstar);
-\draw[->, gray!30] (pstar) -- (dp);
-\draw[->] (dp) -- (pstar2);
-\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar2);
-\draw[->, gray!30] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
-\draw[->, gray!30] (dp2) -- (y);
-\draw[->, gray!30] (pstar2) -- (dp2);
-\draw[->, gray!30] (cd) -- (dp);
-\draw[->, gray!30] (cd) -- (dp2);
-\draw[->, gray!30] (cd) -- (y);
-\draw[->, gray!30, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->] (pstar2) -- (dp2);
+\draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
 \end{tikzpicture}
@@
 
-Even if survival prior to cancer diagnosis does not introduce bias over time, we can only measure ADRD over follow-up in the individuals who survive long enough to have an ADRD diagnosis, as represented in Figure 4.
-In this causal diagram we include one more node of death over time that comes after the exposure, represented as $D_{t+1}$, this acts as a competing event of $Y_{t+1}$ because if a participant dies prior to ADRD diagnosis, death prevents observing ADRD at future time-points.
-Furthermore, since $D_{t+1}$ and $Y_{t+1}$ are events related to aging and its consequences, $C$ represents the shared causes of both events such as cardiovascular conditions.
+To summarize, the complexity of the causal structure that describes the effect of Pin1 through the proxy of cancer diagnosis in the risk of ADRD illustrate the potential sources of bias, as observed in Figure 6.
+Even so, this is a simplified version since we omitted the time varying nature of all nodes in the graph and other sources of measurement error, and thus are ignoring how feedback loops between them further complicate interpretability and identificability.
 
-```{tikz, echo = FALSE, fig.cap = "Death as a competing event for dementia over follow-up"}
+```{tikz, echo = FALSE, fig.cap = "Direct effect of Pin1 in the risk of ADRD"}
 \usetikzlibrary{arrows}
 \begin{tikzpicture}[node distance=2cm]
-\node (p) {$P$};
+\node (p) {$P_{t-1}$};
 \node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
-\node [right of = p, gray!30] (pstar) {$P^*_t$};
-\node [right of = pstar] (dp) {$D_t$};
-\node [right of = dp, gray!30] (pstar2) {$P^*_{t+1}$};
-\node [right of = pstar2] (dp2) {$D_{t+1}$};
+\node [below of = p, yshift = 1cm] (pstar) {$P^*_{t-1}$};
+\node [right of = p] (dp) {$D_t$};
+\node [right of = dp] (p2) {$P_t$};
+\node [below of = p2, yshift = 1cm] (pstar2) {$P^*_t$};
+\node [right of = p2] (dp2) {$D_{t+1}$};
 \node [right of = dp2] (y) {$Y_{t+1}$};
-\node [below of = dp2, yshift = 1cm](cd) {$C$};
-\draw[->, gray!30] (pstar2) -- (dp2);
-\draw[->, gray!30] (p) to (pstar);
-\draw[->, gray!30] (pstar) -- (dp);
-\draw[->, gray!30] (dp) -- (pstar2);
-\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (pstar2);
-\draw[->, gray!30, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
+\node [below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$C$};
+\draw[->] (p) to (pstar);
 \draw[->] (l) -- (p);
 \draw[->] (l) to (dp);
+\draw[->] (pstar) -- (dp);
+\draw[->] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
+\draw[->] (dp) -- (p2);
+\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->] (p2) to (pstar2);
+\draw[->] (dp2) -- (y);
+\draw[->] (cd) -- (dp2);
+\draw[->] (cd) -- (y);
+\draw[->] (l) -- (p2);
 \draw[->] (l) to [out=0,in=90, looseness=0.4] (dp2);
 \draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
-\draw[->] (dp2) -- (y);
-\draw[->] (cd) -- (dp);
-\draw[->] (cd) -- (dp2);
-\draw[->] (cd) -- (y);
+\draw[->] (pstar2) -- (dp2);
+\draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
 \end{tikzpicture}
@@
 
-In the scenario where $P$ represented the targeted-drug for Pin1, and if this drug has no systemic beneficial or harmful side-effects such as that there is no arrow between $P$ and $D_{t+1}$, a total effect would quantify the effect of $P_0$ on $Y_{t+1}$ that does not include any pathway mediated through $D_{t+1}$[@young2020].
+We now turn to an application where we show how these challenges translate in analytic decisions, and we will show ways to mitigate or better understand to the best of the available data's abilities.
+Nevertheless, we acknowledge that cancer diagnosis represents a complex and heterogenous health condition that exceeds the representation of Pin1 expression, thus, at best we may test the sharp null hypothesis rather than estimating the causal effect of Pin1.
+In this section we conduct an analysis of the cancer-ADRD association that is structured to the best of the available data's abilities to inform the possible effect of Pin1 on all cause-dementia.
 
-However, in the context of cancer diagnosis we cannot rule out it's effect on death, as observed in Figure 5.
-In this case a total effect of $P$ in $Y_{t+1}$ would include the causal pathway mediated by the effect of cancer diagnosis in mortality, which may translate into an inverse association[29].
-To isolate the direct effect of $P$ in $Y_{t+1}$ through measurement of $P^*_t$ we have at least two alternatives.
-We can either imagine a causal question where we decompose the effect of cancer by the different mechanisms that affect dementia and death separately[@stensrud2020].
-Alternatively, we can define an scenario where death could have been prevented.
-The latter is defined as the controlled direct effect, where death is considered as a censoring event and it relies on the assumption that we have measured all $C$ to block the pathway $Y_{t+1} \leftarrow C  \rightarrow D_{t+1} \rightarrow P^*_t \rightarrow P_0$.
-Previous studies have defined death as a censoring event [ref], although failed to explicitly define how to relax the independent censoring assumption and while adjusting for confounders for dementia may represent partially the adjustment for the competing event of death [@young2020], adjusting for baseline covariates may not be sufficient.
-Furthermore interpretation regarding the prevention of death was not explicitly portrait.
+## 4. Application to the Rotterdam Study
 
-```{tikz, echo = FALSE, fig.cap = "Direct effect"}
-\usetikzlibrary{arrows}
-\begin{tikzpicture}[node distance=2cm]
-\node (p) {$P$};
-\node [gray!30, above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
-\node [gray!30, right of = p] (pstar) {$P^*_t$};
-\node [gray!30, right of = pstar] (dp) {$D_t$};
-\node [gray!30, right of = dp] (pstar2) {$P^*_{t+1}$};
-\node [gray!30, right of = pstar2] (dp2) {$D_{t+1}$};
-\node [gray!30, right of = dp2] (y) {$Y_{t+1}$};
-\node [gray!30, below of = dp2, yshift = 1cm](cd) {$C$};
-\draw[->, gray!30] (l) -- (p);
-\draw[->, gray!30] (l) to (dp);
-\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (dp2);
-\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->, gray!30] (p) to (pstar);
-\draw[->, gray!30] (pstar) -- (dp);
-\draw[->, gray!30] (dp) -- (pstar2);
-\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (pstar2);
-\draw[->, gray!30] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
-\draw[->, gray!30] (dp2) -- (y);
-\draw[->, gray!30] (pstar2) -- (dp2);
-\draw[->, gray!30] (cd) -- (dp);
-\draw[->, gray!30] (cd) -- (dp2);
-\draw[->, gray!30] (cd) -- (y);
-\draw[->, gray!30, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
-\end{tikzpicture}
-```
+We use data collected in the Rotterdam Study, a population-based prospective cohort study among persons living in the Ommoord district in Rotterdam, the Netherlands.
+Recruitment and initial assessments were held between 1990 and 1993; it was later extended between 2000 and 2001 consisting of individuals who had reached the age of 55 years or who had moved into the study area.
+Participants from first subcohort had follow-up visits between 1993-1995, 1997-1999, 2002-2005, and 2008-2010, second subcohort had follow-up visits between 2004 and 2005, and between 2011 and 2012[@ikram2020].
+All participants had data on incident cancer diagnosis and incident dementia diagnosis through follow-up, collected from medical records of general practitioners (including hospital discharge letters) and through linkage with national registries.
+Date and cause of death was collected on a weekly basis via municipal population registries.
+Data on clinical outcomes was available until 2015.
 
-Finally, if we combine the challenges related to cancer diagnosis as a proxy for Pin1 over-expression, and having death as a competing event of ADRD we observe the complexity of the causal diagram in Figure 5.
-This As mentioned above, this is nevertheless a simplified version since we omitted the time varying nature of all nodes in the graph, and thus are ignoring how feedback loops between them further complicate interpretability and identificability.
-We also ignore further measurement error on $P^*_t$ and $Y_{t+1}$.
+We considered as inclusion criteria being between 60 and 70 years old at study entry, without history of cancer diagnosis, free of cognitive decline or with previous history of dementia.
+Out of 10998 persons who participated at study entry, `r total_n` were considered eligible.
+Time to cancer diagnosis, time to dementia diagnosis and death status was measured for all participants.
+All participants were followed from study entry until dementia diagnosis, death or 20 years after their individual baseline date, whichever occurred first.
 
-We now turn to an application where we show how these challenges, tied to different assumptions lead to different analytic decisions, and we will show ways to mitigate or better understand them given current available data.
-Nevertheless, we acknowledge that given the large gap between $P_0$ and $P^*_t$, not only in terms of time between measurements and potential selection bias, but since cancer diagnosis represents a complex and heterogenous health condition that exceeds the representation of Pin1 expression, at best we may test the sharp null hypothesis rather than estimating the causal effect of Pin1.
+### 4.1. Methods
 
-```{tikz, echo = FALSE, fig.cap = "Direct effect"}
-\usetikzlibrary{arrows}
-\begin{tikzpicture}[node distance=2cm]
-\node (p) {$P$};
-\node [gray!30, above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
-\node [gray!30, right of = p] (pstar) {$P^*_t$};
-\node [gray!30, right of = pstar] (dp) {$D_t$};
-\node [gray!30, right of = dp] (pstar2) {$P^*_{t+1}$};
-\node [gray!30, right of = pstar2] (dp2) {$D_{t+1}$};
-\node [gray!30, right of = dp2] (y) {$Y_{t+1}$};
-\node [gray!30, below of = dp2, yshift = 1cm](cd) {$C$};
-\draw[->, gray!30] (l) -- (p);
-\draw[->, gray!30] (l) to (dp);
-\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (dp2);
-\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->, gray!30] (p) to (pstar);
-\draw[->, gray!30] (pstar) -- (dp);
-\draw[->, gray!30] (dp) -- (pstar2);
-\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (pstar2);
-\draw[->, gray!30] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
-\draw[->, gray!30] (dp2) -- (y);
-\draw[->, gray!30] (pstar2) -- (dp2);
-\draw[->, gray!30] (cd) -- (dp);
-\draw[->, gray!30] (cd) -- (dp2);
-\draw[->, gray!30] (cd) -- (y);
-\draw[->, gray!30, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
-\end{tikzpicture}
-```
+We illustrate the association between cancer and dementia diagnosis under two scenarios.
+For the first scenario (Model A) we defined cancer proxy as _"cancer ever vs.
+never"_.This scenario mimics the definition used in previous studies where cancer was defined as a time-independent exposure[@driver2012; @nudelman2014].
+We continue with a second scenario (Model B) where cancer diagnosis is a time-varying exposure.
+The definition represents the setting where all participants are free of cancer at inclusion and over time some participants will have a cancer diagnosis.
+
+To address confounding we fit inverse probability treatment weights, stabilized and truncated at 99th percentile.
+Weights for model A were defined as the inverse of the probability of cancer diagnosis conditional on baseline covariates such as age at study entry, sex, educational attainment, cohort, smoking status.
+In contrast, for Model B weights recreate a pseudopopulation in which time-varying covariates do not predict the risk of cancer diagnosis at time _t_ [@hernan2000].
+Time varying covariates included: ...
+Further details on modeling specifications and weights assessment are presented as **Supplementary material x**.
+
+To estimate the controlled direct effect of Pin1 in the risk of ADRD, that is in the setting where death could have been prevented, we fit inverse probability censoring weights for death.
+
+in time-varying settings we compared the complement of a weighted Kaplan-Meier survival estimator in participants with incident cancer vs.
+no incident of cancer with time indexed in years.
+The weights in this case are time-varying by follow-up year, defined as a product of the time-fixed IPT weights above and a year-specific inverse probability of censoring (IPC) by death weights.
+For an individual still alive in year t, the time t IPC weight is the product of the inverse probability of surviving in each year prior to t, conditional on measured common causes of death and dementia (that is, variables such as C in Figure 4).
+For an individual who has died by time t, the year t IPC weight is zero.
+We estimated survival probabilities using a logistic regression model for death as a function of baseline and time-varying covariates.
+Baseline covariates included age at study entry, sex, apoe4 status, educational attainment and the time-varying covariates smoking status, systolic blood pressure, BMI and prevalent and incident comorbidities such as: cancer, heart disease, stroke and diabetes.
+We additionally calculated the controlled direct effect considering death as an unconditional independent censoring event (as if there were no arrows from $C$ to $D_{15}$ and $Y_{20}$) for an illustrative purpose.
+
+Estimates of the controlled direct effect at 20 years of follow-up are presented as risk differences (RD), risk ratios (RR) and hazard ratios (HR).
+We note that hazards, unlike risks, inherently condition on surviving both dementia and death, as such they will not have a causal interpretation in this case.
+We present them for comparison against risk ratios.
+
+Since the conditional independent censoring assumption is untestable, we compute Peterson upper and lower bounds to represent: 1) the extreme scenario of independence, that refers to an scenario were those who died would never develop dementia (lower bound) and 2) complete dependency, that refers to an scenario where those who died would have a dementia prior to death (upper bound).
+The lower bound is calculated with the Aalen-Johansen estimator treating death as a competing event, and the upper bound is calculated with the Kaplan Meier estimator for the combined outcome of dementia or death.
+
+We additionally present the causes of death for participants who developed cancer and for participants free of cancer diagnosis over follow-up.
+This information provides insights about the association between cancer diagnosis and death and about the potential misclassification of individuals with no cancer diagnosis (cancer as a cause of death).
+
+All 95% confidence intervals were calculated using percentile-based bootstrapping based on 500 bootstrap samples.
+All analysis were performed using R, code is provided in supplementary material and available in https://github.com/palolili23/2021_cancer_dementia.

### 2021-06-03 06_presentations/melodem2021/index.Rmd
presentantion melodem plus updated dags in manuscript finally yooohooo

--- 2021-04-28 06_presentations/ser2021/index.Rmd
@@
 ---
-title: "Toward a clearer understanding of the inverse association between cancer and dementia"
-subtitle: ""  
+title: "Dissecting the causal question underlying the association between cancer and ADRD"
+subtitle: "An opportunity to understand sources of bias"  
 author: 
@@
 source(here::here("02_R", "05_plots.R"))
+source(here::here("02_R", "05b_plot_causes_death.R"))
 hr_results <- rio::import(here::here("02_R", "hr_results.csv"))
@@
 
-background-image: url(./figs/ser.jpg), url(./figs/erasmus.png)
-background-position: 5% 95%, 95% 5%
+background-image: url(./figs/erasmus.png)
+background-position: 5% 95%
 background-size: 25%, 15%
@@
 
-<br>
-
-## Towards a clearer understanding of the inverse association between cancer and dementia
+## Dissecting the causal question underlying the association between cancer and dementia
 
@@
 
-.right[
+.right[L. Paloma Rojas-Saunero, MD
 
-L. Paloma Rojas-Saunero, MD
+**Epidemiology department**]
 
-**Epidemiology department**
+---
 
-`r icons::fontawesome$brands$twitter` @palolili23
+## Introduction
 
+.pull-left[
+
+Observational studies have found a protective effect of cancer diagnosis in the risk of Alzheimer´s disease or related dementias (ADRD).
+
+These studies lead to discussions of repurposing or augmenting current cancer chemotherapeutics for ADRD
+]
+
+--
+
+.pull-right[ 
+![](figs/forest_plot.PNG)
+]
+
+.footnote[Ospina-Romero M. JAMA Network Open. 2020]
+
+---
+
+## Cancer diagnosis as an exposure
+
+```{r, fig.align='center', out.width= "40%"}
+include_graphics("figs/causal_risks.PNG")
+
+```
+.footnote[L. Frain et al. Alzheimer’s & Dementia. 2017]
+
+---
+
+.pull-left[
+![](figs/pin_knockout.PNG)
+Driver J. et al. (2015)
+]
+
+--
+
+.pull-right[
+
+![](figs/linna2021.PNG)
+<br>
+Lanni C. et al. (2021)
 ]
@@
 ---
-### Introduction
 
-<br><br>
+class: middle, center
 
-- Observational studies have found an inverse association between cancer and dementia risk.
+## How can identify a causal pathways like Pin1 with real-world data?
+
+---
+
+## Aims
+
+- To build a causal directed acyclic graph (DAG) which help us connect the causal question to real-world data on cancer and ADRD and spell out the assumptions we need to identify the effect.
 
@@
 
-- Several genetic and molecular mechanisms have been proposed.
+- To describe the challenges and how they translate into the analytic decisions.
 
@@
 
-- Researchers have raised concerns related to measurement error, unmeasured confounding and selection bias.
+- To exemplify this process in an application using data from the Rotterdam Study.
 
 ---
-
-### Introduction
+### Pin1 targeted-drug and ADRD diagnosis
 
@@
 \begin{tikzpicture}[node distance=2cm]
-\node (p) {$P_0$};
-\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
-\node [right of = p, gray!30] (dp) {$D_t$};
-\node [right of = dp, gray!30] (pstar) {$P^*_t$};
-\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
-\node [right of = d] (y) {$Y_{t+1}$};
-\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
-\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
-\draw[->, gray!30] (p) -- (dp);
-\draw[->, gray!30] (l) -- (p);
-\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
+\node (p) {$P$};
+\node [gray!20,above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
+\node [gray!20, below of = p, yshift = 1cm] (pstar) {$P^*_{t-1}$};
+\node [gray!20, right of = p] (dp) {$D_t$};
+\node [gray!20, right of = dp] (p2) {$P_t$};
+\node [gray!20, below of = p2, yshift = 1cm] (pstar2) {$P^*_t$};
+\node [gray!20, right of = p2] (dp2) {$D_{t+1}$};
+\node [right of = dp2] (y) {$Y_{t+1}$};
+\node [gray!20, below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$C$};
+\draw[->, gray!20] (p) to (pstar);
+\draw[->, gray!20] (p2) to (pstar2);
+\draw[->, gray!20] (l) -- (p);
+\draw[->, gray!20] (l) -- (p2);
+\draw[->, gray!20] (l) to (dp);
+\draw[->, gray!20] (l) to [out=0,in=90, looseness=0.4] (dp2);
+\draw[->, gray!20] (l) to [out=0,in=90, looseness=0.4] (y);
+\draw[->, gray!20] (pstar) -- (dp);
+\draw[->, gray!20] (pstar2) -- (dp2);
+\draw[->, gray!20] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
+\draw[->, gray!20] (dp) -- (p2);
+\draw[->, gray!20] (dp2) -- (y);
+\draw[->, gray!20] (cd) -- (dp2);
+\draw[->, gray!20] (cd) -- (y);
+\draw[->, gray!20, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
 \draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
-\draw[->, gray!30] (dp) -- (pstar);
-\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (pstar);
-\draw[->, gray!30] (pstar) -- (d);
-\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (d);
-\draw[->, gray!30] (d) -- (y);
-\draw[->, gray!30] (c) -- (dp);
-\draw[->, gray!30] (c) -- (pstar);
-\draw[->, gray!30] (cd) -- (d);
-\draw[->, gray!30] (cd) -- (y);
 \end{tikzpicture}
@@
 
-$P_0$ = Pin1 targeted-drug; $Y_{t+1}$ = Dementia diagnosis
+$P$ = Pin1 targeted-drug; $Y_{t+1}$ = Dementia diagnosis
+
+???
+In a setting where this hypothetical Pin1-targeting drug was developed, the best way to understand its effect on dementia risk would be to have a well-conducted randomized trial in which we randomize eligible participants in late midlife (e.g., ages 50-60 years) to receive this drug or not, and closely monitor ADRD over a lengthy follow-up.
 
 ---
-### Pin1 expression as biomarker (Problem 1: _Confounding_)
+### Pin1 expression as biomarker
 
@@
 \begin{tikzpicture}[node distance=2cm]
-\node (p) {$P_0$};
-\node [above of = p, yshift = -1cm, xshift = 1cm](l) {$L$};
-\node [right of = p, gray!30] (dp) {$D_t$};
-\node [right of = dp, gray!30] (pstar) {$P^*_t$};
-\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
-\node [right of = d] (y) {$Y_{t+1}$};
-\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
-\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
-\draw[->, gray!30] (p) -- (dp);
+\node (p) {$P_{t-1}$};
+\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
+\node [gray!20, below of = p, yshift = 1cm] (pstar) {$P^*_{t-1}$};
+\node [gray!20, right of = p] (dp) {$D_t$};
+\node [gray!20, right of = dp] (p2) {$P_t$};
+\node [gray!20, below of = p2, yshift = 1cm] (pstar2) {$P^*_t$};
+\node [gray!20, right of = p2] (dp2) {$D_{t+1}$};
+\node [right of = dp2] (y) {$Y_{t+1}$};
+\node [gray!20, below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$C$};
+\draw[->, gray!20] (p) to (pstar);
+\draw[->, gray!20] (p2) to (pstar2);
 \draw[->] (l) -- (p);
+\draw[->, gray!20] (l) -- (p2);
+\draw[->, gray!20] (l) to (dp);
+\draw[->, gray!20] (l) to [out=0,in=90, looseness=0.4] (dp2);
 \draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->, semithick] (p) to [out=45,in=135, looseness=0.3] (y);
-\draw[->, gray!30] (dp) -- (pstar);
-\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (pstar);
-\draw[->, gray!30] (pstar) -- (d);
-\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (d);
-\draw[->, gray!30] (d) -- (y);
-\draw[->, gray!30] (c) -- (dp);
-\draw[->, gray!30] (c) -- (pstar);
-\draw[->, gray!30] (cd) -- (d);
-\draw[->, gray!30] (cd) -- (y);
+\draw[->, gray!20] (pstar) -- (dp);
+\draw[->, gray!20] (pstar2) -- (dp2);
+\draw[->, gray!20] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
+\draw[->, gray!20] (dp) -- (p2);
+\draw[->, gray!20] (dp2) -- (y);
+\draw[->, gray!20] (cd) -- (dp2);
+\draw[->, gray!20] (cd) -- (y);
+\draw[->, gray!20, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
 \end{tikzpicture}
@@
 
-$P_0$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis; $L$ = Shared causes of $P_0$ and $Y_{t+1}$
+$P_{t-1}$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis; $L$ = Shared causes of $P_{t-1}$ and $Y_{t+1}$
 
 ???
-We might be closer to developing biomarker for Pin1 expression in the future.
-So we can imagine a situation in which Pin1 exposure is measured in a population based cohort study, using stored blood samples.
-Since this would be a observational setting, Pin1 expression and ADRD at time may share causes such as smoking, and that assessing the causal relationship requires adjusting for these confounders.
+Since this drug is not currently available, at best we can use observational data on Pin1 expression measurements.
+For example, suppose that a biomarker test was available to measure Pin1 and we measured this biomarker from (stored) baseline blood samples in a population based-cohort that recruited participants in late midlife.
+In the observational setting, confounding could explain the observed association between Pin1 and ADRD.
 
 ---
-### Cancer diagnosis as the proxy of Pin1 (Problem 2: _Time zero_)
+### Cancer diagnosis as the proxy of Pin1
 
@@
 \begin{tikzpicture}[node distance=2cm]
-\node (p) {$P_0$};
-\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
-\node [right of = p, gray!30] (dp) {$D_t$};
-\node [right of = dp] (pstar) {$P^*_t$};
-\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
-\node [right of = d] (y) {$Y_{t+1}$};
-\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
-\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
-\draw[->, gray!30] (p) -- (dp);
-\draw[->, gray!30] (l) -- (p);
-\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
+\node (p) {$P_{t-1}$};
+\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
+\node [below of = p, yshift = 1cm] (pstar) {$P^*_{t-1}$};
+\node [gray!20, right of = p] (dp) {$D_t$};
+\node [right of = dp] (p2) {$P_t$};
+\node [below of = p2, yshift = 1cm] (pstar2) {$P^*_t$};
+\node [gray!20, right of = p2] (dp2) {$D_{t+1}$};
+\node [right of = dp2] (y) {$Y_{t+1}$};
+\node [gray!20, below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$C$};
+\draw[->] (p) to (pstar);
+\draw[->] (p2) to (pstar2);
+\draw[->] (l) -- (p);
+\draw[->] (l) -- (p2);
+\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
+\draw[->, gray!20] (l) to (dp);
+\draw[->, gray!20] (l) to [out=0,in=90, looseness=0.4] (dp2);
+\draw[->, gray!20] (dp) -- (p2);
+\draw[->, gray!20] (pstar) -- (dp);
+\draw[->, gray!20] (pstar2) -- (dp2);
+\draw[->, gray!20] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
+\draw[->, gray!20] (dp2) -- (y);
+\draw[->, gray!20] (cd) -- (dp2);
+\draw[->, gray!20] (cd) -- (y);
 \draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
-\draw[->, gray!30] (dp) -- (pstar);
-\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
-\draw[->, gray!30] (pstar) -- (d);
-\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (d);
-\draw[->, gray!30] (d) -- (y);
-\draw[->, gray!30] (c) -- (dp);
-\draw[->, gray!30] (c) -- (pstar);
-\draw[->, gray!30] (cd) -- (d);
-\draw[->, gray!30] (cd) -- (y);
+\draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
 \end{tikzpicture}
@@
 
-$P_0$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis; $P^*_t$ = Cancer diagnosis
+$P_{t-1}$ and $P_{t}$= Pin1 over-expression; $P^*_{t-1}$ and $P^*_t$ = Cancer diagnosis; $Y_{t+1}$ = Dementia diagnosis; $L$ = Shared causes;
 
 ???
-At this moment this biomarker is not available, but we know that Pin1 over-expression stimulates tumor growth and for this reason we use cancer diagnosis as the proxy for Pin1.
+But Pin1 is not an available biomarker for population-based research, so at best we can only rely on a proxy for it.
+Because Pin1 over-expression is present in tumors, and tumors are only measured through diagnosis, we may use cancer as a proxy.
+However, unlike measuring Pin1 at the same time to all participants cancer diagnosis is a time-varying process.
+Although this means we would measure the association between cancer diagnosis over time and ADRD in the observed data, we are assuming that the captured effect is only through the pathway that involves Pin1 expression over time.
 
 ---
-### Death prior to cancer diagnosis (Problem 3: _Survival bias_)
+### Time-varying cancer diagnosis and inmortal-time bias
 
@@
 \begin{tikzpicture}[node distance=2cm]
-\node (p) {$P_0$};
-\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
+\node (p) {$P_{t-1}$};
+\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
+\node [below of = p, yshift = 1cm] (pstar) {$P^*_{t-1}$};
 \node [right of = p] (dp) {$D_t$};
-\node [right of = dp] (pstar) {$P^*_t$};
-\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
-\node [right of = d] (y) {$Y_{t+1}$};
-\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
-\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
-\draw[->] (p) -- (dp);
-\draw[->, gray!30] (l) -- (p);
-\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
+\node [right of = dp] (p2) {$P_t$};
+\node [below of = p2, yshift = 1cm] (pstar2) {$P^*_t$};
+\node [gray!20, right of = p2] (dp2) {$D_{t+1}$};
+\node [right of = dp2] (y) {$Y_{t+1}$};
+\node [gray!20, below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$C$};
+\draw[->] (p) to (pstar);
+\draw[->] (p2) to (pstar2);
+\draw[->] (l) -- (p);
+\draw[->] (l) -- (p2);
+\draw[->] (l) to (dp);
+\draw[->] (dp) -- (p2);
+\draw[->, gray!20] (l) to [out=0,in=90, looseness=0.4] (dp2);
+\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
+\draw[->, gray!20] (pstar) -- (dp);
+\draw[->, gray!20] (pstar2) -- (dp2);
+\draw[->, gray!20] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
+\draw[->, gray!20] (dp2) -- (y);
+\draw[->, gray!20] (cd) -- (dp2);
+\draw[->, gray!20] (cd) -- (y);
+\draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
 \draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
-\draw[->] (dp) -- (pstar);
-\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
-\draw[->, gray!30] (pstar) -- (d);
-\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (d);
-\draw[->, gray!30] (d) -- (y);
-\draw[->, gray!30] (c) -- (dp);
-\draw[->, gray!30] (c) -- (pstar);
-\draw[->, gray!30] (cd) -- (d);
-\draw[->, gray!30] (cd) -- (y);
 \end{tikzpicture}
@@
 
-$P_0$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis; $P^*_t$ = Cancer diagnosis; $D_t$ = Death prior to cancer diagnosis
+$P_{t-1}$ and $P_{t}$= Pin1 over-expression; $P^*_{t-1}$ and $P^*_t$ = Cancer diagnosis; $Y_{t+1}$ = Dementia diagnosis; $L$ = Shared causes; $D_t$ = Death prior to $P_{t}$
+
 ???
+A challenge that arises by defining cancer diagnosis as the proxy for Pin1 is defining the time zero, the time where follow-up and screening for ADRD would begin after having the exposure measured.
+The eligibility criteria to join the study may not align with the time cancer diagnosis is measured.
+This situation may introduce inmortal-time bias.
+A participant needs to survive to have a cancer diagnosis over time.
+
+This problem is avoided by recognizing the time-varying nature of cancer diagnosis.
+And previous studies have defined cancer as a tv or time dependant exposure, though only adjusted for baseline covariates.
+
+The price we pay with this approximation is that implicitly, this means that Pin1 would over-express at the time of cancer diagnosis, which is biologically implausible.
 
 ---
-### Death prior to cancer diagnosis (Problem 3: _Survival bias_)
+### Death as a competing event of ADRD
+
+--
+```{tikz, echo = FALSE}
+\usetikzlibrary{arrows}
+\begin{tikzpicture}[node distance=2cm]
+\node [gray!20](p) {$P_{t-1}$};
+\node [gray!20, above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
+\node [gray!20, below of = p, yshift = 1cm] (pstar) {$P^*_{t-1}$};
+\node [gray!20, right of = p] (dp) {$D_t$};
+\node [right of = dp] (p2) {$P_t$};
+\node [gray!20, below of = p2, yshift = 1cm] (pstar2) {$P^*_t$};
+\node [right of = p2] (dp2) {$D_{t+1}$};
+\node [right of = dp2] (y) {$Y_{t+1}$};
+\node [below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$C$};
+\draw[->, gray!20] (p) to (pstar);
+\draw[->, gray!20] (l) -- (p);
+\draw[->, gray!20] (l) to (dp);
+\draw[->, gray!20] (pstar) -- (dp);
+\draw[->, gray!20] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
+\draw[->, gray!20] (dp) -- (p2);
+\draw[->, semithick, gray!20] (p) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->, gray!20] (p2) to (pstar2);
+\draw[->] (dp2) -- (y);
+\draw[->] (cd) -- (dp2);
+\draw[->] (cd) -- (y);
+\draw[->, gray!20] (l) -- (p2);
+\draw[->, gray!20] (l) to [out=0,in=90, looseness=0.4] (dp2);
+\draw[->, gray!20] (l) to [out=0,in=90, looseness=0.4] (y);
+\draw[->, gray!20] (pstar2) -- (dp2);
+\draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
+\end{tikzpicture}
+```
+
+$P_{t}$= Pin1 targeted drug; $D_{t+1}$ = Death prior to $Y_{t+1}$; $Y_{t+1}$ = Dementia diagnosis; $C$ = shared causes
+
+???
+Let's step back for a minute and imagine that we had indeed randomized participants to pin1 drug.
+Even in the context of an RCT, death will act as a competing event since it prevents dementia at future time-points. Talk about the total effect.
+---
+###  Death as a competing event of ADRD
 
@@
 \begin{tikzpicture}[node distance=2cm]
-\node (p) {$P_0$};
-\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
-\node [right of = p] (dp) {$D_t$};
-\node [right of = dp] (pstar) {$P^*_t$};
-\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
-\node [right of = d] (y) {$Y_{t+1}$};
-\node [below of = dp, yshift = 1cm, xshift = 1cm](c) {$C_1$};
-\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
-\draw[->] (p) -- (dp);
-\draw[->, gray!30] (l) -- (p);
-\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
-\draw[->] (dp) -- (pstar);
-\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
-\draw[->, gray!30] (pstar) -- (d);
-\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (d);
-\draw[->, gray!30] (d) -- (y);
-\draw[->] (c) -- (dp);
-\draw[->] (c) -- (pstar);
-\draw[->, gray!30] (cd) -- (d);
-\draw[->, gray!30] (cd) -- (y);
+\node [gray!20](p) {$P_{t-1}$};
+\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
+\node [gray!20, below of = p, yshift = 1cm] (pstar) {$P^*_{t-1}$};
+\node [gray!20, right of = p] (dp) {$D_t$};
+\node [right of = dp] (p2) {$P_t$};
+\node [below of = p2, yshift = 1cm] (pstar2) {$P^*_t$};
+\node [right of = p2] (dp2) {$D_{t+1}$};
+\node [right of = dp2] (y) {$Y_{t+1}$};
+\node [below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$C$};
+\draw[->, gray!20] (p) to (pstar);
+\draw[->, gray!20] (l) -- (p);
+\draw[->, gray!20] (l) to (dp);
+\draw[->, gray!20] (pstar) -- (dp);
+\draw[->, gray!20] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
+\draw[->, gray!20] (dp) -- (p2);
+\draw[->, semithick, gray!20] (p) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->] (p2) to (pstar2);
+\draw[->] (dp2) -- (y);
+\draw[->] (cd) -- (dp2);
+\draw[->] (cd) -- (y);
+\draw[->] (l) -- (p2);
+\draw[->] (l) to [out=0,in=90, looseness=0.4] (dp2);
+\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
+\draw[->] (pstar2) -- (dp2);
+\draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
 \end{tikzpicture}
@@
 
-$P_0$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis; $P^*_t$ = Cancer diagnosis; $D_t$ = Death prior to cancer diagnosis; $C_1$ = Shared causes of $D_t$ and $P^*$
 ???
+Talk about the issue with total effect when we look at the cancer dx.
+Talk about separable effects and the CDE
 
 ---
-### Death prior to dementia diagnosis (Problem 3: _Survival bias_)
+### Causal structure for the effect of Pin1 in ADRD using cancer diagnosis as a proxy for Pin1
 
@@
 \begin{tikzpicture}[node distance=2cm]
-\node (p) {$P_0$};
-\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
-\node [right of = p, gray!30, gray!30] (dp) {$D_t$};
-\node [right of = dp, gray!30] (pstar) {$P^*_t$};
-\node [right of = pstar] (d) {$D_{t+1}$};
-\node [right of = d] (y) {$Y_{t+1}$};
-\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
-\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
-\draw[->, gray!30] (p) -- (dp);
-\draw[->, gray!30] (l) -- (p);
-\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
+\node (p) {$P_{t-1}$};
+\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
+\node [below of = p, yshift = 1cm] (pstar) {$P^*_{t-1}$};
+\node [right of = p] (dp) {$D_t$};
+\node [right of = dp] (p2) {$P_t$};
+\node [below of = p2, yshift = 1cm] (pstar2) {$P^*_t$};
+\node [right of = p2] (dp2) {$D_{t+1}$};
+\node [right of = dp2] (y) {$Y_{t+1}$};
+\node [below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$C$};
+\draw[->] (p) to (pstar);
+\draw[->] (l) -- (p);
+\draw[->] (l) to (dp);
+\draw[->] (pstar) -- (dp);
+\draw[->] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
+\draw[->] (dp) -- (p2);
 \draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
-\draw[->, gray!30] (dp) -- (pstar);
-\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (pstar);
-\draw[->, gray!30] (pstar) -- (d);
-\draw[->] (p) to [out=+45,in=+135, looseness=0.3] (d);
-\draw[->] (d) -- (y);
-\draw[->, gray!30] (c) -- (dp);
-\draw[->, gray!30] (c) -- (pstar);
-\draw[->, gray!30] (cd) -- (d);
-\draw[->, gray!30] (cd) -- (y);
+\draw[->] (p2) to (pstar2);
+\draw[->] (dp2) -- (y);
+\draw[->] (cd) -- (dp2);
+\draw[->] (cd) -- (y);
+\draw[->] (l) -- (p2);
+\draw[->] (l) to [out=0,in=90, looseness=0.4] (dp2);
+\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
+\draw[->] (pstar2) -- (dp2);
+\draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
 \end{tikzpicture}
@@
 
-$P_0$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis; $D_{t+1}$ = Death prior to dementia diagnosis
+---
+### And there is more we could add...
 
----
-### Death prior to dementia diagnosis (Problem 3: _Survival bias_)
+<br>
+.pull-left[
 
-```{tikz, echo = FALSE}
-\usetikzlibrary{arrows}
-\begin{tikzpicture}[node distance=2cm]
-\node (p) {$P_0$};
-\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
-\node [right of = p, gray!30, gray!30] (dp) {$D_t$};
-\node [right of = dp, gray!30] (pstar) {$P^*_t$};
-\node [right of = pstar] (d) {$D_{t+1}$};
-\node [right of = d] (y) {$Y_{t+1}$};
-\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
-\node [below of = d, yshift = 1cm, xshift = 1cm](cd) {$C_2$};
-\draw[->, gray!30] (p) -- (dp);
-\draw[->, gray!30] (l) -- (p);
-\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
-\draw[->, gray!30] (dp) -- (pstar);
-\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (pstar);
-\draw[->, gray!30] (pstar) -- (d);
-\draw[->] (p) to [out=+45,in=+135, looseness=0.3] (d);
-\draw[->] (d) -- (y);
-\draw[->, gray!30] (c) -- (dp);
-\draw[->, gray!30] (c) -- (pstar);
-\draw[->] (cd) -- (d);
-\draw[->] (cd) -- (y);
-\end{tikzpicture}
-```
+![](https://media.giphy.com/media/fVeAI9dyD5ssIFyOyM/giphy.gif)
+]
 
-$P_0$ = Pin1 over-expression; $Y_{t+1}$ = Dementia diagnosis; $D_{t+1}$ = Death prior to dementia diagnosis; $C_2$ as shared causes of $D_{t+1}$ and $Y_{t+1}$
+--
+
+<br>
+
+- Time-varying confounding loops
+
+- Error terms for cancer diagnosis
+
+- Measurement and error term for ADRD
+
+- Differential or dependent error structures
 
@@
 
-- Participants between 60 and 70 years old at study entry, free of cancer and dementia diagnosis at baseline
+- Participants between 60 and 70 years old at study entry
+
+- Without history of cancer diagnosis and dementia at study entry
 
@@
 
-- Data on cancer diagnosis, dementia diagnosis and death was collected from national registries and clinical records.
+- Cancer diagnosis and dementia diagnosis is measured through follow-up and linkage with national registries.
+
+- Date and cause of death was collected on a weekly basis via municipal population registries
+
+- Data on covariates is measured repeatedly over visits and through linkage to EHR
+
+---
+### Methods
+
+We considered 2 scenarios to define Pin1 over-expression:
 
@@
 
-- Cancer:
+- "Cancer ever vs. never" to replicate previous studies (such as Driver 2012)
 
-  + _Time-fixed ever vs. never_
+--
 
-  + _Time-varying_
+- Pin1 and cancer diagnosis as a time-varying exposure
+
+--
+
+20 years of follow-up or up to time of death/ADRD
 
 ---
-### Application to the Rotterdam Study
+### Methods
 
@@
 
-  + *Age, sex, Apoe4, education*
+  + Time-varying cancer diagnosis: weights represent the product of the inverse probability of being diagnosed with cancer over time conditional on time-varying covariate history.
 
@@
 
-- Inverse probability censoring weights for death (IPCW)
+- Inverse probability of censoring weights for death (IPCW)*
 
-  + *Age, sex, Apoe4, education and time-varying covariates: smoking, systolic blood pressure, BMI and incident cancer, heart disease, stroke and diabetes*
+  + For an individual still alive in year _t_, weights represent the the product of the inverse probability of surviving in each year prior to _t_, conditional on measured common causes of death and dementia.
 
+.footnote[Rojas-Saunero et al. Choosing questions before methods in dementia research with competing events and causal goals.MedRxiv.2021]
+
+---
+
+### Methods
+
+- Risk of dementia and risk ratios are estimated with a weighted Kaplan-Meier estimator.
 --
 
-- Risk of dementia and risk ratios are estimated with a weighted Kaplan-Meier estimator.
+- Assessed extreme scenarios of independence and fit upper and lower bounds:
 
+  + Total independence: those who died would never develop dementia (lower bound). Aalen-Johanssen estimator treating death as a competing event.
+
+  + Complete dependency: those who died would have dementia prior to death (upper bound). The KM estimator for a composite outcome
 --
@@
 
----
-### Bounds
-
-<br><br>
-
-- **1) Lower Bound**: Total independence, that refers to an scenario were those who died would never develop dementia (Death as competing event)
-
---
-
-- **2) Upper Bound**: Complete dependency, that refers to an scenario where those who died would have dementia prior to death (Composite outcome, death + dementia)
+Peterson AV. Bounds for a joint distribution function with fixed sub distribution functions: Application to competing risks. 1976
 
@@
 
+???
+First I will show all results related to cancer as ever vs. never.
+In this graph I present three situations, unadjusted, after weighting for confounding, and after weighting for confounding and censoring for death.
+In each case we have risk ratios and hazard ratios.
+We observe the inverse association in these two settings, and after including weights for censoring it shifts towards above the null.
+Lets recall that the model with censoring weights is our best effort to achieve the conditional independence, but if we observe the bounds that represent extreme scenarios, these bounds cross the null.
+
 --
@@
 
+???
+When we define cancer as a time-varying exposure, we observe a shift towards the null prior to adjustment. An estimates and confidence shift above one by including the weights.
+And again, in our best scenario to achieve conditional independence, if we look at the bounds of extreme independency and dependency, they are far apart in both sides of the null.
+
+Please note that in all cases, we would interpret this results as if death could have been prevented.
+
 ---
-### Conclusions
+### Causes of death
 
-- Defining the underlying mechanism behind cancer diagnosis helps in multiple way.
+```{r echo = FALSE, fig.retina = 3, fig.width = 14}
+causes_death + 
+      theme_xaringan(text_font_size = 20,
+                       title_font_size = 20) +
+     theme(legend.position = "right")
+```
 
-  + Helps identify sources of bias and guides analysis
+---
+### Thoughts so far...
 
-  + Opens conversations about future data collection
+- Being explicit about the causal mechanism is useful:
+
+  + Helps identify sources of bias and guides analysis.
+
+  + Opens conversations about future data collection.
 
@@
 
-- Since death acts as a competing event, we first need to define the estimand of interest.
+- We need to define the estimand of interest in respect to death.
 
-  + The direct effect relies on the independent censoring assumption
+  + The direct effect relies on the independent censoring assumption.
 
@@
 
-  + The separable effects provides a clearer intuition to address this question.
+  + The separable effects can provide a clearer intuition
+
+  + % of people who died in each arm should be always presented.
 
 ---
-class: middle
+### Gracias!   Thank you!
 
-### Acknowledgments
+.pull-left[
+![](https://media.giphy.com/media/XHVmD4RyXgSjd8aUMb/giphy.gif)
+]
 
-- Sonja A. Swanson
+--
+<br>
 
-- Kimberly van der Willik
-
-- Arfan M. Ikram
-
----
-class:middle
-
-### Gracias!   Thank you!
+.pull-right[
 
@@
 `r icons::fontawesome$brands$github` @palolili23
+
+]

### 2021-06-06 04_reports/manuscript.Rmd
changed the time-varying approach including tv-covs to the t2cancer proxy

@@
 title: "Dissecting the causal question underlying the association between cancer and ADRD: an opportunity to understand sources of bias"
-author: "L. Paloma Rojas-Saunero"
+author: ""
 date: ""
-output: [html_document]
+output: [pdf_document]
+always_allow_html: true
 editor_options: 
@@
 library(magrittr)
-library(kableExtra)
+library(flextable)
 source(here::here("02_R", "03_descriptives.R"))
+source(here::here("02_R", "table_2.R")) 
+load(here::here("03_figs", "plots.RData")) 
 source(here::here("02_R", "05_plots.R"))
-source(here::here("02_R", "05b_plot_causes_death.R"))
-# table_results <- rio::import(here::here("02_R", "table_results.csv"))
-# table_results_bounds <- rio::import(here::here("02_R", "table_results_bounds.csv"))
-# models <- read_csv(here::here("04_reports", "models.csv")) %>% select(-6)
 ```
@@
 Moreover, making explicit the causal question is one step toward tying a research study to a question that is relevant to decision-making[@didelez2016;@labrecque2017].
+
 To illustrate the complexities of inferring hypothetical or available treatments’ effects on ADRD from the observed cancer-ADRD association, we focus on a specific question conceptualizing the  Pin1 enzyme as the target of intervention.
@@
 Because Pin1 over-expression is present in tumors, and tumors are only measured through diagnosis, previous studies have considered Pin1 over-expression as the explanation behind the inverse association between cancer diagnosis and ADRD, though Pin1 was not explicitly part of the research question[@ospina2020;@driver2012;@musicco2013;@freedman2016;@bowles2017;@frain2017;@schmidt2017;@sun2020;@ording2020].
-Unlike measuring Pin1 at the same time to all participants (thought this would not necessarily mean this would be the ideal time to measure it, we discuss this point further in the discussion section), cancer diagnosis is a time-varying process.
+Unlike measuring Pin1 at the same time to all participants (thought this would not necessarily mean this would be the ideal time to measure it, we discuss this point further in the discussion section), cancer diagnosis is a collected over time.
 We depict this feature in Figure 2, where $P^*_{t-1}$ and $P^*_t$ represent _cancer diagnosis_ over time, the measured proxy of $P_{t-1}$ and $P_t$ respectively.
@@
 In this way, the time prior to cancer diagnosis is allocated to the non-exposed arm, and the time after cancer diagnosis to the exposed arm [@white2013; @hanson2016; @bowles2017].
-The price we pay with this approximation is that implicitly, this means that Pin1 would overexpress at the time of cancer diagnosis, which is biologically implausible.
+The price we pay with this approximation is that implicitly, this means that Pin1 would over-express at the time of cancer diagnosis, which is biologically implausible.
 Moreover, cancer diagnosis will only be measured in the subset of participants who are alive over follow-up.
@@
 In addition, we added an arrow between $L$ and $D_t$, since covariates such as smoking may affect Pin1 over-expression but also affect the risk of death due to other causes such as COPD.
+Although $L$ is time-varying in nature, we only depict $L$ at one time-point for readability.
 
@@
 
-We now move to the challenge of death as a competing event for ADRD, represented in Figure 5.
-For interpretability we exclude the time varying process of cancer diagnosis and focus on Pin1 (and cancer diagnosis) as it had been measured in all participants at $t$.
+We now move to the challenge of death as a competing event for ADRD, represented in Figure 4.
+For interpretability we exclude the time-varying process of cancer diagnosis and focus on Pin1 (and cancer diagnosis) as it had been measured in all participants at $t$.
 Given that some participants may die prior to having a cancer diagnosis over follow-up, we can only have ADRD over follow-up in the individuals who survive long enough to have an ADRD diagnosis.
-In the causal diagram of Figure 4 we include a node for death after the exposure $P^*t$ has been measured, represented as $D_{t+1}$, this acts as a competing event of $Y_{t+1}$ because if a participant dies prior to ADRD diagnosis, death prevents observing ADRD at future time-points.
+In the causal diagram of Figure 4 we include a node for death after the exposure $P^*t$ has been measured, represented as $D_{t+1}$, this acts as a competing event of $Y_{t+1}$ because if a participant dies by $t + 1$, the participant cannot subsequently develop dementia.
 Furthermore, since $D_{t+1}$ and $Y_{t+1}$ are events related to aging and its consequences, $C$ represents the shared causes of both events such as cardiovascular conditions.
@@
 
-In the scenario where $P_t$ represented the targeted-drug for Pin1, and if this drug had no systemic beneficial or harmful side-effects such as that there is no arrow between $P$ and $D_{t+1}$, a total effect would quantify the effect of $P_t$ on $Y_{t+1}$ that does not include any pathway mediated through $D_{t+1}$[@young2020].
+In the setting where $P_t$ represented the targeted-drug for Pin1, and if this drug had no systemic beneficial or harmful side-effects such as that there is no arrow between $P$ and $D_{t+1}$, a total effect would quantify the effect of $P_t$ on $Y_{t+1}$ that does not include any pathway mediated through $D_{t+1}$[@young2020].
 However, in the context of cancer diagnosis as the proxy for Pin1 over-expression, we cannot rule out the effect of cancer diagnosis on death, represented as the arrow between $P^*_t$ and  $D_{t+1}$.
@@
 The latter is defined as the controlled direct effect, where death is considered as a censoring event and it relies on the assumption that we have measured all $C$ to block the pathway $Y_{t+1} \leftarrow C  \rightarrow D_{t+1} \rightarrow P^*_t \rightarrow P_t$.
-Previous studies have defined death as a censoring event [ref], although failed to explicitly define how to relax the independent censoring assumption.
-Moreover, adjusting for confounders for dementia may represent partially the adjustment for the competing event of death [@young2020], adjusting for baseline covariates may not be sufficient.
+Previous studies have defined death as a censoring event [], although failed to explicitly define how to comply with the independent censoring assumption.
+Moreover, adjusting for confounders for dementia (such as only adjusting for $L$) may represent partially the adjustment for the competing event of death [@geskus], to block the pathway mediated by $C$ we need to consider covariates that change over time and that affect both nodes, such as cardiometabolic conditions.
 
@@
 
-To summarize, the complexity of the causal structure that describes the effect of Pin1 through the proxy of cancer diagnosis in the risk of ADRD illustrate the potential sources of bias, as observed in Figure 6.
-Even so, this is a simplified version since we omitted the time varying nature of all nodes in the graph and other sources of measurement error, and thus are ignoring how feedback loops between them further complicate interpretability and identificability.
+To summarize, the complexity of the causal structure that describes the effect of Pin1 through the proxy of cancer diagnosis in the risk of ADRD illustrate the potential sources of bias, as observed in Figure 5.
+Even so, this is a simplified version since we omitted other sources of measurement error [@whatif2020] and the time-varying nature of all nodes and feedback loops between them which would further complicate interpretability and identificability.
 
@@
 
-## 4. Application to the Rotterdam Study
+## 3. Application to the Rotterdam Study
 
@@
 
-### 4.1. Methods
+### 3.1. Methods
 
 We illustrate the association between cancer and dementia diagnosis under two scenarios.
-For the first scenario (Model A) we defined cancer proxy as _"cancer ever vs.
-never"_.This scenario mimics the definition used in previous studies where cancer was defined as a time-independent exposure[@driver2012; @nudelman2014].
-We continue with a second scenario (Model B) where cancer diagnosis is a time-varying exposure.
-The definition represents the setting where all participants are free of cancer at inclusion and over time some participants will have a cancer diagnosis.
-
+Scenario A replicates the setting that defines cancer proxy as _"cancer ever vs.
+never"_ [@driver2012].
+Scenario B defines cancer diagnosis is a time-varying exposure.
 To address confounding we fit inverse probability treatment weights, stabilized and truncated at 99th percentile.
-Weights for model A were defined as the inverse of the probability of cancer diagnosis conditional on baseline covariates such as age at study entry, sex, educational attainment, cohort, smoking status.
-In contrast, for Model B weights recreate a pseudopopulation in which time-varying covariates do not predict the risk of cancer diagnosis at time _t_ [@hernan2000].
-Time varying covariates included: ...
+In scenario A, weights were defined as the inverse of the probability of cancer diagnosis conditional on baseline covariates such as age at study entry, sex, educational attainment, cohort, smoking status.
+In contrast, for scenario B weights were defined to represent the product of the inverse probability of being diagnosed with cancer over time, conditional on time-varying covariate history.
+This recreates a pseudopopulation in which time-varying covariates do not predict the risk of cancer diagnosis at time _t_ [@hernan2000].
+Time varying covariates included age at study entry, sex, ApoE4 status, educational attainment and the time-varying covariates such as smoking status, systolic blood pressure, BMI  and prevalent and incident comorbidities such as: hypertension and diabetes.
 Further details on modeling specifications and weights assessment are presented as **Supplementary material x**.
 
-To estimate the controlled direct effect of Pin1 in the risk of ADRD, that is in the setting where death could have been prevented, we fit inverse probability censoring weights for death.
+Inverse probability censoring weights for death were calculated to relax the independent censoring assumption.
+In scenario A, weights represent the inverse of the probability of not dying conditional on baseline covariates such as age, educational attainment, ApoE4, and baseline measurements of smoking status, hypertension status, systolic blood pressure, BMI, history of diabetes and cohort.
+For individuals who died, their censoring weight is zero [@whatif2020].
+In scenario B time-varying weights represent the product of the inverse probability of surviving in each year prior to _t_, conditional on the measured shared causes of death and dementia.
+For an individual who has died by time _t_, the year _t_ censoring weight is zero.
+Weights were fitted including the same covariates used to estimate weights for time-varying cancer diagnosis, though we additionally added time-varying cancer, stroke, and heart disease diagnosis as predictors for death.
 
-in time-varying settings we compared the complement of a weighted Kaplan-Meier survival estimator in participants with incident cancer vs.
-no incident of cancer with time indexed in years.
-The weights in this case are time-varying by follow-up year, defined as a product of the time-fixed IPT weights above and a year-specific inverse probability of censoring (IPC) by death weights.
-For an individual still alive in year t, the time t IPC weight is the product of the inverse probability of surviving in each year prior to t, conditional on measured common causes of death and dementia (that is, variables such as C in Figure 4).
-For an individual who has died by time t, the year t IPC weight is zero.
-We estimated survival probabilities using a logistic regression model for death as a function of baseline and time-varying covariates.
-Baseline covariates included age at study entry, sex, apoe4 status, educational attainment and the time-varying covariates smoking status, systolic blood pressure, BMI and prevalent and incident comorbidities such as: cancer, heart disease, stroke and diabetes.
-We additionally calculated the controlled direct effect considering death as an unconditional independent censoring event (as if there were no arrows from $C$ to $D_{15}$ and $Y_{20}$) for an illustrative purpose.
+To estimate the controlled direct effect of Pin1 in the risk of ADRD, we compared the complement of a weighted Kaplan-Meier survival estimator for participants who developed cancer versus those who did not, with time indexed in years.
+The weights are time-varying by follow-up year, defined as a product of the year-specific inverse probability of treatment weights and the inverse probability of censoring by death weights.
+Estimates of the controlled direct effect at 20 years of follow-up are presented as risk differences (RD) and risk ratios (RR).
+All 95% confidence intervals were calculated using percentile-based bootstrapping based on 500 bootstrap samples.
 
-Estimates of the controlled direct effect at 20 years of follow-up are presented as risk differences (RD), risk ratios (RR) and hazard ratios (HR).
-We note that hazards, unlike risks, inherently condition on surviving both dementia and death, as such they will not have a causal interpretation in this case.
-We present them for comparison against risk ratios.
+For illustrative and comparative purposes we calculated hazard ratios (HR).
+Hazards, unlike risks, inherently condition on surviving both dementia and death, as such they will not have a causal interpretation in this case [@young2020].
+We additionally calculated the controlled direct effect considering death as an unconditional independent censoring event (as if there were no arrows from $C$ to $D_{t+1}$ and $Y_{t+1}$).
 
-Since the conditional independent censoring assumption is untestable, we compute Peterson upper and lower bounds to represent: 1) the extreme scenario of independence, that refers to an scenario were those who died would never develop dementia (lower bound) and 2) complete dependency, that refers to an scenario where those who died would have a dementia prior to death (upper bound).
+Since the conditional independent censoring assumption is untestable, we compute Peterson upper and lower bounds [@peterson1976] to represent: 1) the extreme scenario of independence, that refers to an scenario were those who died would never develop dementia (lower bound) and 2) complete dependency, that refers to an scenario where those who died would have a dementia prior to death (upper bound).
 The lower bound is calculated with the Aalen-Johansen estimator treating death as a competing event, and the upper bound is calculated with the Kaplan Meier estimator for the combined outcome of dementia or death.
 
-We additionally present the causes of death for participants who developed cancer and for participants free of cancer diagnosis over follow-up.
-This information provides insights about the association between cancer diagnosis and death and about the potential misclassification of individuals with no cancer diagnosis (cancer as a cause of death).
+All analysis were performed using R, code is provided in supplementary material and available in https://github.com/palolili23/2021_cancer_dementia.
 
-All 95% confidence intervals were calculated using percentile-based bootstrapping based on 500 bootstrap samples.
-All analysis were performed using R, code is provided in supplementary material and available in https://github.com/palolili23/2021_cancer_dementia.
+### 3.2. Results
+
+Participants had a mean age of `r mean_age`, and `r women_prop` were women.
+Further details on participants are presented in Table 1.
+Over follow-up, `r yes_cancer` developed cancer and `r no_cancer` remain free of cancer diagnosis, the median age of cancer diagnosis was `r cancer_time`.
+From the total sample, `r dem_total` had dementia over follow-up and median time to dementia was `r dem_time`.
+Among participants with incident cancer, `r cancer_dem` had dementia diagnosis and `r cancer_death` died over follow-up, `r cancer_alive` remain alive at 20 years since study entry.
+In contrast, among participants free of cancer diagnosis over follow-up, `r no_cancer_dem` were diagnosed with dementia and `r no_cancer_death` died over follow-up, `r no_cancer_alive` were alive at the end of follow-up.
+Figure x.
+shows the magnitude of the proportion of people who died over time.
+
+Results for all scenarios are present in Table 2.
+Had we defined the proxy for Pin1 as _ever cancer_, and relying on death as independent censoring event (unconditional), we observe a significant protective effect of ever having cancer in the risk of dementia [RR: `r complete_table[3,4]`, HR:  `r complete_table[3,5]`].
+This effect did not change after including censoring weights for death [RR: `r complete_table[2,4]`; HR: `r complete_table[2,5]`].
+
+In contrast, had we defined cancer diagnosis as a _time-varying_ proxy for Pin1 over-expression and had we prevented death conditional on covariates, the risk of dementia had participants had a cancer diagnosis over time is higher [RR: `r complete_table[6,4]`; HR: `r complete_table[6,5]`], though confidence intervals cross the null.
+
+\newpage
+
+## Tables
+
+### Table 1: Descriptive characteristics of individuals who had a cancer diagnosis and of those free of cancer diagnosis over follow-up.
+
+```{r}
+tableone_exp <-
+  print(
+    tableone,
+    explain = TRUE,
+    test = FALSE,
+    quote = FALSE,
+    noSpaces = TRUE,
+    printToggle = FALSE
+  )
+
+kableExtra::kable(tableone_exp, booktabs = T) %>%
+  kableExtra::kable_styling(latex_options = c("striped", "hold_position"))
+
+```
+
+\newpage
+
+### Table 2. Risk difference and risk ratio for the risk of dementia
+
+```{r}
+kableExtra::kable(complete_table, booktabs = T) %>%
+  kableExtra::kable_styling(latex_options = c("striped", "hold_position"))
+```
+
+\newpage
+
+## Figures
+
+### Figure 1: Distribution of participants under each health status, by age over follow-up
+
+```{r}
+plot_distribution
+```
+
+\newpage

### 2021-06-06 02_R/04_direct_effect_simplified.R

@@ -122,7 +122,7 @@ death_den <-
   glm(
     competing_plr ~ bs(age_0, 3) + sex + education + apoe4 + 
       as.factor(smoke1) + ht1 + bs(sbp1, 3) + bs(bmi1,3) +
-      as.factor(diabetes_prev) +  cancer_v + cohort,
+      as.factor(diabetes_prev) + cohort + cancer_v,
     data = data_wide,
     family = binomial
   )
@@ -266,53 +266,69 @@ rd_1e  <- km_ever_ipcw_up_bound %>%
 
 # 2. Time-varying cancer -----------------------------------------------------
 
+# 2.1. Weights for t2cancer -----------------------------------------------
 
-# 2.1. IPTW for t-v cancer, with baseline covariates -------------------
-
-smoke1 <- data_wide %>% select(id, smoke1, ht1, sbp1, bmi1)
-
-data_long <- data_long %>% 
-  left_join(smoke1)
-
-cancer_den <-
-  glm(
-    cancer_v ~ bs(age_0, 3) + sex + education + as.factor(smoke1) + cohort,
-    data = data_long,
-    family = binomial
-  )
-
-summary(cancer_den)
-
-cancer_num <- glm(cancer_v ~ 1 , data = data_long)
-
-summary(cancer_num)
-
-data_long <- data_long %>%
-  mutate(
-    p_num = predict(cancer_num, type = "response"),
-    p_denom = predict(cancer_den, type = "response"))
-  
-data_long <- data_long %>%
-  group_by(id) %>% 
-  mutate(
-    sw_cancer = ifelse(cancer_v == 1, p_num/p_denom, (1 - p_num)/(1- p_denom)),
-    # sw_cancer = cumprod(sw_cancer),
-    w_cancer = ifelse(cancer_v == 1, 1/p_denom, 1/(1- p_denom)),
-    # w_cancer = cumprod(w_cancer)
-    ) %>% 
-  ungroup()
+# We want the probability of having a cancer dx at each 
+# time point for each participant who has not yet being diagnosed
 
 data_long %<>% 
-  mutate(sw_cancer_t = ifelse((sw_cancer > quantile(sw_cancer, 0.99)), 
-                            quantile(sw_cancer, 0.99), sw_cancer),
-         w_cancer_t = ifelse((w_cancer > quantile(w_cancer, 0.99)), 
[…]

### 2021-06-06 02_R/04c_bootstraps_ever.R

@@ -58,7 +58,7 @@ km_ever <- function(data, crude = TRUE, ipcw = FALSE) {
         glm(
           competing_plr ~ bs(age_0, 3) + sex + education + apoe4 + 
             as.factor(smoke1) + ht1 + bs(sbp1, 3) + bs(bmi1,3) +
-            as.factor(diabetes_prev) +  cancer_v + cohort,
+            as.factor(diabetes_prev) + cohort + cancer_v,
           data = data,
           family = binomial
         )

### 2021-06-06 02_R/04c_bootstraps_tv.R

@@ -53,39 +53,65 @@ km_tv <- function(data_long, crude = TRUE, ipcw = FALSE) {
   }
   
   else {  
-    cancer_den <-
-      glm(
-        cancer_v ~ bs(age_0, 3) + sex + education + as.factor(smoke1) + cohort,
-        data = data_long,
-        family = binomial
-      )
-    
-    cancer_num <- glm(cancer_v ~ 1 , data = data_long)
-    
-    data_long <- data_long %>%
-      mutate(
-        p_num = predict(cancer_num, type = "response"),
-        p_denom = predict(cancer_den, type = "response"))
-    
-    data_long <- data_long %>%
-      group_by(id) %>% 
-      mutate(
-        sw_cancer = ifelse(cancer_v == 1, p_num/p_denom, (1 - p_num)/(1- p_denom)),
-        # sw_cancer = cumprod(sw_cancer),
-        w_cancer = ifelse(cancer_v == 1, 1/p_denom, 1/(1- p_denom)),
-        # w_cancer = cumprod(w_cancer)
-        ) %>% 
-      ungroup()
     
     data_long %<>% 
-      mutate(sw_cancer_t = ifelse((sw_cancer > quantile(sw_cancer, 0.99)), 
-                                  quantile(sw_cancer, 0.99), sw_cancer),
-             w_cancer_t = ifelse((w_cancer > quantile(w_cancer, 0.99)), 
-                                 quantile(w_cancer, 0.99), w_cancer))
+      group_by(id) %>%
+      mutate(pastcancer = ifelse(cancer_v == 1 & lag(cancer_v) == 0, 0, cancer_v),
+             pastcancer = ifelse(is.na(pastcancer), 0, pastcancer)) %>% 
+      ungroup() 
+    
+    
+    # denominator
+    mod <- glm(cancer_v ~ bs(time, 3) + bs(age_0, 3) + sex + education + apoe4 +
+                 as.factor(smoke) + bs(sbp, 3) + bs(bmi, 3) + 
+                 ht + ht_drug + diab_v + cohort,
+               family = quasibinomial(), data = subset(data_long , 
+                                                       pastcancer == 0))
+    
+    summary(mod)
+    
+    data_long %<>%
+      mutate(pred_a_den = ifelse(pastcancer == 1, 
+                                 1, # the pr of ever having cancer is 1 for those with cancer
+                                 predict(mod, type = 'response'))) 
+    
+    # numerator
[…]

### 2021-06-06 02_R/table_2.R

@@ -1,0 +1,35 @@
+library(tidyverse)
+ever <- rio::import(here::here("02_R", "rd_results_ever.csv"))
+tv <- rio::import(here::here("02_R", "rd_results_tv.csv"))
+
+hr <-rio::import(here::here("02_R", "hr_results.csv")) %>%
+  mutate(Proxy = ifelse(Proxy == "Time to cancer", "Time-varying", Proxy)) %>% 
+  mutate_if(is.numeric, round, 2) %>% 
+  mutate(Proxy = str_replace(Proxy, "Ever vs. Never", "Ever vs. never")) %>% 
+  mutate(`Hazard Ratio` = paste0(point, "(", conf.low, ",", conf.high, ")")) %>% 
+  select(Proxy, model, `Hazard Ratio`)
+
+rr <- ever %>%
+  bind_rows(tv) %>%
+  select(Proxy, model, rr, conf) %>%
+  pivot_wider(names_from = conf,
+              values_from = rr) %>%
+  mutate(contrast = "Risk Ratio") %>%
+  rename(conf.high = conf.higher) %>% 
+  mutate_if(is.numeric, round, 2) %>% 
+  mutate(`Risk Ratio` = paste0(point, " (", conf.low, ",", conf.high, ")")) %>% 
+  select(Proxy, model, `Risk Ratio`)
+
+
+rd <- ever %>%
+  bind_rows(tv) %>%
+  select(Proxy, model, rd, conf) %>%
+  pivot_wider(names_from = conf,
+              values_from = rd) %>%
+  mutate(contrast = "Risk Difference") %>%
+  rename(conf.high = conf.higher) %>% 
+  mutate_if(is.numeric, round, 1) %>% 
+  mutate(`Risk Difference`= paste0(point, " (", conf.low, ",", conf.high, ")")) %>% 
+  select(Proxy, model, `Risk Difference`)
+
+complete_table <- rd %>% left_join(rr) %>% left_join(hr)

### 2021-06-06 06_presentations/melodem2021/index.Rmd
update presentation

@@
   select(1,2,rr) %>%
-  filter(Proxy != "Time to cancer",
+  filter(Proxy != "Time-varing",
          model != "IPTW + IPCW") %>%
@@
   mutate(contrast = "Risk Ratio") %>%
-  rename(conf.high = conf.higher) %>%
-  mutate(Proxy = str_replace(Proxy, "Time-varing", "Time-varying"))
+  rename(conf.high = conf.higher)
 
@@
   mutate(contrast = "Hazard Ratio") %>%
-  filter(Proxy != "Time to cancer") %>%
-  mutate(Proxy = ifelse(
-    Proxy == "Ever vs. Never", "Ever vs. never", "Time-varying")) %>%
   bind_rows(rr) %>%
+  mutate(Proxy = str_replace(Proxy, "Time-varying", "Time to cancer"),
+         Proxy = str_replace(Proxy, "never", "Never")) %>%
   mutate(Proxy = str_to_sentence(Proxy))
@@
 **Epidemiology department**]
+
+---
+
+class: middle
+
+### Acknowledgments
+
+- Sonja A. Swanson
+
+- Kimberly van der Willik
+
+- M. Arfan Ikram
 
@@
 
-  + For an individual still alive in year _t_, weights represent the the product of the inverse probability of surviving in each year prior to _t_, conditional on measured common causes of death and dementia.
+  + For an individual still alive in year _t_, weights represent the product of the inverse probability of surviving in each year prior to _t_, conditional on measured common causes of death and dementia.
 
@@
 
-### Results
+### Preliminary results
 
@@
 results %>%
-  filter(Proxy == "Time-varying") %>%
+  filter(Proxy == "Time to cancer") %>%
   mutate(model = fct_relevel(model, c("Unadjusted", "IPTW", "IPTW + IPCW"))) %>%

### 2021-06-15 04_reports/manuscript.Rmd
latest updates on manuscript

@@
 date: ""
-output: [pdf_document]
+output: [word_document]
 always_allow_html: true
@@
 bibliography: cancer_dementia.bib
+csl: AnD_journal.csl
 ---
@@
 We describe the challenges and how they translate into the analytic decisions.
-Last, we discuss how information on mortality and cause of death can provide insight about the direction of some sources of bias.
 
-## 2. Overview of the causal structure
+## 2. Methods
+
+### 2.1. Overview of the causal structure
 
@@
 
-A challenge that arises by defining cancer diagnosis as the proxy for Pin1 is defining the time zero, the time where follow-up and screening for ADRD would begin after having the exposure measured.
+A challenge that arises by defining cancer diagnosis as the proxy for Pin1 is defining the time zero, the time where eligibility criteria is met, where exposure is measured and the time where the screening for ADRD would begin after having the exposure measured.
 The eligibility criteria to join the study may not align with the time cancer diagnosis is measured.
@@
 For interpretability we exclude the time-varying process of cancer diagnosis and focus on Pin1 (and cancer diagnosis) as it had been measured in all participants at $t$.
-Given that some participants may die prior to having a cancer diagnosis over follow-up, we can only have ADRD over follow-up in the individuals who survive long enough to have an ADRD diagnosis.
+Given that some participants may die prior to ADRD diagnosis, we can only measure ADRD over follow-up in the individuals who survive long enough to have an ADRD diagnosis.
 In the causal diagram of Figure 4 we include a node for death after the exposure $P^*t$ has been measured, represented as $D_{t+1}$, this acts as a competing event of $Y_{t+1}$ because if a participant dies by $t + 1$, the participant cannot subsequently develop dementia.
@@
 
-In the setting where $P_t$ represented the targeted-drug for Pin1, and if this drug had no systemic beneficial or harmful side-effects such as that there is no arrow between $P$ and $D_{t+1}$, a total effect would quantify the effect of $P_t$ on $Y_{t+1}$ that does not include any pathway mediated through $D_{t+1}$[@young2020].
+In the setting where $P_t$ represented the targeted-drug for Pin1, and if this drug had no systemic beneficial or harmful side-effects such as that there is no arrow between $P_t$ and $D_{t+1}$, a total effect would quantify the effect of $P_t$ on $Y_{t+1}$ that does not include any pathway mediated through $D_{t+1}$[@young2020].
 However, in the context of cancer diagnosis as the proxy for Pin1 over-expression, we cannot rule out the effect of cancer diagnosis on death, represented as the arrow between $P^*_t$ and  $D_{t+1}$.
@@
 Previous studies have defined death as a censoring event [], although failed to explicitly define how to comply with the independent censoring assumption.
-Moreover, adjusting for confounders for dementia (such as only adjusting for $L$) may represent partially the adjustment for the competing event of death [@geskus], to block the pathway mediated by $C$ we need to consider covariates that change over time and that affect both nodes, such as cardiometabolic conditions.
+Moreover, adjusting for confounders for dementia (such as only adjusting for $L$) may represent partially the adjustment for the competing event of death [@geskus2016], though to block the pathway mediated by $C$ we need to consider covariates that change over time and that affect both nodes, such as cardiometabolic conditions.
 
@@
 
-We now turn to an application where we show how these challenges translate in analytic decisions, and we will show ways to mitigate or better understand to the best of the available data's abilities.
-Nevertheless, we acknowledge that cancer diagnosis represents a complex and heterogenous health condition that exceeds the representation of Pin1 expression, thus, at best we may test the sharp null hypothesis rather than estimating the causal effect of Pin1.
-In this section we conduct an analysis of the cancer-ADRD association that is structured to the best of the available data's abilities to inform the possible effect of Pin1 on all cause-dementia.
+We now turn to an application where we show how these challenges translate in analytic decisions.
+We will show ways to mitigate or better understand the best of the available data's abilities to inform the possible effect of Pin1 on all cause-dementia.
 
-## 3. Application to the Rotterdam Study
+## 2.2. The Rotterdam Study
 
 We use data collected in the Rotterdam Study, a population-based prospective cohort study among persons living in the Ommoord district in Rotterdam, the Netherlands.
-Recruitment and initial assessments were held between 1990 and 1993; it was later extended between 2000 and 2001 consisting of individuals who had reached the age of 55 years or who had moved into the study area.
+Recruitment and initial assessments were held between 1990 and 1993, a second waive of recruitment was held between 2000 and 2001.
 Participants from first subcohort had follow-up visits between 1993-1995, 1997-1999, 2002-2005, and 2008-2010, second subcohort had follow-up visits between 2004 and 2005, and between 2011 and 2012[@ikram2020].
-All participants had data on incident cancer diagnosis and incident dementia diagnosis through follow-up, collected from medical records of general practitioners (including hospital discharge letters) and through linkage with national registries.
+All participants had data on history of cancer and incident dementia diagnosis through follow-up, collected from medical records of general practitioners (including hospital discharge letters) and through linkage with national registries.
 Date and cause of death was collected on a weekly basis via municipal population registries.
@@
 
-We considered as inclusion criteria being between 60 and 70 years old at study entry, without history of cancer diagnosis, free of cognitive decline or with previous history of dementia.
+We defined as inclusion criteria being between 60 and 70 years old at study entry, without history of cancer diagnosis, free of cognitive decline or with previous history of dementia.
 Out of 10998 persons who participated at study entry, `r total_n` were considered eligible.
@@
 
-### 3.1. Methods
+### 2.3. Statistical Methods
 
@@
 In scenario A, weights were defined as the inverse of the probability of cancer diagnosis conditional on baseline covariates such as age at study entry, sex, educational attainment, cohort, smoking status.
-In contrast, for scenario B weights were defined to represent the product of the inverse probability of being diagnosed with cancer over time, conditional on time-varying covariate history.
+In contrast, for scenario B weights were defined to represent the product of the inverse probability of being diagnosed with cancer over time, conditional on the time-varying covariate history prior to cancer diagnosis.
 This recreates a pseudopopulation in which time-varying covariates do not predict the risk of cancer diagnosis at time _t_ [@hernan2000].
@@
 Inverse probability censoring weights for death were calculated to relax the independent censoring assumption.
-In scenario A, weights represent the inverse of the probability of not dying conditional on baseline covariates such as age, educational attainment, ApoE4, and baseline measurements of smoking status, hypertension status, systolic blood pressure, BMI, history of diabetes and cohort.
+In scenario A, weights represent the inverse of the probability of not dying conditional on cancer diagnosis (ever vs.
+never), baseline covariates such as age, educational attainment, ApoE4, and baseline measurements of smoking status, hypertension status, systolic blood pressure, BMI, history of diabetes and cohort.
 For individuals who died, their censoring weight is zero [@whatif2020].
@@
 For illustrative and comparative purposes we calculated hazard ratios (HR).
-Hazards, unlike risks, inherently condition on surviving both dementia and death, as such they will not have a causal interpretation in this case [@young2020].
+Hazards, unlike risks, inherently condition on surviving both dementia and death, as such they will not have a causal interpretation in this setting [@young2020].
 We additionally calculated the controlled direct effect considering death as an unconditional independent censoring event (as if there were no arrows from $C$ to $D_{t+1}$ and $Y_{t+1}$).
@@
 
-### 3.2. Results
+## 3. Results
 
@@
 In contrast, had we defined cancer diagnosis as a _time-varying_ proxy for Pin1 over-expression and had we prevented death conditional on covariates, the risk of dementia had participants had a cancer diagnosis over time is higher [RR: `r complete_table[6,4]`; HR: `r complete_table[6,5]`], though confidence intervals cross the null.
+In the extreme scenarios of dependence between death and dementia over follow-up, bounds are wide and include the null value ()
+
+## 4. Discussion
+
+Several observational studies have defined “cancer diagnosis” as exposure, although this does not represent a target for potential intervention or a modifiable risk factor.
+Instead, this variable is commonly used to represent a mechanism of interest that could not be measured.
+In this study we define cancer diagnosis as the proxy for Pin1 expression, a potential therapeutic drug-target.
+By explicitly including Pin1 as part of the research question we connect the unmeasured mechanism of interest to the observed data outlining the data generating mechanisms.
+This process allows to identify and disentangle potential sources of bias and can guide analytic decisions.
+We show how results of the association between cancer diagnosis and dementia can change substantially according to alternative, yet explicit, assumptions.
+
+For instance, a key challenge of cancer diagnosis as a proxy for Pin1 is the incapacity of defining a clear time zero[@hernanInmortal2016].
+In the setting where cancer diagnosis is defined as "ever vs.
+never", this definition implicitly introduces immortal time bias, only participants who survived and remain free of dementia over follow-up can be considered for the cancer arm.
+All results pertaining to this definition showed a substantial inverse association between cancer diagnosis and dementia, in contrast to cancer diagnosis as a time-varying process which shifted point estimates close to the null.
+In line with the latter definition, including time-varying covariates is possible with well-known statistical methods such as inverse probability weighting, as opposed to previous studies that only considered baseline information.
+Although we attempt to prevent this bias with statistical methods, we can only fully prevent it by having a clear definition of time-zero.
+This definition does not depend on the collected data nor in analytic decisions.
+It relies on a deeper discussion related to when would be the best moment to measure this variable and to what purpose.
+Thus, we hope that these unsolved questions guide future discussions and data collection efforts.
+
+In contrast, death as competing event is a challenge that has a straightforward solution, which begins by choosing the causal parameters (or estimands) of interest [@young2020, @rojas_medrxiv].
+In this study we chose the controlled direct effect, which represents the effect of Pin1 (or cancer) in a setting where death could have been prevented.
+To address this question, death is defined as a censoring event and relies on the independent censoring assumption.
+As opposed to prior studies that treated censoring as ignorable[@frain2017], we show how point estimates change substantially when we include IPCW for death to relax this assumption.
+In contrast to prior studies who have defined "survival bias" or "competing event bias" and have proposed different estimators to account for them[@ospina2020], we conceptualize bias as not meeting the independent censoring assumption conditional on covariates[@vangeloven2014, @willems2018, @young2020].
+Bounds to assess extreme scenarios of dependence between death and dementia[@peterson1976] illustrate the wide range of possible point estimates that cross the null.
+This shows that even with the effort of adjusting for time-varying covariates, we may be far from meeting this assumption and thus better efforts to measure shared causes of dementia and death are needed.
+
+Pin1 is only one of the several mechanisms proposed about the inverse cancer-dementia relationship.
+However, each mechanism and its connection to collected data may have a different causal representation.
+For example, if the intention was to explain the effect of different chemotherapeutics and how these affect brain, the eligibility criteria would change and the causal graph would represent the downstream of cancer diagnosis.
+Likewise, an alternative causal question related to death as a competing event is the separable effects, which avoid evoking consideration of implausible scenarios that were death is eliminated[@stensrud2020].
+The separable effects would allow the decomposition of Pin1 related mechanisms that affect neurodegeneration and tumor proliferation, that could be confirmed with future studies on modified treatments.
+Last, this case-study extends to other questions that study the effect of one disease in the risk of another disease to understand the biological mechanisms behind them.
+
+Nevertheless, we acknowledge that cancer diagnosis represents a complex and heterogeneous health condition that exceeds the representation of Pin1 expression, thus, at best we may test the sharp null hypothesis rather than estimating the causal effect of Pin1.
 

### 2021-06-15 02_R/04_direct_effect_simplified.R

@@ -327,7 +327,7 @@ data_long %<>%
   )
 
 data_long %>% 
-  ggplot(aes(x = as_factor(time), y = w_cancer_t)) +
+  ggplot(aes(x = as_factor(time), y = sw_cancer_t)) +
   geom_boxplot()
 
 # 2.2. IPCW. Weights on death - dementia ---------------------------------------------

### 2021-06-15 02_R/table_2.R

@@ -33,3 +33,9 @@ rd <- ever %>%
   select(Proxy, model, `Risk Difference`)
 
 complete_table <- rd %>% left_join(rr) %>% left_join(hr)
+
+complete_table <- complete_table %>% 
+  mutate(model = as_factor(model),
+         model = fct_relevel(model, c("Unadjusted", "IPTW", "IPTW + IPCW"))) %>% 
+  group_by(Proxy) %>% 
+  arrange(model)

### 2021-06-16 04_reports/manuscript.Rmd
complete manuscript

@@
 source(here::here("02_R", "05_plots.R"))
+source(here::here("02_R", "05_plots.R"))
 ```
@@
 Many observational studies have consistently found that individuals with cancer have a lower risk of developing of Alzheimer´s disease or related dementias (ADRD) when compared to individuals with no history of cancer [@ma2014;@hanson2016;@vanderwillik2018;@ospina2020].
-These findings have motivated substantial research toward mechanistic explanations, including searching for and hypothesizing that molecular and genetic mechanisms may explain this association[@behrens2009;@harris2014;@nudelman2019;@Papin2020;@driverbiogeront2014;@olson2019;@li2021;@driverpin12015].
+These findings have motivated substantial research toward mechanistic explanations, including molecular and genetic pathways that may explain this association[@behrens2009;@harris2014;@nudelman2019;@Papin2020;@driverbiogeront2014;@olson2019;@li2021;@driverpin12015].
 These research inquiries inevitably lead to discussions of repurposing or augmenting current cancer chemotherapeutics for ADRD [@Snyder2017].
@@
 However, understanding these or other sources of bias first requires making explicit the causal question.
-Moreover, making explicit the causal question is one step toward tying a research study to a question that is relevant to decision-making[@didelez2016;@labrecque2017].
+Moreover, this is one step toward tying a research study to a question that is relevant to decision-making[@didelez2016;@labrecque2017].
 
@@
 To explore how we might learn about this effect using real-world data on cancer and ADRD, we progressively build a causal directed acyclic graph to connect this particular causal question to the observable data and the assumptions we rely on to study the effect.
-We exemplify different scenarios with data collected from the Rotterdam Study, a population-based cohort study.
-We describe the challenges and how they translate into the analytic decisions.
+We exemplify these challenges and how they translate into the analytic decision using data collected from the Rotterdam Study, a population-based cohort study.
 
@@
 
-If this hypothetical Pin1-targeting drug was developed, the best way to understand its effect on dementia risk would be to have a well-conducted randomized trial in which we randomize eligible participants in late midlife (e.g., ages 50-60 years) to receive this drug or not, and closely monitor ADRD over a lengthy follow-up.
+If this hypothetical Pin1-targeting drug was developed, the best way to understand its effect on ADRD risk would be to have a well-conducted randomized trial in which we randomize eligible participants in late midlife (e.g., ages 50-60 years) to receive this drug or not, and closely monitor ADRD over a lengthy follow-up.
 Since this drug is not currently available, at best we can use observational data on Pin1 expression measurements.
-For example, suppose that a biomarker test was available to measure Pin1 and we measured this biomarker from (stored) baseline blood samples in a population based-cohort that recruited participants in late midlife.
+For example, suppose that a biomarker test was available to measure Pin1 and we measured this biomarker in stored baseline blood samples from late-midlife participants recruited for a population based-cohort.
 
 In the observational setting, confounding could explain the observed association between Pin1 and ADRD.
-In the directed acyclic graph [@whatif2020] of Figure 1, we present Pin1 expression measured at time $t$ as $P_{t-1}$ and ADRD diagnosis over time as $Y_{t+1}$.
+In the directed acyclic graph [@whatif2020] of Figure 1, we present Pin1 expression as $P_{t-1}$ and ADRD diagnosis over time as $Y_{t+1}$.
 Both nodes may share causes $L$, and to assess the causal relationship between $P_{t-1}$ and $Y_{t+1}$ we would require adjusting for $L$.
 Previous studies have described age, sex, educational level and race/ethnicity as the minimal adjusting set of covariates[@ospina2020].
-However, environmental and behavioral factors such as smoking, which are known to cause microenvironmental changes such as inflammation and changes in tissue remodeling, may translate into Pin1 over-expression and are also related to the development of ADRD[ref].
+However, environmental and behavioral factors such as smoking may translate into Pin1 over-expression and are also related to the development of ADRD[ref].
 
@@
 Currently, Pin1 expression is not an available biomarker for population-based research, so at best we can only rely on a proxy for it.
-Because Pin1 over-expression is present in tumors, and tumors are only measured through diagnosis, previous studies have considered Pin1 over-expression as the explanation behind the inverse association between cancer diagnosis and ADRD, though Pin1 was not explicitly part of the research question[@ospina2020;@driver2012;@musicco2013;@freedman2016;@bowles2017;@frain2017;@schmidt2017;@sun2020;@ording2020].
+Because Pin1 over-expression is present in tumors, and tumors are only measured through diagnosis, previous studies have considered Pin1 over-expression as the explanation behind the inverse association between cancer diagnosis and ADRD, though Pin1 was not explicitly part of the research question[@driver2012;@musicco2013;@freedman2016;@bowles2017;@frain2017;@schmidt2017;@sun2020;@ording2020].
 Unlike measuring Pin1 at the same time to all participants (thought this would not necessarily mean this would be the ideal time to measure it, we discuss this point further in the discussion section), cancer diagnosis is a collected over time.
@@
 
-A challenge that arises by defining cancer diagnosis as the proxy for Pin1 is defining the time zero, the time where eligibility criteria is met, where exposure is measured and the time where the screening for ADRD would begin after having the exposure measured.
-The eligibility criteria to join the study may not align with the time cancer diagnosis is measured.
-This situation may introduce inmortal-time bias [@hernanInmortal2016].
+A challenge that arises by choosing cancer diagnosis as the proxy for Pin1 is defining the time zero, the time where eligibility criteria is met, where exposure is measured and the time where the screening for ADRD would begin after having the exposure measured[@hernanInmortal2016].
+The eligibility criteria to participate in a trial for Pin-targeting drug will not necessarily share the same criteria to be recruited for the cohort setting.
+The latter may not align with the time cancer diagnosis is measured since it happens over follow-up and this situation may introduce inmortal-time bias[@hernanInmortal2016].
 For example, a study performed using data from the Framingham study [@driver2012] defined the exposed group with cancer as those participants with prevalent or incident cancer diagnosis (alternatively defined as "ever cancer" [@hanson2016]).
@@
 In this way, the time prior to cancer diagnosis is allocated to the non-exposed arm, and the time after cancer diagnosis to the exposed arm [@white2013; @hanson2016; @bowles2017].
-The price we pay with this approximation is that implicitly, this means that Pin1 would over-express at the time of cancer diagnosis, which is biologically implausible.
+The price we pay with this approximation is that implicitly, this means that Pin1 would over-express at the time of cancer diagnosis and not before, which is biologically implausible.
 Moreover, cancer diagnosis will only be measured in the subset of participants who are alive over follow-up.
-Thus we included death prior to cancer diagnosis as $D_t$ and an arrow between $D_t$ and $P_t$ that represents a deterministic association such as that $P_t$ is only observed if $D_t$ is zero in Figure 3.
-In addition, we added an arrow between $L$ and $D_t$, since covariates such as smoking may affect Pin1 over-expression but also affect the risk of death due to other causes such as COPD.
+Thus, we included death prior to cancer diagnosis as $D_t$ and an arrow between $D_t$ and $P_t$ that represents a deterministic association such as that $P_t$ is only observed if $D_t$ is zero in Figure 3.
+In addition, we added an arrow between $L$ and $D_t$, since covariates such as smoking may affect Pin1 over-expression but also affect the risk of death due to other causes such as from chronic obstructive pulmonary disease.
 Although $L$ is time-varying in nature, we only depict $L$ at one time-point for readability.
@@
 We now move to the challenge of death as a competing event for ADRD, represented in Figure 4.
-For interpretability we exclude the time-varying process of cancer diagnosis and focus on Pin1 (and cancer diagnosis) as it had been measured in all participants at $t$.
+For interpretability we exclude the time-varying process of cancer diagnosis and focus on Pin1 (and cancer diagnosis) as it had been measured in all participants at time $t$.
 Given that some participants may die prior to ADRD diagnosis, we can only measure ADRD over follow-up in the individuals who survive long enough to have an ADRD diagnosis.
-In the causal diagram of Figure 4 we include a node for death after the exposure $P^*t$ has been measured, represented as $D_{t+1}$, this acts as a competing event of $Y_{t+1}$ because if a participant dies by $t + 1$, the participant cannot subsequently develop dementia.
+In the causal diagram of Figure 4 we include a node for death after the exposure $P^*_t$ has been measured, represented as $D_{t+1}$ and which acts as a competing event of $Y_{t+1}$ because if a participant dies by $t + 1$, the participant cannot subsequently develop dementia.
 Furthermore, since $D_{t+1}$ and $Y_{t+1}$ are events related to aging and its consequences, $C$ represents the shared causes of both events such as cardiovascular conditions.
@@
 The latter is defined as the controlled direct effect, where death is considered as a censoring event and it relies on the assumption that we have measured all $C$ to block the pathway $Y_{t+1} \leftarrow C  \rightarrow D_{t+1} \rightarrow P^*_t \rightarrow P_t$.
-Previous studies have defined death as a censoring event [], although failed to explicitly define how to comply with the independent censoring assumption.
-Moreover, adjusting for confounders for dementia (such as only adjusting for $L$) may represent partially the adjustment for the competing event of death [@geskus2016], though to block the pathway mediated by $C$ we need to consider covariates that change over time and that affect both nodes, such as cardiometabolic conditions.
+Previous studies have defined death as a censoring event [@frain2017], although failed to explicitly define how to comply with the independent censoring assumption.
+Moreover, adjusting for confounders for dementia (such as only adjusting for $L$) will be insufficient to block the pathway mediated by death, time-varying covariates that represent $C$, such as cardiovascular conditions should be considered.
 
@@
 To summarize, the complexity of the causal structure that describes the effect of Pin1 through the proxy of cancer diagnosis in the risk of ADRD illustrate the potential sources of bias, as observed in Figure 5.
-Even so, this is a simplified version since we omitted other sources of measurement error [@whatif2020] and the time-varying nature of all nodes and feedback loops between them which would further complicate interpretability and identificability.
+Even so, this is a simplified version since we omitted other sources of measurement error and the time-varying nature of all nodes and feedback loops between them which would further complicate interpretability[@whatif2020].
 
@@
 
-We now turn to an application where we show how these challenges translate in analytic decisions.
+We now turn to an application where we show how these challenges translate into analytic decisions.
 We will show ways to mitigate or better understand the best of the available data's abilities to inform the possible effect of Pin1 on all cause-dementia.
@@
 Participants from first subcohort had follow-up visits between 1993-1995, 1997-1999, 2002-2005, and 2008-2010, second subcohort had follow-up visits between 2004 and 2005, and between 2011 and 2012[@ikram2020].
-All participants had data on history of cancer and incident dementia diagnosis through follow-up, collected from medical records of general practitioners (including hospital discharge letters) and through linkage with national registries.
+All participants had data on history of cancer and dementia and incident diagnosis through follow-up until 2015, collected from medical records of general practitioners (including hospital discharge letters) and through linkage with national registries.
 Date and cause of death was collected on a weekly basis via municipal population registries.
-Data on clinical outcomes was available until 2015.
 
@@
 never"_ [@driver2012].
-Scenario B defines cancer diagnosis is a time-varying exposure.
+Scenario B defines cancer diagnosis as time-varying.
 To address confounding we fit inverse probability treatment weights, stabilized and truncated at 99th percentile.
@@
 This recreates a pseudopopulation in which time-varying covariates do not predict the risk of cancer diagnosis at time _t_ [@hernan2000].
-Time varying covariates included age at study entry, sex, ApoE4 status, educational attainment and the time-varying covariates such as smoking status, systolic blood pressure, BMI  and prevalent and incident comorbidities such as: hypertension and diabetes.
+Baseline covariates included age at study entry, sex, ApoE4 status, educational attainment and the time-varying covariates such as smoking status, systolic blood pressure, BMI  and prevalent and incident comorbidities such as: hypertension and diabetes.
 Further details on modeling specifications and weights assessment are presented as **Supplementary material x**.
@@
 In scenario A, weights represent the inverse of the probability of not dying conditional on cancer diagnosis (ever vs.
-never), baseline covariates such as age, educational attainment, ApoE4, and baseline measurements of smoking status, hypertension status, systolic blood pressure, BMI, history of diabetes and cohort.
+never) and baseline covariates such as age, educational attainment, ApoE4, and baseline measurements of smoking status, hypertension status, systolic blood pressure, BMI, history of diabetes and cohort.
 For individuals who died, their censoring weight is zero [@whatif2020].
 In scenario B time-varying weights represent the product of the inverse probability of surviving in each year prior to _t_, conditional on the measured shared causes of death and dementia.
-For an individual who has died by time _t_, the year _t_ censoring weight is zero.
+For an individual who has died by time _t_, the year _t_ censoring weight is zero [@young2020].
 Weights were fitted including the same covariates used to estimate weights for time-varying cancer diagnosis, though we additionally added time-varying cancer, stroke, and heart disease diagnosis as predictors for death.
@@
 In contrast, among participants free of cancer diagnosis over follow-up, `r no_cancer_dem` were diagnosed with dementia and `r no_cancer_death` died over follow-up, `r no_cancer_alive` were alive at the end of follow-up.
-Figure x.
-shows the magnitude of the proportion of people who died over time.
+Figure 6 shows the proportion of participants in each possible status over follow-up.
 
@@
 Had we defined the proxy for Pin1 as _ever cancer_, and relying on death as independent censoring event (unconditional), we observe a significant protective effect of ever having cancer in the risk of dementia [RR: `r complete_table[3,4]`, HR:  `r complete_table[3,5]`].
-This effect did not change after including censoring weights for death [RR: `r complete_table[2,4]`; HR: `r complete_table[2,5]`].
-
+This effect shifts towards the null after including censoring weights for death [RR: `r complete_table[5,4]`; HR: `r complete_table[5,5]`].
 In contrast, had we defined cancer diagnosis as a _time-varying_ proxy for Pin1 over-expression and had we prevented death conditional on covariates, the risk of dementia had participants had a cancer diagnosis over time is higher [RR: `r complete_table[6,4]`; HR: `r complete_table[6,5]`], though confidence intervals cross the null.
-In the extreme scenarios of dependence between death and dementia over follow-up, bounds are wide and include the null value ()
+In addition, the extreme scenarios of dependence between death and dementia over follow-up represented as bounds are wide and include the null value (`r bounds_tv_ipw`).
 
@@
 In this study we define cancer diagnosis as the proxy for Pin1 expression, a potential therapeutic drug-target.
-By explicitly including Pin1 as part of the research question we connect the unmeasured mechanism of interest to the observed data outlining the data generating mechanisms.
-This process allows to identify and disentangle potential sources of bias and can guide analytic decisions.
+By explicitly including Pin1 as part of the research question we connect the unmeasured mechanism of interest to the observed data outlining the data generation process.
+This practice helps to identify and disentangle potential sources of bias and can guide analytic decisions.
 We show how results of the association between cancer diagnosis and dementia can change substantially according to alternative, yet explicit, assumptions.
@@
 In the setting where cancer diagnosis is defined as "ever vs.
-never", this definition implicitly introduces immortal time bias, only participants who survived and remain free of dementia over follow-up can be considered for the cancer arm.
-All results pertaining to this definition showed a substantial inverse association between cancer diagnosis and dementia, in contrast to cancer diagnosis as a time-varying process which shifted point estimates close to the null.
-In line with the latter definition, including time-varying covariates is possible with well-known statistical methods such as inverse probability weighting, as opposed to previous studies that only considered baseline information.
+never", this definition implicitly introduces immortal time bias.
+All results pertaining to this definition showed a substantial inverse association between cancer diagnosis and dementia in contrast to cancer diagnosis as a time-varying process which shifted point estimates close to the null.
+In line with the latter definition, including time-varying covariates is possible with well-known statistical methods such as inverse probability weighting, as opposed to previous studies that only considered baseline information[lots of refs].
 Although we attempt to prevent this bias with statistical methods, we can only fully prevent it by having a clear definition of time-zero.
 This definition does not depend on the collected data nor in analytic decisions.
-It relies on a deeper discussion related to when would be the best moment to measure this variable and to what purpose.
+It relies on a deeper discussion related to when would be the best moment to measure this biomarker and to what purpose.
 Thus, we hope that these unsolved questions guide future discussions and data collection efforts.
 
-In contrast, death as competing event is a challenge that has a straightforward solution, which begins by choosing the causal parameters (or estimands) of interest [@young2020, @rojas_medrxiv].
-In this study we chose the controlled direct effect, which represents the effect of Pin1 (or cancer) in a setting where death could have been prevented.
-To address this question, death is defined as a censoring event and relies on the independent censoring assumption.
+On the other hand, death as competing event is a challenge that has a straightforward solution, which begins by choosing the causal parameters (or estimand) of interest [@young2020, @rojas_medrxiv].
+In this study we chose the controlled direct effect, which represents the effect of Pin1 (or cancer) in a setting where death could have been prevented and relies on the independent censoring assumption since death is treated as a censoring event.
 As opposed to prior studies that treated censoring as ignorable[@frain2017], we show how point estimates change substantially when we include IPCW for death to relax this assumption.
@@
 This shows that even with the effort of adjusting for time-varying covariates, we may be far from meeting this assumption and thus better efforts to measure shared causes of dementia and death are needed.
+In addition, presenting the proportion of participants in each status over time, and the proportion of participants that died prior to ADRD diagnosis in each arm will always provide insights and improve transparency of results and interpretations.
 
 Pin1 is only one of the several mechanisms proposed about the inverse cancer-dementia relationship.
-However, each mechanism and its connection to collected data may have a different causal representation.
+We acknowledge that cancer diagnosis represents a complex and heterogeneous health condition that exceeds the representation of Pin1 expression, thus, at best we may test the sharp null hypothesis rather than estimating the causal effect of Pin1.
+Furthermore, alternative mechanisms underlying cancer diagnosis and its connection to collected data may be represented in different causal representations.
 For example, if the intention was to explain the effect of different chemotherapeutics and how these affect brain, the eligibility criteria would change and the causal graph would represent the downstream of cancer diagnosis.
-Likewise, an alternative causal question related to death as a competing event is the separable effects, which avoid evoking consideration of implausible scenarios that were death is eliminated[@stensrud2020].
-The separable effects would allow the decomposition of Pin1 related mechanisms that affect neurodegeneration and tumor proliferation, that could be confirmed with future studies on modified treatments.
-Last, this case-study extends to other questions that study the effect of one disease in the risk of another disease to understand the biological mechanisms behind them.
-
-Nevertheless, we acknowledge that cancer diagnosis represents a complex and heterogeneous health condition that exceeds the representation of Pin1 expression, thus, at best we may test the sharp null hypothesis rather than estimating the causal effect of Pin1.
+For this reason, we believe is crucial to have a discussion related to cancer diagnosis as an exposure and how we can shift efforts to collect more suitable data.
+Furthermore, we consider that this case-study extends to other questions that study the effect of one disease in the risk of another disease to understand the biological mechanisms behind them.
 
@@
 
-```{r}
+```{r, echo = FALSE}
 tableone_exp <-
@@
 
-kableExtra::kable(tableone_exp, booktabs = T) %>%
-  kableExtra::kable_styling(latex_options = c("striped", "hold_position"))
+Characteristics <- rownames(tableone_exp)
+
+tableone_exp%>% as_tibble() %>% 
+  cbind(Characteristics) %>% select(Characteristics, everything()) %>% 
+  mutate(
+    Characteristics = str_replace(Characteristics, "smoke1", "Smoking status"),
+    Characteristics = str_replace(Characteristics, "sex = ", ""),
+    Characteristics = str_replace(Characteristics, "age_0", "Age at baseline"),
+    Characteristics = str_replace(Characteristics, "education", "Educational attainment"),
+    Characteristics = str_replace(Characteristics, "apoe4", "ApoE4"),
+    Characteristics = str_replace(Characteristics, "bmi1", "Body Mass Index"),
+    Characteristics = str_replace(Characteristics, "sbp1", "Systolic blood pressure (mmHg)"),
+    Characteristics = str_replace(Characteristics, "ht1 = ", ""),
+    Characteristics = str_replace(Characteristics, "hd_prev = ", ""),
+    Characteristics = str_replace(Characteristics, "hd_v = ", ""),
+    Characteristics = str_replace(Characteristics, "diab_v = ", ""),
+    Characteristics = str_replace(Characteristics, "Unknown", "Unknown history of diabetes"),
+    Characteristics = str_replace(Characteristics, "stroke_prev = ", ""),
+    Characteristics = str_replace(Characteristics, "stroke_v = ", ""),
+  ) %>% 
+  filter(!Characteristics %in% c("oh1 (mean (SD))", "diabetes_prev (%)", "cancer_v = No incident cancer (%)")) %>% 
+  flextable() %>% 
+  autofit()
 
@@
 ```{r}
-kableExtra::kable(complete_table, booktabs = T) %>%
-  kableExtra::kable_styling(latex_options = c("striped", "hold_position"))
+complete_table %>% 
+  flextable() %>% 
+  autofit()
 ```
@@
 
-### Figure 1: Distribution of participants under each health status, by age over follow-up
+### Figure 6: Distribution of participants under each health status, by age over follow-up
 
-```{r}
+```{r, fig.width= 12, fig.height=8}
 plot_distribution

### 2021-06-16 02_R/table_2.R

@@ -39,3 +39,9 @@ complete_table <- complete_table %>%
          model = fct_relevel(model, c("Unadjusted", "IPTW", "IPTW + IPCW"))) %>% 
   group_by(Proxy) %>% 
   arrange(model)
+
+bounds <- rio::import(here::here("02_R", "table_results_bounds.csv"))
+
+bounds_tv_ipw <- bounds %>% slice(4,6) %>% select(model, rr) %>% pull(rr)
+
+bounds_tv_ipw <- paste0("RR: ", bounds_tv_ipw[1], ", ", bounds_tv_ipw[2])

### 2021-07-13 04_reports/dags.Rmd
its been a while (new edits to Sonja's edits)

@@
 title: "Dags"
-output:
-  xaringan::moon_reader
+output: word_document
 ---
-
-```{tikz, echo = FALSE, fig.cap = "Pin1 as biomarker and risk of dementia"}
+## Current DAG
+```{tikz, echo = FALSE}
 \usetikzlibrary{arrows}
 \begin{tikzpicture}[node distance=2cm]
-\node (p) {$P_0$};
-\node [above of = p, yshift = -1cm, xshift = 1cm](l) {$L$};
-\node [right of = p, gray!30] (dp) {$D_t$};
-\node [right of = dp, gray!30] (pstar) {$P^*_t$};
-\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
-\node [right of = d] (y) {$Y_{t+1}$};
-\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
-\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
-\draw[->, gray!30] (p) -- (dp);
+\node (p) {$P_{t-1}$};
+\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
+\node [below of = p, yshift = 1cm] (pstar) {$P^*_{t-1}$};
+\node [right of = p] (dp) {$D_t$};
+\node [right of = dp] (p2) {$P_t$};
+\node [below of = p2, yshift = 1cm] (pstar2) {$P^*_t$};
+\node [right of = p2] (dp2) {$D_{t+1}$};
+\node [right of = dp2] (y) {$Y_{t+1}$};
+\node [below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$C$};
+\draw[->] (p) to (pstar);
 \draw[->] (l) -- (p);
+\draw[->] (l) to (dp);
+\draw[->] (pstar) -- (dp);
+\draw[->] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
+\draw[->] (dp) -- (p2);
+\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->] (p2) to (pstar2);
+\draw[->] (dp2) -- (y);
+\draw[->] (cd) -- (dp2);
+\draw[->] (cd) -- (y);
+\draw[->] (l) -- (p2);
+\draw[->] (l) to [out=0,in=90, looseness=0.4] (dp2);
 \draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->, semithick] (p) to [out=45,in=135, looseness=0.3] (y);
-\draw[->, gray!30] (dp) -- (pstar);
-\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (pstar);
-\draw[->, gray!30] (pstar) -- (d);
-\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (d);
-\draw[->, gray!30] (d) -- (y);
-\draw[->, gray!30] (c) -- (dp);
-\draw[->, gray!30] (c) -- (pstar);
-\draw[->, gray!30] (cd) -- (d);
-\draw[->, gray!30] (cd) -- (y);
+\draw[->] (pstar2) -- (dp2);
+\draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
+\end{tikzpicture}
+```
+
+---
+## Changing time and L
+```{tikz, echo = FALSE}
+\usetikzlibrary{arrows}
+\begin{tikzpicture}[node distance=2cm]
+\node (p) {$P_t$};
+\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L_1$};
+\node [below of = p, yshift = 1cm] (pstar) {$P^*_t$};
+\node [right of = p] (dp) {$D_{t+1}$};
+\node [right of = dp] (p2) {$P_{t+1}$};
+\node [below of = p2, yshift = 1cm] (pstar2) {$P^*_{t+1}$};
+\node [right of = p2] (dp2) {$D_{t+k}$};
+\node [right of = dp2] (y) {$Y_{t+k}$};
+\node [below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$L_2$};
+\draw[->] (p) to (pstar);
+\draw[->] (l) -- (p);
+\draw[->] (l) to (dp);
+\draw[->] (pstar) -- (dp);
+\draw[->] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
+\draw[->] (dp) -- (p2);
+\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->] (p2) to (pstar2);
+\draw[->] (dp2) -- (y);
+\draw[->] (cd) -- (dp2);
+\draw[->] (cd) -- (y);
+\draw[->] (l) -- (p2);
+\draw[->] (l) to [out=0,in=90, looseness=0.4] (dp2);
+\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
+\draw[->] (pstar2) -- (dp2);
+\draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
+\end{tikzpicture}
+```
+
+---
+## Changing time, L and proxy
+```{tikz, echo = FALSE}
+\usetikzlibrary{arrows}
+\begin{tikzpicture}[node distance=2cm]
+\node (p) {$P_t$};
+\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L_1$};
+\node [below of = p, yshift = 1cm] (pstar) {$C_t$};
+\node [right of = p] (dp) {$D_{t+1}$};
+\node [right of = dp] (p2) {$P_{t+1}$};
+\node [below of = p2, yshift = 1cm] (pstar2) {$C_{t+1}$};
+\node [right of = p2] (dp2) {$D_{t+k}$};
+\node [right of = dp2] (y) {$Y_{t+k}$};
+\node [below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$L_2$};
+\draw[->] (p) to (pstar);
+\draw[->] (l) -- (p);
+\draw[->] (l) to (dp);
+\draw[->] (pstar) -- (dp);
+\draw[->] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
+\draw[->] (dp) -- (p2);
+\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->] (p2) to (pstar2);
+\draw[->] (dp2) -- (y);
+\draw[->] (cd) -- (dp2);
+\draw[->] (cd) -- (y);
+\draw[->] (l) -- (p2);
+\draw[->] (l) to [out=0,in=90, looseness=0.4] (dp2);
+\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
+\draw[->] (pstar2) -- (dp2);
+\draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
 \end{tikzpicture}
@@
 
-```{tikz, echo = FALSE, fig.cap = "Cancer diagnosis as proxy for Pin1 expression"}
+## Changing time, L and proxy, plus L to C
+```{tikz, echo = FALSE}
 \usetikzlibrary{arrows}
 \begin{tikzpicture}[node distance=2cm]
-\node (p) {$P_0$};
-\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
-\node [right of = p, gray!30] (dp) {$D_t$};
-\node [right of = dp] (pstar) {$P^*_t$};
-\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
-\node [right of = d] (y) {$Y_{t+1}$};
-\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
-\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
-\draw[->, gray!30] (p) -- (dp);
-\draw[->, gray!30] (l) -- (p);
-\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
+\node (p) {$P_t$};
+\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L_1$};
+\node [below of = p, yshift = 1cm] (pstar) {$C_t$};
+\node [right of = p] (dp) {$D_{t+1}$};
+\node [right of = dp] (p2) {$P_{t+1}$};
+\node [below of = p2, yshift = 1cm] (pstar2) {$C_{t+1}$};
+\node [right of = p2] (dp2) {$D_{t+k}$};
+\node [right of = dp2] (y) {$Y_{t+k}$};
+\node [below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$L_2$};
+\draw[->] (p) to (pstar);
+\draw[->] (l) -- (p);
+\draw[->] (l) -- (pstar);
+\draw[->] (l) to (dp);
+\draw[->] (pstar) -- (dp);
+\draw[->] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
+\draw[->] (dp) -- (p2);
 \draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
-\draw[->, gray!30] (dp) -- (pstar);
-\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
-\draw[->, gray!30] (pstar) -- (d);
-\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (d);
-\draw[->, gray!30] (d) -- (y);
-\draw[->, gray!30] (c) -- (dp);
-\draw[->, gray!30] (c) -- (pstar);
-\draw[->, gray!30] (cd) -- (d);
-\draw[->, gray!30] (cd) -- (y);
+\draw[->] (p2) to (pstar2);
+\draw[->] (dp2) -- (y);
+\draw[->] (cd) -- (dp2);
+\draw[->] (cd) -- (y);
+\draw[->] (l) -- (p2);
+\draw[->] (l) to [out=-45,in=-180, looseness=0.4] (pstar2);
+\draw[->] (l) to [out=0,in=90, looseness=0.4] (dp2);
+\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
+\draw[->] (pstar2) -- (dp2);
+\draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
 \end{tikzpicture}
 ```
-
----
-
-```{tikz, echo = FALSE, fig.cap = "Survival bias when treating cancer diagnosis as the proxy for Pin1 expression"}
-
-\usetikzlibrary{arrows}
-\begin{tikzpicture}[node distance=2cm]
-\node (p) {$P_0$};
-\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
-\node [right of = p] (dp) {$D_t$};
-\node [right of = dp] (pstar) {$P^*_t$};
-\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
-\node [right of = d] (y) {$Y_{t+1}$};
-\node [below of = dp, yshift = 1cm, xshift = 1cm](c) {$C_1$};
-\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
-\draw[->] (p) -- (dp);
-\draw[->, gray!30] (l) -- (p);
-\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
-\draw[->] (dp) -- (pstar);
-\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
-\draw[->, gray!30] (pstar) -- (d);
-\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (d);
-\draw[->] (d) -- (y);
-\draw[->] (c) -- (dp);
-\draw[->] (c) -- (pstar);
-\draw[->, gray!30] (cd) -- (d);
-\draw[->, gray!30] (cd) -- (y);
-\end{tikzpicture}
-```
-
----
-
-```{tikz, echo = FALSE, fig.cap = "Death as a competing event for dementia over follow-up"}
-\usetikzlibrary{arrows}
-\begin{tikzpicture}[node distance=2cm]
-\node (p) {$P_0$};
-\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
-\node [right of = p, gray!30, gray!30] (dp) {$D_t$};
-\node [right of = dp, gray!30] (pstar) {$P^*_t$};
-\node [right of = pstar] (d) {$D_{t+1}$};
-\node [right of = d] (y) {$Y_{t+1}$};
-\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
-\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C_2$};
-\draw[->, gray!30] (p) -- (dp);
-\draw[->, gray!30] (l) -- (p);
-\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
-\draw[->, gray!30] (dp) -- (pstar);
-\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (pstar);
-\draw[->, gray!30] (pstar) -- (d);
-\draw[->, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (d);
-\draw[->] (d) -- (y);
-\draw[->, gray!30] (c) -- (dp);
-\draw[->, gray!30] (c) -- (pstar);
-\draw[->, gray!30] (cd) -- (d);
-\draw[->, gray!30] (cd) -- (y);
-\end{tikzpicture}
-```
-
----
-
-```{tikz, echo = FALSE, fig.cap = "Direct effect"}
-\usetikzlibrary{arrows}
-\begin{tikzpicture}[node distance=2cm]
-\node (p) {$P_0$};
-\node [above of = p, yshift = -1cm, xshift = 1cm, gray!30](l) {$L$};
-\node [right of = p, gray!30, gray!30] (dp) {$D_t$};
-\node [right of = dp] (pstar) {$P^*_t$};
-\node [right of = pstar] (d) {$D_{t+1}$};
-\node [right of = d] (y) {$Y_{t+1}$};
-\node [below of = dp, yshift = 1cm, xshift = 1cm, gray!30](c) {$C_1$};
-\node [below of = d, yshift = 1cm, xshift = 1cm](cd) {$C_2$};
-\draw[->, gray!30] (p) -- (dp);
-\draw[->, gray!30] (l) -- (p);
-\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
-\draw[->, gray!30] (dp) -- (pstar);
-\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
-\draw[->] (pstar) -- (d);
-\draw[->] (p) to [out=+45,in=+135, looseness=0.3] (d);
-\draw[->] (d) -- (y);
-\draw[->, gray!30] (c) -- (dp);
-\draw[->, gray!30] (c) -- (pstar);
-\draw[->] (cd) -- (d);
-\draw[->] (cd) -- (y);
-\end{tikzpicture}
-```
-
----
-
-```{tikz, echo = FALSE, fig.cap = "Direct effect"}
-\usetikzlibrary{arrows}
-\begin{tikzpicture}[node distance=2cm]
-\node (p) {$P_0$};
-\node [above of = p, yshift = -1cm, xshift = 1cm](l) {$L$};
-\node [right of = p] (dp) {$D_t$};
-\node [right of = dp] (pstar) {$P^*_t$};
-\node [right of = pstar] (d) {$D_{t+1}$};
-\node [right of = d] (y) {$Y_{t+1}$};
-\node [below of = dp, yshift = 1cm, xshift = 1cm](c) {$C_1$};
-\node [below of = d, yshift = 1cm, xshift = 1cm](cd) {$C_2$};
-\draw[->] (p) -- (dp);
-\draw[->] (l) -- (p);
-\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
-\draw[->] (dp) -- (pstar);
-\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
-\draw[->] (pstar) -- (d);
-\draw[->] (p) to [out=+45,in=+135, looseness=0.3] (d);
-\draw[->] (d) -- (y);
-\draw[->] (c) -- (dp);
-\draw[->] (c) -- (pstar);
-\draw[->] (cd) -- (d);
-\draw[->] (cd) -- (y);
-\end{tikzpicture}
-```

### 2021-07-13 02_R/03_descriptives.R

@@ -31,7 +31,7 @@ total_n <- dim(data_wide)[1]
 total_cancer <- data_wide %>% 
   count(cancer_v) %>% 
   mutate(prop = round(100*n/sum(n),0)) %>% 
-  mutate(cancer_count = paste0(prop, "%", " (n = ", n, ")")) %>% 
+  mutate(cancer_count = paste0(prop, "%", " (n=", n,")")) %>% 
   select(cancer_count, cancer_v)
   
 no_cancer <- total_cancer %>% 
@@ -46,7 +46,7 @@ total_status <- data_wide %>%
   group_by(cancer_v) %>% 
   count(status) %>% 
   mutate(prop = round(100*n/sum(n),0)) %>% 
-  mutate(total = paste0(prop, "%", " (n = ", n, ")")) %>% 
+  mutate(total = paste0(prop, "%", " (n =", n,")")) %>% 
   select(cancer_v, status, total)
 
 cancer_dem <- total_status %>% filter(cancer_v == 1, status == "Cancer and Dementia") %>% pull(3)
@@ -62,12 +62,12 @@ cancer_time <- data_long %>%
   summarize(m = median(age_scale),
             l = quantile(age_scale, 0.25),
             u = quantile(age_scale, 0.75)) %>% 
-  mutate(temp = paste0(m, " (IQR: ", l, "-", u, ")")) %>% 
+  mutate(temp = paste0(m, " (IQR:", l, "-", u,")")) %>% 
   pull(temp)
 
 dem_total <- data_wide %>% count(outcome_plr) %>% 
   mutate(prop = round(100*n/sum(n),0)) %>% 
-  mutate(temp = paste0(prop, "%", " (n = ", n, ")")) %>%
+  mutate(temp = paste0(prop, "%", " (n=", n,")")) %>%
   filter(outcome_plr == 1) %>% 
   pull(temp)
   
@@ -76,23 +76,23 @@ dem_time <- data_long %>%
   summarize(m = median(age_scale),
             l = quantile(age_scale, 0.25),
             u = quantile(age_scale, 0.75)) %>% 
-  mutate(temp = paste0(m, " (IQR: ", l, "-", u, ")")) %>% 
+  mutate(temp = paste0(m, " (IQR:", l, "-", u,")")) %>% 
   pull(temp)
 
 
 ## Baseline covs
 
 mean_age <- data_wide %>% 
-  summarize(a = round(mean(age_0),  2),
[…]

### 2021-07-13 02_R/table_2.R

@@ -6,7 +6,7 @@ hr <-rio::import(here::here("02_R", "hr_results.csv")) %>%
   mutate(Proxy = ifelse(Proxy == "Time to cancer", "Time-varying", Proxy)) %>% 
   mutate_if(is.numeric, round, 2) %>% 
   mutate(Proxy = str_replace(Proxy, "Ever vs. Never", "Ever vs. never")) %>% 
-  mutate(`Hazard Ratio` = paste0(point, "(", conf.low, ",", conf.high, ")")) %>% 
+  mutate(`Hazard Ratio` = paste0(point, " (", conf.low, ",", conf.high, ")")) %>% 
   select(Proxy, model, `Hazard Ratio`)
 
 rr <- ever %>%
@@ -35,9 +35,15 @@ rd <- ever %>%
 complete_table <- rd %>% left_join(rr) %>% left_join(hr)
 
 complete_table <- complete_table %>% 
-  mutate(model = as_factor(model),
-         model = fct_relevel(model, c("Unadjusted", "IPTW", "IPTW + IPCW"))) %>% 
-  group_by(Proxy) %>% 
+  mutate(Proxy = ifelse(Proxy == "Ever vs. never", "A", "B")) %>%
+  rename(Scenario = Proxy) %>% 
+  mutate(model = 
+           case_when(model == "IPTW" ~ "With weights for confounding",
+                     model == "IPTW + IPCW" ~ "With weights for confounding and for censoring for death",
+                     TRUE ~ model),
+         model = as_factor(model),
+         model = fct_relevel(model, c("Unadjusted", "With weights for confounding", "With weights for confounding and for censoring for death"))) %>% 
+  group_by(Scenario) %>% 
   arrange(model)
 
 bounds <- rio::import(here::here("02_R", "table_results_bounds.csv"))
@@ -45,3 +51,5 @@ bounds <- rio::import(here::here("02_R", "table_results_bounds.csv"))
 bounds_tv_ipw <- bounds %>% slice(4,6) %>% select(model, rr) %>% pull(rr)
 
 bounds_tv_ipw <- paste0("RR: ", bounds_tv_ipw[1], ", ", bounds_tv_ipw[2])
+
+

### 2021-07-15 04_reports/manuscript.Rmd
cleaning

@@
 
-Many observational studies have consistently found that individuals with cancer have a lower risk of developing of Alzheimer´s disease or related dementias (ADRD) when compared to individuals with no history of cancer [@ma2014;@hanson2016;@vanderwillik2018;@ospina2020].
+Many observational studies have consistently found that individuals with cancer have a lower risk of developing Alzheimer´s disease and related dementias (ADRD) when compared to individuals with no history of cancer [@ma2014;@hanson2016;@vanderwillik2018;@ospina2020].
 These findings have motivated substantial research toward mechanistic explanations, including molecular and genetic pathways that may explain this association[@behrens2009;@harris2014;@nudelman2019;@Papin2020;@driverbiogeront2014;@olson2019;@li2021;@driverpin12015].
-These research inquiries inevitably lead to discussions of repurposing or augmenting current cancer chemotherapeutics for ADRD [@Snyder2017].
+These research inquiries inevitably lead to discussions of repurposing or augmenting current cancer chemotherapeutic for ADRD [@Snyder2017].
 
@@
 However, understanding these or other sources of bias first requires making explicit the causal question.
-Moreover, this is one step toward tying a research study to a question that is relevant to decision-making[@didelez2016;@labrecque2017].
+Moreover, enumerating an explicit causal question is one step toward tying a research study to a question that is relevant to decision-making[@didelez2016;@labrecque2017].
 
@@
 
-To explore how we might learn about this effect using real-world data on cancer and ADRD, we progressively build a causal directed acyclic graph to connect this particular causal question to the observable data and the assumptions we rely on to study the effect.
+To explore how we might learn about this effect using real-world data on cancer and ADRD, we progressively build a causal directed acyclic graph (DAG) to connect this particular causal question to the observable data and the assumptions we rely on to study the effect.
 We exemplify these challenges and how they translate into the analytic decision using data collected from the Rotterdam Study, a population-based cohort study.
@@
 In the observational setting, confounding could explain the observed association between Pin1 and ADRD.
-In the directed acyclic graph [@whatif2020] of Figure 1, we present Pin1 expression as $P_{t-1}$ and ADRD diagnosis over time as $Y_{t+1}$.
-Both nodes may share causes $L$, and to assess the causal relationship between $P_{t-1}$ and $Y_{t+1}$ we would require adjusting for $L$.
-Previous studies have described age, sex, educational level and race/ethnicity as the minimal adjusting set of covariates[@ospina2020].
-However, environmental and behavioral factors such as smoking may translate into Pin1 over-expression and are also related to the development of ADRD[ref].
+In the causal diagram [@whatif2020] in Figure 1, we present Pin1 expression as $P_{t}$ and ADRD diagnosis by time $t+k$ as $Y_{t+k}$.
+Both may share causes $L_1$, and to assess the causal relationship between $P_{t}$ and $Y_{t+k}$ we would thus adjust for $L_1$.
+Previous ADRD studies have described age, sex, educational level and race/ethnicity as a minimal adjusting set of covariates[@ospina2020].
+However, environmental and behavioral factors such as smoking may translate into Pin1 over-expression[@tan2010] and are also related to the development of ADRD[@dementia_lancet].
 
@@
 \begin{tikzpicture}[node distance=2cm]
-\node (p) {$P_{t-1}$};
-\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
-\node [gray!30, below of = p, yshift = 1cm] (pstar) {$P^*_{t-1}$};
-\node [gray!30, right of = p] (dp) {$D_t$};
-\node [gray!30, right of = dp] (p2) {$P_t$};
-\node [gray!30, below of = p2, yshift = 1cm] (pstar2) {$P^*_t$};
-\node [gray!30, right of = p2] (dp2) {$D_{t+1}$};
-\node [right of = dp2] (y) {$Y_{t+1}$};
-\node [gray!30, below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$C$};
-\draw[->, gray!30] (p) to (pstar);
-\draw[->, gray!30] (p2) to (pstar2);
+\node (p) {$P_t$};
+\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L_1$};
+\node [gray!30, below of = p, yshift = 1cm] (pstar) {$C_t$};
+\node [gray!30, right of = p] (dp) {$D_{t+1}$};
+\node [gray!30, right of = dp] (p2) {$P_{t+1}$};
+\node [gray!30, below of = p2, yshift = 1cm] (pstar2) {$C_{t+1}$};
+\node [gray!30, right of = p2] (dp2) {$D_{t+k}$};
+\node [right of = dp2] (y) {$Y_{t+k}$};
+\node [gray!30, below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$L_2$};
+\draw[gray!30, ->] (p) to (pstar);
+\draw[gray!30, ->] (l) to (dp);
+\draw[gray!30, ->] (pstar) -- (dp);
+\draw[gray!30, ->] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
+\draw[gray!30, ->] (dp) -- (p2);
+\draw[gray!30,->] (p2) to (pstar2);
+\draw[gray!30, ->] (dp2) -- (y);
+\draw[gray!30, ->] (cd) -- (dp2);
+\draw[gray!30, ->] (cd) -- (y);
+\draw[gray!30, ->] (l) -- (p2);
+\draw[gray!30, ->] (l) to [out=0,in=90, looseness=0.4] (dp2);
+\draw[gray!30, ->] (pstar2) -- (dp2);
+\draw[gray!30, ->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
 \draw[->] (l) -- (p);
-\draw[->, gray!30] (l) -- (p2);
-\draw[->, gray!30] (l) to (dp);
-\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (dp2);
-\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->, gray!30] (pstar) -- (dp);
-\draw[->, gray!30] (pstar2) -- (dp2);
-\draw[->, gray!30] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
-\draw[->, gray!30] (dp) -- (p2);
-\draw[->, gray!30] (dp2) -- (y);
-\draw[->, gray!30] (cd) -- (dp2);
-\draw[->, gray!30] (cd) -- (y);
-\draw[->, gray!30, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
-\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
 \end{tikzpicture}
@@
 Unlike measuring Pin1 at the same time to all participants (thought this would not necessarily mean this would be the ideal time to measure it, we discuss this point further in the discussion section), cancer diagnosis is a collected over time.
-We depict this feature in Figure 2, where $P^*_{t-1}$ and $P^*_t$ represent _cancer diagnosis_ over time, the measured proxy of $P_{t-1}$ and $P_t$ respectively.
+We depict this feature in Figure 2, where $C_t$ and $C_{t+1}$ represent _cancer diagnosis_ over time, the measured proxy of $P_{t}$ and $P_{t+1}$ respectively.
 Although this means we would measure the association between cancer diagnosis over time and ADRD in the observed data, we are assuming that the captured effect is only through the pathway that involves Pin1 expression over time.
+Furthermore, even if we have measurements of $C_{t+1}$ and $Y_{t+k}$ and some subset of $L_1$, our question focused on $P_t$ remains.
 
@@
 \begin{tikzpicture}[node distance=2cm]
-\node (p) {$P_{t-1}$};
-\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
-\node [below of = p, yshift = 1cm] (pstar) {$P^*_{t-1}$};
-\node [gray!30, right of = p] (dp) {$D_t$};
-\node [right of = dp] (p2) {$P_t$};
-\node [below of = p2, yshift = 1cm] (pstar2) {$P^*_t$};
-\node [gray!30, right of = p2] (dp2) {$D_{t+1}$};
-\node [right of = dp2] (y) {$Y_{t+1}$};
-\node [gray!30, below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$C$};
+\node (p) {$P_t$};
+\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L_1$};
+\node [below of = p, yshift = 1cm] (pstar) {$C_t$};
+\node [gray!30, right of = p] (dp) {$D_{t+1}$};
+\node [right of = dp] (p2) {$P_{t+1}$};
+\node [below of = p2, yshift = 1cm] (pstar2) {$C_{t+1}$};
+\node [gray!30, right of = p2] (dp2) {$D_{t+k}$};
+\node [right of = dp2] (y) {$Y_{t+k}$};
+\node [gray!30, below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$L_2$};
+\draw[gray!30, ->] (l) to (dp);
+\draw[gray!30, ->] (pstar) -- (dp);
+\draw[gray!30, ->] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
+\draw[gray!30, ->] (dp) -- (p2);
+\draw[gray!30, ->] (dp2) -- (y);
+\draw[gray!30, ->] (cd) -- (dp2);
+\draw[gray!30, ->] (cd) -- (y);
+\draw[gray!30, ->] (l) to [out=0,in=90, looseness=0.4] (dp2);
+\draw[gray!30, ->] (pstar2) -- (dp2);
 \draw[->] (p) to (pstar);
+\draw[->] (l) -- (p);
+\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
 \draw[->] (p2) to (pstar2);
-\draw[->] (l) -- (p);
 \draw[->] (l) -- (p2);
 \draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->, gray!30] (l) to (dp);
-\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (dp2);
-\draw[->, gray!30] (dp) -- (p2);
-\draw[->, gray!30] (pstar) -- (dp);
-\draw[->, gray!30] (pstar2) -- (dp2);
-\draw[->, gray!30] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
-\draw[->, gray!30] (dp2) -- (y);
-\draw[->, gray!30] (cd) -- (dp2);
-\draw[->, gray!30] (cd) -- (y);
-\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
 \draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
@@
 A challenge that arises by choosing cancer diagnosis as the proxy for Pin1 is defining the time zero, the time where eligibility criteria is met, where exposure is measured and the time where the screening for ADRD would begin after having the exposure measured[@hernanInmortal2016].
-The eligibility criteria to participate in a trial for Pin-targeting drug will not necessarily share the same criteria to be recruited for the cohort setting.
+Emulating the eligibility criteria to participate in a trial for Pin-targeting drug will not necessarily be possible in a cohort setting recruiting participants for discordant reasons.
 The latter may not align with the time cancer diagnosis is measured since it happens over follow-up and this situation may introduce inmortal-time bias[@hernanInmortal2016].
@@
 
-This problem is avoided by recognizing the time-varying nature of cancer diagnosis.
+This problem is partly avoided by recognizing the time-varying nature of cancer diagnosis.
 Several studies have considered cancer diagnosis as a time-dependent exposure.
@@
 Moreover, cancer diagnosis will only be measured in the subset of participants who are alive over follow-up.
-Thus, we included death prior to cancer diagnosis as $D_t$ and an arrow between $D_t$ and $P_t$ that represents a deterministic association such as that $P_t$ is only observed if $D_t$ is zero in Figure 3.
-In addition, we added an arrow between $L$ and $D_t$, since covariates such as smoking may affect Pin1 over-expression but also affect the risk of death due to other causes such as from chronic obstructive pulmonary disease.
-Although $L$ is time-varying in nature, we only depict $L$ at one time-point for readability.
+Thus, we included death prior to cancer diagnosis as $D_{t+1}$ and an arrow between $D_{t+1}$ and $P_{t+1}$ that represents a deterministic association such as that $P_{t+1}$ is only observed if $D_{t+1}$ is zero in Figure 3.
+In addition, we added an arrow between $L_1$ and $D_{t+1}$, since covariates such as smoking may affect Pin1 over-expression but also affect the risk of death due to other causes such as from chronic obstructive pulmonary disease.
+Although $L_1$ can be time-varying in nature, we only depict $L_1$ at one time-point for readability.
 
@@
 \begin{tikzpicture}[node distance=2cm]
-\node (p) {$P_{t-1}$};
-\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
-\node [below of = p, yshift = 1cm] (pstar) {$P^*_{t-1}$};
-\node [right of = p] (dp) {$D_t$};
-\node [right of = dp] (p2) {$P_t$};
-\node [below of = p2, yshift = 1cm] (pstar2) {$P^*_t$};
-\node [gray!30, right of = p2] (dp2) {$D_{t+1}$};
-\node [right of = dp2] (y) {$Y_{t+1}$};
-\node [gray!30, below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$C$};
+\node (p) {$P_t$};
+\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L_1$};
+\node [below of = p, yshift = 1cm] (pstar) {$C_t$};
+\node [right of = p] (dp) {$D_{t+1}$};
+\node [right of = dp] (p2) {$P_{t+1}$};
+\node [below of = p2, yshift = 1cm] (pstar2) {$C_{t+1}$};
+\node [gray!30, right of = p2] (dp2) {$D_{t+k}$};
+\node [right of = dp2] (y) {$Y_{t+k}$};
+\node [gray!30, below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$L_2$};
+\draw[gray!30, ->] (pstar) -- (dp);
+\draw[gray!30, ->] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
+\draw[gray!30, ->] (dp2) -- (y);
+\draw[gray!30, ->] (cd) -- (dp2);
+\draw[gray!30, ->] (cd) -- (y);
+\draw[gray!30, ->] (l) to [out=0,in=90, looseness=0.4] (dp2);
+\draw[gray!30, ->] (pstar2) -- (dp2);
 \draw[->] (p) to (pstar);
-\draw[->] (p2) to (pstar2);
 \draw[->] (l) -- (p);
-\draw[->] (l) -- (p2);
 \draw[->] (l) to (dp);
 \draw[->] (dp) -- (p2);
-\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (dp2);
+\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->] (p2) to (pstar2);
+\draw[->] (l) -- (p2);
 \draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->, gray!30] (pstar) -- (dp);
-\draw[->, gray!30] (pstar2) -- (dp2);
-\draw[->, gray!30] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
-\draw[->, gray!30] (dp2) -- (y);
-\draw[->, gray!30] (cd) -- (dp2);
-\draw[->, gray!30] (cd) -- (y);
 \draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
-\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
 \end{tikzpicture}
@@
 We now move to the challenge of death as a competing event for ADRD, represented in Figure 4.
-For interpretability we exclude the time-varying process of cancer diagnosis and focus on Pin1 (and cancer diagnosis) as it had been measured in all participants at time $t$.
+For interpretability we exclude the time-varying process of cancer diagnosis and focus on Pin1 (and cancer diagnosis) as it had been measured in all participants at time $t+1$.
 Given that some participants may die prior to ADRD diagnosis, we can only measure ADRD over follow-up in the individuals who survive long enough to have an ADRD diagnosis.
-In the causal diagram of Figure 4 we include a node for death after the exposure $P^*_t$ has been measured, represented as $D_{t+1}$ and which acts as a competing event of $Y_{t+1}$ because if a participant dies by $t + 1$, the participant cannot subsequently develop dementia.
-Furthermore, since $D_{t+1}$ and $Y_{t+1}$ are events related to aging and its consequences, $C$ represents the shared causes of both events such as cardiovascular conditions.
-We also include an arrow between $L$ and $D_{t+1}$ following the argument discussed previously.
+In the causal diagram of Figure 4 we include a node for death after the exposure $C_{t+1}$ has been measured, represented as $D_{t+k}$ and which acts as a competing event of $Y_{t+k}$ because if a participant dies by $t + k$, the participant cannot subsequently develop dementia.
+Furthermore, since $D_{t+k}$ and $Y_{t+k}$ are events related to aging and its consequences, $L_2$ represents the shared causes of both events such as cardiovascular conditions.
+We also include an arrow between $L_1$ and $D_{t+1}$ following the argument discussed previously.
 
-In the setting where $P_t$ represented the targeted-drug for Pin1, and if this drug had no systemic beneficial or harmful side-effects such as that there is no arrow between $P_t$ and $D_{t+1}$, a total effect would quantify the effect of $P_t$ on $Y_{t+1}$ that does not include any pathway mediated through $D_{t+1}$[@young2020].
-However, in the context of cancer diagnosis as the proxy for Pin1 over-expression, we cannot rule out the effect of cancer diagnosis on death, represented as the arrow between $P^*_t$ and  $D_{t+1}$.
-In this setting, a total effect of $P_t$ in $Y_{t+1}$ would include the causal pathway mediated by the effect of cancer diagnosis in mortality, which may translate into an inverse association[@young2020].
+In the setting where $P_{t+1}$ represented the targeted-drug for Pin1, and if this drug had no systemic beneficial or harmful side-effects such as that there is no arrow between $P_{t+1}$ and $D_{t+k}$, a total effect would quantify the effect of $P_{t+1}$ on $Y_{t+k}$ that does not include any pathway mediated through $D_{t+k}$[@young2020].
+However, in the context of cancer diagnosis as the proxy for Pin1 over-expression, we cannot rule out the effect of cancer diagnosis on death, represented as the arrow between $C_{t+1}$ and $D_{t+k}$.
+In this setting, a total effect of $P_{t+1}$ in $Y_{t+k}$ would include the causal pathway mediated by the effect of cancer diagnosis in mortality, which may translate into an inverse association[@young2020].
 
-To isolate the direct effect of $P_t$ in $Y_{t+1}$ through measurement of $P^*_t$ we have at least two alternatives of causal questions we can ask.
+To isolate the direct effect of $P_{t+1}$ in $Y_{t+k}$ through measurement of $C_{t+1}$ we have at least two alternatives of causal questions we can ask.
 We can either imagine a causal question where we decompose the effect of cancer by the different mechanisms that affect dementia and death separately[@stensrud2020].
 Alternatively, we can define an scenario where death could have been prevented.
-The latter is defined as the controlled direct effect, where death is considered as a censoring event and it relies on the assumption that we have measured all $C$ to block the pathway $Y_{t+1} \leftarrow C  \rightarrow D_{t+1} \rightarrow P^*_t \rightarrow P_t$.
+The latter is defined as the controlled direct effect, where death is considered as a censoring event and it relies on the assumption that we have measured all $L$ to block the pathway $Y_{t+k} \leftarrow L_2  \rightarrow D_{t+k} \rightarrow C_{t+1} \rightarrow P_{t+1}$.
 Previous studies have defined death as a censoring event [@frain2017], although failed to explicitly define how to comply with the independent censoring assumption.
-Moreover, adjusting for confounders for dementia (such as only adjusting for $L$) will be insufficient to block the pathway mediated by death, time-varying covariates that represent $C$, such as cardiovascular conditions should be considered.
+Moreover, adjusting for confounders for dementia (such as only adjusting for $L$) will be insufficient to block the pathway mediated by death, time-varying covariates that represent $L_2$, such as cardiovascular conditions should be considered.
 
@@
 \begin{tikzpicture}[node distance=2cm]
-\node [gray!30](p) {$P_{t-1}$};
-\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
-\node [gray!30, below of = p, yshift = 1cm] (pstar) {$P^*_{t-1}$};
-\node [gray!30, right of = p] (dp) {$D_t$};
-\node [right of = dp] (p2) {$P_t$};
-\node [below of = p2, yshift = 1cm] (pstar2) {$P^*_t$};
-\node [right of = p2] (dp2) {$D_{t+1}$};
-\node [right of = dp2] (y) {$Y_{t+1}$};
-\node [below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$C$};
-\draw[->, gray!30] (p) to (pstar);
-\draw[->, gray!30] (l) -- (p);
-\draw[->, gray!30] (l) to (dp);
-\draw[->, gray!30] (pstar) -- (dp);
-\draw[->, gray!30] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
-\draw[->, gray!30] (dp) -- (p2);
-\draw[->, semithick, gray!30] (p) to [out=+45,in=+135, looseness=0.3] (y);
+\node [gray!30](p) {$P_t$};
+\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L_1$};
+\node [gray!30, below of = p, yshift = 1cm] (pstar) {$C_t$};
+\node [gray!30, right of = p] (dp) {$D_{t+1}$};
+\node [right of = dp] (p2) {$P_{t+1}$};
+\node [below of = p2, yshift = 1cm] (pstar2) {$C_{t+1}$};
+\node [right of = p2] (dp2) {$D_{t+k}$};
+\node [right of = dp2] (y) {$Y_{t+k}$};
+\node [below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$L_2$};
+\draw[gray!30, ->] (p) to (pstar);
+\draw[gray!30, ->] (l) -- (p);
+\draw[gray!30, ->] (l) to (dp);
+\draw[gray!30, ->] (pstar) -- (dp);
+\draw[gray!30,->] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
+\draw[gray!30, ->] (dp) -- (p2);
+\draw[gray!30, ->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
 \draw[->] (p2) to (pstar2);
@@
 To summarize, the complexity of the causal structure that describes the effect of Pin1 through the proxy of cancer diagnosis in the risk of ADRD illustrate the potential sources of bias, as observed in Figure 5.
-Even so, this is a simplified version since we omitted other sources of measurement error and the time-varying nature of all nodes and feedback loops between them which would further complicate interpretability[@whatif2020].
+Even so, this is a simplified version since we omitted additional arrows from $L_1$ to $C_t$ and $C_{t+2}$ for brevity, as well as other sources of measurement error and the time-varying nature of all nodes and feedback loops between them, which would further complicate interpretability[@whatif2020].
 
@@
 \begin{tikzpicture}[node distance=2cm]
-\node (p) {$P_{t-1}$};
-\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
-\node [below of = p, yshift = 1cm] (pstar) {$P^*_{t-1}$};
-\node [right of = p] (dp) {$D_t$};
-\node [right of = dp] (p2) {$P_t$};
-\node [below of = p2, yshift = 1cm] (pstar2) {$P^*_t$};
-\node [right of = p2] (dp2) {$D_{t+1}$};
-\node [right of = dp2] (y) {$Y_{t+1}$};
-\node [below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$C$};
+\node (p) {$P_t$};
+\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L_1$};
+\node [below of = p, yshift = 1cm] (pstar) {$C_t$};
+\node [right of = p] (dp) {$D_{t+1}$};
+\node [right of = dp] (p2) {$P_{t+1}$};
+\node [below of = p2, yshift = 1cm] (pstar2) {$C_{t+1}$};
+\node [right of = p2] (dp2) {$D_{t+k}$};
+\node [right of = dp2] (y) {$Y_{t+k}$};
+\node [below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$L_2$};
 \draw[->] (p) to (pstar);
@@
 We now turn to an application where we show how these challenges translate into analytic decisions.
-We will show ways to mitigate or better understand the best of the available data's abilities to inform the possible effect of Pin1 on all cause-dementia.
+We will show ways to mitigate or better understand bias to the best of the available data's abilities in an attempt to inform the possible effect of Pin1 on all cause-dementia.
 
-## 2.2. The Rotterdam Study
+### 2.2. The Rotterdam Study
 
 We use data collected in the Rotterdam Study, a population-based prospective cohort study among persons living in the Ommoord district in Rotterdam, the Netherlands.
-Recruitment and initial assessments were held between 1990 and 1993, a second waive of recruitment was held between 2000 and 2001.
-Participants from first subcohort had follow-up visits between 1993-1995, 1997-1999, 2002-2005, and 2008-2010, second subcohort had follow-up visits between 2004 and 2005, and between 2011 and 2012[@ikram2020].
-All participants had data on history of cancer and dementia and incident diagnosis through follow-up until 2015, collected from medical records of general practitioners (including hospital discharge letters) and through linkage with national registries.
+Recruitment and initial assessments were held between 1990 and 1993, a second wave of recruitment was held between 2000 and 2001.
+Participants from the first subcohort had follow-up visits between 1993-1995, 1997-1999, 2002-2005, and 2008-2010, while the second subcohort had follow-up visits between 2004 and 2005, and between 2011 and 2012[@ikram2020].
+All participants had data on history of cancer and dementia and incident diagnosis through follow-up until 31 of December, 2015, collected from medical records of general practitioners (including hospital discharge letters) and through linkage with national registries.
 Date and cause of death was collected on a weekly basis via municipal population registries.
+These ascertainment methods imply that the Rotterdam Study has functionally no loss to follow-up with respect to dementia diagnosis and death.
 
-We defined as inclusion criteria being between 60 and 70 years old at study entry, without history of cancer diagnosis, free of cognitive decline or with previous history of dementia.
+Eligibility criteria included: ages 60 to 70 years at study entry; no history of cancer diagnosis, no history of dementia diagnosis; and free of cognitive decline (defined by a Mini Mental Score <26).
 Out of 10998 persons who participated at study entry, `r total_n` were considered eligible.
@@
 All participants were followed from study entry until dementia diagnosis, death or 20 years after their individual baseline date, whichever occurred first.
+Given that participants of the second wave were followed for 15 years on average, the 20 year-risk of dementia includes extrapolation of the risk for the second cohort, under the assumption that participants from second cohort would have had a similar distribution of dementia and mortality than the first cohort, between year 15 and 20.
 
@@
 
-We illustrate the association between cancer and dementia diagnosis under two scenarios.
+We illustrate the association between cancer and dementia diagnosis under two scenarios, the first of which replicates a common analytic strategy and the second which mitigates some (but not all) the biases described above.
 Scenario A replicates the setting that defines cancer proxy as _"cancer ever vs.
-never"_ [@driver2012].
-Scenario B defines cancer diagnosis as time-varying.
-To address confounding we fit inverse probability treatment weights, stabilized and truncated at 99th percentile.
-In scenario A, weights were defined as the inverse of the probability of cancer diagnosis conditional on baseline covariates such as age at study entry, sex, educational attainment, cohort, smoking status.
-In contrast, for scenario B weights were defined to represent the product of the inverse probability of being diagnosed with cancer over time, conditional on the time-varying covariate history prior to cancer diagnosis.
-This recreates a pseudopopulation in which time-varying covariates do not predict the risk of cancer diagnosis at time _t_ [@hernan2000].
-Baseline covariates included age at study entry, sex, ApoE4 status, educational attainment and the time-varying covariates such as smoking status, systolic blood pressure, BMI  and prevalent and incident comorbidities such as: hypertension and diabetes.
-Further details on modeling specifications and weights assessment are presented as **Supplementary material x**.
+never"_ [@driver2012], meaning we compare dementia outcomes in individuals who ever develop cancer during follow-up to those who were not observed to develop cancer during follow-up.
+Scenario B defines cancer diagnosis as time-varying, meaning ....
+To address confounding, we fit inverse probability treatment weights, stabilized and truncated at the 99th percentile.
+In Scenario A, weights were defined as the inverse of the probability of cancer diagnosis conditional on baseline covariates such as age at study entry, sex, educational attainment, cohort, smoking status.
+In contrast, for Scenario B weights were defined to represent the product of the inverse probability of being diagnosed with cancer over time, conditional on the time-varying covariate history prior to cancer diagnosis [@hernan2000].
+Baseline covariates included age at study entry, sex, ApoE4 status, educational attainment and the time-varying covariates such as smoking status, systolic blood pressure, BMI  and prevalent and incident hypertension and diabetes.
+Further details on modeling specifications and weights assessment are presented as **Supplementary material**.
 
 Inverse probability censoring weights for death were calculated to relax the independent censoring assumption.
-In scenario A, weights represent the inverse of the probability of not dying conditional on cancer diagnosis (ever vs.
+In Scenario A, weights represent the inverse of the probability of not dying conditional on cancer diagnosis (ever vs.
 never) and baseline covariates such as age, educational attainment, ApoE4, and baseline measurements of smoking status, hypertension status, systolic blood pressure, BMI, history of diabetes and cohort.
 For individuals who died, their censoring weight is zero [@whatif2020].
-In scenario B time-varying weights represent the product of the inverse probability of surviving in each year prior to _t_, conditional on the measured shared causes of death and dementia.
+In Scenario B time-varying weights represent the product of the inverse probability of surviving in each year prior to _t_, conditional on the measured shared causes of death and dementia.
 For an individual who has died by time _t_, the year _t_ censoring weight is zero [@young2020].
@@
 For illustrative and comparative purposes we calculated hazard ratios (HR).
-Hazards, unlike risks, inherently condition on surviving both dementia and death, as such they will not have a causal interpretation in this setting [@young2020].
-We additionally calculated the controlled direct effect considering death as an unconditional independent censoring event (as if there were no arrows from $C$ to $D_{t+1}$ and $Y_{t+1}$).
+Hazards, unlike risks, inherently condition on surviving both dementia and death, as such a causal interpretation is problematic [@young2020].
 
@@
 
-All analysis were performed using R, code is provided in supplementary material and available in https://github.com/palolili23/2021_cancer_dementia.
+All analysis were performed using R; code is provided in supplementary material and available in https://github.com/palolili23/2021_cancer_dementia.
 
@@
 
-Participants had a mean age of `r mean_age`, and `r women_prop` were women.
-Further details on participants are presented in Table 1.
-Over follow-up, `r yes_cancer` developed cancer and `r no_cancer` remain free of cancer diagnosis, the median age of cancer diagnosis was `r cancer_time`.
-From the total sample, `r dem_total` had dementia over follow-up and median time to dementia was `r dem_time`.
-Among participants with incident cancer, `r cancer_dem` had dementia diagnosis and `r cancer_death` died over follow-up, `r cancer_alive` remain alive at 20 years since study entry.
-In contrast, among participants free of cancer diagnosis over follow-up, `r no_cancer_dem` were diagnosed with dementia and `r no_cancer_death` died over follow-up, `r no_cancer_alive` were alive at the end of follow-up.
-Figure 6 shows the proportion of participants in each possible status over follow-up.
+Participants had a mean age of `r mean_age` years, and `r women_prop` were women (Table 1).
+Over follow-up, `r yes_cancer` developed cancer, with a median age of cancer diagnosis at `r cancer_time`.
+From the total sample, `r dem_total` had dementia over follow-up with a median age of at `r dem_time` years.
+Among participants with incident cancer, `r cancer_dem` had dementia diagnosis, `r cancer_death` died over follow-up, and `r cancer_alive` remain alive and dementia-free at 20 years since study entry.
+In contrast, among participants free of cancer diagnosis over follow-up, `r no_cancer_dem` were diagnosed with dementia, `r no_cancer_death` died over follow-up, and `r no_cancer_alive` were alive and dementia-free at the end of follow-up.
+**Figure 6 shows the proportion of participants in each possible status over follow-up.**
 
 Results for all scenarios are present in Table 2.
-Had we defined the proxy for Pin1 as _ever cancer_, and relying on death as independent censoring event (unconditional), we observe a significant protective effect of ever having cancer in the risk of dementia [RR: `r complete_table[3,4]`, HR:  `r complete_table[3,5]`].
-This effect shifts towards the null after including censoring weights for death [RR: `r complete_table[5,4]`; HR: `r complete_table[5,5]`].
-In contrast, had we defined cancer diagnosis as a _time-varying_ proxy for Pin1 over-expression and had we prevented death conditional on covariates, the risk of dementia had participants had a cancer diagnosis over time is higher [RR: `r complete_table[6,4]`; HR: `r complete_table[6,5]`], though confidence intervals cross the null.
-In addition, the extreme scenarios of dependence between death and dementia over follow-up represented as bounds are wide and include the null value (`r bounds_tv_ipw`).
+Using Scenario A’s analytic design and without adjusting for confounding or selection bias due to conditioning on death, we observe a protective association with a risk ratio (RR, 95% Confidence interval) of `r complete_table[3,4]` and a hazard ratio [HR, (95% Confidence interval)] of `r complete_table[3,5]`.
+Though adjusting for measured confounding only minimally changes the observed association, the association is closer to the null after including censoring weights for death [RR: `r complete_table[5,4]`; HR: `r complete_table[5,5]`].
+In contrast, using Scenario B's analytic design, the fully adjusted model results in a RR of RR: `r complete_table[6,4]` and a HR of `r complete_table[6,5]`, though confidence intervals cross the null.
+~~In addition, the extreme scenarios of dependence between death and dementia over follow-up represented as bounds are wide and include the null value (`r bounds_tv_ipw`).~~
 
@@
 Instead, this variable is commonly used to represent a mechanism of interest that could not be measured.
-In this study we define cancer diagnosis as the proxy for Pin1 expression, a potential therapeutic drug-target.
+In this study we describe a particular research question of investigating a potential therapeutic drug-target of Pin1 expression, and discuss using cancer diagnosis as a proxy.
 By explicitly including Pin1 as part of the research question we connect the unmeasured mechanism of interest to the observed data outlining the data generation process.
 This practice helps to identify and disentangle potential sources of bias and can guide analytic decisions.
-We show how results of the association between cancer diagnosis and dementia can change substantially according to alternative, yet explicit, assumptions.
+We showed how estimates can change substantially according to alternative, yet explicit, assumptions.
 
@@
 In the setting where cancer diagnosis is defined as "ever vs.
-never", this definition implicitly introduces immortal time bias.
-All results pertaining to this definition showed a substantial inverse association between cancer diagnosis and dementia in contrast to cancer diagnosis as a time-varying process which shifted point estimates close to the null.
-In line with the latter definition, including time-varying covariates is possible with well-known statistical methods such as inverse probability weighting, as opposed to previous studies that only considered baseline information[lots of refs].
+never", this definition inntroduces immortal time bias.
+All results pertaining to this definition had an inverse association between cancer diagnosis and dementia, while results that did not introduce this particular form of immortal-time bias had point estimates close to the null.
 Although we attempt to prevent this bias with statistical methods, we can only fully prevent it by having a clear definition of time-zero.
@@
 
-On the other hand, death as competing event is a challenge that has a straightforward solution, which begins by choosing the causal parameters (or estimand) of interest [@young2020, @rojas_medrxiv].
-In this study we chose the controlled direct effect, which represents the effect of Pin1 (or cancer) in a setting where death could have been prevented and relies on the independent censoring assumption since death is treated as a censoring event.
-As opposed to prior studies that treated censoring as ignorable[@frain2017], we show how point estimates change substantially when we include IPCW for death to relax this assumption.
-In contrast to prior studies who have defined "survival bias" or "competing event bias" and have proposed different estimators to account for them[@ospina2020], we conceptualize bias as not meeting the independent censoring assumption conditional on covariates[@vangeloven2014, @willems2018, @young2020].
+On the other hand, death as competing event is a challenge that has some straightforward strategies, beginning first and foremost by choosing the causal parameter (or estimand) of interest [@young2020, @rojas_medrxiv].
+In this study we chose the controlled direct effect, which represents the effect of Pin1 (or cancer) in a setting where death due to cancer and other causes could have been prevented, yet without an explicit intervention, which is yet ambiguous.
+This is different than conceiving a drug-target that increases Pin1 expression only brain tissue, with no side effects that could increase cancer risk (and thus, death due to cancer).
+As opposed to prior studies that implicitly address a direct effect, and who define censoring as ignorable[@frain2017], we show how point estimates change substantially when we include weights for death to relax the independent censoring assumption.
+In contrast to prior studies that conceptualize bias around estimators or measures of effect (rates vs.
+cumulative incidence)[@ospina2020], we define bias as failing to relax the independent censoring assumption conditional on covariates[@vangeloven2014, @willems2018, @young2020] when the estimand of interest is the controlled direct effect.
 Bounds to assess extreme scenarios of dependence between death and dementia[@peterson1976] illustrate the wide range of possible point estimates that cross the null.
 This shows that even with the effort of adjusting for time-varying covariates, we may be far from meeting this assumption and thus better efforts to measure shared causes of dementia and death are needed.
-In addition, presenting the proportion of participants in each status over time, and the proportion of participants that died prior to ADRD diagnosis in each arm will always provide insights and improve transparency of results and interpretations.
+In addition, presenting the proportion of participants in each status over time, and the proportion of participants that died prior to ADRD diagnosis in each arm provides insights and improves transparency of results and interpretations.
 
 Pin1 is only one of the several mechanisms proposed about the inverse cancer-dementia relationship.
-We acknowledge that cancer diagnosis represents a complex and heterogeneous health condition that exceeds the representation of Pin1 expression, thus, at best we may test the sharp null hypothesis rather than estimating the causal effect of Pin1.
+Certainly, cancer diagnosis represents a complex and heterogeneous health condition that exceeds the representation of Pin1 expression.
 Furthermore, alternative mechanisms underlying cancer diagnosis and its connection to collected data may be represented in different causal representations.
-For example, if the intention was to explain the effect of different chemotherapeutics and how these affect brain, the eligibility criteria would change and the causal graph would represent the downstream of cancer diagnosis.
-For this reason, we believe is crucial to have a discussion related to cancer diagnosis as an exposure and how we can shift efforts to collect more suitable data.
-Furthermore, we consider that this case-study extends to other questions that study the effect of one disease in the risk of another disease to understand the biological mechanisms behind them.
+To understand how – if at all – the cancer-dementia association informs the potential effects of any other mechanism or treatment strategy would require deep consideration of each causal question anew.
+For example, if researchers used this association to inform the possible effects of different chemotherapeutics on cognitive decline among patients undergoing treatment for cancer, different challenges would arise for mapping the observed association to the hypothetical randomized trial underlying this research question.
+Notably, that target trial, unlike the one considered here, would include cancer diagnosis as part of the eligibility, and thus researchers would need to instead grapple with how using data on persons without cancer is useful and useable [@huitfeldt2016].
+Because each question requires thinking about bias anew and each question brings its own set of challenges and opportunities, we believe it is crucial to have the discussion about the question or set of questions that are relevant first, as this can also inform efforts to collect more suitable data.
+Furthermore, we underscore that this is just one case study for how observed associations between two diseases or health states may be disentangled to more transparently unveil possible mechanisms (and sources of bias) behind them.
+
+\newpage
+
+## References
+
+<div id="refs"></div>
 
@@
 ```{r, echo = FALSE}
-tableone_exp <-
-  print(
-    tableone,
-    explain = TRUE,
-    test = FALSE,
-    quote = FALSE,
-    noSpaces = TRUE,
-    printToggle = FALSE
-  )
-
-Characteristics <- rownames(tableone_exp)
-
-tableone_exp%>% as_tibble() %>% 
-  cbind(Characteristics) %>% select(Characteristics, everything()) %>% 
-  mutate(
-    Characteristics = str_replace(Characteristics, "smoke1", "Smoking status"),
-    Characteristics = str_replace(Characteristics, "sex = ", ""),
-    Characteristics = str_replace(Characteristics, "age_0", "Age at baseline"),
-    Characteristics = str_replace(Characteristics, "education", "Educational attainment"),
-    Characteristics = str_replace(Characteristics, "apoe4", "ApoE4"),
-    Characteristics = str_replace(Characteristics, "bmi1", "Body Mass Index"),
-    Characteristics = str_replace(Characteristics, "sbp1", "Systolic blood pressure (mmHg)"),
-    Characteristics = str_replace(Characteristics, "ht1 = ", ""),
-    Characteristics = str_replace(Characteristics, "hd_prev = ", ""),
-    Characteristics = str_replace(Characteristics, "hd_v = ", ""),
-    Characteristics = str_replace(Characteristics, "diab_v = ", ""),
-    Characteristics = str_replace(Characteristics, "Unknown", "Unknown history of diabetes"),
-    Characteristics = str_replace(Characteristics, "stroke_prev = ", ""),
-    Characteristics = str_replace(Characteristics, "stroke_v = ", ""),
-  ) %>% 
-  filter(!Characteristics %in% c("oh1 (mean (SD))", "diabetes_prev (%)", "cancer_v = No incident cancer (%)")) %>% 
+tableone_complete %>% 
   flextable() %>% 
@@
 
-### Table 2. Risk difference and risk ratio for the risk of dementia
+### Table 2. 20-year risk differences, risk ratio, and hazard ratios for dementia risk by cancer proxy and model adjustment
 
@@
   flextable() %>% 
-  autofit()
+  autofit() %>% 
+  add_footer_lines("Scenario A represents the setting where cancer is defined as 'ever vs. never'; Scenario B represents the setting where cancer is defined as a time-varying proxy")
 ```
@@
 
-## Figures
+\newpage
 
-### Figure 6: Distribution of participants under each health status, by age over follow-up
+## Supplementary material
+
+### A. Modeling description
+
+### B. Distribution of participants under each health status, by age over follow-up
 
@@
 
-\newpage
+### Cumulative incidence plots

### 2021-07-15 02_R/04_crude_net_risks.R

@@ -1,313 +1,0 @@
-library(ggplot2)
-library(tidyverse)
-library(rio)
-library(wesanderson)
-library(survival)
-library(survminer)
-library(broom)
-library(splines)
-library(magrittr)
-
-# Import data -------------------------------------------------------------
-source(here::here("02_R", "04b_auxiliary_fx.R"))
-
-data_long <-
-  import(here::here("01_data", "clean_data", "dementia_long.RData"))
-
-data_wide <-
-  data_long %>%
-  group_by(id) %>% 
-  slice(n()) %>% 
-  ungroup()
-
-data_long %<>% 
-  group_by(id) %>% 
-  mutate(
-    fuptime = time - 1,
-    tstart = lag(fuptime),
-    tstart = ifelse(is.na(tstart), -1, tstart)) %>% 
-  ungroup()
-
-## idea of how numbers look like over follow-up
-
-data_long %<>% 
-  mutate(both_outcomes = 
-           case_when(outcome_plr == 1 ~ 1,
-                     competing_plr == 1 ~2,
-                     TRUE ~ 0))
-
-data_wide %<>%
-  mutate(dementia_efu =
-           case_when(outcome_plr == 1 ~ 1,
-                     competing_plr == 1 ~ 2,
-                     TRUE ~ 0))
-
-data_wide %<>%
-  mutate(t2dem_efu =
-           ifelse(t2dem >= 240, 240, t2dem))
-
-
-# 0. Weights for death as competing event of dementia -------------------------------------------------------
-
-death_den <- glm(competing_plr ~ 
-                   cancer_v + bs(time,3) + bs(age_0,3) + sex + education + apoe4 + 
-                   bs(sbp,3) + bs(bmi,3 ) = bs(oh,3 )+ hd_v + stroke_v + diab_v, 
-                 data = data_long, family = quasibinomial)
-
-summary(death_den)
-
-death_den %>% broom::tidy() %>% View()
-
-death_num <- glm(competing_plr ~ 1, data = data_long, family = binomial)
-
-data_long$p_denom = predict(death_den, data_long, type = "response")
-
-data_long$p_num = predict(death_num, data_long, type = "response")
-
-data_long %<>%
-  group_by(id) %>%
-  mutate(w = cumprod(1 - p_num) / cumprod(1 - p_denom),
-  ) %>% 
-  ungroup()
-
-data_long %<>% 
[…]

### 2021-07-15 02_R/02_long.R
extending to 20 yf instead of 19, added lag values

@@ -17,8 +17,8 @@ data_long <- data %>%
          sex, age_0, education, apoe4, 
          hd_prev, hd_date, diabetes_prev, diabetes_date, stroke_prev, 
          stroke_date) %>% 
-  filter(t2death >= 1) %>% 
   mutate(t2death_y = round(t2death_y,0)) %>% 
+  filter(t2death_y >= 1) %>% 
   group_by(id) %>% 
   slice(rep(1:n(), each = t2death_y + 2)) %>% #repeat rows by #years until death
   mutate(year = year(e1),
@@ -129,6 +129,28 @@ data_long_2015 %<>%
     death_v = ifelse(is.na(death_v),0, death_v)) %>% 
   ungroup()
 
+data_long_2015 %<>% 
+  group_by(id) %>% 
+  mutate(cancer_lag = lag(cancer_v),
+         cancer_lag = ifelse(is.na(cancer_lag), 0, cancer_lag),
+         stroke_lag = lag(stroke_v),
+         stroke_lag = ifelse(is.na(stroke_lag), 0, stroke_lag),
+         hd_lag = lag(hd_v),
+         hd_lag = ifelse(is.na(hd_lag), 0, hd_lag),
+         diab_lag = lag(diab_v),
+         diab_lag = ifelse(is.na(diab_lag), 0, diab_lag),
+         sbp_lag = lag(sbp),
+         sbp_lag = ifelse(is.na(sbp_lag), sbp, sbp_lag),
+         bmi_lag = lag(bmi),
+         bmi_lag = ifelse(is.na(bmi_lag), bmi, bmi_lag),
+         smoke_lag = lag(smoke),
+         smoke_lag = ifelse(is.na(smoke_lag), smoke, smoke_lag),
+         ht_lag = lag(ht),
+         ht_lag = ifelse(is.na(ht_lag), ht, ht_lag),
+         ht_drug_lag = lag(ht_drug),
+         ht_drug_lag = ifelse(is.na(ht_drug_lag), ht_drug, ht_drug_lag),
+         ) %>% 
+  ungroup()
 
 data_long_2015 %>%
   select(
@@ -147,6 +169,7 @@ data_long_2015 %>%
     death_year,
     t2death_y
   ) %>% View()
+
 #### Create a "status" variable
 
 data_long_2015 %<>% 
@@ -178,7 +201,7 @@ data_long_dem <- data_long_2015 %>%
   mutate(end_dementia_death = ifelse(!is.na(dementia_date), 
[…]

### 2021-07-15 02_R/02_long.R
need to exclude anyone who dies at time 1 (n=4) else I need to reconceptualize things as months

@@ -213,7 +213,12 @@ data_long_dem %<>%
     outcome_plr = ifelse(end_dementia_death == year & dementia == 1, 1, 0),
     competing_plr = ifelse(!is.na(dementia_date), 0, death_v)) %>%
   ungroup() %>%
+  filter(!time == 0 & !competing_plr == 1) %>%
   arrange(id, time)
+
+id2 <- data_long_dem %>% 
+  filter(time == 0 & competing_plr == 1) %>% 
+  pull(id)
 
 data_long_dem %>% 
   count(outcome_plr, competing_plr)

### 2021-07-15 02_R/04_direct_effect_simplified.R
change all models to lags, export 3rd model for A and B

@@ -9,7 +9,7 @@ library(WeightIt)
 library(cobalt)
 
 # Import data -------------------------------------------------------------
-source(here::here("02_R", "04b_auxiliary_fx.R"))
+source(here::here("02_R", "04_auxiliary_fx.R"))
 
 data_long <-
   import(here::here("01_data", "clean_data", "dementia_long.RData"))
@@ -280,8 +280,8 @@ data_long %<>%
 
 # denominator
 mod <- glm(cancer_v ~ bs(time, 3) + bs(age_0, 3) + sex + education + apoe4 +
-             as.factor(smoke) + bs(sbp, 3) + bs(bmi, 3) + 
-             ht + ht_drug + diab_v + cohort,
+             as.factor(smoke_lag) + bs(sbp_lag, 3) + bs(bmi_lag, 3) + 
+             ht_lag + ht_drug_lag + diab_lag + cohort,
            family = quasibinomial(), data = subset(data_long , 
                                                    pastcancer == 0))
 
@@ -333,21 +333,23 @@ data_long %>%
 # 2.2. IPCW. Weights on death - dementia ---------------------------------------------
 
 death_den <- glm(
-  competing_plr ~ cancer_v + bs(time, 3) + bs(age_0, 3) + sex + education + apoe4 +
-    as.factor(smoke) + bs(sbp, 3) + bs(bmi, 3) + 
-    ht + ht_drug + hd_v + stroke_v + diab_v + cohort,
+  competing_plr ~ cancer_v + cancer_lag + bs(time, 3) + bs(age_0, 3) + sex + education + apoe4 +
+    as.factor(smoke_lag) + bs(sbp_lag, 3) + bs(bmi_lag, 3) + 
+    ht_lag + ht_drug_lag + hd_lag + stroke_lag + diab_lag + cohort,
   data = data_long,
   family = quasibinomial
 )
 
-summary(death_den)
+broom::tidy(death_den, exponentiate = TRUE)
 
 death_num <-
   glm(
-    competing_plr ~ cancer_v + bs(time, 3) + cohort,
+    competing_plr ~ cancer_lag + bs(time, 3) + cohort,
     data = data_long,
     family = binomial
   )
+
+broom::tidy(death_num, exponentiate = TRUE)
 
 data_long$p_denom = predict(death_den, data_long, type = "response")
 
@@ -512,3 +514,7 @@ plots <- list(
   plot_1b, plot_1c, plot_3b, plot_3c)
[…]

### 2021-07-15 02_R/04c_bootstraps_tv.R
all models use laggued covariates rather than what was observed at that time point

@@ -9,7 +9,7 @@ library(WeightIt)
 library(cobalt)
 
 # Import data -------------------------------------------------------------
-source(here::here("02_R", "04b_auxiliary_fx.R"))
+source(here::here("02_R", "04_auxiliary_fx.R"))
 
 data_long <-
   import(here::here("01_data", "clean_data", "dementia_long.RData"))
@@ -63,8 +63,8 @@ km_tv <- function(data_long, crude = TRUE, ipcw = FALSE) {
     
     # denominator
     mod <- glm(cancer_v ~ bs(time, 3) + bs(age_0, 3) + sex + education + apoe4 +
-                 as.factor(smoke) + bs(sbp, 3) + bs(bmi, 3) + 
-                 ht + ht_drug + diab_v + cohort,
+                 as.factor(smoke_lag) + bs(sbp_lag, 3) + bs(bmi_lag, 3) + 
+                 ht_lag + ht_drug_lag + diab_lag + cohort,
                family = quasibinomial(), data = subset(data_long , 
                                                        pastcancer == 0))
     
@@ -113,9 +113,9 @@ km_tv <- function(data_long, crude = TRUE, ipcw = FALSE) {
     
     if (ipcw != FALSE) {
       death_den <- glm(
-        competing_plr ~ cancer_v + bs(time, 3) + bs(age_0, 3) + sex + education + apoe4 +
-          as.factor(smoke) + bs(sbp, 3) + bs(bmi, 3) + 
-          ht + ht_drug + hd_v + stroke_v + diab_v + cohort,
+        competing_plr ~ cancer_v + cancer_lag + bs(time, 3) + bs(age_0, 3) + sex + education + apoe4 +
+          as.factor(smoke_lag) + bs(sbp_lag, 3) + bs(bmi_lag, 3) + 
+          ht_lag + ht_drug_lag + hd_lag + stroke_lag + diab_lag + cohort,
         data = data_long,
         family = quasibinomial
       )
@@ -124,7 +124,7 @@ km_tv <- function(data_long, crude = TRUE, ipcw = FALSE) {
       
       death_num <-
         glm(
-          competing_plr ~ cancer_v + bs(time, 3) + cohort,
+          competing_plr ~ cancer_lag + bs(time, 3) + cohort,
           data = data_long,
           family = binomial
         )

### 2021-07-19 04_reports/manuscript.Rmd
many many updates

@@
 ---
-title: "Dissecting the causal question underlying the association between cancer and ADRD: an opportunity to understand sources of bias"
+title: "Towards a clearer causal question underlying the association between cancer and ADRD"
 author: ""
@@
 
-Many observational studies have consistently found that individuals with cancer have a lower risk of developing Alzheimer´s disease and related dementias (ADRD) when compared to individuals with no history of cancer [@ma2014;@hanson2016;@vanderwillik2018;@ospina2020].
+Many observational studies have consistently found that individuals with cancer have a lower risk of developing dementia when compared to individuals with no history of cancer [@ma2014;@hanson2016;@vanderwillik2018;@ospina2020].
 These findings have motivated substantial research toward mechanistic explanations, including molecular and genetic pathways that may explain this association[@behrens2009;@harris2014;@nudelman2019;@Papin2020;@driverbiogeront2014;@olson2019;@li2021;@driverpin12015].
@@
 
-If this hypothetical Pin1-targeting drug was developed, the best way to understand its effect on ADRD risk would be to have a well-conducted randomized trial in which we randomize eligible participants in late midlife (e.g., ages 50-60 years) to receive this drug or not, and closely monitor ADRD over a lengthy follow-up.
+If this hypothetical Pin1-targeting drug was developed, the best way to understand its effect on ADRD risk would be to have a well-conducted randomized trial in which we randomize eligible participants in late midlife (e.g., ages 50-60 years) to receive this drug or not, and closely monitor dementia diagnosis over a lengthy follow-up.
 Since this drug is not currently available, at best we can use observational data on Pin1 expression measurements.
@@
 
-In the observational setting, confounding could explain the observed association between Pin1 and ADRD.
-In the causal diagram [@whatif2020] in Figure 1, we present Pin1 expression as $P_{t}$ and ADRD diagnosis by time $t+k$ as $Y_{t+k}$.
+In the observational setting, confounding could explain the observed association between Pin1 and dementia.
+In the causal diagram [@whatif2020] in Figure 1, we present Pin1 expression as $P_{t}$ and dementia diagnosis by time $t+k$ as $Y_{t+k}$.
 Both may share causes $L_1$, and to assess the causal relationship between $P_{t}$ and $Y_{t+k}$ we would thus adjust for $L_1$.
-Previous ADRD studies have described age, sex, educational level and race/ethnicity as a minimal adjusting set of covariates[@ospina2020].
-However, environmental and behavioral factors such as smoking may translate into Pin1 over-expression[@tan2010] and are also related to the development of ADRD[@dementia_lancet].
+Previous dementia studies have described age, sex, educational level and race/ethnicity as a minimal adjusting set of covariates[@ospina2020].
+However, environmental and behavioral factors such as smoking may translate into Pin1 over-expression[@tan2010] and are also related to the development of dementia[@dementia_lancet].
 
-```{tikz, echo = FALSE, fig.cap = "Pin1 as biomarker and risk of dementia"}
-\usetikzlibrary{arrows}
-\begin{tikzpicture}[node distance=2cm]
-\node (p) {$P_t$};
-\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L_1$};
-\node [gray!30, below of = p, yshift = 1cm] (pstar) {$C_t$};
-\node [gray!30, right of = p] (dp) {$D_{t+1}$};
-\node [gray!30, right of = dp] (p2) {$P_{t+1}$};
-\node [gray!30, below of = p2, yshift = 1cm] (pstar2) {$C_{t+1}$};
-\node [gray!30, right of = p2] (dp2) {$D_{t+k}$};
-\node [right of = dp2] (y) {$Y_{t+k}$};
-\node [gray!30, below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$L_2$};
-\draw[gray!30, ->] (p) to (pstar);
-\draw[gray!30, ->] (l) to (dp);
-\draw[gray!30, ->] (pstar) -- (dp);
-\draw[gray!30, ->] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
-\draw[gray!30, ->] (dp) -- (p2);
-\draw[gray!30,->] (p2) to (pstar2);
-\draw[gray!30, ->] (dp2) -- (y);
-\draw[gray!30, ->] (cd) -- (dp2);
-\draw[gray!30, ->] (cd) -- (y);
-\draw[gray!30, ->] (l) -- (p2);
-\draw[gray!30, ->] (l) to [out=0,in=90, looseness=0.4] (dp2);
-\draw[gray!30, ->] (pstar2) -- (dp2);
-\draw[gray!30, ->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
-\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
-\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->] (l) -- (p);
-\end{tikzpicture}
-```
+< Figure 1 >
 
 Currently, Pin1 expression is not an available biomarker for population-based research, so at best we can only rely on a proxy for it.
-Because Pin1 over-expression is present in tumors, and tumors are only measured through diagnosis, previous studies have considered Pin1 over-expression as the explanation behind the inverse association between cancer diagnosis and ADRD, though Pin1 was not explicitly part of the research question[@driver2012;@musicco2013;@freedman2016;@bowles2017;@frain2017;@schmidt2017;@sun2020;@ording2020].
-Unlike measuring Pin1 at the same time to all participants (thought this would not necessarily mean this would be the ideal time to measure it, we discuss this point further in the discussion section), cancer diagnosis is a collected over time.
+Because Pin1 over-expression is present in tumors, and tumors are only measured through diagnosis, previous studies have considered Pin1 over-expression as the explanation behind the inverse association between cancer diagnosis and dementia, though Pin1 was not explicitly part of the research question[@driver2012;@musicco2013;@freedman2016;@bowles2017;@frain2017;@schmidt2017;@sun2020;@ording2020].
+Unlike measuring Pin1 at the same time to all participants (thought this would not necessarily mean this would be the ideal time to measure it, we discuss this point further in the discussion section), cancer diagnosis is a collected over follow-up.
 We depict this feature in Figure 2, where $C_t$ and $C_{t+1}$ represent _cancer diagnosis_ over time, the measured proxy of $P_{t}$ and $P_{t+1}$ respectively.
@@
 
-```{tikz, echo = FALSE, fig.cap = "Cancer diagnosis as proxy for Pin1 expression"}
-\usetikzlibrary{arrows}
-\begin{tikzpicture}[node distance=2cm]
-\node (p) {$P_t$};
-\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L_1$};
-\node [below of = p, yshift = 1cm] (pstar) {$C_t$};
-\node [gray!30, right of = p] (dp) {$D_{t+1}$};
-\node [right of = dp] (p2) {$P_{t+1}$};
-\node [below of = p2, yshift = 1cm] (pstar2) {$C_{t+1}$};
-\node [gray!30, right of = p2] (dp2) {$D_{t+k}$};
-\node [right of = dp2] (y) {$Y_{t+k}$};
-\node [gray!30, below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$L_2$};
-\draw[gray!30, ->] (l) to (dp);
-\draw[gray!30, ->] (pstar) -- (dp);
-\draw[gray!30, ->] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
-\draw[gray!30, ->] (dp) -- (p2);
-\draw[gray!30, ->] (dp2) -- (y);
-\draw[gray!30, ->] (cd) -- (dp2);
-\draw[gray!30, ->] (cd) -- (y);
-\draw[gray!30, ->] (l) to [out=0,in=90, looseness=0.4] (dp2);
-\draw[gray!30, ->] (pstar2) -- (dp2);
-\draw[->] (p) to (pstar);
-\draw[->] (l) -- (p);
-\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
-\draw[->] (p2) to (pstar2);
-\draw[->] (l) -- (p2);
-\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
-\end{tikzpicture}
-```
+< Figure 2 >
 
-A challenge that arises by choosing cancer diagnosis as the proxy for Pin1 is defining the time zero, the time where eligibility criteria is met, where exposure is measured and the time where the screening for ADRD would begin after having the exposure measured[@hernanInmortal2016].
+A challenge that arises when choosing cancer diagnosis as the proxy for Pin1 is defining the time zero, the time where eligibility criteria is met, where exposure is measured and the time where the screening for dementia would begin after having the exposure measured[@hernanInmortal2016].
 Emulating the eligibility criteria to participate in a trial for Pin-targeting drug will not necessarily be possible in a cohort setting recruiting participants for discordant reasons.
@@
 This problem is partly avoided by recognizing the time-varying nature of cancer diagnosis.
-Several studies have considered cancer diagnosis as a time-dependent exposure.
-In this way, the time prior to cancer diagnosis is allocated to the non-exposed arm, and the time after cancer diagnosis to the exposed arm [@white2013; @hanson2016; @bowles2017].
+Several studies have considered cancer diagnosis as a time-dependent exposure [@white2013; @hanson2016; @bowles2017].
 The price we pay with this approximation is that implicitly, this means that Pin1 would over-express at the time of cancer diagnosis and not before, which is biologically implausible.
 Moreover, cancer diagnosis will only be measured in the subset of participants who are alive over follow-up.
-Thus, we included death prior to cancer diagnosis as $D_{t+1}$ and an arrow between $D_{t+1}$ and $P_{t+1}$ that represents a deterministic association such as that $P_{t+1}$ is only observed if $D_{t+1}$ is zero in Figure 3.
+Thus, in Figure 3 we included death prior to cancer diagnosis as $D_{t+1}$ and an arrow between $D_{t+1}$ and $P_{t+1}$ that represents a deterministic association such as that $P_{t+1}$ is only observed if $D_{t+1}$ is zero.
 In addition, we added an arrow between $L_1$ and $D_{t+1}$, since covariates such as smoking may affect Pin1 over-expression but also affect the risk of death due to other causes such as from chronic obstructive pulmonary disease.
@@
 
-```{tikz, echo = FALSE, fig.cap = "Time-varying cancer diagnosis and inmortal-time bias"}
-\usetikzlibrary{arrows}
-\begin{tikzpicture}[node distance=2cm]
-\node (p) {$P_t$};
-\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L_1$};
-\node [below of = p, yshift = 1cm] (pstar) {$C_t$};
-\node [right of = p] (dp) {$D_{t+1}$};
-\node [right of = dp] (p2) {$P_{t+1}$};
-\node [below of = p2, yshift = 1cm] (pstar2) {$C_{t+1}$};
-\node [gray!30, right of = p2] (dp2) {$D_{t+k}$};
-\node [right of = dp2] (y) {$Y_{t+k}$};
-\node [gray!30, below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$L_2$};
-\draw[gray!30, ->] (pstar) -- (dp);
-\draw[gray!30, ->] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
-\draw[gray!30, ->] (dp2) -- (y);
-\draw[gray!30, ->] (cd) -- (dp2);
-\draw[gray!30, ->] (cd) -- (y);
-\draw[gray!30, ->] (l) to [out=0,in=90, looseness=0.4] (dp2);
-\draw[gray!30, ->] (pstar2) -- (dp2);
-\draw[->] (p) to (pstar);
-\draw[->] (l) -- (p);
-\draw[->] (l) to (dp);
-\draw[->] (dp) -- (p2);
-\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
-\draw[->] (p2) to (pstar2);
-\draw[->] (l) -- (p2);
-\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
-\end{tikzpicture}
-```
+< Figure 3 >
 
-We now move to the challenge of death as a competing event for ADRD, represented in Figure 4.
+Another challenge to adress is related to death as a competing event for dementia, represented in Figure 4.
 For interpretability we exclude the time-varying process of cancer diagnosis and focus on Pin1 (and cancer diagnosis) as it had been measured in all participants at time $t+1$.
-Given that some participants may die prior to ADRD diagnosis, we can only measure ADRD over follow-up in the individuals who survive long enough to have an ADRD diagnosis.
+Given that some participants may die prior to dementia diagnosis, we can only measure dementia over follow-up in the individuals who survive long enough to have a dementia diagnosis.
 In the causal diagram of Figure 4 we include a node for death after the exposure $C_{t+1}$ has been measured, represented as $D_{t+k}$ and which acts as a competing event of $Y_{t+k}$ because if a participant dies by $t + k$, the participant cannot subsequently develop dementia.
@@
 We also include an arrow between $L_1$ and $D_{t+1}$ following the argument discussed previously.
+
+< Figure 4 >
 
@@
 
-```{tikz, echo = FALSE, fig.cap = "Death as a competing event"}
-\usetikzlibrary{arrows}
-\begin{tikzpicture}[node distance=2cm]
-\node [gray!30](p) {$P_t$};
-\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L_1$};
-\node [gray!30, below of = p, yshift = 1cm] (pstar) {$C_t$};
-\node [gray!30, right of = p] (dp) {$D_{t+1}$};
-\node [right of = dp] (p2) {$P_{t+1}$};
-\node [below of = p2, yshift = 1cm] (pstar2) {$C_{t+1}$};
-\node [right of = p2] (dp2) {$D_{t+k}$};
-\node [right of = dp2] (y) {$Y_{t+k}$};
-\node [below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$L_2$};
-\draw[gray!30, ->] (p) to (pstar);
-\draw[gray!30, ->] (l) -- (p);
-\draw[gray!30, ->] (l) to (dp);
-\draw[gray!30, ->] (pstar) -- (dp);
-\draw[gray!30,->] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
-\draw[gray!30, ->] (dp) -- (p2);
-\draw[gray!30, ->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
-\draw[->] (p2) to (pstar2);
-\draw[->] (dp2) -- (y);
-\draw[->] (cd) -- (dp2);
-\draw[->] (cd) -- (y);
-\draw[->] (l) -- (p2);
-\draw[->] (l) to [out=0,in=90, looseness=0.4] (dp2);
-\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->] (pstar2) -- (dp2);
-\draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
-\end{tikzpicture}
-```
-
 To summarize, the complexity of the causal structure that describes the effect of Pin1 through the proxy of cancer diagnosis in the risk of ADRD illustrate the potential sources of bias, as observed in Figure 5.
@@
 
-```{tikz, echo = FALSE, fig.cap = "Direct effect of Pin1 in the risk of ADRD"}
-\usetikzlibrary{arrows}
-\begin{tikzpicture}[node distance=2cm]
-\node (p) {$P_t$};
-\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L_1$};
-\node [below of = p, yshift = 1cm] (pstar) {$C_t$};
-\node [right of = p] (dp) {$D_{t+1}$};
-\node [right of = dp] (p2) {$P_{t+1}$};
-\node [below of = p2, yshift = 1cm] (pstar2) {$C_{t+1}$};
-\node [right of = p2] (dp2) {$D_{t+k}$};
-\node [right of = dp2] (y) {$Y_{t+k}$};
-\node [below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$L_2$};
-\draw[->] (p) to (pstar);
-\draw[->] (l) -- (p);
-\draw[->] (l) to (dp);
-\draw[->] (pstar) -- (dp);
-\draw[->] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
-\draw[->] (dp) -- (p2);
-\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
-\draw[->] (p2) to (pstar2);
-\draw[->] (dp2) -- (y);
-\draw[->] (cd) -- (dp2);
-\draw[->] (cd) -- (y);
-\draw[->] (l) -- (p2);
-\draw[->] (l) to [out=0,in=90, looseness=0.4] (dp2);
-\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
-\draw[->] (pstar2) -- (dp2);
-\draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
-\end{tikzpicture}
-```
+< Figure 5 >
 
@@
 Participants from the first subcohort had follow-up visits between 1993-1995, 1997-1999, 2002-2005, and 2008-2010, while the second subcohort had follow-up visits between 2004 and 2005, and between 2011 and 2012[@ikram2020].
-All participants had data on history of cancer and dementia and incident diagnosis through follow-up until 31 of December, 2015, collected from medical records of general practitioners (including hospital discharge letters) and through linkage with national registries.
-Date and cause of death was collected on a weekly basis via municipal population registries.
-These ascertainment methods imply that the Rotterdam Study has functionally no loss to follow-up with respect to dementia diagnosis and death.
+All participants had data on history of cancer and dementia and incident diagnosis, collected from medical records of general practitioners (including hospital discharge letters) and through linkage with national registries.
+Date and cause of death was collected via municipal population registries.
+These ascertainment methods imply that the Rotterdam Study has functionally no loss to follow-up with respect to dementia diagnosis and death
 
@@
 All participants were followed from study entry until dementia diagnosis, death or 20 years after their individual baseline date, whichever occurred first.
-Given that participants of the second wave were followed for 15 years on average, the 20 year-risk of dementia includes extrapolation of the risk for the second cohort, under the assumption that participants from second cohort would have had a similar distribution of dementia and mortality than the first cohort, between year 15 and 20.
+Given that participants from the second subcohort were followed for 15 years, we assume that they would have had a similar distribution of dementia risk and mortality as the first subcohort, between year 15 and 20.
 
@@
 never"_ [@driver2012], meaning we compare dementia outcomes in individuals who ever develop cancer during follow-up to those who were not observed to develop cancer during follow-up.
-Scenario B defines cancer diagnosis as time-varying, meaning ....
+Scenario B defines cancer diagnosis as time-varying meaning that time prior to cancer diagnosis is allocated to the non-exposed arm, and the time after cancer diagnosis to the exposed arm.
 To address confounding, we fit inverse probability treatment weights, stabilized and truncated at the 99th percentile.
 In Scenario A, weights were defined as the inverse of the probability of cancer diagnosis conditional on baseline covariates such as age at study entry, sex, educational attainment, cohort, smoking status.
-In contrast, for Scenario B weights were defined to represent the product of the inverse probability of being diagnosed with cancer over time, conditional on the time-varying covariate history prior to cancer diagnosis [@hernan2000].
-Baseline covariates included age at study entry, sex, ApoE4 status, educational attainment and the time-varying covariates such as smoking status, systolic blood pressure, BMI  and prevalent and incident hypertension and diabetes.
-Further details on modeling specifications and weights assessment are presented as **Supplementary material**.
+In contrast, for Scenario B, weights were defined to represent the product of the inverse probability of being diagnosed with cancer over time, conditional on the time-varying covariate history prior to cancer diagnosis [@hernan2000].
+Baseline covariates included age at study entry, sex, ApoE4 status, educational attainment and the time-varying covariates such as smoking status, systolic blood pressure, body mass index and prevalent and incident hypertension and diabetes.
 
@@
 For an individual who has died by time _t_, the year _t_ censoring weight is zero [@young2020].
-Weights were fitted including the same covariates used to estimate weights for time-varying cancer diagnosis, though we additionally added time-varying cancer, stroke, and heart disease diagnosis as predictors for death.
+Weights were fitted including the same covariates used to estimate weights for time-varying cancer diagnosis, though we additionally added time-varying cancer, stroke, and heart disease diagnosis as predictors for death.Further details on modeling specifications and weights assessment are presented as **Supplementary material**.
 
@@
 The weights are time-varying by follow-up year, defined as a product of the year-specific inverse probability of treatment weights and the inverse probability of censoring by death weights.
-Estimates of the controlled direct effect at 20 years of follow-up are presented as risk differences (RD) and risk ratios (RR).
+Estimates of the controlled direct effect are presented as 20-year risk differences (RD) and risk ratios (RR).
 All 95% confidence intervals were calculated using percentile-based bootstrapping based on 500 bootstrap samples.
@@
 Since the conditional independent censoring assumption is untestable, we compute Peterson upper and lower bounds [@peterson1976] to represent: 1) the extreme scenario of independence, that refers to an scenario were those who died would never develop dementia (lower bound) and 2) complete dependency, that refers to an scenario where those who died would have a dementia prior to death (upper bound).
-The lower bound is calculated with the Aalen-Johansen estimator treating death as a competing event, and the upper bound is calculated with the Kaplan Meier estimator for the combined outcome of dementia or death.
+The lower bound is calculated with the Aalen-Johansen estimator treating death as a competing event, and the upper bound is calculated with the Kaplan-Meier estimator for the combined outcome of dementia or death.
 
@@
 In contrast, among participants free of cancer diagnosis over follow-up, `r no_cancer_dem` were diagnosed with dementia, `r no_cancer_death` died over follow-up, and `r no_cancer_alive` were alive and dementia-free at the end of follow-up.
-**Figure 6 shows the proportion of participants in each possible status over follow-up.**
+The proportion of participants in each possible status over follow-up and leading causes of death for both arms are presented as Figure 1 an 2 in Supplementary Material.
 
@@
 In contrast, using Scenario B's analytic design, the fully adjusted model results in a RR of RR: `r complete_table[6,4]` and a HR of `r complete_table[6,5]`, though confidence intervals cross the null.
+The cumulative incidence curves for dementia for participants free of cancer diagnosis and with incident cancer diagnosis, under scenarios A and B, reflect that the difference between groups changes over time and flips direction when considering cancer as a time-varying proxy.
 ~~In addition, the extreme scenarios of dependence between death and dementia over follow-up represented as bounds are wide and include the null value (`r bounds_tv_ipw`).~~
@@
 In the setting where cancer diagnosis is defined as "ever vs.
-never", this definition inntroduces immortal time bias.
+never", this definition introduces immortal time bias.
 All results pertaining to this definition had an inverse association between cancer diagnosis and dementia, while results that did not introduce this particular form of immortal-time bias had point estimates close to the null.
@@
 This is different than conceiving a drug-target that increases Pin1 expression only brain tissue, with no side effects that could increase cancer risk (and thus, death due to cancer).
-As opposed to prior studies that implicitly address a direct effect, and who define censoring as ignorable[@frain2017], we show how point estimates change substantially when we include weights for death to relax the independent censoring assumption.
-In contrast to prior studies that conceptualize bias around estimators or measures of effect (rates vs.
-cumulative incidence)[@ospina2020], we define bias as failing to relax the independent censoring assumption conditional on covariates[@vangeloven2014, @willems2018, @young2020] when the estimand of interest is the controlled direct effect.
+As opposed to prior studies that implicitly address a direct effect, and who define censoring as ignorable[@frain2017], we show how point estimates change substantially when we include weights for death to relax the independent censoring assumption [@vangeloven2014, @willems2018, @young2020] when the estimand of interest is the controlled direct effect.
 Bounds to assess extreme scenarios of dependence between death and dementia[@peterson1976] illustrate the wide range of possible point estimates that cross the null.
 This shows that even with the effort of adjusting for time-varying covariates, we may be far from meeting this assumption and thus better efforts to measure shared causes of dementia and death are needed.
-In addition, presenting the proportion of participants in each status over time, and the proportion of participants that died prior to ADRD diagnosis in each arm provides insights and improves transparency of results and interpretations.
+In addition, presenting the proportion of participants that died prior to ADRD diagnosis in each arm, as well as the proportion of participants in each status over time, improves transparency and puts in evidence the limitations of the data to answer this question.
 
@@
 Certainly, cancer diagnosis represents a complex and heterogeneous health condition that exceeds the representation of Pin1 expression.
-Furthermore, alternative mechanisms underlying cancer diagnosis and its connection to collected data may be represented in different causal representations.
-To understand how – if at all – the cancer-dementia association informs the potential effects of any other mechanism or treatment strategy would require deep consideration of each causal question anew.
+To understand how – if at all – the cancer-dementia association informs the potential effects of any other mechanism or treatment strategy, and its connection to collected data, we may require different causal representations.
 For example, if researchers used this association to inform the possible effects of different chemotherapeutics on cognitive decline among patients undergoing treatment for cancer, different challenges would arise for mapping the observed association to the hypothetical randomized trial underlying this research question.
@@
 
-\newpage
+### Figures
+
+```{tikz, echo = FALSE, fig.cap = "Pin1 as biomarker and risk of dementia"}
+\usetikzlibrary{arrows}
+\begin{tikzpicture}[node distance=2cm]
+\node (p) {$P_t$};
+\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L_1$};
+\node [gray!30, below of = p, yshift = 1cm] (pstar) {$C_t$};
+\node [gray!30, right of = p] (dp) {$D_{t+1}$};
+\node [gray!30, right of = dp] (p2) {$P_{t+1}$};
+\node [gray!30, below of = p2, yshift = 1cm] (pstar2) {$C_{t+1}$};
+\node [gray!30, right of = p2] (dp2) {$D_{t+k}$};
+\node [right of = dp2] (y) {$Y_{t+k}$};
+\node [gray!30, below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$L_2$};
+\draw[gray!30, ->] (p) to (pstar);
+\draw[gray!30, ->] (l) to (dp);
+\draw[gray!30, ->] (pstar) -- (dp);
+\draw[gray!30, ->] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
+\draw[gray!30, ->] (dp) -- (p2);
+\draw[gray!30,->] (p2) to (pstar2);
+\draw[gray!30, ->] (dp2) -- (y);
+\draw[gray!30, ->] (cd) -- (dp2);
+\draw[gray!30, ->] (cd) -- (y);
+\draw[gray!30, ->] (l) -- (p2);
+\draw[gray!30, ->] (l) to [out=0,in=90, looseness=0.4] (dp2);
+\draw[gray!30, ->] (pstar2) -- (dp2);
+\draw[gray!30, ->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
+\draw[->] (l) -- (p);
+\end{tikzpicture}
+```
+
+**Figure 1.
+Pin1 as biomarker and risk of dementia.** Causal directed acyclic graph where Pin1 at time $t$ is represented as $P_t$, Dementia at time $t+k$ is represented as $Y_{t+k}$.
+$L_1$ represents shared causes of Pin1 and Dementia, such as smoking.
+To isolate the effect of $P_t$ and $Y_{y+k}$ we need to block the backdoor path $Y_{t+k} \leftarrow L_1 \rightarrow P_t$.
+
+```{tikz, echo = FALSE, fig.cap = " Incident cancer diagnosis as proxy for Pin1 expression"}
+\usetikzlibrary{arrows}
+\begin{tikzpicture}[node distance=2cm]
+\node (p) {$P_t$};
+\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L_1$};
+\node [below of = p, yshift = 1cm] (pstar) {$C_t$};
+\node [gray!30, right of = p] (dp) {$D_{t+1}$};
+\node [right of = dp] (p2) {$P_{t+1}$};
+\node [below of = p2, yshift = 1cm] (pstar2) {$C_{t+1}$};
+\node [gray!30, right of = p2] (dp2) {$D_{t+k}$};
+\node [right of = dp2] (y) {$Y_{t+k}$};
+\node [gray!30, below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$L_2$};
+\draw[gray!30, ->] (l) to (dp);
+\draw[gray!30, ->] (pstar) -- (dp);
+\draw[gray!30, ->] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
+\draw[gray!30, ->] (dp) -- (p2);
+\draw[gray!30, ->] (dp2) -- (y);
+\draw[gray!30, ->] (cd) -- (dp2);
+\draw[gray!30, ->] (cd) -- (y);
+\draw[gray!30, ->] (l) to [out=0,in=90, looseness=0.4] (dp2);
+\draw[gray!30, ->] (pstar2) -- (dp2);
+\draw[->] (p) to (pstar);
+\draw[->] (l) -- (p);
+\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->] (p2) to (pstar2);
+\draw[->] (l) -- (p2);
+\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
+\draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
+\end{tikzpicture}
+```
+
+**Figure 2.
+Incident cancer diagnosis as proxy for Pin1 expression.** This causal graph is an expansion of Figure 1, includes Pin1 expression at time $P_{t+1}$.
+Incident cancer diagnosis is represented as $C_t$ and $C_{t+1}$.
+To isolate the effect of $P_t$ and $P_{t+1}$ on $Y_{y+k}$ we need to block the backdoor path $Y_{t+k} \leftarrow L_1 \rightarrow P_t$ and $Y_{t+k} \leftarrow L_1 \rightarrow P_{t+1}$.
+Although we represent $L_1$ as a single node for readability, $L_1$ is time-varying too.
+
+```{tikz, echo = FALSE, fig.cap = "Time-varying cancer diagnosis and inmortal-time bias"}
+\usetikzlibrary{arrows}
+\begin{tikzpicture}[node distance=2cm]
+\node (p) {$P_t$};
+\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L_1$};
+\node [below of = p, yshift = 1cm] (pstar) {$C_t$};
+\node [right of = p] (dp) {$D_{t+1}$};
+\node [right of = dp] (p2) {$P_{t+1}$};
+\node [below of = p2, yshift = 1cm] (pstar2) {$C_{t+1}$};
+\node [gray!30, right of = p2] (dp2) {$D_{t+k}$};
+\node [right of = dp2] (y) {$Y_{t+k}$};
+\node [gray!30, below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$L_2$};
+\draw[gray!30, ->] (pstar) -- (dp);
+\draw[gray!30, ->] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
+\draw[gray!30, ->] (dp2) -- (y);
+\draw[gray!30, ->] (cd) -- (dp2);
+\draw[gray!30, ->] (cd) -- (y);
+\draw[gray!30, ->] (l) to [out=0,in=90, looseness=0.4] (dp2);
+\draw[gray!30, ->] (pstar2) -- (dp2);
+\draw[->] (p) to (pstar);
+\draw[->] (l) -- (p);
+\draw[->] (l) to (dp);
+\draw[->] (dp) -- (p2);
+\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->] (p2) to (pstar2);
+\draw[->] (l) -- (p2);
+\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
+\draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
+\end{tikzpicture}
+```
+
+**Figure 3.
+Time-varying cancer diagnosis and inmortal-time bias.**  This causal graph is an expansion of Figure 2, where $D_{t+1}$ represents death at time $t+1$.
+Cancer diagnosis at $t+1$ can only be measured among those who are alive at $t+1$.
+
+```{tikz, echo = FALSE, fig.cap = "Death as a competing event"}
+\usetikzlibrary{arrows}
+\begin{tikzpicture}[node distance=2cm]
+\node [gray!30](p) {$P_t$};
+\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L_1$};
+\node [gray!30, below of = p, yshift = 1cm] (pstar) {$C_t$};
+\node [gray!30, right of = p] (dp) {$D_{t+1}$};
+\node [right of = dp] (p2) {$P_{t+1}$};
+\node [below of = p2, yshift = 1cm] (pstar2) {$C_{t+1}$};
+\node [right of = p2] (dp2) {$D_{t+k}$};
+\node [right of = dp2] (y) {$Y_{t+k}$};
+\node [below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$L_2$};
+\draw[gray!30, ->] (p) to (pstar);
+\draw[gray!30, ->] (l) -- (p);
+\draw[gray!30, ->] (l) to (dp);
+\draw[gray!30, ->] (pstar) -- (dp);
+\draw[gray!30,->] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
+\draw[gray!30, ->] (dp) -- (p2);
+\draw[gray!30, ->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->] (p2) to (pstar2);
+\draw[->] (dp2) -- (y);
+\draw[->] (cd) -- (dp2);
+\draw[->] (cd) -- (y);
+\draw[->] (l) -- (p2);
+\draw[->] (l) to [out=0,in=90, looseness=0.4] (dp2);
+\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
+\draw[->] (pstar2) -- (dp2);
+\draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
+\end{tikzpicture}
+```
+
+**Figure 4.
+Death as competing event of dementia.** In this graph we only focus attention to the exposure as if it was measured for all at time $P_{t+1}$.
+We include $D_{t+k}$ as death at time $t+k$, $L_2$ represents possible shared causes of dementia and death (such as cardiovascular comorbidities).
+Pin1 may affect the risk of death through cancer diagnosis, represented as an arrow between $C_{t+1}$ and  $D_{t+k}$.
+To isolate the direct effect of $P_{t+1}$ and $Y_{y+k}$, we have to block the backdoor pathway from $Y_{t+k} \leftarrow L_2 \rightarrow D_{t+k} \leftarrow C_{t+1} \rightarrow P_{t+1}$ and $Y_{t+k} \rightarrow D_{t+k} \leftarrow L_1 \rightarrow P_{t+1}$
+
+```{tikz, echo = FALSE, fig.cap = "Direct effect of Pin1 in the risk of ADRD"}
+\usetikzlibrary{arrows}
+\begin{tikzpicture}[node distance=2cm]
+\node (p) {$P_t$};
+\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L_1$};
+\node [below of = p, yshift = 1cm] (pstar) {$C_t$};
+\node [right of = p] (dp) {$D_{t+1}$};
+\node [right of = dp] (p2) {$P_{t+1}$};
+\node [below of = p2, yshift = 1cm] (pstar2) {$C_{t+1}$};
+\node [right of = p2] (dp2) {$D_{t+k}$};
+\node [right of = dp2] (y) {$Y_{t+k}$};
+\node [below of = dp2, yshift = 1cm, xshift = +1cm](cd) {$L_2$};
+\draw[->] (p) to (pstar);
+\draw[->] (l) -- (p);
+\draw[->] (l) to (dp);
+\draw[->] (pstar) -- (dp);
+\draw[->] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
+\draw[->] (dp) -- (p2);
+\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
+\draw[->] (p2) to (pstar2);
+\draw[->] (dp2) -- (y);
+\draw[->] (cd) -- (dp2);
+\draw[->] (cd) -- (y);
+\draw[->] (l) -- (p2);
+\draw[->] (l) to [out=0,in=90, looseness=0.4] (dp2);
+\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
+\draw[->] (pstar2) -- (dp2);
+\draw[->, semithick] (p2) to [out=+45,in=+135, looseness=0.3] (y);
+\end{tikzpicture}
+```
+
+**Figure 5. Causal graph for effect of Pin1 through the proxy of cancer diagnosis in the risk of dementia, including all challenges discussed through Figures 1 to 4.**
+
+```{r}
+knitr::include_graphics(here::here("03_figs", "cif_plot.tiff"))
+```
 
@@
 
-```{r, fig.width= 12, fig.height=8}
-plot_distribution
+```{r}
+knitr::include_graphics(here::here("03_figs", "plot_distribution.tiff"))
 ```
 
-### Cumulative incidence plots
+### C. Causes of death for participants who died prior to dementia diagnosis
+
+```{r}
+knitr::include_graphics(here::here("03_figs", "causes_death.tiff"))
+```
+
+Causes of death were defined by ICD10 codes...xxx

### 2021-07-19 02_R/02_long.R

@@ -213,7 +213,7 @@ data_long_dem %<>%
     outcome_plr = ifelse(end_dementia_death == year & dementia == 1, 1, 0),
     competing_plr = ifelse(!is.na(dementia_date), 0, death_v)) %>%
   ungroup() %>%
-  filter(!time == 0 & !competing_plr == 1) %>%
+  filter(!(time == 0 & competing_plr == 1)) %>%
   arrange(id, time)
 
 id2 <- data_long_dem %>% 

### 2021-07-19 02_R/03_descriptives.R

@@ -46,7 +46,7 @@ total_status <- data_wide %>%
   group_by(cancer_v) %>% 
   count(status) %>% 
   mutate(prop = round(100*n/sum(n),0)) %>% 
-  mutate(total = paste0(prop, "%", " (n =", n,")")) %>% 
+  mutate(total = paste0(prop, "%", " (n=", n,")")) %>% 
   select(cancer_v, status, total)
 
 cancer_dem <- total_status %>% filter(cancer_v == 1, status == "Cancer and Dementia") %>% pull(3)
@@ -138,7 +138,8 @@ data_tableone %<>%
          stroke_prev = as_factor(stroke_prev),
          stroke_v = ifelse(stroke_v == 1, "Incident stroke", "No incident stroke"),
          stroke_v = as_factor(stroke_v),
-         cancer_v = ifelse(cancer_v == 1, "Incident cancer", "No incident cancer"))
+         cancer_v = ifelse(cancer_v == 1, "Incident cancer", "No incident cancer"),
+         cancer_v = as_factor(cancer_v))
 
 
 
@@ -146,7 +147,8 @@ num <- c("age_0","bmi1", "oh1", "sbp1" )
 
 cat <- myvars[!myvars %in% num]
 
-tableone <- CreateTableOne(vars = myvars, data = data_tableone, factorVars = cat, strata = "cancer_v")
+library(tableone)
+tableone <- CreateTableOne(vars = myvars, data = data_tableone, factorVars = cat)
 
 
 tableone_exp <-
@@ -178,8 +180,8 @@ tableone_complete <- tableone_exp %>% as_tibble() %>%
     Characteristics = str_replace(Characteristics, "Unknown", "Unknown history of diabetes"),
     Characteristics = str_replace(Characteristics, "stroke_prev = ", ""),
     Characteristics = str_replace(Characteristics, "stroke_v = ", ""),
+    Characteristics = str_replace(Characteristics, "cancer_v = ", ""),
   ) %>% 
-  filter(!Characteristics %in% c("oh1 (mean (SD))", "diabetes_prev (%)", "cancer_v = No incident cancer (%)")) %>% 
-  filter(!Characteristics %in% str_detect(Characteristics, "No history of diabetes")) %>%  ## Not working :S
-  rename(`Incident cancer during follow-up` = `Incident cancer`,
-         `No incident cancer observed during follow-up` = `No incident cancer`)
[…]

### 2021-07-19 02_R/04_auxiliary_fx.R

@@ -1,0 +1,129 @@
+  mytheme <- theme_minimal(base_family = "serif") +
+  theme(
+    panel.grid.major.x = element_blank(),
+    panel.grid.minor.x = element_blank(),
+    axis.line = element_line(color = "black"),
+    strip.text = element_text(size = 12, color = "black"),
+    plot.caption = element_text(hjust = 0))
+
+
+
+risks_cif <- function(model){
+  model %>% 
+    broom::tidy() %>%
+    filter(state == "1") %>%
+    group_by(strata) %>% 
+    slice(n()) %>% 
+    select(estimate, strata) %>% 
+    pivot_wider(names_from = strata, values_from = estimate) %>% 
+    mutate(rd = .[[2]] - .[[1]],
+           rr = .[[2]] / .[[1]]) %>% 
+    mutate_at(c(1:3), ~.*100)
+}
+
+
+
+plot_cif <- function(model, title, ...){
+  
+  label_exp <- model %>% 
+    tidy() %>% slice(1) %>% pull(strata)
+  
+  tidy <- model %>% 
+    broom::tidy() %>% 
+    filter(state == "1") %>%
+    select(time, strata, estimate, conf.high, conf.low) %>%
+    rename(CIF = estimate) %>% 
+    mutate(strata = ifelse(strata == label_exp, "Free of cancer", "Incident cancer"))
+  
+  plot <- tidy %>% 
+    ggplot(aes(time, CIF, group = strata)) +
+    geom_line(aes(color = strata), size = 0.7) +
+    geom_ribbon(aes(ymin = conf.low, ymax = conf.high) ,alpha = 0.2) +
+    scale_color_manual(values = c("#011A5E", "#e4a803")) +
+    scale_y_continuous(limits = c(0, 0.70)) +
+    labs(
+      title = paste0(title),
+      color = NULL,
+      y = NULL) +
+    theme_minimal() +
+    theme(legend.position = "bottom")
+  
+  return(plot)
+  
+}
+
+
+# Km ----------------------------------------------------------------------
+
+plot_km <- function(model, title, ...){
+  
+  label_exp <- model %>% 
+    tidy() %>% slice(1) %>% pull(strata)
+  
+  tidy <- model %>% 
+    broom::tidy() %>% 
+    transmute(
+      time = time, 
+      strata = ifelse(strata == label_exp, 
[…]

### 2021-07-19 02_R/04_direct_effect_simplified.R

@@ -20,7 +20,7 @@ data_wide <-
 data_long %<>% 
   group_by(id) %>% 
   mutate(
-    fuptime = time - 1,
+    fuptime = time,
     tstart = lag(fuptime),
     tstart = ifelse(is.na(tstart), -1, tstart)) %>% 
   ungroup()
@@ -333,7 +333,7 @@ data_long %>%
 # 2.2. IPCW. Weights on death - dementia ---------------------------------------------
 
 death_den <- glm(
-  competing_plr ~ cancer_v + cancer_lag + bs(time, 3) + bs(age_0, 3) + sex + education + apoe4 +
+  competing_plr ~ cancer_lag + bs(time, 3) + bs(age_0, 3) + sex + education + apoe4 +
     as.factor(smoke_lag) + bs(sbp_lag, 3) + bs(bmi_lag, 3) + 
     ht_lag + ht_drug_lag + hd_lag + stroke_lag + diab_lag + cohort,
   data = data_long,
@@ -346,7 +346,7 @@ death_num <-
   glm(
     competing_plr ~ cancer_lag + bs(time, 3) + cohort,
     data = data_long,
-    family = binomial
+    family = quasibinomial
   )
 
 broom::tidy(death_num, exponentiate = TRUE)
@@ -382,6 +382,7 @@ data_long %<>%
          sw_weights_both = sw_cancer * sw_death,
          sw_weights_both_t = sw_cancer_t * sw_death_t)
 
+data_long %>% filter(tstart >= fuptime) %>% select(id, time, tstart, fuptime)
 
 # 3.a. KM unadjusted ------------------------------------------------------
 

### 2021-07-19 02_R/04b_auxiliary_fx.R

@@ -1,129 +1,0 @@
-  mytheme <- theme_minimal(base_family = "serif") +
-  theme(
-    panel.grid.major.x = element_blank(),
-    panel.grid.minor.x = element_blank(),
-    axis.line = element_line(color = "black"),
-    strip.text = element_text(size = 12, color = "black"),
-    plot.caption = element_text(hjust = 0))
-
-
-
-risks_cif <- function(model){
-  model %>% 
-    broom::tidy() %>%
-    filter(state == "1") %>%
-    group_by(strata) %>% 
-    slice(n()) %>% 
-    select(estimate, strata) %>% 
-    pivot_wider(names_from = strata, values_from = estimate) %>% 
-    mutate(rd = .[[2]] - .[[1]],
-           rr = .[[2]] / .[[1]]) %>% 
-    mutate_at(c(1:3), ~.*100)
-}
-
-
-
-plot_cif <- function(model, title, ...){
-  
-  label_exp <- model %>% 
-    tidy() %>% slice(1) %>% pull(strata)
-  
-  tidy <- model %>% 
-    broom::tidy() %>% 
-    filter(state == "1") %>%
-    select(time, strata, estimate, conf.high, conf.low) %>%
-    rename(CIF = estimate) %>% 
-    mutate(strata = ifelse(strata == label_exp, "Free of cancer", "Incident cancer"))
-  
-  plot <- tidy %>% 
-    ggplot(aes(time, CIF, group = strata)) +
-    geom_line(aes(color = strata), size = 0.7) +
-    geom_ribbon(aes(ymin = conf.low, ymax = conf.high) ,alpha = 0.2) +
-    scale_color_manual(values = c("#011A5E", "#e4a803")) +
-    scale_y_continuous(limits = c(0, 0.70)) +
-    labs(
-      title = paste0(title),
-      color = NULL,
-      y = NULL) +
-    theme_minimal() +
-    theme(legend.position = "bottom")
-  
-  return(plot)
-  
-}
-
-
-# Km ----------------------------------------------------------------------
-
-plot_km <- function(model, title, ...){
-  
-  label_exp <- model %>% 
-    tidy() %>% slice(1) %>% pull(strata)
-  
-  tidy <- model %>% 
-    broom::tidy() %>% 
-    transmute(
-      time = time, 
-      strata = ifelse(strata == label_exp, 
[…]

### 2021-07-19 02_R/04c_bootstraps_ever.R

@@ -1,4 +1,4 @@
-source(here::here("02_R", "04b_auxiliary_fx.R"))
+source(here::here("02_R", "04_auxiliary_fx.R"))
 
 data_wide <-
   import(here::here("01_data", "clean_data", "wide_after_truncation.RData"))

### 2021-07-19 02_R/04c_bootstraps_tv.R

@@ -20,7 +20,7 @@ data_wide <-
 data_long %<>% 
   group_by(id) %>% 
   mutate(
-    fuptime = time - 1,
+    fuptime = time,
     tstart = lag(fuptime),
     tstart = ifelse(is.na(tstart), -1, tstart)) %>% 
   ungroup()
@@ -113,7 +113,7 @@ km_tv <- function(data_long, crude = TRUE, ipcw = FALSE) {
     
     if (ipcw != FALSE) {
       death_den <- glm(
-        competing_plr ~ cancer_v + cancer_lag + bs(time, 3) + bs(age_0, 3) + sex + education + apoe4 +
+        competing_plr ~ cancer_lag + bs(time, 3) + bs(age_0, 3) + sex + education + apoe4 +
           as.factor(smoke_lag) + bs(sbp_lag, 3) + bs(bmi_lag, 3) + 
           ht_lag + ht_drug_lag + hd_lag + stroke_lag + diab_lag + cohort,
         data = data_long,

### 2021-07-19 02_R/04x_extra_models.R

@@ -1,54 +1,0 @@
-w.out1 <-
-  weightit(
-    cancer_v ~ bs(age_0, 3) + sex + education + apoe4 + as.factor(smoke1) +
-      bs(sbp1, 3) + bs(bmi1, 3) + ht1 + as.factor(diabetes_prev) + cohort,
-    data = data_wide,
-    stabilize = TRUE,
-    estimand = "ATE",
-    method = "ps"
-  )
-w.out1
-
-# 1.2. IPCW for death, based on baseline covariates ----------------------------
-
-w.out_death <-
-  weightit(
-    competing_plr ~ bs(age_0, 3) + sex + education + apoe4 + 
-      as.factor(smoke1) + ht1 + bs(sbp1, 3) + bs(bmi1,3) +
-      as.factor(diabetes_prev) +  cancer_v + cohort,
-    data = data_wide,
-    stabilize = TRUE,
-    estimand = "ATE",
-    method = "ps"
-  )
-
-
-# 2. Time-varying cancer -----------------------------------------------------
-
-cancer_den <-
-  glm(
-    cancer_v ~ bs(age_0, 3) + sex + education + apoe4 + as.factor(smoke1) + ht1 + 
-      bs(sbp1, 3) + bs(bmi1, 3) + as.factor(diabetes_prev) + cohort,
-    data = data_long,
-    family = binomial
-  )
-
-
-# 2.2. IPCW. Weights on death - dementia ---------------------------------------------
-
-death_den <- glm(
-  competing_plr ~ cancer_v + bs(time, 3) + bs(age_0, 3) + sex + education + apoe4 +
-    as.factor(smoke1) + ht1 + bs(sbp1, 3) + bs(bmi1, 3) + as.factor(diabetes_prev) +
-    as.factor(smoke) + bs(sbp, 3) + bs(bmi, 3) + ht + ht_drug + hd_v + stroke_v + diab_v + cohort,
-  data = data_long,
-  family = quasibinomial
-)
-
-death_num <-
-  glm(
-    competing_plr ~ cancer_v + bs(time, 3) + bs(age_0, 3) + sex + education + apoe4 +
-      as.factor(smoke1) + ht1 + bs(sbp1, 3) + bs(bmi1, 3) + as.factor(diabetes_prev) + cohort,
-    data = data_long,
-    family = binomial
-  )
-

### 2021-07-19 02_R/05_plots.R

@@ -41,7 +41,7 @@ data_dem <- import(here::here("01_data", "clean_data", "dementia_long.RData"))
 plot_distribution <- data_long %>% 
   group_by(age_scale) %>% 
   count(status) %>%
-  mutate(prop = round(100*n/sum(n)),0) %>% 
+  mutate(prop = round(100*n/sum(n),2)) %>% 
   ungroup() %>% 
   filter(age_scale <= 90) %>% 
   ggplot(aes(age_scale, prop, fill = status)) +
@@ -52,15 +52,13 @@ plot_distribution <- data_long %>%
     fill = NULL,
     y = NULL,
     x = 'Age over follow-up') +
-  theme_minimal() +
-  theme(legend.position = "bottom",
-        legend.text = element_text(size=10)) +
-  theme(
-    strip.text.x = element_text(size = 10),
-        # strip.background = element_blank(),
+  theme_minimal(base_family = "serif") +
+  theme(legend.position = "left",
+        legend.text = element_text(size=12)) +
+  theme(strip.text.x = element_text(size = 11),
         strip.background = element_rect(fill=NA),
-        axis.text=element_text(size=10),
-        axis.title=element_text(size=10))
+        axis.text=element_text(size=12),
+        axis.title=element_text(size=12))
 
 # data_wide %>% 
 #   ggplot(aes(age_0)) +
@@ -115,3 +113,82 @@ plot_distribution <- data_long %>%
 #   facet_wrap(.~cancer)
 # 
 
+
+
+ggsave(filename = "plot_distribution.tiff",
+       plot = plot_distribution,
+       path = here::here("03_figs"),
+       device = "tiff",
+       width = 8,
+       height = 4.1,
+       dpi = "retina")
+
+
+# CIR curves --------------------------------------------------------------
+load(here::here("02_R", "models.RData")) 
+
+model1 <- models %>% pluck(1) %>% broom::tidy() %>% 
+  transmute(
+    time = time,
+    strata = strata,
+    cif = 1 - estimate,
+    conf.low2 = 1 - conf.low,
+    conf.high2 = 1 - conf.high
+  ) %>%
+  rename(conf.high = conf.low2,
+         conf.low = conf.high2)
+
+model2 <- models %>% pluck(2) %>% broom::tidy() %>% 
+  transmute(
+    time = time,
+    strata = strata,
+    cif = 1 - estimate,
[…]

### 2021-07-19 02_R/05b_plot_causes_death.R

@@ -33,18 +33,25 @@ causes_death <- count_cancer_death %>%
   geom_col() +
   ggthemes::scale_fill_tableau() +
   coord_flip() + 
-  theme_minimal() +
-  labs(y = "% of participants that died",
+  labs(y = "% of participants that died prior to a dementia diagnosis",
        x = NULL,
        fill = NULL) +
+  theme_minimal(base_family = "serif") +
   theme(legend.position = "bottom",
-        legend.text = element_text(size=10)) +
-  theme(
-    strip.text.x = element_text(size = 10),
-    # strip.background = element_blank(),
-    strip.background = element_rect(fill=NA),
-    axis.text=element_text(size=10),
-    axis.title=element_text(size=10))
+        legend.text = element_text(size=12)) +
+  theme(strip.text.x = element_text(size = 11),
+        strip.background = element_rect(fill=NA),
+        axis.text=element_text(size=12),
+        axis.title=element_text(size=12))
+
+ggsave(filename = "causes_death.tiff",
+       plot = causes_death,
+       path = here::here("03_figs"),
+       device = "tiff",
+       width = 8,
+       height = 4.1,
+       dpi = "retina")
+
 
 count_dementia_death <- data %>% 
   filter(competing_plr == 0) %>% 

### 2021-09-24 04_reports/manuscript.Rmd
Latest draft, including Sonja's, Kim's and Arfan's updates

@@
 ---
-title: "Towards a clearer causal question underlying the association between cancer and ADRD"
+title: "Towards a clearer causal question underlying the association between cancer and dementia"
 author: ""
@@
 These findings have motivated substantial research toward mechanistic explanations, including molecular and genetic pathways that may explain this association[@behrens2009;@harris2014;@nudelman2019;@Papin2020;@driverbiogeront2014;@olson2019;@li2021;@driverpin12015].
-These research inquiries inevitably lead to discussions of repurposing or augmenting current cancer chemotherapeutic for ADRD [@Snyder2017].
+These research questions have led to discussions of repurposing or augmenting current cancer treatments, including chemotherapy, for dementia [@Snyder2017].
 
-Nevertheless, inferring any treatment or mechanistic effects from the observed cancer-ADRD inverse association is not straightforward.
+Nevertheless, inferring any treatment or mechanistic effects from the observed cancer-dementia inverse association is not straightforward.
 Researchers have raised concerns related to the competing event of death, unmeasured confounding, and ascertainment error that could explain these results[@driverbiogeront2014;@ganguli2015].
-However, understanding these or other sources of bias first requires making explicit the causal question.
+However, understanding these or other sources of bias first requires making explicit a causal question.
 Moreover, enumerating an explicit causal question is one step toward tying a research study to a question that is relevant to decision-making[@didelez2016;@labrecque2017].
 
-To illustrate the complexities of inferring hypothetical or available treatments’ effects on ADRD from the observed cancer-ADRD association, we focus on a specific question conceptualizing the Pin1 enzyme as the target of intervention.
+To illustrate the complexities of inferring hypothetical or available cancer treatments’ effects on dementia from the observed cancer-dementia association, we focus on a specific question conceptualizing the Pin1 enzyme as the target of intervention.
 Previous animal studies have shown that Pin1 enzyme over-expression promotes tumorigenesis, while its down-regulation is attributed to mechanisms that contribute to neurodegeneration and amyloid deposition[@driverpin12015;@angelucci2017;@li2021].
-If we one day could develop a drug that increases Pin1 expression specifically in brain tissue in hopes of preventing dementia, we could pose the question as: _What is the effect of this Pin1-targeting drug on the risk of ADRD over time compared to standard treatments?_
+If we – hypothetically speaking –  one day could develop a drug that increases Pin1 expression specifically in brain tissue in hopes of preventing dementia, we could pose the question as: _What is the effect of this Pin1-targeting drug on the risk of dementia over time compared to standard treatments?_
 
-To explore how we might learn about this effect using real-world data on cancer and ADRD, we progressively build a causal directed acyclic graph (DAG) to connect this particular causal question to the observable data and the assumptions we rely on to study the effect.
-We exemplify these challenges and how they translate into the analytic decision using data collected from the Rotterdam Study, a population-based cohort study.
+To explore how we might learn about this effect using real-world data on cancer and dementia, we progressively build a causal directed acyclic graph (DAG) to connect this question to the observable data and the assumptions we rely on to study the effect.
+We exemplify these challenges and how they translate into the analytic decisions using data collected from the Rotterdam Study, a population-based cohort study.
+In our discussion, we describe how other investigators may expand this exercise for other possible causal questions, including repurposing existing chemotherapy regimens or identifying other novel drug targets.
 
@@
 
-If this hypothetical Pin1-targeting drug was developed, the best way to understand its effect on ADRD risk would be to have a well-conducted randomized trial in which we randomize eligible participants in late midlife (e.g., ages 50-60 years) to receive this drug or not, and closely monitor dementia diagnosis over a lengthy follow-up.
+If this hypothetical Pin1-targeting drug was developed, the best way to understand its effect on dementia risk would be to have a well-conducted randomized trial in which we randomize eligible participants in late midlife (e.g., ages 50-60 years) to receive this drug or not, and closely monitor dementia diagnosis over a lengthy follow-up.
 Since this drug is not currently available, at best we can use observational data on Pin1 expression measurements.
@@
 Previous dementia studies have described age, sex, educational level and race/ethnicity as a minimal adjusting set of covariates[@ospina2020].
-However, environmental and behavioral factors such as smoking may translate into Pin1 over-expression[@tan2010] and are also related to the development of dementia[@dementia_lancet].
+However, environmental and behavioral factors such as smoking may translate into Pin1 over-expression[@tan2010] and are also related to the development of dementia[@dementia_lancet] and should therefore be adjusted for.
+Although $L_1$ can be time-varying in nature, we only depict $L_1$ at one time-point for readability.
 
@@
 Currently, Pin1 expression is not an available biomarker for population-based research, so at best we can only rely on a proxy for it.
-Because Pin1 over-expression is present in tumors, and tumors are only measured through diagnosis, previous studies have considered Pin1 over-expression as the explanation behind the inverse association between cancer diagnosis and dementia, though Pin1 was not explicitly part of the research question[@driver2012;@musicco2013;@freedman2016;@bowles2017;@frain2017;@schmidt2017;@sun2020;@ording2020].
-Unlike measuring Pin1 at the same time to all participants (thought this would not necessarily mean this would be the ideal time to measure it, we discuss this point further in the discussion section), cancer diagnosis is a collected over follow-up.
+Because Pin1 over-expression is present in tumors, and tumors are only diagnosed through biopsies, previous studies have speculated that Pin1 over-expression partly explains the inverse association between cancer diagnosis and dementia, though Pin1 was not explicitly part of the research question [@driver2012;@musicco2013;@freedman2016;@bowles2017;@frain2017;@schmidt2017;@sun2020;@ording2020].
+Unlike measuring Pin1 at the same time for all participants (though this would not necessarily mean this would be the ideal time to measure it, we discuss this point further in the discussion section), cancer diagnosis is collected during follow-up.
 We depict this feature in Figure 2, where $C_t$ and $C_{t+1}$ represent _cancer diagnosis_ over time, the measured proxy of $P_{t}$ and $P_{t+1}$ respectively.
-Although this means we would measure the association between cancer diagnosis over time and ADRD in the observed data, we are assuming that the captured effect is only through the pathway that involves Pin1 expression over time.
-Furthermore, even if we have measurements of $C_{t+1}$ and $Y_{t+k}$ and some subset of $L_1$, our question focused on $P_t$ remains.
+Although this means we would measure the association between cancer diagnosis over time and dementia in the observed data, we are assuming that the captured effect is only through the pathway that involves Pin1 expression over time.
+That is, we only have measurements of $C_{t+1}$ and $Y_{t+k}$ and some subset of $L_1$, but our question remains focused on the effect of $P_t$.
+
+We note that this means we already are changing our question from one of estimating the effect of Pin1 to one that is testing the sharp causal null hypothesis that Pin1 has no effect on dementia in any individuals.
+That is, an association between cancer and dementia is at best evidence against a null effect, and it would take substantial and unknown assumptions about the relationship between Pin1 and cancer diagnosis to describe how the magnitude of a cancer-dementia association is related to the likely magnitude of a Pin1-dementia effect.
 
@@
 
-A challenge that arises when choosing cancer diagnosis as the proxy for Pin1 is defining the time zero, the time where eligibility criteria is met, where exposure is measured and the time where the screening for dementia would begin after having the exposure measured[@hernanInmortal2016].
-Emulating the eligibility criteria to participate in a trial for Pin-targeting drug will not necessarily be possible in a cohort setting recruiting participants for discordant reasons.
-The latter may not align with the time cancer diagnosis is measured since it happens over follow-up and this situation may introduce inmortal-time bias[@hernanInmortal2016].
-For example, a study performed using data from the Framingham study [@driver2012] defined the exposed group with cancer as those participants with prevalent or incident cancer diagnosis (alternatively defined as "ever cancer" [@hanson2016]).
+Another challenge that arises when choosing cancer diagnosis as the proxy for Pin1 is defining the time zero, i.e.
+the time when eligibility criteria are met, “treatment” is assigned, and screening for dementia would begin[@hernanInmortal2016].Emulating the eligibility criteria to participate in a trial for Pin1-targeting drugs will not necessarily be possible in a cohort setting recruiting participants for discordant reasons.
+Moreover, the latter may not align with the moment at which cancer diagnosis is measured since it happens continuously over follow-up and this situation may introduce immortal-time bias[@hernanInmortal2016].
+For example, a study performed using data from the Framingham Heart Study [@driver2012] defined the exposed group with cancer as those participants with prevalent or incident cancer diagnosis (alternatively defined as "ever cancer" [@hanson2016]).
 This meant that a participant who had cancer diagnosis over follow-up contributed all their person-time to the cancer arm, including the time prior to the cancer diagnosis.
-By defining the exposure after the time of inclusion to the cohort, only participants who remain alive and free of ADRD diagnosis over follow-up can be defined as "ever cancer" [@anderson1983, @hernanInmortal2016].
+By defining the exposure after the time of inclusion to the cohort, only participants who remain alive and free of dementia diagnosis over follow-up can be defined as "ever cancer" [@anderson1983, @hernanInmortal2016].
 
-This problem is partly avoided by recognizing the time-varying nature of cancer diagnosis.
-Several studies have considered cancer diagnosis as a time-dependent exposure [@white2013; @hanson2016; @bowles2017].
-The price we pay with this approximation is that implicitly, this means that Pin1 would over-express at the time of cancer diagnosis and not before, which is biologically implausible.
+This problem is partly mitigated by recognizing the time-varying nature of cancer diagnosis.
+Some studies have considered cancer diagnosis as a time-dependent exposure [@white2013; @hanson2016; @bowles2017].
+The price we pay for this approximation is that implicitly, this means that Pin1 would over-express at the time of cancer diagnosis and not before, which is biologically implausible.
+The time between first biological changes that eventually can lead to cancer and cancer manifestation can range between five and forty years[@nadler2013].
 Moreover, cancer diagnosis will only be measured in the subset of participants who are alive over follow-up.
@@
 In addition, we added an arrow between $L_1$ and $D_{t+1}$, since covariates such as smoking may affect Pin1 over-expression but also affect the risk of death due to other causes such as from chronic obstructive pulmonary disease.
-Although $L_1$ can be time-varying in nature, we only depict $L_1$ at one time-point for readability.
 
@@
 
-Another challenge to adress is related to death as a competing event for dementia, represented in Figure 4.
+Another challenge to address is related to death as a competing event for dementia, represented in Figure 4.
 For interpretability we exclude the time-varying process of cancer diagnosis and focus on Pin1 (and cancer diagnosis) as it had been measured in all participants at time $t+1$.
@@
 However, in the context of cancer diagnosis as the proxy for Pin1 over-expression, we cannot rule out the effect of cancer diagnosis on death, represented as the arrow between $C_{t+1}$ and $D_{t+k}$.
-In this setting, a total effect of $P_{t+1}$ in $Y_{t+k}$ would include the causal pathway mediated by the effect of cancer diagnosis in mortality, which may translate into an inverse association[@young2020].
+In this setting, a total effect of $P_{t+1}$ in $Y_{t+k}$ includes the indirect causal pathway mediated by the effect of cancer diagnosis on mortality, which may translate into an inverse association[@young2020].
 
 To isolate the direct effect of $P_{t+1}$ in $Y_{t+k}$ through measurement of $C_{t+1}$ we have at least two alternatives of causal questions we can ask.
-We can either imagine a causal question where we decompose the effect of cancer by the different mechanisms that affect dementia and death separately[@stensrud2020].
-Alternatively, we can define an scenario where death could have been prevented.
-The latter is defined as the controlled direct effect, where death is considered as a censoring event and it relies on the assumption that we have measured all $L$ to block the pathway $Y_{t+k} \leftarrow L_2  \rightarrow D_{t+k} \rightarrow C_{t+1} \rightarrow P_{t+1}$.
-Previous studies have defined death as a censoring event [@frain2017], although failed to explicitly define how to comply with the independent censoring assumption.
-Moreover, adjusting for confounders for dementia (such as only adjusting for $L$) will be insufficient to block the pathway mediated by death, time-varying covariates that represent $L_2$, such as cardiovascular conditions should be considered.
+One, we can imagine a causal question where we decompose the total effect of cancer by the different pathways that affect dementia and death separately[@stensrud2020].
+Two, we can define a scenario where death could have been prevented.
+The latter is defined as the controlled direct effect, where death is considered as an independent censoring event by relying on the assumption that we have measured all shared causes $L_2$ to block the pathway $Y_{t+k} \leftarrow L_2 \rightarrow D_{t+k} \leftarrow C_{t+1} \leftarrow P_{t+1}$.
+Previous studies have defined death as a censoring event [@frain2017], although failed to explicitly define  the independent censoring assumption and did not consider its plausibility.
+Moreover, adjusting for time-fixed shared causes between dementia and death may be insufficient to block this pathway, and therefore time-varying measurements of $L_2$ should be considered.
 
-To summarize, the complexity of the causal structure that describes the effect of Pin1 through the proxy of cancer diagnosis in the risk of ADRD illustrate the potential sources of bias, as observed in Figure 5.
+To summarize, the complexity of the causal structure that describes the effect of Pin1 on dementia risk while proxying Pin1 with cancer diagnosis illustrates the potential sources of bias, as observed in Figure 5.
 Even so, this is a simplified version since we omitted additional arrows from $L_1$ to $C_t$ and $C_{t+2}$ for brevity, as well as other sources of measurement error and the time-varying nature of all nodes and feedback loops between them, which would further complicate interpretability[@whatif2020].
@@
 We now turn to an application where we show how these challenges translate into analytic decisions.
-We will show ways to mitigate or better understand bias to the best of the available data's abilities in an attempt to inform the possible effect of Pin1 on all cause-dementia.
+We will show ways to mitigate or better understand bias to the best of the available data's abilities in an attempt to inform the possible effect of Pin1 on dementia risk.
 
@@
 Recruitment and initial assessments were held between 1990 and 1993, a second wave of recruitment was held between 2000 and 2001.
-Participants from the first subcohort had follow-up visits between 1993-1995, 1997-1999, 2002-2005, and 2008-2010, while the second subcohort had follow-up visits between 2004 and 2005, and between 2011 and 2012[@ikram2020].
+Participants from the first sub-cohort had follow-up visits between 1993-1995, 1997-1999, 2002-2005, and 2008-2010, while the second sub-cohort had follow-up visits between 2004 and 2005, and between 2011 and 2012[@ikram2020].
 All participants had data on history of cancer and dementia and incident diagnosis, collected from medical records of general practitioners (including hospital discharge letters) and through linkage with national registries.
@@
 All participants were followed from study entry until dementia diagnosis, death or 20 years after their individual baseline date, whichever occurred first.
-Given that participants from the second subcohort were followed for 15 years, we assume that they would have had a similar distribution of dementia risk and mortality as the first subcohort, between year 15 and 20.
+Given that participants from the second sub-cohort were followed for 15 years, we assume that they would have had a similar distribution of dementia risk and mortality as the first subcohort, between year 15 and 20.
 
@@
 Scenario A replicates the setting that defines cancer proxy as _"cancer ever vs.
-never"_ [@driver2012], meaning we compare dementia outcomes in individuals who ever develop cancer during follow-up to those who were not observed to develop cancer during follow-up.
+never"_[@driver2012], meaning we compare dementia outcomes in individuals who ever develop cancer during follow-up to those who were not observed to develop cancer during follow-up.
 Scenario B defines cancer diagnosis as time-varying meaning that time prior to cancer diagnosis is allocated to the non-exposed arm, and the time after cancer diagnosis to the exposed arm.
@@
 In Scenario A, weights were defined as the inverse of the probability of cancer diagnosis conditional on baseline covariates such as age at study entry, sex, educational attainment, cohort, smoking status.
-In contrast, for Scenario B, weights were defined to represent the product of the inverse probability of being diagnosed with cancer over time, conditional on the time-varying covariate history prior to cancer diagnosis [@hernan2000].
+In contrast, for Scenario B, weights were defined to represent the product of the inverse probability of being diagnosed with cancer over time, conditional on the time-varying covariate history prior to cancer diagnosis[@hernan2000].
 Baseline covariates included age at study entry, sex, ApoE4 status, educational attainment and the time-varying covariates such as smoking status, systolic blood pressure, body mass index and prevalent and incident hypertension and diabetes.
 
-Inverse probability censoring weights for death were calculated to relax the independent censoring assumption.
+Inverse probability censoring weights for death were calculated, assuming independent censoring conditional on measured covariates.
 In Scenario A, weights represent the inverse of the probability of not dying conditional on cancer diagnosis (ever vs.
 never) and baseline covariates such as age, educational attainment, ApoE4, and baseline measurements of smoking status, hypertension status, systolic blood pressure, BMI, history of diabetes and cohort.
-For individuals who died, their censoring weight is zero [@whatif2020].
+For individuals who died, their censoring weight is zero[@whatif2020].
 In Scenario B time-varying weights represent the product of the inverse probability of surviving in each year prior to _t_, conditional on the measured shared causes of death and dementia.
-For an individual who has died by time _t_, the year _t_ censoring weight is zero [@young2020].
-Weights were fitted including the same covariates used to estimate weights for time-varying cancer diagnosis, though we additionally added time-varying cancer, stroke, and heart disease diagnosis as predictors for death.Further details on modeling specifications and weights assessment are presented as **Supplementary material**.
+For an individual who has died by time _t_, the year _t_ censoring weight is zero[@young2020].
+Weights were fitted including the same covariates used to estimate weights for time-varying cancer diagnosis, though we additionally added time-varying cancer, stroke, and heart disease diagnosis as predictors for death.
+Further details on modeling specifications and weights assessment are presented as **Supplementary material**.
 
-To estimate the controlled direct effect of Pin1 in the risk of ADRD, we compared the complement of a weighted Kaplan-Meier survival estimator for participants who developed cancer versus those who did not, with time indexed in years.
+To estimate the controlled direct effect of Pin1 on the risk of dementia, we compared the complement of a weighted Kaplan-Meier survival estimator for participants who developed cancer versus those who did not, with time indexed in years.
 The weights are time-varying by follow-up year, defined as a product of the year-specific inverse probability of treatment weights and the inverse probability of censoring by death weights.
@@
 
-For illustrative and comparative purposes we calculated hazard ratios (HR).
-Hazards, unlike risks, inherently condition on surviving both dementia and death, as such a causal interpretation is problematic [@young2020].
+Estimates are presented as 20-year risk differences (RD) and risk ratios (RR).
+All 95% confidence intervals were calculated using percentile-based bootstrapping based on 500 bootstrap samples.
+For illustrative and comparative purposes, we also calculated hazard ratios (HR).
+Hazards, unlike risks, inherently condition on surviving both dementia and death, and as such a causal interpretation is problematic[@young2020].
 
-Since the conditional independent censoring assumption is untestable, we compute Peterson upper and lower bounds [@peterson1976] to represent: 1) the extreme scenario of independence, that refers to an scenario were those who died would never develop dementia (lower bound) and 2) complete dependency, that refers to an scenario where those who died would have a dementia prior to death (upper bound).
+Since the conditional independent censoring assumption is untestable, we compute Peterson upper and lower bounds[@peterson1976] to represent: (1) the extreme scenario of independence, that refers to a scenario were those who died would never develop dementia (lower bound) and (2) complete dependency, that refers to an scenario where those who died would have a dementia diagnosis prior to death (upper bound).
 The lower bound is calculated with the Aalen-Johansen estimator treating death as a competing event, and the upper bound is calculated with the Kaplan-Meier estimator for the combined outcome of dementia or death.
 
-All analysis were performed using R; code is provided in supplementary material and available in https://github.com/palolili23/2021_cancer_dementia.
+All analysis were performed using R. Code is provided in supplementary material and available in https://github.com/palolili23/2021_cancer_dementia.
 
@@
 Over follow-up, `r yes_cancer` developed cancer, with a median age of cancer diagnosis at `r cancer_time`.
-From the total sample, `r dem_total` had dementia over follow-up with a median age of at `r dem_time` years.
-Among participants with incident cancer, `r cancer_dem` had dementia diagnosis, `r cancer_death` died over follow-up, and `r cancer_alive` remain alive and dementia-free at 20 years since study entry.
+From the total sample, `r dem_total` were diagnosed with dementia over follow-up with a median age of at `r dem_time` years.
+Among participants with incident cancer, `r cancer_dem` had dementia diagnosis, `r cancer_death` died over follow-up, and `r cancer_alive` remained alive and dementia-free at 20 years since study entry.
 In contrast, among participants free of cancer diagnosis over follow-up, `r no_cancer_dem` were diagnosed with dementia, `r no_cancer_death` died over follow-up, and `r no_cancer_alive` were alive and dementia-free at the end of follow-up.
-The proportion of participants in each possible status over follow-up and leading causes of death for both arms are presented as Figure 1 an 2 in Supplementary Material.
+The proportion of participants in each possible status over follow-up and leading causes of death for both arms are presented as Figure 1 and 2 in Supplementary Material.
 
-Results for all scenarios are present in Table 2.
+Results for all scenarios are presented in Table 2.
 Using Scenario A’s analytic design and without adjusting for confounding or selection bias due to conditioning on death, we observe a protective association with a risk ratio (RR, 95% Confidence interval) of `r complete_table[3,4]` and a hazard ratio [HR, (95% Confidence interval)] of `r complete_table[3,5]`.
-Though adjusting for measured confounding only minimally changes the observed association, the association is closer to the null after including censoring weights for death [RR: `r complete_table[5,4]`; HR: `r complete_table[5,5]`].
+Though adjusting for measured confounding only minimally changed the observed association, the association is closer to the null after including censoring weights for death [RR: `r complete_table[5,4]`; HR: `r complete_table[5,5]`].
 In contrast, using Scenario B's analytic design, the fully adjusted model results in a RR of RR: `r complete_table[6,4]` and a HR of `r complete_table[6,5]`, though confidence intervals cross the null.
-The cumulative incidence curves for dementia for participants free of cancer diagnosis and with incident cancer diagnosis, under scenarios A and B, reflect that the difference between groups changes over time and flips direction when considering cancer as a time-varying proxy.
-~~In addition, the extreme scenarios of dependence between death and dementia over follow-up represented as bounds are wide and include the null value (`r bounds_tv_ipw`).~~
+The cumulative incidence curves for dementia for participants free of cancer diagnosis and with incident cancer diagnosis, under Scenarios A and B, reflect that the difference between groups changes over time and flips direction when considering cancer as a time-varying proxy.
+Bounds (rather than point estimates) were wide and covered the null **(lower bound RR 0.39; upper bound RR 2.08).” (`r bounds_tv_ipw`).**
 
@@
 
-Several observational studies have defined “cancer diagnosis” as exposure, although this does not represent a target for potential intervention or a modifiable risk factor.
+Several observational studies have defined “cancer diagnosis” as an exposure, although this does not represent a target for potential intervention or a modifiable risk factor.
 Instead, this variable is commonly used to represent a mechanism of interest that could not be measured.
@@
 never", this definition introduces immortal time bias.
-All results pertaining to this definition had an inverse association between cancer diagnosis and dementia, while results that did not introduce this particular form of immortal-time bias had point estimates close to the null.
+All results pertaining to this definition had an inverse association between cancer diagnosis and dementia, while results that did not introduce this particular form of immortal-time bias had point estimates closer to the null.
 Although we attempt to prevent this bias with statistical methods, we can only fully prevent it by having a clear definition of time-zero.
-This definition does not depend on the collected data nor in analytic decisions.
-It relies on a deeper discussion related to when would be the best moment to measure this biomarker and to what purpose.
+Importantly, this definition does not depend on the collected data nor in analytic decisions: it relies on a deeper discussion related to when would be the best moment to potentially intervene on this biomarker and to what purpose.
 Thus, we hope that these unsolved questions guide future discussions and data collection efforts.
 
-On the other hand, death as competing event is a challenge that has some straightforward strategies, beginning first and foremost by choosing the causal parameter (or estimand) of interest [@young2020, @rojas_medrxiv].
-In this study we chose the controlled direct effect, which represents the effect of Pin1 (or cancer) in a setting where death due to cancer and other causes could have been prevented, yet without an explicit intervention, which is yet ambiguous.
-This is different than conceiving a drug-target that increases Pin1 expression only brain tissue, with no side effects that could increase cancer risk (and thus, death due to cancer).
-As opposed to prior studies that implicitly address a direct effect, and who define censoring as ignorable[@frain2017], we show how point estimates change substantially when we include weights for death to relax the independent censoring assumption [@vangeloven2014, @willems2018, @young2020] when the estimand of interest is the controlled direct effect.
+On the other hand, death as a competing event is a challenge that has some straightforward strategies, beginning first and foremost by choosing the causal parameter (or estimand) of interest [@young2020, @rojas_medrxiv].
+In this study we chose the controlled direct effect, which represents the effect of Pin1 (or cancer) in a setting where death due to cancer and other causes could have been prevented, yet without an explicit intervention, which makes it ambiguous.
+This is different than conceiving a drug-target that increases Pin1 expression only in brain tissue, with no side effects that could increase cancer risk (and thus, death due to cancer).
+As opposed to prior studies that implicitly address a direct effect, and that define censoring as ignorable[@frain2017], we show how point estimates change substantially when we include weights for death to relax the independent censoring assumption [@vangeloven2014, @willems2018, @young2020] when the estimand of interest is the controlled direct effect.
 Bounds to assess extreme scenarios of dependence between death and dementia[@peterson1976] illustrate the wide range of possible point estimates that cross the null.
-This shows that even with the effort of adjusting for time-varying covariates, we may be far from meeting this assumption and thus better efforts to measure shared causes of dementia and death are needed.
-In addition, presenting the proportion of participants that died prior to ADRD diagnosis in each arm, as well as the proportion of participants in each status over time, improves transparency and puts in evidence the limitations of the data to answer this question.
+This shows that even with the effort of adjusting for time-varying covariates, we may be far from meeting this assumption and thus better efforts to measure shared (time-varying) causes of dementia and death are needed.
+In addition, presenting the proportion of participants that died prior to dementia diagnosis in each arm, as well as the proportion of participants in each status over time, improves transparency and puts in evidence the limitations of the data to answer this question.
 
-Pin1 is only one of the several mechanisms proposed about the inverse cancer-dementia relationship.
+Pin1 is only one of the several mechanisms proposed by other investigators as an explanation for the inverse cancer-dementia relationship.
 Certainly, cancer diagnosis represents a complex and heterogeneous health condition that exceeds the representation of Pin1 expression.
 To understand how – if at all – the cancer-dementia association informs the potential effects of any other mechanism or treatment strategy, and its connection to collected data, we may require different causal representations.
-For example, if researchers used this association to inform the possible effects of different chemotherapeutics on cognitive decline among patients undergoing treatment for cancer, different challenges would arise for mapping the observed association to the hypothetical randomized trial underlying this research question.
+For example, if researchers used this association to inform the possible effects of different chemotherapeutics on cognitive decline among patients undergoing treatment for cancer, different challenges would arise for mapping the observed association to the hypothetical randomized trial underlying that new research question.
 Notably, that target trial, unlike the one considered here, would include cancer diagnosis as part of the eligibility, and thus researchers would need to instead grapple with how using data on persons without cancer is useful and useable [@huitfeldt2016].
-Because each question requires thinking about bias anew and each question brings its own set of challenges and opportunities, we believe it is crucial to have the discussion about the question or set of questions that are relevant first, as this can also inform efforts to collect more suitable data.
-Furthermore, we underscore that this is just one case study for how observed associations between two diseases or health states may be disentangled to more transparently unveil possible mechanisms (and sources of bias) behind them.
+Each question requires thinking about bias anew and each question brings its own set of challenges and opportunities.
+
+We underscore that our study is just one case study for how observed associations between two diseases or health states may be disentangled to more transparently unveil possible mechanisms (and sources of bias) behind them, and how to maximize what we can learn about the potential mechanisms given available or readily-collected data.
+We hope to stimulate researchers first have a discussion about the exact research question, rather than trying to replicate the cancer and dementia relation in their dataset.
+Such discussion is needed to obtain meaningful conclusions.
 
@@
 $L_1$ represents shared causes of Pin1 and Dementia, such as smoking.
-To isolate the effect of $P_t$ and $Y_{y+k}$ we need to block the backdoor path $Y_{t+k} \leftarrow L_1 \rightarrow P_t$.
+To isolate the effect of $P_t$ on $Y_{y+k}$ we need to block the backdoor path $Y_{t+k} \leftarrow L_1 \rightarrow P_t$.
 
@@
 **Figure 3.
-Time-varying cancer diagnosis and inmortal-time bias.**  This causal graph is an expansion of Figure 2, where $D_{t+1}$ represents death at time $t+1$.
+Time-varying cancer diagnosis and immortal-time bias.**  This causal graph is an expansion of Figure 2, where $D_{t+1}$ represents death at time $t+1$.
 Cancer diagnosis at $t+1$ can only be measured among those who are alive at $t+1$.
@@
 Pin1 may affect the risk of death through cancer diagnosis, represented as an arrow between $C_{t+1}$ and  $D_{t+k}$.
-To isolate the direct effect of $P_{t+1}$ and $Y_{y+k}$, we have to block the backdoor pathway from $Y_{t+k} \leftarrow L_2 \rightarrow D_{t+k} \leftarrow C_{t+1} \rightarrow P_{t+1}$ and $Y_{t+k} \rightarrow D_{t+k} \leftarrow L_1 \rightarrow P_{t+1}$
+To isolate the direct effect of $P_{t+1}$ and $Y_{y+k}$, we have to block the backdoor pathway from $Y_{t+k} \leftarrow L_2 \rightarrow D_{t+k} \leftarrow C_{t+1} \leftarrow P_{t+1}$ and $Y_{t+k} \leftarrow D_{t+k} \leftarrow L_1 \rightarrow P_{t+1}$
 
-```{tikz, echo = FALSE, fig.cap = "Direct effect of Pin1 in the risk of ADRD"}
+```{tikz, echo = FALSE, fig.cap = "Direct effect of Pin1 in the risk of dementia"}
 \usetikzlibrary{arrows}
@@
 
+**Figure 6.
+Risk of dementia under elimination of death over 20 years of follow-up by cancer diagnosis status.** Panel A.
+represents the scenario where cancer was considered as “ever vs.
+never”.
+Panel B.
+represents the scenario where cancer was considered as a time-varying proxy.
+
 ## Supplementary material
 
-### A. Modeling description
+### 1. Modeling description
 
-### B. Distribution of participants under each health status, by age over follow-up
+### 2. Figure 1. Distribution of participants under each health status, by age over follow-up
 
@@
 
-### C. Causes of death for participants who died prior to dementia diagnosis
+### 3. Figure 2. Causes of death for participants who died prior to dementia diagnosis
 
@@
 ```
-
-Causes of death were defined by ICD10 codes...xxx

### 2021-09-24 02_R/03_descriptives.R

@@ -83,7 +83,7 @@ dem_time <- data_long %>%
 ## Baseline covs
 
 mean_age <- data_wide %>% 
-  summarize(temp = round(mean(age_0),  2)) %>% 
+  summarize(temp = round(mean(age_0),  1)) %>% 
   #           b = round(sd(age_0),  2)) %>%
   # mutate(temp = paste0(a, " (SD:", b, ")")) %>% 
   pull(temp)

### 2021-10-21 04_reports/manuscript.Rmd
updates

@@
 All participants were followed from study entry until dementia diagnosis, death or 20 years after their individual baseline date, whichever occurred first.
-Given that participants from the second sub-cohort were followed for 15 years, we assume that they would have had a similar distribution of dementia risk and mortality as the first subcohort, between year 15 and 20.
+Given that participants from the second sub-cohort were followed for 15 years, we assume that they would have had a similar distribution of dementia risk and mortality as the first subcohort, between year 15 and 20 of follow-up.
 

### 2021-10-26 04_reports/manuscript.Rmd
version to submit to Epidemiology

@@
 bibliography: cancer_dementia.bib
-csl: AnD_journal.csl
+csl: epidemiology.csl
 ---
@@
 
-## 1. Introduction
+## Abstract
+
+**Background:** Several observational studies have described an inverse association between cancer diagnosis and dementia.
+Several biological mechanisms and sources of bias have been proposed.
+Since bias cannot be assessed without a clear causal question, we propose to study the controlled direct effect of the protein Pin1 on the risk of dementia.
+With this case-study, we will outline the needed assumptions and potential sources of bias and exemplify how these translate into the analytic decisions under the guidance of causal directed acyclic graphs.
+**Methods:** We used data from the Rotterdam Study, a population-based cohort.
+We illustrate the association between cancer diagnosis (as proxy for Pin1) and dementia diagnosis under two scenarios: (A) cancer proxy as "cancer ever vs.
+never"; (B) cancer diagnosis as time-varying proxy.
+We address confounding and censoring for death with inverse probability weights and contrast the complement of a weighted Kaplan-Meier survival estimator at 20-years of follow-up.
+**Results:** Out of 3634 participants, 899 (25%) were diagnosed with cancer, of whom 53 (6%) had dementia, and 567 (63%) died.
+Among those without cancer, 15% (411) were diagnosed with dementia, and 667 (24%) died over follow-up.
+In Scenario A, the risk ratio (and 95%CI) after adjustment for confounding was 0.70 (0.49, 0.93); and with additional weights for death 0.91 (0.65, 1.19).
+In Scenario B, the risk ratio was 0.98 (0.71,1.19) and 1.05 (0.79, 1.29), respectively.
+**Conclusion**: Being explicit about the underlying mechanism of interest and defining a clear causal contrast is key to maximize what we can learn about this association given available or readily-collected data, detecting and preventing potential biases.
+
+\newpage
+
+## Towards a clearer causal question underlying the association between cancer and dementia
 
@@
 Weights were fitted including the same covariates used to estimate weights for time-varying cancer diagnosis, though we additionally added time-varying cancer, stroke, and heart disease diagnosis as predictors for death.
-Further details on modeling specifications and weights assessment are presented as **Supplementary material**.
+Further details on modeling specifications and weights assessment are presented in the **eAppendix**.
 
@@
 In contrast, among participants free of cancer diagnosis over follow-up, `r no_cancer_dem` were diagnosed with dementia, `r no_cancer_death` died over follow-up, and `r no_cancer_alive` were alive and dementia-free at the end of follow-up.
-The proportion of participants in each possible status over follow-up and leading causes of death for both arms are presented as Figure 1 and 2 in Supplementary Material.
+The proportion of participants in each possible status over follow-up and leading causes of death for both arms are presented as eFigure 1 and 2 in the eAppendix.
 
@@
 The cumulative incidence curves for dementia for participants free of cancer diagnosis and with incident cancer diagnosis, under Scenarios A and B, reflect that the difference between groups changes over time and flips direction when considering cancer as a time-varying proxy.
-Bounds (rather than point estimates) were wide and covered the null **(lower bound RR 0.39; upper bound RR 2.08).” (`r bounds_tv_ipw`).**
+Bounds (rather than point estimates) were wide and covered the null (Lower bound RR: 0.39; Upper bound RR: 2.08).”**
 
@@
 This is different than conceiving a drug-target that increases Pin1 expression only in brain tissue, with no side effects that could increase cancer risk (and thus, death due to cancer).
-As opposed to prior studies that implicitly address a direct effect, and that define censoring as ignorable[@frain2017], we show how point estimates change substantially when we include weights for death to relax the independent censoring assumption [@vangeloven2014, @willems2018, @young2020] when the estimand of interest is the controlled direct effect.
+As opposed to prior studies that implicitly address a direct effect, and that define censoring as ignorable[@frain2017], we show how point estimates change substantially when we include weights for death to relax the independent censoring assumption [@weave2012, @vangeloven2014, @willems2018, @young2020] when the estimand of interest is the controlled direct effect.
 Bounds to assess extreme scenarios of dependence between death and dementia[@peterson1976] illustrate the wide range of possible point estimates that cross the null.
@@
 Such discussion is needed to obtain meaningful conclusions.
-
-\newpage
-
-## References
-
-<div id="refs"></div>
 
@@
 $L_1$ represents shared causes of Pin1 and Dementia, such as smoking.
-To isolate the effect of $P_t$ on $Y_{y+k}$ we need to block the backdoor path $Y_{t+k} \leftarrow L_1 \rightarrow P_t$.
+To isolate the effect of $P_t$ on $Y_{t+k}$ we need to block the backdoor path $Y_{t+k} \leftarrow L_1 \rightarrow P_t$.
+Gray nodes and arrows, which represent additional data-generating mechanisms, will be described progressively in the following figures.
 
@@
 Incident cancer diagnosis is represented as $C_t$ and $C_{t+1}$.
-To isolate the effect of $P_t$ and $P_{t+1}$ on $Y_{y+k}$ we need to block the backdoor path $Y_{t+k} \leftarrow L_1 \rightarrow P_t$ and $Y_{t+k} \leftarrow L_1 \rightarrow P_{t+1}$.
+To isolate the effect of $P_t$ and $P_{t+1}$ on $Y_{t+k}$ we need to block the backdoor path $Y_{t+k} \leftarrow L_1 \rightarrow P_t$ and $Y_{t+k} \leftarrow L_1 \rightarrow P_{t+1}$.
 Although we represent $L_1$ as a single node for readability, $L_1$ is time-varying too.
@@
 Time-varying cancer diagnosis and immortal-time bias.**  This causal graph is an expansion of Figure 2, where $D_{t+1}$ represents death at time $t+1$.
-Cancer diagnosis at $t+1$ can only be measured among those who are alive at $t+1$.
+Cancer diagnosis at $t+1$ can only be measured among those who are alive at $t+1$
 
@@
 Death as competing event of dementia.** In this graph we only focus attention to the exposure as if it was measured for all at time $P_{t+1}$.
-We include $D_{t+k}$ as death at time $t+k$, $L_2$ represents possible shared causes of dementia and death (such as cardiovascular comorbidities).
+We include $D_{t+k}$ as death at time $t+k$ since $Y_{t+1}$ is only observable when participants are alive at $t+k$.
+$L_2$ represents possible shared causes of dementia and death (such as cardiovascular comorbidities).
 Pin1 may affect the risk of death through cancer diagnosis, represented as an arrow between $C_{t+1}$ and $D_{t+k}$.
-To isolate the direct effect of $P_{t+1}$ and $Y_{y+k}$, we have to block the backdoor pathway from $Y_{t+k} \leftarrow L_2 \rightarrow D_{t+k} \leftarrow C_{t+1} \leftarrow P_{t+1}$ and $Y_{t+k} \leftarrow D_{t+k} \leftarrow L_1 \rightarrow P_{t+1}$
+To isolate the direct effect of $P_{t+1}$ and $Y_{t+k}$, we have to block the backdoor pathway from $Y_{t+k} \leftarrow L_2 \rightarrow D_{t+k} \leftarrow C_{t+1} \leftarrow P_{t+1}$ and $Y_{t+k} \leftarrow D_{t+k} \leftarrow L_1 \rightarrow P_{t+1}$
 
@@
 
-## Supplementary material
+\newpage
 
-### 1. Modeling description
+## References
 
-### 2. Figure 1. Distribution of participants under each health status, by age over follow-up
-
-```{r}
-knitr::include_graphics(here::here("03_figs", "plot_distribution.tiff"))
-```
-
-### 3. Figure 2. Causes of death for participants who died prior to dementia diagnosis
-
-```{r}
-knitr::include_graphics(here::here("03_figs", "causes_death.tiff"))
-```
+<div id="refs"></div>

### 2021-10-26 04_reports/supplemental.Rmd
version to submit to Epidemiology

---
title: "supplemental"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Supplementary material

### 1. Modeling description

### 2. Figure 1. Distribution of participants under each health status, by age over follow-up

```{r}
knitr::include_graphics(here::here("03_figs", "plot_distribution.tiff"))
```

### 3. Figure 2. Causes of death for participants who died prior to dementia diagnosis

```{r}
knitr::include_graphics(here::here("03_figs", "causes_death.tiff"))
```

### 2021-11-03 04_reports/manuscript.Rmd
need to fix all references and (.) order; footnote on tables; add 0 decimal; bold eappendix
submitted manuscript!

@@
 With this case-study, we will outline the needed assumptions and potential sources of bias and exemplify how these translate into the analytic decisions under the guidance of causal directed acyclic graphs.
-**Methods:** We used data from the Rotterdam Study, a population-based cohort.
-We illustrate the association between cancer diagnosis (as proxy for Pin1) and dementia diagnosis under two scenarios: (A) cancer proxy as "cancer ever vs.
-never"; (B) cancer diagnosis as time-varying proxy.
-We address confounding and censoring for death with inverse probability weights and contrast the complement of a weighted Kaplan-Meier survival estimator at 20-years of follow-up.
+
+**Methods:**We used data from the Rotterdam Study, a population-based cohort.
+We estimate the association between cancer diagnosis (as proxy for Pin1) and dementia diagnosis using two different proxy methods and with confounding and censoring for death addressed with inverse probability weights.
+We estimate and compare the complements of a weighted Kaplan-Meier survival estimator at 20-years of follow-up.
+
 **Results:** Out of 3634 participants, 899 (25%) were diagnosed with cancer, of whom 53 (6%) had dementia, and 567 (63%) died.
 Among those without cancer, 15% (411) were diagnosed with dementia, and 667 (24%) died over follow-up.
-In Scenario A, the risk ratio (and 95%CI) after adjustment for confounding was 0.70 (0.49, 0.93); and with additional weights for death 0.91 (0.65, 1.19).
-In Scenario B, the risk ratio was 0.98 (0.71,1.19) and 1.05 (0.79, 1.29), respectively.
-**Conclusion**: Being explicit about the underlying mechanism of interest and defining a clear causal contrast is key to maximize what we can learn about this association given available or readily-collected data, detecting and preventing potential biases.
+Depending on the confounding and selection bias control, and the way in which cancer was used as a time-varying proxy, the risk ratio ranged from 0.70 (0.49, 0.93) to 1.05 (0.79, 1.29).
+
+**Conclusion:** Being explicit about the underlying mechanism of interest and defining a clear causal contrast is key to maximize what we can learn about this association given available or readily-collected data, and to defining, detecting, and preventing potential biases.
 
@@
 
-Many observational studies have consistently found that individuals with cancer have a lower risk of developing dementia when compared to individuals with no history of cancer [@ma2014;@hanson2016;@vanderwillik2018;@ospina2020].
+Many observational studies have consistently found that individuals with cancer have a lower risk of developing dementia when compared to individuals with no history of cancer[@ma2014;@hanson2016;@vanderwillik2018;@ospina2020].
 These findings have motivated substantial research toward mechanistic explanations, including molecular and genetic pathways that may explain this association[@behrens2009;@harris2014;@nudelman2019;@Papin2020;@driverbiogeront2014;@olson2019;@li2021;@driverpin12015].
-These research questions have led to discussions of repurposing or augmenting current cancer treatments, including chemotherapy, for dementia [@Snyder2017].
+These research questions have led to discussions of repurposing or augmenting current cancer treatments, including chemotherapy, for dementia[@Snyder2017].
 
@@
 In the observational setting, confounding could explain the observed association between Pin1 and dementia.
-In the causal diagram [@whatif2020] in Figure 1, we present Pin1 expression as $P_{t}$ and dementia diagnosis by time $t+k$ as $Y_{t+k}$.
+In the causal diagram[@whatif2020] in Figure 1, we present Pin1 expression as $P_{t}$ and dementia diagnosis by time $t+k$ as $Y_{t+k}$.
 Both may share causes $L_1$, and to assess the causal relationship between $P_{t}$ and $Y_{t+k}$ we would thus adjust for $L_1$.
@@
 Moreover, the latter may not align with the moment at which cancer diagnosis is measured since it happens continuously over follow-up and this situation may introduce immortal-time bias[@hernanInmortal2016].
-For example, a study performed using data from the Framingham Heart Study [@driver2012] defined the exposed group with cancer as those participants with prevalent or incident cancer diagnosis (alternatively defined as "ever cancer" [@hanson2016]).
+For example, a study performed using data from the Framingham Heart Study[@driver2012] defined the exposed group with cancer as those participants with prevalent or incident cancer diagnosis (alternatively defined as "ever cancer"[@hanson2016]).
 This meant that a participant who had cancer diagnosis over follow-up contributed all their person-time to the cancer arm, including the time prior to the cancer diagnosis.
-By defining the exposure after the time of inclusion to the cohort, only participants who remain alive and free of dementia diagnosis over follow-up can be defined as "ever cancer" [@anderson1983, @hernanInmortal2016].
+By defining the exposure after the time of inclusion to the cohort, only participants who remain alive and free of dementia diagnosis over follow-up can be defined as "ever cancer"[@anderson1983, @hernanInmortal2016].
 
 This problem is partly mitigated by recognizing the time-varying nature of cancer diagnosis.
-Some studies have considered cancer diagnosis as a time-dependent exposure [@white2013; @hanson2016; @bowles2017].
+Some studies have considered cancer diagnosis as a time-dependent exposure[@white2013; @hanson2016; @bowles2017].
 The price we pay for this approximation is that implicitly, this means that Pin1 would over-express at the time of cancer diagnosis and not before, which is biologically implausible.
@@
 The latter is defined as the controlled direct effect, where death is considered as an independent censoring event by relying on the assumption that we have measured all shared causes $L_2$ to block the pathway $Y_{t+k} \leftarrow L_2 \rightarrow D_{t+k} \leftarrow C_{t+1} \leftarrow P_{t+1}$.
-Previous studies have defined death as a censoring event [@frain2017], although failed to explicitly define  the independent censoring assumption and did not consider its plausibility.
+Previous studies have defined death as a censoring event[@frain2017], although failed to explicitly define  the independent censoring assumption and did not consider its plausibility.
 Moreover, adjusting for time-fixed shared causes between dementia and death may be insufficient to block this pathway, and therefore time-varying measurements of $L_2$ should be considered.
@@
 
-On the other hand, death as a competing event is a challenge that has some straightforward strategies, beginning first and foremost by choosing the causal parameter (or estimand) of interest [@young2020, @rojas_medrxiv].
+On the other hand, death as a competing event is a challenge that has some straightforward strategies, beginning first and foremost by choosing the causal parameter (or estimand) of interest[@young2020; @rojas_medrxiv].
 In this study we chose the controlled direct effect, which represents the effect of Pin1 (or cancer) in a setting where death due to cancer and other causes could have been prevented, yet without an explicit intervention, which makes it ambiguous.
 This is different than conceiving a drug-target that increases Pin1 expression only in brain tissue, with no side effects that could increase cancer risk (and thus, death due to cancer).
-As opposed to prior studies that implicitly address a direct effect, and that define censoring as ignorable[@frain2017], we show how point estimates change substantially when we include weights for death to relax the independent censoring assumption [@weave2012, @vangeloven2014, @willems2018, @young2020] when the estimand of interest is the controlled direct effect.
+As opposed to prior studies that implicitly address a direct effect, and that define censoring as ignorable[@frain2017], we show how point estimates change substantially when we include weights for death to relax the independent censoring assumption[@weave2012; @vangeloven2014; @willems2018; @young2020] when the estimand of interest is the controlled direct effect.
 Bounds to assess extreme scenarios of dependence between death and dementia[@peterson1976] illustrate the wide range of possible point estimates that cross the null.
@@
 For example, if researchers used this association to inform the possible effects of different chemotherapeutics on cognitive decline among patients undergoing treatment for cancer, different challenges would arise for mapping the observed association to the hypothetical randomized trial underlying that new research question.
-Notably, that target trial, unlike the one considered here, would include cancer diagnosis as part of the eligibility, and thus researchers would need to instead grapple with how using data on persons without cancer is useful and useable [@huitfeldt2016].
+Notably, that target trial, unlike the one considered here, would include cancer diagnosis as part of the eligibility, and thus researchers would need to instead grapple with how using data on persons without cancer is useful and useable[@huitfeldt2016].
 Each question requires thinking about bias anew and each question brings its own set of challenges and opportunities.
@@
 
-### Table 1: Descriptive characteristics of individuals who had a cancer diagnosis and of those free of cancer diagnosis over follow-up.
+### Table 1: Descriptive characteristics of individuals included in this study.
+
+<br>
 
@@
 **Figure 1.
-Pin1 as biomarker and risk of dementia.** Causal directed acyclic graph where Pin1 at time $t$ is represented as $P_t$, Dementia at time $t+k$ is represented as $Y_{t+k}$.
-$L_1$ represents shared causes of Pin1 and Dementia, such as smoking.
+Causal directed acyclic graph highlighting confounding of the potential effect of Pin1 on risk of dementia.** Pin1 at time $t$ is represented as $P_t$; dementia at time $t+k$ is represented as $Y_{t+k}$; $L_1$ represents shared causes of Pin1 and dementia, such as smoking.
 To isolate the effect of $P_t$ on $Y_{t+k}$ we need to block the backdoor path $Y_{t+k} \leftarrow L_1 \rightarrow P_t$.
-Gray nodes and arrows, which represent additional data-generating mechanisms, will be described progressively in the following figures.
+Gray nodes and arrows are described progressively in Figures 1-5.
 
@@
 **Figure 2.
-Incident cancer diagnosis as proxy for Pin1 expression.** This causal graph is an expansion of Figure 1, includes Pin1 expression at time $P_{t+1}$.
-Incident cancer diagnosis is represented as $C_t$ and $C_{t+1}$.
+Causal directed acyclic graph highlighting the use of incident cancer diagnosis as a proxy for Pin1 expression.** Pin1 at time $t$ and $t+1$ are represented as $P_t$ and $P_{t+1}$, respectively; dementia at time $t+k$ is represented as $Y_{t+k}$; $L_1$ represents shared causes of Pin1 and dementia; incident cancer diagnosis at time $t$ and $t+1$ are represented as $C_t$ and $C_{t+1}$.
 To isolate the effect of $P_t$ and $P_{t+1}$ on $Y_{t+k}$ we need to block the backdoor path $Y_{t+k} \leftarrow L_1 \rightarrow P_t$ and $Y_{t+k} \leftarrow L_1 \rightarrow P_{t+1}$.
 Although we represent $L_1$ as a single node for readability, $L_1$ is time-varying too.
+Gray nodes and arrows are described progressively in Figures 1-5.
 
@@
 **Figure 3.
-Time-varying cancer diagnosis and immortal-time bias.**  This causal graph is an expansion of Figure 2, where $D_{t+1}$ represents death at time $t+1$.
-Cancer diagnosis at $t+1$ can only be measured among those who are alive at $t+1$
+Causal directed acyclic graph highlighting the time-varying nature of cancer diagnosis and immortal time bias.**  Pin1 at time $t$ and $t+1$ are represented as $P_t$ and $P_{t+1}$, respectively; dementia at time $t+k$ is represented as $Y_{t+k}$; $L_1$ represents shared causes of Pin1 and dementia; incident cancer diagnosis at time $t$ and $t+1$ are represented as $C_t$ and $C_{t+1}$.
+$D_{t+1}$ represents death at time $t+1$, cancer diagnosis at $t+1$ can only be measured among those who are alive at $t+1$.
+Gray nodes and arrows are described progressively in Figures 1-5.
 
@@
 **Figure 4.
-Death as competing event of dementia.** In this graph we only focus attention to the exposure as if it was measured for all at time $P_{t+1}$.
+Causal directed acyclic graph highlighting death as a competing event of dementia** Pin1 at time $t$ and $t+1$ are represented as $P_t$ and $P_{t+1}$, respectively; dementia at time $t+k$ is represented as $Y_{t+k}$; $L_1$ represents shared causes of Pin1 and dementia; incident cancer diagnosis at time $t$ and $t+1$ are represented as $C_t$ and $C_{t+1}$; $D_{t+1}$ represents death at time $t+1$.
+In this graph we only focus attention to the exposure as if it was measured for all at time $P_{t+1}$.
 We include $D_{t+k}$ as death at time $t+k$ since $Y_{t+1}$ is only observable when participants are alive at $t+k$.
@@
 Pin1 may affect the risk of death through cancer diagnosis, represented as an arrow between $C_{t+1}$ and $D_{t+k}$.
-To isolate the direct effect of $P_{t+1}$ and $Y_{t+k}$, we have to block the backdoor pathway from $Y_{t+k} \leftarrow L_2 \rightarrow D_{t+k} \leftarrow C_{t+1} \leftarrow P_{t+1}$ and $Y_{t+k} \leftarrow D_{t+k} \leftarrow L_1 \rightarrow P_{t+1}$
+To isolate the direct effect of $P_{t+1}$ and $Y_{t+k}$, we have to block the backdoor pathway from $Y_{t+k} \leftarrow L_2 \rightarrow D_{t+k} \leftarrow C_{t+1} \leftarrow P_{t+1}$ and $Y_{t+k} \leftarrow D_{t+k} \leftarrow L_1 \rightarrow P_{t+1}$.
+Gray nodes and arrows are described progressively in Figures 1-5.
 
@@
 
-**Figure 5. Causal graph for effect of Pin1 through the proxy of cancer diagnosis in the risk of dementia, including all challenges discussed through Figures 1 to 4.**
+**Figure 5.
+Causal directed acyclic graph depicting multiple challenges to using the proxy of cancer diagnosis to study the effect of Pin1 on risk of dementia.** Pin1 at time $t$ and $t+1$ are represented as $P_t$ and $P_{t+1}$, respectively; dementia at time $t+k$ is represented as $Y_{t+k}$; $L_1$ represents shared causes of Pin1 and dementia; incident cancer diagnosis at time $t$ and $t+1$ are represented as $C_t$ and $C_{t+1}$; $D_{t+1}$ and $D_{t+k}$ represents death at time $t+1$ and $t+k$; $L_2$ represents possible shared causes of dementia and death.
+The distinct challenges were highlighted separately in Figures 1-4.
 
